---
title: "Diagnostics & Model Selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diagnostics & Model Selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

Occupancy models simultaneously estimate two processes: occupancy (is a
species present at a site?) and detection (given presence, is it observed on
a given visit?). This dual-process structure makes diagnostics both essential
and non-trivial. Misspecification in either the occupancy or detection
submodel can bias parameters in *both* submodels, because the likelihood
entangles the two. Traditional residual plots, designed for models where
the response is observed directly, do not apply when the latent state
(true presence) is never seen.

INLAocc addresses this with simulation-based diagnostics. The idea is
simple: if the fitted model is correct, data simulated from it
should look like the observed data. Any systematic discrepancy points to a
specific kind of misspecification. This approach requires no distributional
assumptions beyond the model itself, and it works for any occupancy model
structure (single-season, multi-season, spatial, or multi-species).

This vignette covers the full diagnostic and model selection toolkit:
the one-call diagnostic panel (`checkModel()`), individual hypothesis tests
for uniformity, dispersion, outliers, and zero inflation, spatial and
temporal autocorrelation diagnostics, DHARMa integration, posterior
predictive checks, model comparison via information criteria and
cross-validation, model averaging, and identifiability checks. For model
fitting basics, see `vignette("quickstart")`; for spatial models
specifically, see `vignette("spatial-models")`.

## The simulation-based approach

All simulation-based diagnostics in INLAocc follow the same logic. Given
fitted parameters $\hat{\boldsymbol{\theta}}$:

1. Simulate latent occupancy states
   $z_i^{(r)} \sim \text{Bernoulli}(\hat{\psi}_i)$ for replicate
   $r = 1, \ldots, R$.
2. Simulate detection histories
   $y_{ij}^{(r)} \sim \text{Bernoulli}(z_i^{(r)} \cdot \hat{p}_{ij})$.
3. Compute a test statistic $T(\mathbf{y}^{(r)})$ on each simulated
   dataset.
4. Compare the observed statistic $T(\mathbf{y})$ to the distribution
   $\{T(\mathbf{y}^{(r)})\}_{r=1}^R$.

If $T(\mathbf{y})$ falls in the tails of the simulated distribution, the
model is misspecified in the direction that $T$ measures. Different choices
of $T$ target different aspects of model fit: variance (dispersion), tail
behaviour (outliers), the number of all-zero sites (zero inflation), or the
full distribution (PIT residuals).

This approach is well suited to occupancy models because the standard diagnostic toolkit breaks down when the response is latent. In a logistic regression, you observe 0s and 1s and can compute residuals directly. In an occupancy model, you observe detection histories, not occupancy states. A site with zero detections might be unoccupied or might be occupied with low detection probability. There is no way to compute a residual for occupancy without first deciding whether each site is occupied, which is the quantity you are trying to estimate. Simulation-based diagnostics avoid this circularity. They do not require residuals from the occupancy process at all. Instead, they simulate the full generative model: first draw occupancy states from the fitted occupancy probabilities, then draw detection histories from those states and the fitted detection probabilities. The simulated detection histories are compared to the observed detection histories, and any discrepancy is evaluated in terms of the detection data, which are observed.

The second advantage is generality. The test statistic $T$ can be anything: a variance ratio, a count of zeros, a spatial autocorrelation coefficient, a quantile of the distribution. The reference distribution for $T$ comes from the model's own simulations, not from a theoretical distribution that assumes a particular error structure. This means the same diagnostic framework works for single-season models, multi-season models, spatial models, and multi-species models without modification. The only requirement is that the model can simulate from itself, which any properly specified occupancy model can do.

## Setup: fit a model to work with

We simulate data with known parameters and fit a model that we will use
throughout the vignette:

```{r fit-example}
library(INLAocc)

sim <- simulate_occu(N = 300, J = 4,
                     beta_occ = c(0.5, -0.8, 0.8),
                     beta_det = c(0, -0.5),
                     seed = 42)
fit <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim$data, verbose = 0)
```

Because the data-generating model matches the fitted model, all diagnostics
should pass. When they do not, the sections below explain what each failure
means and how to fix it.

## One-call diagnostic panel

`checkModel()` produces a four-panel diagnostic plot that gives a quick
visual overview of model adequacy:

```{r checkmodel}
checkModel(fit)
```

The four panels are:

1. **QQ plot of PIT residuals.** Probability Integral Transform (PIT)
   residuals transform each observed value through the fitted CDF. Under the
   true model, these residuals follow a Uniform(0, 1) distribution, so the
   QQ plot should track the diagonal. An S-shaped curve indicates
   overdispersion (more variance than the model predicts). An inverted S
   indicates underdispersion. A systematic shift away from the diagonal
   suggests location bias: the model is consistently over- or
   under-predicting. Pay attention to where the deviations occur. Departures at the tails (the upper-right and lower-left corners of the QQ plot) indicate problems with extreme values: outlier sites that the model cannot explain, or heavy-tailed variation in the data. Departures in the middle of the plot point to problems with the mean structure: the model is getting the central tendency wrong, which usually means a missing covariate or a misspecified functional form.

2. **Residuals vs fitted.** Deviance residuals plotted against fitted
   detection probabilities. This should be a formless cloud with no trends.
   A funnel shape (residuals spreading out as fitted values increase) indicates heteroscedasticity, meaning the variance of the detection process is not constant across sites. A curved trend suggests a
   missing covariate or a wrong functional form (e.g., a linear term where a
   quadratic is needed). A step pattern, where residuals jump at a particular fitted value, suggests a threshold effect in the data that the model's smooth link function cannot capture.

3. **Dispersion histogram.** The variance of site-level detection counts is
   computed for each simulated dataset, producing a reference distribution.
   The red vertical line marks the observed variance. A ratio greater than
   about 1.5 suggests overdispersion; below about 0.7 suggests
   underdispersion. Overdispersion in occupancy models most often signals unmodelled heterogeneity: missing random effects, unaccounted spatial structure, or a covariate that groups sites into categories with different detection rates. Before adding complexity, check whether a single missing covariate explains the excess variance.

4. **Correlogram.** If site coordinates are available, this panel shows
   Moran's I at increasing numbers of neighbours. A flat line near zero
   means no residual spatial autocorrelation. Peaks at small neighbour
   counts suggest local spatial structure that the model is not capturing.

`checkModel()` returns its results invisibly, so you can extract numerical
summaries:

```{r checkmodel-extract}
diag <- checkModel(fit)
diag$ks_p       # KS test p-value for PIT residuals
diag$disp_ratio # observed / expected dispersion ratio
```

## Individual diagnostic tests

Each test below returns an `htest` object with a `$p.value` that you can
use programmatically.

### Uniformity (PIT residuals)

```{r uniformity}
testUniformity(fit)
```

A high p-value means the observed data are consistent with the fitted model
across the entire distribution. This is the most general diagnostic: it
detects any systematic departure from the model, though it has less power
than targeted tests for specific problems.

### Dispersion

```{r dispersion}
testDispersion(fit)
```

A dispersion ratio near 1 with a high p-value indicates correctly specified
variance. Overdispersion (ratio substantially above 1) in occupancy models
most often points to missing covariates, unmodelled heterogeneity among
sites, or residual spatial structure.

### Outliers

```{r outliers}
testOutliers(fit)
```

Sites are flagged as outliers if they fall outside the simulation envelope.
A similar number of outlier sites to the expected count is unremarkable. A large excess of
outliers points to heavy-tailed variation that the model does not account
for.

### Zero inflation

```{r zero-inflation}
testZeroInflation(fit)
```

"Zeros" here means sites where the species was never detected across all
visits. The test is one-sided: it asks whether there are *more* all-zero
sites than the model predicts. A significant result suggests either that
true occupancy is lower than estimated (the occupancy submodel is
misspecified) or that an additional zero-generating process is at work, such
as a habitat suitability threshold that makes some sites structurally
unsuitable.

In occupancy models, all-zero detection histories have two possible causes: the site is truly unoccupied, or the species is present but was never detected. The zero-inflation test evaluates whether the observed number of all-zero sites exceeds what the model predicts after accounting for both sources. A significant result means the model is producing too few zeros in simulation, which typically indicates that some sites are structurally unsuitable for the species. For example, a forest bird survey that includes open grassland sites will show excess zeros if the model lacks a forest cover covariate. The model assigns moderate occupancy probability to grassland sites (because it has no information to exclude them), but the species is never there. Adding a habitat suitability covariate usually resolves the zero inflation.

## Spatial diagnostics

### Moran's I

Residual spatial autocorrelation means the model is not capturing all
spatial structure in the data. To test for it, first simulate spatially
structured data and fit two models: one that ignores space, one that
accounts for it:

```{r moran-setup}
sim_sp <- simulate_occu(N = 300, J = 4, spatial_range = 0.3, seed = 123)

# Non-spatial fit (ignores spatial structure)
fit_nospatial <- occu(~ occ_x1, ~ det_x1, data = sim_sp$data, verbose = 0)
moranI(fit_nospatial)
```

Significant Moran's I confirms that the non-spatial model leaves spatial
structure in the residuals. Now fit a spatial model:

```{r moran-spatial}
fit_sp <- occu(~ occ_x1, ~ det_x1, data = sim_sp$data,
               spatial = sim_sp$data$coords, verbose = 0)
moranI(fit_sp)
```

Non-significant Moran's I confirms that the spatial random effect has absorbed
the spatial structure. Two weighting schemes are available:

- `weights = "inverse"` (default): inverse-distance weights across all pairs
  of sites. Sensitive to broad-scale spatial structure.
- `weights = "knn"` with `k = 10` (or another value): k nearest neighbours.
  Sensitive to local spatial structure.

```{r moran-knn}
moranI(fit_sp, weights = "knn", k = 10)
```

### Semivariogram

```{r variogram}
variogram(fit_sp)
```

The semivariogram plots semivariance against distance. A flat line at the
sill means no remaining spatial trend in the residuals. An increasing curve
that does not flatten suggests the SPDE mesh is too coarse or the fitted
range parameter is too short. See `vignette("spatial-models")` for guidance
on mesh construction and range priors.

## Temporal diagnostics

For multi-season models, the Durbin-Watson test checks for residual lag-1
autocorrelation across seasons:

```{r durbin-watson}
sim_t <- simTOcc(N = 100, J = 3, n_seasons = 5, seed = 300)
fit_t <- occu(~ 1, ~ 1, data = sim_t$data, temporal = "ar1", verbose = 0)

durbinWatson(fit_t)
```

The DW statistic ranges from 0 to 4. A value near 2 indicates no
autocorrelation. Values near 0 suggest positive autocorrelation (successive
residuals are similar), and values near 4 suggest negative autocorrelation
(successive residuals alternate in sign). If the AR(1) temporal structure
is adequate, the DW statistic should be close to 2.

## DHARMa integration

If [DHARMa](https://CRAN.R-project.org/package=DHARMa) is installed,
`dharma()` converts INLAocc's simulation-based residuals into a DHARMa
residual object. This gives access to DHARMa's full suite of diagnostic
tests and plots:

```{r dharma}
res <- dharma(fit)
plot(res)

# Any DHARMa test works on the result
DHARMa::testDispersion(res)
DHARMa::testSpatialAutocorrelation(res, x = sim$data$coords[, 1],
                                        y = sim$data$coords[, 2])
```

The advantage of this integration is interoperability: if you already use
DHARMa for GLMs or GAMs, you can apply the same workflow to occupancy
models. INLAocc handles the occupancy-specific simulation (accounting for
the latent state) and DHARMa handles the visualization and testing.

## Posterior predictive checks

`ppcOccu()` implements Bayesian posterior predictive checks (Gelman et al.
2013). It computes a fit statistic on both the observed data and on
datasets replicated from the posterior, then reports a Bayesian p-value:

```{r ppc}
ppc <- ppcOccu(fit, fit.stat = "freeman-tukey", group = 1)
ppc$bayesian.p
```

A Bayesian p-value near 0.5 indicates good fit; the model generates data
whose fit statistic distribution is centred on the observed value. Values
near 0 or 1 indicate systematic misfit. The `fit.stat` argument accepts
`"freeman-tukey"` or `"chi-squared"`. The `group` argument controls
aggregation: `group = 1` computes the statistic by site (sensitive to
site-level misfit), `group = 2` computes it by visit (sensitive to
visit-level misfit).

The choice of fit statistic affects what kinds of misfit you detect. Freeman-Tukey is sensitive to both the location and spread of the distribution, making it a good default for overall model assessment. Chi-squared emphasizes individual observations that deviate strongly from their expected values, so it is better at flagging specific problem sites. For a well-specified model, both statistics give similar Bayesian p-values. If Freeman-Tukey passes but chi-squared fails, look for a small number of outlier sites rather than a systematic problem. Use Freeman-Tukey as the default; switch to chi-squared when you suspect that a few individual sites are driving the misfit.

## Model comparison

### Information criteria

Fit a set of candidate models and compare them:

```{r compare}
m1 <- occu(~ 1, ~ 1, data = sim$data, verbose = 0)
m2 <- occu(~ occ_x1, ~ det_x1, data = sim$data, verbose = 0)
m3 <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim$data, verbose = 0)

compare_models(null = m1, simple = m2, full = m3, criterion = "waic")
```

The `delta` column gives the difference in WAIC from the best model.
`weight` gives the Akaike-style model weight, interpretable as the
probability that each model is the best approximating model given the
candidate set. WAIC (Widely Applicable Information Criterion) is
recommended for Bayesian models because it uses the full posterior rather
than a point estimate. AIC and BIC are also available via
`criterion = "aic"` and `criterion = "bic"`.

WAIC, AIC, and BIC answer related but distinct questions. WAIC uses the full posterior predictive distribution to evaluate model fit, which means it accounts for parameter uncertainty. AIC uses only the maximum likelihood estimate (or in Bayesian contexts, the posterior mode) and adds a penalty equal to the number of parameters. For models with simple fixed-effect structures and flat priors, WAIC and AIC give nearly identical rankings. The differences appear when the model includes informative priors or complex random effects. In those cases, AIC undercounts the effective complexity because it does not account for the prior's regularizing effect, while WAIC captures this through the effective number of parameters $p_{\text{WAIC}}$.

BIC penalizes complexity more heavily than AIC, with the penalty growing as $\log(N)$ instead of a constant 2 per parameter. For large sample sizes, BIC strongly favors simpler models. This is useful when parsimony matters more than predictive accuracy, for instance when selecting a minimal model for population monitoring where the same model will be applied across many years and regions. In practice for occupancy models: use WAIC as the default criterion, AIC as a cross-check (if they disagree, investigate why), and BIC when you need the simplest defensible model. When all three criteria select the same model, you can be confident in the ranking.

### WAIC decomposition

`waicOccu()` decomposes WAIC into contributions from the occupancy and
detection submodels:

```{r waic-detail}
w <- waicOccu(m3)
w
```

The `pD` column gives the effective number of parameters for each component.
If `pD` is much larger than the actual number of parameters, the model may
be overfitting. Consider simplifying the covariate structure or adding
regularizing priors.

### K-fold cross-validation

For out-of-sample predictive assessment, pass `k.fold` during fitting:

```{r kfold}
fit_cv <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim$data, k.fold = 5, verbose = 0)
fit_cv$k.fold
```

This holds out 1/k of sites at a time, refits on the remaining data, and
evaluates predictions on the held-out sites. K-fold cross-validation is more
reliable than information criteria for small datasets or models with complex
random effect structures, but it is k times slower because the model must be
refit for each fold.

## Model averaging

When no single model clearly dominates (for example, when the top two
models have similar WAIC weights), averaging predictions across models
avoids conditioning on a potentially wrong model:

```{r averaging}
avg <- modelAverage(null = m1, simple = m2, full = m3, criterion = "waic")

avg$weights       # model weights used for averaging
avg$psi_hat[1:5]  # model-averaged occupancy estimates (first 5 sites)
avg$psi_se[1:5]   # unconditional standard errors (first 5 sites)
```

The unconditional standard errors follow Burnham & Anderson (2002,
eq. 4.9). They account for both within-model uncertainty (how precisely each
model estimates occupancy) and between-model uncertainty (how much the
models disagree). This avoids the false precision that comes from
conditioning on a single selected model.

## Prior specification

INLAocc uses penalised complexity (PC) priors for random effects and spatial components. The default settings work well for most applications, but understanding how to adjust them is important when defaults produce unexpected results.

### Fixed effects

The default prior on all fixed-effect coefficients is $\mathcal{N}(0, 2.72)$ (precision 0.367), following Gelman et al. (2008) for logistic regression. On the logit scale, this gives a 95% prior interval of approximately $(-3.3, 3.3)$, which corresponds to occupancy probabilities from about 0.04 to 0.96. This is weakly informative: it rules out extreme values but lets the data dominate.

To change fixed-effect priors:

```{r prior-fixed}
prior <- occu_priors(
  beta.normal = list(mean = 0, var = 1),    # tighter on occupancy
  alpha.normal = list(mean = 0, var = 2.72)  # default on detection
)
fit_prior <- occu(~ occ_x1, ~ det_x1, data = sim$data, priors = prior)
```

### Random effects

PC priors on random effect precision have two parameters: a scale $u$ and a tail probability $\alpha$, interpreted as $P(\sigma > u) = \alpha$. The default is $P(\sigma > 1) = 0.05$, meaning there is a 5% prior probability that the random effect standard deviation exceeds 1 on the logit scale. This shrinks toward the base model (no random effect) unless the data provide evidence otherwise.

When to adjust:

- **Few groups (< 5)**: The variance estimate is poorly identified. A tighter prior (e.g., $P(\sigma > 0.5) = 0.01$) prevents unreasonably large variance estimates.

- **Many groups (> 50)**: The data strongly inform the variance. A looser prior (e.g., $P(\sigma > 2) = 0.1$) avoids over-shrinkage.

- **Strong prior knowledge**: If previous studies show between-region SD of about 0.5 on the logit scale, set $P(\sigma > 0.5) = 0.5$ to centre the prior there.

### Spatial components

The SPDE spatial field uses PC priors on the range and marginal standard deviation, set via `occu_spatial()` or `spde.args`. The range prior $P(\text{range} < r_0) = p$ controls the minimum expected spatial correlation distance. The variance prior $P(\sigma > s_0) = p$ controls the maximum expected spatial variation.

### Checking prior influence

If a posterior is nearly identical to its prior, the data are not informing that parameter. Compare:

```{r prior-check}
# Compare prior vs posterior for the occupancy intercept
# Default prior: N(0, 2.72) -> SD = sqrt(2.72) = 1.65
prior_sd <- sqrt(2.72)
posterior_sd <- fit$occ_fit$summary.fixed["(Intercept)", "sd"]
cat("Prior SD:", round(prior_sd, 2), "\n")
cat("Posterior SD:", round(posterior_sd, 2), "\n")
cat("Ratio (posterior/prior):", round(posterior_sd / prior_sd, 2), "\n")
if (posterior_sd / prior_sd > 0.9) {
  cat("Warning: posterior SD is close to prior SD",
      "-- data may not inform this parameter\n")
} else {
  cat("Data is informing this parameter (posterior is tighter than prior)\n")
}
```

Prior-posterior overlap is a useful diagnostic for individual parameters. If the posterior distribution for a coefficient is nearly identical to the prior, the data contain little information about that parameter. A concrete way to check this: compare the posterior standard deviation to the prior standard deviation. With the default $\mathcal{N}(0, 2.72)$ prior, the prior SD is 1.65. If a coefficient's posterior SD is also around 1.65, the data have not narrowed the estimate beyond what the prior already specified. The posterior mean may differ from zero, but if the width has not changed, the estimate is driven by the prior rather than by the data. This is not necessarily a problem if the prior is well justified, but it means the data do not support inference about that particular effect.

For random effects, the analogous check compares the posterior distribution of the variance (or precision) to the PC prior. The PC prior places most of its mass near zero variance, favoring the simpler model without random effects. If the posterior for the random effect variance sits right on top of the prior, the data cannot distinguish the random effect from noise. This happens when the number of groups is small (fewer than 5 or 6) or when the between-group variance is genuinely small relative to within-group variance. In either case, the random effect is not contributing to the model, and you should consider dropping it.

## Choosing the number of latent factors

### Why factor count matters

Multi-species occupancy models estimate separate occupancy and detection parameters for every species in a community. With $S$ species, the number of pairwise species correlations in occupancy grows as $S(S-1)/2$. For a 20-species community, that is 190 covariance parameters --- far too many to estimate directly from typical survey data. Latent factor models solve this by assuming that species correlations arise from a small number of shared latent gradients. Each species loads onto $k$ latent factors, and the $S \times S$ covariance matrix is approximated by the product $\Lambda \Lambda^\top$, where $\Lambda$ is the $S \times k$ matrix of factor loadings. This reduces the covariance parameterization from $O(S^2)$ to $O(Sk)$, which is tractable when $k \ll S$.

The number of latent factors $k$ controls this dimension reduction. It is a model selection decision analogous to choosing the number of principal components in a PCA: too few factors miss the dominant species associations, biasing occupancy estimates for species whose co-occurrence patterns are not represented; too many factors overfit to sampling noise in the detection histories, inflating variance and slowing computation without improving ecological insight.

Choosing $k$ matters for three reasons. First, it determines how well the model captures species correlations, which directly affects the precision of occupancy estimates for rare species that borrow information from common ones. Second, it controls the effective number of parameters and therefore the risk of overfitting. Third, it determines computational cost, because each additional factor adds $S$ loading parameters and $N$ latent scores that must be estimated.

### A worked example

We simulate a community of 15 species surveyed at 150 sites with 4 repeat visits, then fit latent factor models with $k = 1, 2, 3, 5,$ and $8$ factors. The simulation uses the default community-level coefficient distributions, which produce a mix of common and rare species with moderate inter-species correlation.

```{r factor-setup}
sim_ms <- simMsOcc(N = 150, J = 4, n_species = 15, seed = 42)
```

Now we fit a sequence of models, each with a different number of latent factors, and record WAIC and computation time for each:

```{r factor-fit-sequence}
k_values <- c(1, 2, 3, 5, 8)

results <- lapply(k_values, function(k) {
  t0 <- proc.time()
  fit_k <- occu(~ occ_x1, ~ det_x1, data = sim_ms$data,
                multispecies = TRUE, n.factors = k, verbose = 0)
  elapsed <- (proc.time() - t0)["elapsed"]
  waic_k <- waicOccu(fit_k)$WAIC
  list(k = k, waic = waic_k, time = elapsed, fit = fit_k)
})

factor_table <- data.frame(
  k       = sapply(results, `[[`, "k"),
  WAIC    = sapply(results, `[[`, "waic"),
  time_s  = round(sapply(results, `[[`, "time"), 1)
)
factor_table$delta_WAIC <- factor_table$WAIC - min(factor_table$WAIC)
factor_table
```

The table shows WAIC, the WAIC difference from the best model ($\Delta$WAIC), and wall-clock time in seconds. We can visualize this as an elbow plot:

```{r factor-elbow-plot, fig.height = 4}
library(ggplot2)

ggplot(factor_table, aes(x = k, y = WAIC)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 3) +
  labs(x = "Number of latent factors (k)", y = "WAIC") +
  theme_minimal() +
  theme(
    plot.background = element_rect(fill = "transparent", colour = NA),
    panel.background = element_rect(fill = "transparent", colour = NA)
  )
```

The pattern to look for is an elbow: WAIC improves steeply as $k$ increases from 1 to some value, then flattens or begins to increase. The point where the curve flattens is the number of factors that captures the dominant species associations without overfitting. In this simulated example with 15 species and no true latent factor structure (species correlations arise only from shared covariates), a small number of factors (typically 2 or 3) is sufficient, and additional factors provide negligible improvement.

### Examining factor loadings

After selecting a value of $k$, inspect the estimated factor loadings to check whether the factors have ecological interpretability. The loadings matrix $\Lambda$ shows how strongly each species associates with each latent gradient:

```{r factor-loadings}
best_idx <- which.min(factor_table$WAIC)
best_fit <- results[[best_idx]]$fit

lam <- best_fit$lambda
rownames(lam) <- best_fit$species_names
colnames(lam) <- paste0("Factor_", seq_len(ncol(lam)))
round(lam, 3)
```

Species with loadings of the same sign on a given factor tend to co-occur; species with opposite signs tend to segregate. A factor where most species load near zero is not contributing meaningful structure and can likely be dropped. In practice, you are looking for factors where a clear subset of species loads strongly (absolute loading above 0.3) and the remaining species are near zero. Such a factor represents a shared ecological gradient --- for example, a preference for wet versus dry habitats, or a response to canopy cover.

If the loadings matrix shows no interpretable pattern even at the optimal $k$ by WAIC, this does not mean the factors are useless. Latent factors serve two purposes: capturing interpretable species associations and improving prediction by borrowing strength across species. A model can achieve the second goal without achieving the first. Uninterpretable factors that improve WAIC are still doing useful work: they absorb residual covariance that would otherwise inflate standard errors.

### Rules of thumb for choosing $k$

**Starting point.** A reasonable initial guess is $k = \lfloor S / 2 \rfloor$, where $S$ is the number of species. Then search downward by comparing WAIC. For the 15-species community above, this suggests starting at $k = 7$ and checking whether simpler models perform as well. This starting point is deliberately generous; it is easier to simplify a model that works than to discover that you need more factors after fitting everything with too few.

**Small communities (5--10 species).** With few species, the number of pairwise correlations is modest ($\leq 45$), and 1--3 factors usually suffice. The data rarely contain enough information to estimate more factors reliably. If $k = 1$ and $k = 2$ give similar WAIC, prefer $k = 1$ for interpretability.

**Medium communities (10--30 species).** This is the range where factor models provide the most benefit. The covariance matrix is too large to estimate freely, but there are enough species that shared gradients can be identified. Expect 2--5 factors to be optimal in most cases. If the community includes distinct guilds (e.g., forest specialists and grassland specialists), you may need one factor per guild.

**Large communities (30+ species).** With many species, more factors may be needed to capture the full correlation structure. However, the marginal gain from each additional factor decreases, and computational cost increases. Rarely are more than 8--10 factors justified in practice. For communities of 50+ species, consider whether a hierarchical grouping (e.g., by functional guild or taxonomic family) might be more appropriate than a single large factor model.

**Data richness.** The number of sites and visits constrains how many factors the data can support. As a rough guide, you need at least $10 \cdot k$ sites with detections spread across multiple species to estimate $k$ factors reliably. A dataset with 50 sites and 3 visits per site can support at most 2--3 factors regardless of the number of species. Adding more factors with sparse data leads to loadings that are poorly identified: the posterior concentrations around the prior rather than around a data-driven estimate.

**Rare species.** Communities with many rare species (detected at fewer than 10% of sites) benefit from factor models because the factors borrow information from common species to improve estimates for rare ones. However, rare species also have the weakest signal for estimating their own loadings. If most species are rare, prefer fewer factors. The loadings for rare species will be dominated by the prior regardless of $k$, and adding factors mainly helps the common species.

### When more factors help versus when they overfit

More factors help when the true species correlation structure is complex, for example when the community spans multiple habitat types and species respond differently to each. In such cases, a single factor cannot capture the pattern of some species co-occurring in forests while others co-occur in grasslands, and separate factors are needed for each habitat axis. The WAIC improvement from adding the second or third factor is substantial (typically $> 10$ WAIC units) and consistent across different random seeds.

More factors overfit when the true correlations are simple or when the data are sparse. Overfitting manifests as WAIC that does not improve (or gets worse) as $k$ increases, and as loadings that are small in magnitude for most species on the added factors. A more subtle sign is instability: if you refit the model with a different random seed and the loadings on the last factor change sign or reorder, that factor is fitting noise.

A practical diagnostic is to check whether the per-species WAIC decomposition changes with $k$:

```{r factor-per-species-waic}
# Compare per-species WAIC between k=2 and k=5
fit_k2 <- results[[which(factor_table$k == 2)]]$fit
fit_k5 <- results[[which(factor_table$k == 5)]]$fit

sp_waic_k2 <- waicOccu(fit_k2, by.sp = TRUE)
sp_waic_k5 <- waicOccu(fit_k5, by.sp = TRUE)

delta_sp <- data.frame(
  species    = sp_waic_k2$species,
  WAIC_k2    = round(sp_waic_k2$WAIC, 1),
  WAIC_k5    = round(sp_waic_k5$WAIC, 1),
  difference = round(sp_waic_k5$WAIC - sp_waic_k2$WAIC, 1)
)
delta_sp
```

If the additional factors help only one or two species and make no difference for the rest, the improvement is fragile; it may reflect idiosyncratic noise in those species rather than genuine shared structure. A reliable improvement from adding factors should be distributed across multiple species.

### Computational cost

Each latent factor adds $S$ loading parameters and $N$ factor scores. The dominant cost in INLAocc's latent factor models is the per-species INLA fits, which scale with the number of sites and the complexity of the covariate structure. The factor extraction itself (via PCA on the fitted logit-occupancy matrix) is fast and adds negligible overhead.

The main computational scaling is therefore:

- **Per-species fits**: $O(S)$ independent INLA calls, each scaling with $N$ and the number of covariates. This is the bottleneck and can be parallelized via `options(INLAocc.cores = 4)`.
- **Factor extraction**: $O(N \cdot S \cdot k)$ for the PCA, which is negligible relative to the INLA fits.
- **Memory**: The factor loadings matrix and scores add $O(S \cdot k + N \cdot k)$ storage, which is small.

In practice, the wall-clock time scales approximately linearly with $S$ (since each species is fit independently) and is nearly independent of $k$. The factor count affects the post-processing step but not the dominant per-species fitting cost. This means that the computational argument for fewer factors is about preventing overfitting, not about speed. You can compare the timing column in the table above to confirm that fitting with $k = 8$ factors is not substantially slower than fitting with $k = 2$ factors for this dataset.

For large communities where per-species fitting is the bottleneck, enable parallel fitting:

```{r factor-parallel, eval = FALSE}
options(INLAocc.cores = 4)
fit_par <- occu(~ occ_x1, ~ det_x1, data = sim_ms$data,
                multispecies = TRUE, n.factors = 3, verbose = 0)
```

### Checking whether factor count changes ecological conclusions

The final and most important check: does the choice of $k$ affect the ecological conclusions you would draw? If the community-level covariate effects are stable across different values of $k$, the factor count is a modelling detail rather than a scientific decision. If they change, you need to understand why.

Compare the community-level occupancy effect of the covariate across different values of $k$:

```{r factor-coef-stability}
coef_table <- do.call(rbind, lapply(results, function(r) {
  cf <- coef(r$fit, process = "occupancy")
  # Community-level mean coefficients
  data.frame(
    k         = r$k,
    intercept = mean(cf[grepl("Intercept", names(cf))]),
    occ_x1    = mean(cf[grepl("occ_x1", names(cf))])
  )
}))
coef_table
```

If the estimated effect of `occ_x1` is 0.4 with $k = 2$ and 0.5 with $k = 5$, the conclusions are the same (moderate positive effect). If it is 0.4 with $k = 2$ and $-0.1$ with $k = 5$, the additional factors are absorbing variation that was previously attributed to the covariate, and you need to investigate further, because the covariate effect may be confounded with a latent gradient.

A stable set of conclusions across a range of $k$ values is reassuring. It means the covariate effects are identified by the data regardless of how the residual species correlations are modelled. Instability means the covariate effects depend on the correlation structure, which is a sign that more data or a different model structure is needed to separate the two.

### Summary of the factor selection workflow

1. **Simulate or assemble data.** Ensure the multi-species data object is correctly formatted via `simMsOcc()` or `occu_format_ms()`.

2. **Fit a sequence.** Start at $k = \lfloor S / 2 \rfloor$ and fit models for $k = 1, 2, 3, \ldots$ up to that value.

3. **Compare WAIC.** Plot the elbow curve. Select the $k$ where WAIC stops improving ($\Delta$WAIC $< 2$ relative to the next simpler model).

4. **Inspect loadings.** Check whether the selected factors have ecological interpretability. Uninterpretable factors that improve WAIC are acceptable but deserve scrutiny.

5. **Check per-species WAIC.** Verify that the improvement from additional factors is distributed across species, not driven by one or two.

6. **Test conclusion stability.** Compare community-level covariate effects across values of $k$. If conclusions are stable, report the simplest adequate model. If they are not, investigate the source of instability before proceeding.

## Identifiability

Before interpreting coefficients, verify that the model is identifiable:

```{r identifiability}
checkIdentifiability(fit)
```

This function flags three kinds of problems:

- **Flat posteriors**: a parameter whose posterior is essentially the same as
  its prior, meaning the data provide no information about it.
- **Extreme correlations**: pairs of parameters with posterior correlation
  above 0.95 in absolute value, indicating that the data cannot separate
  their effects.
- **Boundary estimates**: coefficients at or near the edge of the parameter
  space (e.g., detection probability near 0 or 1).

Identifiability problems are common in occupancy models when detection
probability is very low (few detections per site), when the same covariate
appears in both submodels, or when the number of visits per site is small.
Remedies include increasing survey effort, reducing model complexity, using
informative priors on detection parameters, or fixing one submodel while
varying the other.

## A diagnostic workflow

A practical checklist for diagnosing a fitted occupancy model:

1. **`checkModel(fit)`**: visual overview of the four key diagnostics.

2. **`testUniformity(fit)`**: overall calibration of the model.

3. **`testDispersion(fit)`**: variance structure.

4. **`testZeroInflation(fit)`**: excess all-zero detection histories.

5. **`moranI(fit)`**: residual spatial autocorrelation (if site
   coordinates are available).
6. **`durbinWatson(fit)`**: residual temporal autocorrelation (if
   multi-season model).
7. **`compare_models(...)`**: select the best model from candidates.

8. **`ppcOccu(fit)`**: Bayesian posterior predictive validation.

9. **`checkIdentifiability(fit)`**: parameter identifiability.

Steps 1--4 diagnose the current model. Step 5--6 check for structure the
model may be missing. Steps 7--8 evaluate the model against alternatives
and against its own assumptions. Step 9 ensures the parameters you plan to
interpret are actually informed by the data.

## When diagnostics fail

Common diagnostic failures and their typical remedies:

- **Significant Moran's I.** The model is not capturing spatial structure.
  Add a spatial random effect via the `spatial` argument to `occu()`. See
  `vignette("spatial-models")`.

  For example, a model for a bird species that uses only climate covariates (temperature, precipitation) will show significant spatial autocorrelation if the species also depends on land cover and land cover is spatially clustered. The climate variables explain some of the spatial pattern, but the residual pattern in land cover leaks into the residuals. Adding a land cover covariate may resolve the autocorrelation without needing a spatial random effect. If it does not, the remaining spatial signal is from unmeasured processes, and a spatial term is needed.

- **Overdispersion (ratio >> 1).** More variance than the model predicts.
  Consider adding site-level random effects, additional covariates, or a
  spatial term. Overdispersion often signals missing heterogeneity rather
  than a wrong distribution.

  A common scenario: a detection model with a single observer-skill covariate, but the survey was conducted by three observers with very different skill levels. If the covariate only captures average experience (years of fieldwork) but not observer-specific quirks (one observer is unusually good at auditory detection, another is color-blind), the residual variation in detection rates across sites will exceed what the model predicts. Adding observer as a random effect, or including observer-specific covariates, will absorb this extra variation.

- **Excess zeros.** More all-zero sites than expected. This can indicate
  that the occupancy submodel needs additional covariates (some sites are
  truly unoccupied for reasons the model does not capture) or that a
  zero-inflated occupancy structure is needed.

  Consider a survey of a forest-dependent species where the sampling design includes sites at the edge of the species' range with marginal habitat. The model, using elevation and canopy cover as covariates, assigns moderate occupancy probability to these marginal sites. But some of them are simply too small or too isolated to support a population, and no combination of the available covariates can separate them from the suitable sites. The result is an excess of all-zero detection histories. Adding a patch size or connectivity covariate, if available, can fix this. If no such covariate exists, a zero-inflated model structure explicitly accounts for the two sources of zeros.

- **Non-uniform PIT residuals.** The model is systematically mis-calibrated.
  Check the functional form of covariates. Try adding polynomial or
  spline terms if a covariate's effect may be nonlinear. Also check for
  missing covariates.

  A typical case is a unimodal relationship between occupancy and elevation: the species is most common at intermediate elevations and absent at both extremes. A linear elevation term forces a monotonic relationship, which under-predicts occupancy at intermediate sites and over-predicts at the extremes. The PIT residuals will show a systematic S-shape or reverse S-shape. Adding a quadratic term (elevation + elevation^2) usually resolves this.

- **Bayesian p-value near 0 or 1.** The model fundamentally cannot
  reproduce the observed data. This usually requires rethinking the model
  structure rather than tweaking covariates: consider whether the occupancy
  or detection process is specified correctly, whether there is unmodelled
  heterogeneity, or whether the link function is appropriate.

  This failure is qualitatively different from the others. A Bayesian p-value near zero means no set of parameter values under the current model structure can generate data that resemble the observed data. This happens, for instance, when detection probability varies strongly across visits (due to weather or season) but the model uses a constant detection intercept with no visit-level covariates. The model cannot produce the observed pattern of detections regardless of how it sets its parameters.

- **Identifiability warning.** The data do not support the current model
  complexity. Reduce the number of covariates, increase survey effort
  (more sites or more visits per site), use informative priors on detection
  parameters, or fix one submodel to a simpler form while varying the
  other. For multi-species models (`vignette("multi-species")`),
  shared detection parameters across species can improve identifiability.

  A frequent cause is including the same covariate in both the occupancy and detection submodels. For instance, if "distance to road" appears in both submodels, the model must decide whether sites far from roads have low occupancy or low detection (because they are harder to access). With limited data, both explanations fit equally well, and the posterior for that covariate in both submodels becomes wide and uninformative. The fix is to decide, based on ecological reasoning, which submodel the covariate belongs in, and remove it from the other.

## Convergence troubleshooting

When the EM algorithm does not converge within `max_iter` iterations, the model reports `converged = FALSE`. Common causes and fixes:

- **Too many parameters for the data.** Simplify the model: fewer covariates, fewer random effects, no spatial term. Fit the simplest model first and add complexity incrementally.

- **Near-complete separation.** A covariate perfectly predicts occupancy or detection. Check `summary(fit)` for coefficients with very large absolute values (> 5) and wide credible intervals. Remove or transform the problematic covariate.

- **Low detection.** When naive detection is below 0.10, the E-step weights for undetected sites are all near the prior, and the EM has little information to update. Increase survey effort, simplify the detection model, or use informative priors on detection.

- **Oscillation.** The EM alternates between two parameter values without settling. Increase the damping parameter via `control = list(damping = 0.5)` or higher. Damping slows convergence but prevents oscillation.

Use `verbose = 2` to monitor per-iteration convergence:

```{r verbose}
fit_debug <- occu(~ occ_x1, ~ det_x1, data = sim$data, verbose = 2)
```

This prints the maximum change in $\psi$ and $p$ at each iteration, the current damping value, and whether detection has been frozen. Look for: (a) delta values that decrease monotonically (healthy convergence), (b) delta values that oscillate (increase damping), or (c) delta values that plateau above `tol` (simplify model).

For details on the EM algorithm and its tuning parameters, see `vignette("algorithm-details")`.
