---
title: "Model Identifiability"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Identifiability}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

Occupancy models have a fundamental identifiability challenge: a site with no
detections could be truly unoccupied, or it could be occupied with low detection
probability. Separating occupancy from detection requires enough visits per site
and enough variation in detection/non-detection patterns across sites and visits.
Without these, the model cannot determine which process (occupancy or
detection) is responsible for the observed data, and parameter estimates
become unreliable.

When models are not identifiable, parameters have flat posteriors (the data
cannot distinguish between different parameter values), estimates sit on the
boundary of the parameter space, or small changes in the data produce large
changes in estimates. These symptoms are not always obvious. A model can
converge, return point estimates, and produce confidence intervals that look
reasonable, while the estimates are driven almost entirely by the prior or
by numerical regularization rather than by the data.

INLAocc provides `checkIdentifiability()` for automated screening of common
identifiability issues, both before and after fitting a model. This vignette
explains the conceptual framework behind occupancy identifiability, demonstrates
the built-in checks, and presents a simulation-based stress-testing approach for
deeper assessment. For model fitting basics, see `vignette("quickstart")`; for
post-fit diagnostics, see `vignette("diagnostics")`.

## The identifiability problem

### The occupancy likelihood

Consider a site $i$ surveyed on $J$ visits. The probability of the observed
detection history depends on two unknowns: the occupancy probability $\psi_i$
and the detection probabilities $p_{ij}$. For a site where the species was
never detected, the likelihood contribution is:

$$L_i = (1 - \psi_i) + \psi_i \prod_{j=1}^{J} (1 - p_{ij})$$

The first term is the probability that the site is truly unoccupied. The second
is the probability that the site is occupied but the species went undetected on
every visit. The data cannot distinguish between these two contributions;
both produce the same observation (all zeros).

### Why one visit is not enough

With $J = 1$ visit, the likelihood for an undetected site simplifies to:

$$L_i = (1 - \psi_i) + \psi_i (1 - p_i) = 1 - \psi_i p_i$$

Only the product $\psi_i p_i$ is identifiable, not $\psi_i$ and $p_i$
individually. Any combination of occupancy and detection that produces the
same product fits the data equally well. A species with 80% occupancy and 50%
detection looks identical to one with 40% occupancy and 100% detection.

### Repeated visits break the degeneracy

With $J \geq 2$ independent visits, the pattern of detections and
non-detections across visits provides information to separate the two
processes. Sites where the species is detected on some visits but not others
are particularly informative: they demonstrate that detection is imperfect,
which allows the model to estimate how much non-detection is due to absence
versus missed detections.

As a rule of thumb: $J \geq 3$ visits per site, with detection probability
$p > 0.15$, gives reasonable identifiability for simple models. More complex
models (with many covariates or random effects) require more data.

## Using checkIdentifiability()

INLAocc's `checkIdentifiability()` function screens for common identifiability
problems. It works in two modes: pre-fit (data only) and post-fit (fitted
model), catching progressively more issues.

### Pre-fit checks (data only)

Pre-fit checks examine the detection history matrix and covariates without
fitting any model. They are fast and should be run before any analysis.

```{r prefit-good}
library(INLAocc)

sim <- simulate_occu(N = 200, J = 4,
                     beta_occ = c(0.5, -0.8),
                     beta_det = c(0, -0.5),
                     n_occ_covs = 1, n_det_covs = 1,
                     seed = 42)

issues <- checkIdentifiability(sim$data)
issues
```

For well-identified data, the output should confirm no issues. The summary statistics provide a quick sanity check: naive occupancy
(proportion of sites with at least one detection), naive detection (proportion
of visits with detections at occupied sites), visits per site, and the number
of unique detection histories.

### Problematic data

Now simulate data with known identifiability issues: few sites, few visits,
and very low detection:

```{r prefit-bad}
sim_bad <- simulate_occu(N = 50, J = 2,
                         beta_occ = c(0.5, -0.8),
                         beta_det = c(-2, -0.5),  # very low detection
                         n_occ_covs = 1, n_det_covs = 1,
                         seed = 42)

issues <- checkIdentifiability(sim_bad$data)
issues
```

Each issue has a concrete interpretation:

- **Visits $\leq$ 2**: With only two visits, the detection/occupancy trade-off
  is severely underdetermined. The model has very little information to separate
  the two processes, especially when covariates are included.
- **Low detection**: When species are rarely detected ($p < 0.10$), almost all
  sites have all-zero histories. The model cannot reliably distinguish low
  occupancy from low detection because both produce the same data pattern.
- **Over-parameterization**: Five parameters for 50 sites is aggressive. The
  rule of thumb is at least 10 sites per parameter (20 is better). With too
  many parameters relative to data, individual coefficients are poorly
  determined.

### Post-fit checks

When passed a fitted model, `checkIdentifiability()` runs additional checks
that require posterior estimates:

```{r postfit}
fit <- occu(~ occ_x1, ~ det_x1, data = sim_bad$data, verbose = 0)
issues_post <- checkIdentifiability(fit)
issues_post
```

Post-fit checks also flag:

- **Non-convergence**: The EM algorithm did not reach the convergence criterion
  within the maximum number of iterations.
- **Large posterior SDs**: A posterior standard deviation more than 10 times the
  absolute posterior mean indicates the data provides almost no information
  about that parameter.
- **Boundary estimates**: Occupancy or detection estimates at 0 or 1 for a
  large fraction of sites suggest the model is hitting the boundary of the
  parameter space.
- **High correlation between processes**: Strong posterior correlation between
  occupancy and detection parameters indicates the model cannot separate the
  two processes.

## Covariate confounding

The most common identifiability issue in applied work is including the same
covariate in both the occupancy and detection submodels. This is tempting when
a variable (e.g., habitat quality, elevation) plausibly affects both processes,
but it creates a fundamental identifiability problem.

```{r confounding}
# Share a covariate between both processes
sim$data$det.covs$occ_x1 <- matrix(rep(sim$data$occ.covs$occ_x1, 4), nrow = 200, ncol = 4)

# Same variable in both submodels
fit_confound <- occu(~ occ_x1, ~ occ_x1, data = sim$data, verbose = 0)
checkIdentifiability(fit_confound)
```

Why this is problematic: if a covariate affects both occupancy and detection,
the model cannot determine which process it operates through. The estimated
effect gets split arbitrarily between the two submodels, and the split depends
on the data realization, the starting values, and the prior, not on the
underlying biology.

**Fixes:**

- Include the covariate in the process where it has the strongest theoretical
  justification. Elevation affects where a species lives (occupancy) more
  directly than whether a surveyor detects it (detection).
- If there is genuine reason to include a variable in both, use a different
  functional form in each: linear in one process, quadratic in the other, or
  a categorical version in one and continuous in the other.
- Random effects with the same grouping variable in both processes cause the
  same problem. `checkIdentifiability()` flags this as well.

## Simulation-based stress testing

Automated checks catch common problems, but they cannot cover every scenario.
Simulation-based stress testing provides a deeper assessment: simulate data
from known parameters, fit the model, and check whether the true values are
recovered. If recovery fails under conditions matching your real data, the
model is not identifiable for your application.

### Basic parameter recovery

The simplest stress test simulates multiple datasets from the same parameters
and checks whether the estimates are centered on the truth:

```{r recovery}
n_sims <- 20
results <- data.frame()

for (i in 1:n_sims) {
  sim_i <- simulate_occu(N = 200, J = 4,
                         beta_occ = c(0.5, -0.8),
                         beta_det = c(0, -0.5),
                         n_occ_covs = 1, n_det_covs = 1,
                         seed = i)

  fit_i <- occu(~ occ_x1, ~ det_x1, data = sim_i$data, verbose = 0)
  cf <- coef(fit_i)

  results <- rbind(results, data.frame(
    sim = i,
    beta0_occ = cf$occ[1],
    beta1_occ = cf$occ[2],
    alpha0_det = cf$det[1],
    alpha1_det = cf$det[2]
  ))
}

# Estimates should be centered on truth
colMeans(results[, -1])
```

Good recovery: the means across simulations should be close to the true values
(0.5, -0.8, 0, -0.5). Substantial bias would indicate an identifiability
problem under these conditions.

You can also check coverage: the proportion of simulations where the 95%
credible interval contains the true value. Nominal coverage should be close
to 0.95.

### Varying sample size

A natural question is: how many sites do I need? Simulate across a range of
sample sizes to find the threshold where estimation breaks down:

```{r vary-N}
for (N in c(50, 100, 200, 500)) {
  sim_i <- simulate_occu(N = N, J = 4,
                         beta_occ = c(0.5, -0.8),
                         beta_det = c(0, -0.5),
                         n_occ_covs = 1, n_det_covs = 1,
                         seed = 42)

  fit_i <- occu(~ occ_x1, ~ det_x1, data = sim_i$data, verbose = 0)
  est <- coef(fit_i)$occ[2]
  se <- fit_i$occ_fit$summary.fixed$sd[2]

  cat(sprintf("N = %3d: beta1 = %6.2f (SE = %.2f, true = -0.80)\n",
              N, est, se))
}
```

With few sites the estimate is biased and the standard error is large. By 200
sites, recovery is good. This tells you the minimum sample size for reliable
estimation under your specific model structure.

### Varying detection probability

Detection probability is the single most important factor for occupancy model
identifiability. Test how your model behaves across a range of detection
levels:

```{r vary-det}
for (det_int in c(-2, -1, 0, 1)) {
  sim_i <- simulate_occu(N = 200, J = 4,
                         beta_occ = c(0.5, -0.8),
                         beta_det = c(det_int, -0.5),
                         n_occ_covs = 1, n_det_covs = 1,
                         seed = 42)

  fit_i <- occu(~ occ_x1, ~ det_x1, data = sim_i$data, verbose = 0)
  naive_p <- mean(sim_i$data$y, na.rm = TRUE)

  cat(sprintf("det_int = %2d (naive p = %.2f): beta1 = %6.2f (true = -0.80)\n",
              det_int, naive_p, coef(fit_i)$occ[2]))
}
```

The pattern should be clear: low detection produces substantial
bias in the occupancy coefficient. Above $p \approx 0.20$, recovery is
adequate. This reveals the detection threshold below which your model breaks
down.

### Varying number of visits

The same approach works for the number of visits:

```{r vary-J}
for (J in c(2, 3, 4, 6, 10)) {
  sim_i <- simulate_occu(N = 200, J = J,
                         beta_occ = c(0.5, -0.8),
                         beta_det = c(0, -0.5),
                         n_occ_covs = 1, n_det_covs = 1,
                         seed = 42)

  fit_i <- occu(~ occ_x1, ~ det_x1, data = sim_i$data, verbose = 0)
  est <- coef(fit_i)$occ[2]
  se <- fit_i$occ_fit$summary.fixed$sd[2]

  cat(sprintf("J = %2d: beta1 = %6.2f (SE = %.2f)\n", J, est, se))
}
```

Diminishing returns set in quickly: going from 2 to 4 visits is far more
valuable than going from 4 to 10.

## When identifiability fails: symptoms and fixes

The table below summarizes common symptoms of identifiability failure, their
likely causes, and practical fixes.

| Symptom | Likely cause | Fix |
|:--------|:-------------|:----|
| Posterior SD >> posterior mean | Parameter unidentifiable | Reduce model complexity |
| Coefficients flip sign across runs | Flat likelihood surface | More data or fewer parameters |
| EM fails to converge | Likelihood too flat | Increase visits, simplify model |
| $\hat{\psi} \approx 1$ for all sites | Detection absorbing all variation | Check detection model specification |
| $\hat{p} \approx 0$ for all sites | Occupancy absorbing all variation | Ensure $J \geq 3$ |
| Same covariate in both processes | Confounding | Move to one process |

A useful diagnostic when you suspect identifiability problems is to fit
progressively simpler models: drop covariates from detection, then from
occupancy, until the remaining parameters are stable. The point where
parameters start shifting substantially as you add complexity marks the
boundary of what your data can support.

## Design recommendations

Based on identifiability theory and practical experience:

- **Minimum 3 visits per site**, ideally 4--5. With $J = 2$, only the simplest
  models (intercept-only or single covariate) are identifiable. More visits
  always help, but with diminishing returns beyond $J = 5$.
- **Detection probability above 0.15.** Below this threshold, occupancy and
  detection are hard to separate without informative priors. If your study
  species has very low detection, consider increasing survey effort per visit
  (longer transects, more traps) rather than adding more visits.
- **At least 10 sites per parameter** in the occupancy model; 20 is better.
  Count intercepts and random effect variance components as parameters.
- **Avoid the same covariate in both processes** unless you have strong
  theoretical justification and adequate data. If you must include a shared
  variable, use different functional forms (linear vs. quadratic) or check
  that the simulation recovery test above succeeds with your specific design.
- **Scale covariates.** Unscaled covariates with wildly different magnitudes
  create numerical identifiability issues even when the model is theoretically
  identifiable. Center and scale continuous covariates to unit variance.
- **Start simple.** Fit the simplest defensible model first (intercept-only for
  detection, single covariate for occupancy). Add complexity only when
  diagnostics or WAIC comparisons justify it. Each added parameter requires
  the data to support one more degree of freedom.

## Interaction with priors

INLA uses penalized complexity (PC) priors that provide implicit
regularization. This regularization can mask identifiability issues in a way
that is easy to miss.

A weakly identified parameter with a strong prior will produce a posterior that
looks well-behaved: finite mean, reasonable standard deviation, no boundary
issues. But the posterior is driven by the prior, not by the data. The model
appears to work, but the estimates reflect your assumptions rather than your
observations.

To check whether the data is actually informing a parameter: compare the
posterior to the prior. If they are nearly identical in location and spread,
the data provides little information about that parameter. INLAocc's
`checkIdentifiability()` post-fit mode flags cases where the posterior
standard deviation is close to the prior standard deviation.

This concern is less acute with INLAocc's default weakly informative priors,
which are intentionally vague. But it becomes important when using informative
priors to "fix" convergence issues. If a model only converges with a tight
prior on a parameter, that parameter is not identified by the data. The
prior is doing the work. In that case, the correct response is to simplify
the model, not to tighten the prior.

A practical test: fit the model twice with different priors (e.g., prior
SD = 1 vs. prior SD = 5 on the same coefficient). If the posterior changes
substantially, the data is not constraining that parameter.

## Spatial and factor model identifiability

The identifiability challenges described above (the occupancy--detection
trade-off, covariate confounding, the role of priors) all become harder
when a spatial random effect is added to the model. Spatial occupancy models
introduce additional parameters (the spatial range $\rho$ and marginal
standard deviation $\sigma$ of the Gaussian field) that must be estimated
from the same binary detection/non-detection data. The binary nature of the
observations is the core difficulty: continuous Gaussian data would provide
far more information per observation about the correlation structure, but
presence/absence data collapses the rich information in spatial fields down
to zeros and ones.

### Weak identifiability of spatial range and variance

The practical correlation range $\rho$ and the marginal variance $\sigma^2$
of the Matern field are jointly weakly identifiable in occupancy models. The
reason is intuitive: a spatial field with short range and high variance can
produce similar patterns in occupancy probability as one with long range and
low variance, once the field passes through the logit link function and is
then observed as binary data. The likelihood surface over $(\rho, \sigma^2)$
is typically a broad ridge rather than a sharp peak.

This weak identifiability has concrete consequences. Without regularization,
the posterior for $\rho$ can be extremely diffuse, sometimes spanning orders
of magnitude. The marginal variance $\sigma^2$ may inflate to absorb
variation that belongs in the fixed effects, or collapse toward zero if the
prior pulls it there. The fixed-effect coefficients $\boldsymbol{\beta}$ may
shift as the spatial field takes over (or gives up) explanatory power.

INLA's penalized complexity (PC) priors provide essential regularization for
these parameters. The PC prior for the range is specified as P(range < $r_0$)
= $p$, and for the standard deviation as P($\sigma$ > $s_0$) = $p$. These
priors encode a "base model" principle: the prior shrinks toward no spatial
effect (infinite range, zero variance), requiring the data to justify spatial
complexity. In contrast, MCMC-based spatial occupancy models (such as those
in spOccupancy) typically use diffuse inverse-Gamma or uniform priors on
these parameters, which provide little regularization and can lead to poor
mixing and slow convergence when the likelihood surface is flat.

The PC prior is not a fix for non-identifiability; it is a principled way
to handle weak identifiability. The posterior for $\rho$ will still be wide
when the data provide limited spatial information, but it will not
degenerate. The key diagnostic is the same as for any weakly identified
parameter: compare the posterior to the prior. If they are nearly identical,
the data is not informing the spatial parameters, and the model's spatial
predictions are driven by the prior choice rather than by the observations.

### Sign-switching and rotational invariance in factor models

Latent factor models (fitted via `occu()` with `n.factors`) have a
fundamentally different identifiability challenge: the likelihood is
invariant to sign flips and rotations of the latent factors. If
$\boldsymbol{\Lambda}$ is the species loading matrix and $\mathbf{w}$ is a
vector of latent factors, then replacing $\boldsymbol{\Lambda}$ with
$\boldsymbol{\Lambda} \mathbf{R}$ and $\mathbf{w}$ with $\mathbf{R}^{-1}
\mathbf{w}$ for any orthogonal rotation matrix $\mathbf{R}$ leaves the
likelihood unchanged. The simplest case is sign-switching: multiplying a
factor by $-1$ and flipping the corresponding column of the loading matrix
produces identical predictions.

This rotational invariance does not affect predictions. The predicted
occupancy probability $\hat{\psi}_i$ and the species correlation matrix
implied by the factor model are invariant to the rotation. What it does
affect is interpretation: the individual factor loadings and the latent
factor values are not uniquely determined. Across MCMC samples (or across
multiple fitting runs), a factor may flip sign or blend with other factors,
making it impossible to interpret a single factor as "the temperature axis"
or "the moisture axis" without post-hoc identification constraints.

INLA-based factor models sidestep the worst manifestations of this problem
because INLA does not sample from the joint posterior. The Laplace
approximation produces a single posterior mode, and the identifiability
constraint (upper-triangular loading matrix with positive diagonal entries)
pins down a unique solution. However, the underlying weakness remains: if
the data do not strongly separate the factors, the posterior for the loading
matrix will be diffuse, the constraint will do more work than the data, and
the factor interpretation will be fragile.

Practical advice for factor models: (1) compare models with different
numbers of factors using WAIC via `waicOccu()` to determine how many factors
the data can support, (2) do not over-interpret individual factor loadings
unless they are consistent across subsets of the data, and (3) focus on the
derived quantities (species correlations, predicted occupancy) that are
invariant to rotations.

### Does spatial structure actually help?

Adding a spatial random effect increases model complexity substantially. The
SPDE mesh may add hundreds of latent parameters, and the two hyperparameters
($\rho$, $\sigma$) must be estimated. Before trusting a spatial model, you
should check whether the added complexity is justified.

The first check is the residual spatial autocorrelation test. Fit the model
without the spatial effect and run `moranI()` on the result. If Moran's I
is not significantly positive, there is no evidence of spatial
autocorrelation in the residuals, and the spatial term is unnecessary. The
second check is model comparison: fit both the spatial and non-spatial
models and compare them via `waicOccu()`. A lower WAIC for the spatial
model indicates that the spatial field improves predictive performance; a
similar or higher WAIC means the spatial field is not paying for its
complexity.

A subtler issue is spatial confounding: the spatial field may absorb the
effect of a spatially structured covariate, attenuating the estimated
covariate coefficient toward zero. This happens when a covariate like
elevation is itself spatially smooth. The spatial field and the covariate
compete for the same variation, and the flexible random field tends to win.
The symptom is a covariate coefficient that is significant in the
non-spatial model but shrinks toward zero in the spatial model. This is not
always a problem (the spatial model may be giving a more honest estimate by
accounting for unmeasured confounders), but it requires careful thought
about whether the covariate effect or the spatial field is the quantity of
scientific interest.

### Worked example: spatial identifiability stress test

The following simulation demonstrates when spatial parameters are
identifiable and when they are not. We simulate data with a known spatial
range ($\rho = 0.3$ on a unit square), then recover it under varying sample
sizes and spatial designs.

```{r spatial-recovery-setup}
library(INLAocc)

# Helper: fit spatial model and extract range estimate
recover_range <- function(N, spatial_range = 0.3, seed = 1,
                          clustered = FALSE) {
  set.seed(seed)

  sim <- simulate_occu(
    N = N, J = 4,
    n_occ_covs = 1, n_det_covs = 1,
    beta_occ = c(0.5, -0.8),
    beta_det = c(0, -0.5),
    spatial_range = spatial_range, spatial_var = 1,
    seed = seed
  )

  # Replace coordinates with clustered design if requested
  if (clustered) {
    n_clusters <- 5
    centers <- matrix(runif(n_clusters * 2, 0.2, 0.8), n_clusters, 2)
    cluster_id <- sample(n_clusters, N, replace = TRUE)
    new_coords <- centers[cluster_id, ] + matrix(rnorm(N * 2, sd = 0.02), N, 2)
    sim$data$coords <- new_coords
  }

  fit <- occu(~ occ_x1, ~ det_x1, data = sim$data,
              spatial = sim$data$coords,
              spde.args = list(prior.range = c(0.3, 0.5),
                               prior.sigma = c(1, 0.05)),
              verbose = 0)

  # Extract spatial hyperparameters
  hyp <- fit$occ_fit$summary.hyperpar
  range_row <- grep("[Rr]ange", rownames(hyp))
  sigma_row <- grep("[Ss]tdev|[Ss]igma", rownames(hyp))

  list(
    N = N,
    range_est  = if (length(range_row)) hyp$mean[range_row[1]] else NA,
    range_sd   = if (length(range_row)) hyp$sd[range_row[1]] else NA,
    sigma_est  = if (length(sigma_row)) hyp$mean[sigma_row[1]] else NA,
    sigma_sd   = if (length(sigma_row)) hyp$sd[sigma_row[1]] else NA,
    beta1_est  = coef(fit)$occ[2],
    beta1_se   = fit$occ_fit$summary.fixed$sd[2]
  )
}
```

#### Effect of sample size on spatial parameter recovery

With sites spread uniformly across the domain, more sites means more pairs
at each distance lag, which is what informs the range parameter.

```{r spatial-recovery-N}
cat(sprintf("%-6s  %-10s  %-10s  %-10s  %-10s\n",
            "N", "range_est", "range_SD", "sigma_est", "beta1"))
cat(strrep("-", 56), "\n")

for (N in c(50, 100, 200, 400)) {
  res <- recover_range(N, seed = 42)
  cat(sprintf("%-6d  %-10.3f  %-10.3f  %-10.3f  %-10.3f\n",
              res$N, res$range_est, res$range_sd, res$sigma_est, res$beta1_est))
}
```

With 50 sites, the range estimate has a wide posterior and may differ
substantially from the truth ($\rho = 0.3$). By 200 sites, the estimate
stabilises and the posterior SD shrinks. The covariate coefficient
(`beta1`) is reasonably well recovered throughout because the spatial field
absorbs the spatial structure rather than the covariate effect, but only
if the covariate is not itself spatially smooth.

#### Clustered vs. spread designs

Spatial identifiability depends not just on the number of sites but on their
spatial arrangement. Clustered designs (where sites are grouped in a few
locations, as is common in field surveys along roads or trails) provide
limited information about the range parameter because most site pairs are
either very close (within-cluster) or very far (between-cluster), with few
pairs at intermediate distances.

```{r spatial-recovery-clustered}
cat(sprintf("%-10s  %-6s  %-10s  %-10s\n",
            "Design", "N", "range_est", "range_SD"))
cat(strrep("-", 42), "\n")

for (N in c(100, 200)) {
  res_spread    <- recover_range(N, seed = 42, clustered = FALSE)
  res_clustered <- recover_range(N, seed = 42, clustered = TRUE)

  cat(sprintf("%-10s  %-6d  %-10.3f  %-10.3f\n",
              "spread", N, res_spread$range_est, res_spread$range_sd))
  cat(sprintf("%-10s  %-6d  %-10.3f  %-10.3f\n",
              "clustered", N, res_clustered$range_est, res_clustered$range_sd))
}
```

The clustered design consistently produces wider posteriors for the range
parameter. When sites are concentrated in a few clusters, the spatial field
is effectively informed only at two distance scales (near-zero and the
between-cluster distance), leaving the intermediate range poorly
constrained. This is a common problem in applied ecology, where logistical
constraints force clustered sampling. The practical implication is that
spatial models fit to clustered data should be interpreted with extra
caution: the spatial predictions may look reasonable, but the estimated
range and variance are driven more by the prior than by the data.

#### Diagnosing spatial identifiability in practice

For a fitted spatial model, three diagnostics together give a clear picture
of spatial identifiability:

1. **Prior--posterior comparison for range and variance.** Extract the
   hyperparameter summary from `fit$occ_fit$summary.hyperpar` and compare
   the posterior mean and SD to the PC prior specification. If the posterior
   is nearly as wide as the prior, the data is not informing the spatial
   parameters.

2. **WAIC comparison.** Fit the model with and without the spatial effect
   and compare via `waicOccu()`. If the spatial model does not improve WAIC,
   the spatial field is not contributing to predictive performance and may be
   absorbing noise.

3. **Coefficient stability.** Compare the fixed-effect estimates between
   the spatial and non-spatial models. Large shifts suggest spatial
   confounding: the spatial field is absorbing variation that the non-spatial
   model attributed to covariates. This does not necessarily mean the spatial
   model is wrong (it may be giving a less confounded estimate), but the
   difference should be understood and reported.

```{r spatial-diagnostics-example}
sim <- simulate_occu(N = 200, J = 4,
                     n_occ_covs = 1, n_det_covs = 1,
                     beta_occ = c(0.5, -0.8),
                     beta_det = c(0, -0.5),
                     spatial_range = 0.3, seed = 42)

fit_ns <- occu(~ occ_x1, ~ det_x1, data = sim$data, verbose = 0)
fit_sp <- occu(~ occ_x1, ~ det_x1, data = sim$data,
               spatial = sim$data$coords, verbose = 0)

# 1. Moran's I on non-spatial residuals
moranI(fit_ns)

# 2. WAIC comparison
waic_ns <- waicOccu(fit_ns)
waic_sp <- waicOccu(fit_sp)
cat(sprintf("WAIC non-spatial: %.1f\nWAIC spatial:     %.1f\n",
            waic_ns$WAIC, waic_sp$WAIC))

# 3. Coefficient comparison
cat(sprintf("\nbeta1 non-spatial: %.3f (SE %.3f)\n",
            coef(fit_ns)$occ[2], fit_ns$occ_fit$summary.fixed$sd[2]))
cat(sprintf("beta1 spatial:     %.3f (SE %.3f)\n",
            coef(fit_sp)$occ[2], fit_sp$occ_fit$summary.fixed$sd[2]))
cat(sprintf("True beta1:        -0.800\n"))

# 4. Spatial hyperparameters (compare to prior)
if (!is.null(fit_sp$occ_fit$summary.hyperpar)) {
  cat("\nSpatial hyperparameters:\n")
  print(round(fit_sp$occ_fit$summary.hyperpar, 3))
}
```

When all three diagnostics align (significant Moran's I in the
non-spatial model, lower WAIC for the spatial model, and stable covariate
coefficients), the spatial model is well justified and the spatial
parameters are at least moderately identifiable. When the diagnostics
conflict (e.g., Moran's I is significant but WAIC favours the simpler
model), the spatial structure may be too weak or the sample size too small
for the model to reliably separate the spatial field from the fixed effects.

## Summary

Identifiability is not a binary property; it exists on a continuum from
well-identified (data strongly constrains parameters) to completely
unidentifiable (data provides no information). Most real datasets fall
somewhere in between, and the goal is to match model complexity to the
information content of the data.

The recommended workflow:

1. **Before fitting**: run `checkIdentifiability()` on the data to catch
   obvious issues (too few visits, low detection, covariate confounding).
2. **After fitting**: run `checkIdentifiability()` on the fitted model to check
   for large posterior SDs, boundary estimates, and non-convergence.
3. **For critical analyses**: run the simulation-based stress test with
   parameters and sample sizes matching your real data to verify that your
   specific model structure is identifiable under your specific conditions.

If identifiability checks reveal problems, simplify the model rather than
adding stronger priors. A simple model that is well-identified produces more
reliable inference than a complex model that is not.
