---
title: "Integrated Occupancy Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrated Occupancy Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

Many species are monitored simultaneously by multiple programs that differ in
protocol, effort, and spatial coverage. A state wildlife agency may run
structured point-count surveys at a fixed set of sites, while a network of
camera traps operates year-round at a partially overlapping set of locations,
and citizen-science platforms like eBird accumulate opportunistic checklists
across the broader landscape. Each data source captures the same underlying
quantity --- whether the species occupies a site --- but filters it through a
different detection process. Point counts detect vocalisations during brief
windows; cameras detect physical presence over extended periods; eBird
checklists reflect variable observer skill and effort. Analysing any single
source in isolation discards the information contained in the others.

Integrated occupancy models solve this problem by combining all sources into a
single inference. The true occupancy state
$z_i$ at site $i$ is shared across every data source. What differs is how
likely each source is to detect the species when it is present. By modelling
each source's detection process separately while linking them through a common
occupancy layer, the model borrows strength across programs. Sites surveyed by
multiple sources are particularly informative, because detections (or
non-detections) from one source constrain what the other source's data imply
about occupancy.

INLAocc fits integrated occupancy models through the `integrated = TRUE`
argument in `occu()`. The EM algorithm iterates over a shared occupancy
posterior while fitting source-specific detection models via INLA, exactly
paralleling the single-source workflow. This vignette covers the model
formulation, the data structure required, simulation, model fitting, and
extensions to spatial and multi-species integrated models. It assumes
familiarity with single-species models; see `vignette("quickstart")` for an
introduction to `occu()` and the INLAocc data format.


## The integrated occupancy model

For $D$ data sources observing the same set of sites, the integrated occupancy
model has a shared occupancy layer and source-specific detection layers.

**Occupancy.** At each of $N$ unique sites, the latent occupancy state is:

$$z_i \sim \text{Bernoulli}(\psi_i), \quad \text{logit}(\psi_i) = \mathbf{x}_i^\top \boldsymbol{\beta}$$

where $\mathbf{x}_i$ is a vector of site-level covariates and $\boldsymbol{\beta}$
the occupancy coefficients. This is identical to the standard single-source
model --- every source shares the same $\psi_i$.

**Detection.** Each source $d = 1, \ldots, D$ has its own detection process.
Source $d$ surveys a subset of sites (not necessarily all $N$), and at each
surveyed site $i$, it makes $J_d$ visits. The observation model is:

$$y_{ijd} \mid z_i \sim \text{Bernoulli}(z_i \cdot p_{ijd}), \quad \text{logit}(p_{ijd}) = \mathbf{w}_{ijd}^\top \boldsymbol{\alpha}_d$$

where $\mathbf{w}_{ijd}$ are visit-level covariates specific to source $d$,
and $\boldsymbol{\alpha}_d$ is a source-specific detection coefficient vector.
Each source has its own intercept, its own covariates, and its own regression
coefficients --- reflecting the fact that, say, a camera trap and a point count
have fundamentally different detection mechanisms.

**E-step.** The EM algorithm updates the posterior probability that each site is
occupied, conditional on all observed data. For a site surveyed by sources
$\mathcal{D}_i \subseteq \{1, \ldots, D\}$:

$$P(z_i = 1 \mid \mathbf{y}) = \frac{\psi_i \prod_{d \in \mathcal{D}_i} \prod_{j=1}^{J_d} p_{ijd}^{y_{ijd}} (1 - p_{ijd})^{1 - y_{ijd}}}{\psi_i \prod_{d \in \mathcal{D}_i} \prod_{j=1}^{J_d} p_{ijd}^{y_{ijd}} (1 - p_{ijd})^{1 - y_{ijd}} + (1 - \psi_i) \, \mathbb{I}[\mathbf{y}_{i \cdot} = \mathbf{0}]}$$

The product runs over all sources that surveyed site $i$ and all visits within
each source. For sites with at least one detection from any source, the
posterior is 1 (the species is present). For sites with zero
detections across all sources, the posterior depends on $\psi_i$ and the
detection probabilities from every source. Each additional source of
non-detection further lowers the posterior
occupancy probability, giving more precise estimates of true absence than any
single source alone.


## Data structure

Integrated models require data in a specific list format. The key difference
from standard models is that detection data (`y`) and detection covariates
(`det.covs`) are lists of length $D$, one element per source, and a `sites`
mapping tells the model which rows of the shared occupancy covariates correspond
to each source's sites.

```{r data-structure, eval = FALSE}
data <- list(
  y = list(
    point_counts = matrix(..., nrow = 80, ncol = 4),    # 80 sites, 4 visits
    cameras      = matrix(..., nrow = 60, ncol = 10)     # 60 sites, 10 occasions
  ),
  occ.covs = data.frame(
    elevation = rnorm(110),
    forest    = runif(110)
  ),
  det.covs = list(
    list(date = matrix(rnorm(80 * 4), 80, 4)),           # source 1 covariates
    list(effort = matrix(rnorm(60 * 10), 60, 10))        # source 2 covariates
  ),
  sites = list(1:80, 51:110),
  coords = matrix(runif(110 * 2), ncol = 2)              # all unique sites
)
```

Each component plays a specific role:

- **`y`**: a list of $D$ detection matrices. Source $d$ has an $N_d \times J_d$
  matrix where rows are sites and columns are visits. Sources can have different
  numbers of sites and visits --- the only requirement is that each matrix has
  rows matching the corresponding entry in `sites`.

- **`occ.covs`**: a data.frame with one row per unique site ($N = 110$ in this
  example). All occupancy covariates live here, shared across sources.

- **`det.covs`**: a list of $D$ lists. Each inner list contains named matrices
  of visit-level detection covariates for that source. Source 1 has a `date`
  matrix with 80 rows and 4 columns; source 2 has an `effort` matrix with 60
  rows and 10 columns.

- **`sites`**: a list of $D$ integer vectors. Each vector maps the rows of the
  corresponding `y` matrix to rows of `occ.covs`. In this example, source 1
  covers `occ.covs` rows 1--80 and source 2 covers rows 51--110. Sites 51--80
  appear in both vectors --- these 30 overlap sites are surveyed by both
  programs.

- **`coords`**: an $N \times 2$ matrix of coordinates for all unique sites.
  Required only for spatial models.

The overlap sites are particularly valuable. They are the only sites where the
model directly observes the same occupancy state through two different detection
lenses, which is what allows it to calibrate the detection parameters across
sources.


## Simulating integrated data

INLAocc provides `simIntOcc()` to generate integrated occupancy datasets with
known parameters, which is essential for validating model performance and
understanding the data structure.

```{r simulate}
library(INLAocc)

sim <- simIntOcc(
  N_total  = 150,
  n_data   = 2,
  J        = c(4, 3),
  n_shared = 20,
  beta_occ = c(0.5, -0.8),
  beta_det = list(c(0, -0.5), c(-0.3, 0.2)),
  seed     = 42
)

str(sim$data)
```

The simulator creates a dataset with 150 unique sites. Source 1 surveys 85
sites with 4 visits each, using occupancy coefficients $\boldsymbol{\beta} =
(0.5, -0.8)$ and detection coefficients $\boldsymbol{\alpha}_1 = (0, -0.5)$.
Source 2 surveys a different set of 85 sites with 3 visits, with detection
coefficients $\boldsymbol{\alpha}_2 = (-0.3, 0.2)$. Twenty sites appear in
both sources, providing the overlap needed to calibrate detection across
programs. The occupancy formula is shared --- both sources inform the same
$\psi_i$.


## Fitting the integrated model

Fitting an integrated model uses the same `occu()` interface, with
`integrated = TRUE`:

```{r fit-basic}
fit <- occu(~ occ_x1, ~ det_x1, data = sim$data, integrated = TRUE, verbose = 0)
summary(fit)
```

Several things to note in the output:

- The occupancy estimates integrate information from both sources. The estimated
  coefficients should be close to the true values (0.5 and -0.8), reflecting the
  combined power of 150 sites observed through two detection processes.

- Source 1 and source 2 should have visibly different detection characteristics,
  matching the simulation parameters.

- The 20 overlap sites are doing the heavy lifting for detection calibration.
  At those sites, the model sees the same $z_i$ filtered through two different
  detection lenses, which pins down the relative detection rates.


## Source-specific detection formulas

When a single detection formula is provided, it is applied to all sources. But
sources often have different covariates --- point counts have observer ID and
wind speed, while cameras have trigger sensitivity and battery voltage. In
INLAocc, you can pass a list of formulas, one per source:

```{r source-formulas, eval = FALSE}
fit2 <- occu(
  ~ occ_x1,
  list(~ date + duration, ~ effort),
  data = sim$data,
  integrated = TRUE,
  verbose = 0
)
summary(fit2)
```

The first formula applies to source 1, the second to source 2. The occupancy
formula is always shared --- there is no mechanism for source-specific occupancy
effects, because occupancy is a property of the site, not of the observation
method.

This is the recommended approach when sources genuinely differ in what drives
detection. Start with a shared detection formula to establish a baseline, then
try source-specific formulas and compare via WAIC (see below). If the
source-specific model has lower WAIC despite the additional parameters, the
detection processes are sufficiently different to warrant separate treatment.


## Spatial integrated models

Integrated models extend naturally to spatial settings. Adding a spatial
random field to the occupancy layer accounts for unmeasured environmental
variation that is spatially structured:

```{r fit-spatial}
fit_sp <- occu(
  ~ occ_x1, ~ det_x1,
  data    = sim$data,
  integrated = TRUE,
  spatial = sim$data$coords,
  verbose = 0
)
summary(fit_sp)
```

The spatial field is shared across all data sources --- it captures spatial
variation in occupancy that is common regardless of how the species is detected.
This is conceptually correct: spatial autocorrelation in occupancy arises from
environmental processes (habitat connectivity, dispersal limitation, unmeasured
gradients) that operate independently of the observation method.

The spatial integrated model works best
when sources have complementary spatial coverage. Structured surveys may cover a
dense grid in a core area, while citizen-science data fill in the periphery. The
shared spatial field allows information to flow between regions dominated by
different sources, smoothing the occupancy surface across the entire study area.

For details on mesh construction, PC priors for the spatial range and variance,
and occupancy mapping, see `vignette("spatial-models")`.


## Multi-species integrated models

Community-level integrated models combine multiple species and multiple data
sources under a single hierarchical framework. Species-level occupancy and
detection parameters are drawn from community distributions, and each source
retains its own detection process.

```{r fit-multispecies, eval = FALSE}
sim_ms <- simIntMsOcc(
  N_total   = 150,
  n_data    = 2,
  J         = c(4, 3),
  n_species = 8,
  n_shared  = 20,
  seed      = 100
)

fit_ms <- occu(
  ~ occ_x1, ~ det_x1,
  data         = sim_ms$data,
  multispecies = TRUE,
  integrated   = TRUE,
  verbose      = 0
)
summary(fit_ms)
```

This fits 8 species simultaneously across 2 data sources. Each species has its
own occupancy and detection parameters, but all species share the community-level
hyperparameters (mean and variance of the occupancy and detection coefficient
distributions). Rare species that appear in only one data source borrow strength
from the community mean and from the overlap sites observed across sources.

Multi-species integrated models are particularly useful for biodiversity
monitoring programs that combine structured surveys (few species observed
reliably) with broad-coverage citizen science (many species observed
inconsistently). The hierarchical structure stabilises estimates for the long
tail of rarely detected species.


## WAIC for integrated models

Model comparison via WAIC works as expected:

```{r waic}
w <- waicOccu(fit)
w
```

The WAIC is decomposed by component: one entry for the occupancy layer and one
for each detection source. This decomposition reveals which part of the model
drives complexity. Typically, the occupancy component contributes the most to
both the effective number of parameters ($p_D$) and the overall WAIC, while the
source with fewer visits and sites is the simplest component.

Use the total WAIC to compare competing integrated models --- for instance, a
model with shared versus source-specific detection formulas, or a non-spatial
versus spatial occupancy layer. The per-component breakdown helps diagnose
where a model is over- or under-fitting.


## When to use integrated models

Integrated models are appropriate when multiple survey programs cover the same
species in overlapping regions with different observation protocols:

- **Structured + citizen science.** Structured surveys provide high detection
  reliability at a small number of sites, while citizen-science platforms like
  eBird or iNaturalist offer broad spatial coverage with variable and often
  lower detection probability. Combining them yields better occupancy estimates
  than either source alone.

- **Multiple sensor types.** Camera traps, acoustic recorders, and visual
  surveys detect different behaviours (movement, vocalisation, presence) and
  operate on different temporal scales. An integrated model acknowledges these
  differences explicitly.

- **Historical + modern data.** Older surveys using different methods
  (e.g., mist netting vs. point counts) can be combined with contemporary data,
  provided the occupancy state can be treated as shared within a season.

- **Multi-scale monitoring.** Regional monitoring programs with sparse but
  standardised coverage can be integrated with local intensive studies to
  improve estimates at both scales.


### When NOT to use integrated models

- **No spatial overlap.** If two sources cover completely disjoint regions with
  no shared sites, the model cannot calibrate detection across sources. Fit
  them separately and combine predictions post hoc.

- **One source has no detections.** A source with zero detections across all
  sites and visits contributes no information about detection and only adds
  parameters. Drop it.

- **Identical detection processes.** If two sources use the same protocol,
  observers, and equipment, they have the same detection process and should
  simply be pooled into a single detection matrix with more visits. Fitting
  them as separate sources wastes degrees of freedom on estimating differences
  that do not exist.


## Practical guidance

**Overlap is critical.** Sites surveyed by multiple sources are the anchor
points that let the model learn how detection differs across programs. With zero
overlap, the model must extrapolate detection calibration entirely from the
occupancy structure, which is fragile. Aim for at least 10--20 shared sites.
More overlap is always better, but even a modest number of shared sites
substantially improves identifiability.

**Unequal contributions.** Sources with more visits or higher per-visit
detection probability contribute more to the occupancy estimates. This is
desirable --- it means the model automatically up-weights the more informative
source rather than treating all sources equally.

**Start simple.** Begin with a shared detection formula across sources. This
baseline model has fewer parameters and is easier to diagnose. Then try
source-specific detection formulas and compare via WAIC. Only add
source-specific terms when the data support them.

**Scaling considerations.** The total number of site-visit combinations grows
as $\sum_d N_d \times J_d$, which can become large with many sources. The EM
algorithm handles this efficiently because each detection source is fit as a
separate INLA call, but memory requirements scale with the total data volume.
For very large datasets, consider reducing the number of visits (e.g.,
collapsing weekly camera occasions into monthly summaries) before fitting.

**Convergence.** Integrated models typically require more EM iterations than
single-source models because the E-step must reconcile information from
multiple detection processes. Monitor convergence using the EM trace
(available via `plot(fit, type = "convergence")`) and increase `max_iter` if
needed. Warm-starting from a simpler model (e.g., the non-spatial version)
can accelerate convergence for spatial integrated models.
