---
title: "Multi-Species & Community Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Species & Community Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

Ecological communities are assemblages of species that share a landscape.
Understanding how multiple species respond to the same environmental gradients,
compete for resources, or partition habitat is central to community ecology.
Single-species models treat each species in isolation, discarding the information
contained in co-occurrence patterns and shared environmental filtering. When the
goal is to understand community structure rather than a single focal species, a
joint modelling framework is required.

Multi-species occupancy models address this by fitting all species
simultaneously under a hierarchical framework. Species-level parameters are
drawn from a shared community distribution, so rare species with few detections
borrow strength from the data-rich common species. This shrinkage toward the
community mean stabilises estimates that would otherwise be unreliable, while
still allowing abundant species to express idiosyncratic responses. Beyond
improved estimation, multi-species models yield community-level summaries ---
average covariate effects, species richness corrected for imperfect detection,
and latent environmental drivers that no single measured covariate captures.

This vignette covers community occupancy models, parallel fitting across
species, latent factor models for species correlations, spatial latent factors,
joint species distribution models (JSDMs), species richness estimation, trait
analysis, and temporal multi-species models. It assumes familiarity with
single-species occupancy models; see `vignette("quickstart")` for an
introduction to `occu()` and the INLAocc data format.


## The multi-species occupancy model

For species $s = 1, \ldots, S$ at sites $i = 1, \ldots, N$ with visits
$j = 1, \ldots, J_i$:

**Ecological process (occupancy):**

$$z_{is} \sim \text{Bernoulli}(\psi_{is}), \quad \text{logit}(\psi_{is}) = \mathbf{x}_i^\top \boldsymbol{\beta}_s$$

**Observation process (detection):**

$$y_{ijs} \mid z_{is} \sim \text{Bernoulli}(z_{is} \cdot p_{ijs}), \quad \text{logit}(p_{ijs}) = \mathbf{w}_{ij}^\top \boldsymbol{\alpha}_s$$

**Community-level hierarchy:**

$$\boldsymbol{\beta}_s \sim \mathcal{N}(\boldsymbol{\mu}_\beta, \boldsymbol{\Sigma}_\beta)$$

$$\boldsymbol{\alpha}_s \sim \mathcal{N}(\boldsymbol{\mu}_\alpha, \boldsymbol{\Sigma}_\alpha)$$

The species-level regression coefficients $\boldsymbol{\beta}_s$ and
$\boldsymbol{\alpha}_s$ are drawn from shared community distributions. This is
the borrowing-strength mechanism: a species detected at only a handful of sites
has its estimates pulled toward the community mean $\boldsymbol{\mu}_\beta$,
preventing the extreme coefficient values that maximum-likelihood estimation
would produce with sparse data.


## Simulating multi-species data

INLAocc provides `simMsOcc()` for generating multi-species detection/non-detection
datasets with known parameters:

```{r sim-data}
library(INLAocc)

sim <- simMsOcc(N = 200, J = 4, n_species = 10, seed = 42)
```

The returned object contains:

- `sim$data$y` --- a 3D detection array with dimensions species $\times$ sites
  $\times$ visits.
- `sim$data$occ_x1` --- a site-level covariate shared across species.
- `sim$data$det_x1` --- a visit-level detection covariate.
- `sim$params` --- the true species-level coefficients drawn from the community
  distribution, useful for parameter recovery checks.

Each species has its own occupancy and detection coefficients, but these are
drawn from the same community-level normal distribution. The resulting dataset
contains a mix of common (high $\psi$) and rare (low $\psi$) species.


## Community occupancy model

Fitting a community model requires only the `multispecies = TRUE` flag. The
formula interface is identical to single-species models:

```{r comm-fit}
fit <- occu(~ occ_x1, ~ det_x1, data = sim$data, multispecies = TRUE, verbose = 0)
summary(fit)
```

The community-level means represent the response of an "average species" in the
assemblage. Per-species estimates show how each species deviates from this
average. Species with few detections have their estimates shrunk more strongly
toward the community mean, which prevents overfitting to sparse data.

### Posterior predictive checks

After fitting any multi-species model, verify that the fitted model can reproduce the observed data:

```{r ppc-community, eval = FALSE}
ppc <- ppcOccu(fit, fit.stat = "freeman-tukey", group = 1)
ppc$bayesian.p
```

A Bayesian p-value near 0.5 for the community model means the fitted occupancy and detection parameters, pooled across species, generate data consistent with the observations. Species-level checks can also be run by extracting individual species fits.


## Parallel fitting

Each species is fitted independently under the community model, making
parallelization embarrassingly parallel. Set the number of cores via
`options()`:

```{r parallel, eval = FALSE}
options(INLAocc.cores = 4)
fit <- occu(~ occ_x1, ~ det_x1, data = sim$data, multispecies = TRUE, verbose = 0)
```

Wall time scales approximately as $S / \text{cores} \times t_{\text{single}}$,
where $t_{\text{single}}$ is the time for one species. For assemblages with
more than 20 species, parallelization provides substantial speedups. The
community-level hyperparameters are estimated after all species fits complete.


## Species richness

Site-level species richness corrected for imperfect detection is computed from
the posterior occupancy probabilities:

```{r richness}
r <- richness(fit)
head(r)
```

Richness at site $i$ is computed as $\hat{R}_i = \sum_{s=1}^S \hat{z}_{is}$,
the sum of estimated occupancy states across species. This accounts for species
that are present but went undetected during surveys. The `n_detected` column
gives the naive count (species detected at least once), which is always less
than or equal to the model-based estimate. The difference quantifies the
detection shortfall at each site.


## Per-species analysis

Extract species-level coefficients and summaries:

```{r per-species, eval = FALSE}
coef(fit)                  # all species in one table
confint(fit, species = 3)  # CIs for species 3
plot(fit)                  # coefficient forest plot
```

The forest plot displays species-level coefficients with 95% credible intervals,
ordered by effect size. It shows
which species respond most strongly to covariates and which deviate from the
community mean. Species whose intervals do not overlap zero show strong
evidence of a covariate effect; species with wide intervals and estimates
near zero are those most influenced by shrinkage.

### Prediction at new sites

Multi-species predictions at new covariate values work the same as single-species prediction:

```{r predict-ms, eval = FALSE}
new_sites <- data.frame(occ_x1 = seq(-2, 2, by = 0.5))
preds <- predict(fit, X.0 = new_sites, type = "occupancy")
```

The result contains species-level predicted occupancy at each new site. To compute predicted richness at the new sites:

```{r predict-richness, eval = FALSE}
# Sum predicted occupancy across species at each new site
predicted_richness <- rowSums(preds$psi.0$mean)
data.frame(occ_x1 = new_sites$occ_x1, richness = round(predicted_richness, 1))
```

This shows how community richness changes along the `occ_x1` gradient, reflecting the community-level covariate effect.


## Latent factor model

The community model assumes species are independent conditional on measured
covariates. In practice, unmeasured environmental gradients and biotic
interactions induce residual species correlations. Latent factor models capture
these by augmenting the linear predictor with shared latent variables:

$$\text{logit}(\psi_{is}) = \mathbf{x}_i^\top \boldsymbol{\beta}_s + \boldsymbol{\lambda}_s^\top \mathbf{f}_i$$

where $\mathbf{f}_i \sim \mathcal{N}(\mathbf{0}, \mathbf{I}_k)$ are $k$ latent
factors at site $i$, and $\boldsymbol{\lambda}_s$ is a vector of species-specific
factor loadings. The latent factors represent unmeasured environmental axes;
the loadings describe how each species responds to them.

The loadings induce a residual species correlation structure:

$$\boldsymbol{\Sigma}_{ss'} \propto \boldsymbol{\lambda}_s^\top \boldsymbol{\lambda}_{s'}$$

Species with similar loading vectors tend to co-occur; species with opposing
loadings are negatively associated.

### Fitting

Specify the number of factors via `n.factors`:

```{r lf-fit}
fit_lf <- occu(~ occ_x1, ~ det_x1, data = sim$data,
               multispecies = TRUE, n.factors = 3, verbose = 0)
summary(fit_lf)
```

### Factor selection via WAIC

The number of latent factors is a model choice. Fit a sequence and compare
using WAIC:

```{r lf-waic, eval = FALSE}
fits <- lapply(1:5, function(k) {
  occu(~ occ_x1, ~ det_x1, data = sim$data,
       multispecies = TRUE, n.factors = k, verbose = 0)
})
sapply(fits, function(f) waicOccu(f)$waic)
```

Select the number of factors where WAIC stops improving. A rule of thumb is to
start with $k = \lfloor S / 2 \rfloor$ and search downward.

### Species correlation matrix

The estimated loadings can be used to reconstruct the residual species
correlation matrix:

```{r lf-cor, eval = FALSE}
lambda <- fit_lf$lambda
sp_cov <- lambda %*% t(lambda)
sp_cor <- cov2cor(sp_cov + diag(nrow(sp_cov)))
round(sp_cor[1:5, 1:5], 2)
```

Positive correlations indicate species pairs that co-occur more often than
expected from measured covariates alone; negative correlations suggest spatial
segregation or competitive exclusion.

### Posterior predictive check for factor model

Check whether adding latent factors improves model adequacy:

```{r ppc-factor, eval = FALSE}
ppc_comm <- ppcOccu(fit, fit.stat = "freeman-tukey", group = 1)
ppc_lf <- ppcOccu(fit_lf, fit.stat = "freeman-tukey", group = 1)
c(community = ppc_comm$bayesian.p, latent_factor = ppc_lf$bayesian.p)
```

Both p-values should be near 0.5. If the community model showed a p-value near 0 or 1 while the factor model did not, that would confirm the latent factors are capturing structure the community model misses.

### Prediction with latent factors

Prediction from a latent factor model accounts for species correlations. At a new site, the predicted occupancy of each species incorporates both the covariate effects and the latent factor structure:

```{r predict-lf, eval = FALSE}
new_sites <- data.frame(occ_x1 = c(-1, 0, 1))
preds_lf <- predict(fit_lf, X.0 = new_sites, type = "occupancy")
```

The factor loadings determine how species co-occur at the prediction sites. Species with correlated loadings will have correlated predicted occupancy probabilities, even at sites not in the original data.


## Spatial latent factor model

When species share spatially structured environmental drivers, each latent
factor can be given its own spatial random field. This captures spatially
structured species co-occurrence patterns that a non-spatial latent factor
model would attribute to global correlations:

```{r sf-fit, eval = FALSE}
sim_sp <- simMsOcc(N = 200, J = 4, n_species = 10,
                   spatial_range = 0.3, seed = 123)

fit_sf <- occu(~ occ_x1, ~ det_x1, data = sim_sp$data,
               multispecies = TRUE, n.factors = 3,
               spatial = sim_sp$data$coords, verbose = 0)
```

Each of the $k$ latent factors receives its own SPDE spatial field, so the
model can represent species associations that vary across the study region.
This is the multi-species analogue of the single-species spatial model described
in `vignette("spatial-models")`.


## Joint Species Distribution Model (JSDM)

When detection is not a concern --- for example, with atlas data, vegetation
plots, or high-effort presence/absence surveys --- the occupancy observation
model is unnecessary. A JSDM fits the ecological process directly:

```{r jsdm, eval = FALSE}
fit_jsdm <- occu(~ occ_x1, data = sim$data,
                 multispecies = "jsdm", n.factors = 3, verbose = 0)
```

Note the absence of a detection formula. The JSDM assumes perfect detection
($p = 1$), so there is no E-step for the latent occupancy states. This makes
JSDMs substantially faster than occupancy models, especially for large
assemblages. The latent factor structure and species correlation estimation
work identically to the occupancy case described above.


## Community-level random effects

Random effects in the occupancy or detection formula are supported. For
example, a random intercept for habitat type allows occupancy to vary across
habitats while still pooling information at the community level:

```{r re, eval = FALSE}
fit_re <- occu(~ occ_x1 + (1 | habitat), ~ det_x1, data = sim$data,
               multispecies = TRUE, verbose = 0)
ranef(fit_re)
```

Each species gets its own random effect variance, but these are regularised
by the community-level distribution. This is useful when habitat categories
are shared across species but their effects differ in magnitude.


## Multi-species temporal models

For multi-season data with multiple species, combine `multispecies = TRUE` with
a temporal correlation structure:

```{r temporal, eval = FALSE}
sim_mt <- simTMsOcc(N = 50, J = 3, n_species = 5, n_seasons = 4, seed = 100)

fit_mt <- occu(~ 1, ~ 1, data = sim_mt$data,
               multispecies = TRUE, temporal = "ar1", verbose = 0)
```

Each species receives its own AR(1) temporal correlation on the occupancy
intercept, allowing species-specific trends across seasons. The community
distribution pools the temporal autocorrelation parameters, so species with
few detections in some seasons still produce stable trend estimates.


## Model comparison across species

Comparing nested models on a per-species basis reveals which species benefit
from additional covariates:

```{r compare, eval = FALSE}
m1 <- occu(~ 1, ~ 1, data = sim$data, multispecies = TRUE, verbose = 0)
m2 <- occu(~ occ_x1, ~ det_x1, data = sim$data, multispecies = TRUE, verbose = 0)

w1 <- waicOccu(m1, by.sp = TRUE)
w2 <- waicOccu(m2, by.sp = TRUE)

delta <- w1$by_species$waic - w2$by_species$waic
names(delta) <- w1$by_species$species
sort(delta)
```

A positive $\Delta$WAIC means the covariate model fits that species better
than the intercept-only model. Species with large positive values respond
strongly to the covariates. Species with negative $\Delta$WAIC values suggest
the covariates add unnecessary complexity for those species.


## Post-hoc trait analysis

A natural follow-up question is whether species-level coefficient variation
correlates with functional traits. The `postHocLM()` function regresses
estimated species coefficients on a trait matrix:

```{r traits, eval = FALSE}
traits <- data.frame(
  species = names(coef(fit)),
  body_mass = rnorm(10),
  diet = sample(c("insect", "seed"), 10, replace = TRUE)
)

ph <- postHocLM(fit, ~ body_mass + diet, traits = traits)
summary(ph)
```

This analysis answers questions like "do larger-bodied species have higher
occupancy?" or "do insectivores respond differently to habitat covariates than
granivores?" The trait regression is fitted to the posterior means of the
species-level coefficients, with standard errors that account for estimation
uncertainty in those coefficients.


## Practical guidance

- **Start simple.** Fit the community model first (`multispecies = TRUE`
  without latent factors). Add latent factors only if species correlations are
  of scientific interest or if the community model shows poor fit.

- **Factor selection.** Begin with $k = \lfloor S / 2 \rfloor$ and compare
  models using WAIC. In most empirical datasets, 2--5 factors suffice even for
  assemblages of 30+ species.

- **Rare species.** Species with fewer than 5 detections benefit the most from
  community-level pooling. Without the hierarchical structure, their
  coefficients would be dominated by prior or boundary effects.

- **JSDMs vs. occupancy models.** JSDMs are faster because they skip the
  detection E-step, but they assume perfect detection. Use them for atlas-quality
  data or when detection probability is known to be high.

- **Parallelization.** For assemblages with more than 20 species, set
  `options(INLAocc.cores = ...)` to the number of available cores.
  Parallelization is embarrassingly parallel and scales linearly.

- **Spatial factors.** The spatial latent factor model is computationally
  expensive because each factor carries its own SPDE mesh. Reduce $k$ or
  coarsen the mesh (via `mesh.args`) if runtime is prohibitive.

- **Traits.** Use `postHocLM()` for exploratory trait--environment analysis.
  For formal inference on trait--environment interactions, incorporate traits
  directly as predictors of the community-level mean (a fourth-corner model),
  which INLAocc does not currently support.
