---
title: "Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

INLAocc fits hierarchical occupancy models using Integrated Nested Laplace
Approximations (INLA) instead of Markov chain Monte Carlo (MCMC). The package
provides a single entry point --- `occu()` --- that handles single-species,
multi-species, spatial, and temporal occupancy models through a familiar
formula interface.

Occupancy modeling addresses a fundamental problem in ecology: species are
present at some sites but may go undetected on any given survey visit. Without
accounting for this imperfect detection, naive estimates of species occurrence
are biased low. The occupancy model separates two processes --- the ecological
state (is the species truly there?) and the observation process (did we detect
it?) --- to produce unbiased estimates of both occupancy and detection
probability.

INLA replaces MCMC sampling with a deterministic approximation to the
posterior distribution. In practice, this means no burn-in period, no thinning,
no convergence diagnostics, and identical results every time you run the model.
For occupancy models of moderate complexity, this translates to 5--20x
speedups over equivalent MCMC implementations while producing posterior
summaries that are nearly indistinguishable from long MCMC runs.

This vignette walks through the core workflow: simulating data with known
parameters, fitting a basic single-species model, checking the fit with
diagnostics, adding random effects, comparing competing models, and making
predictions. For spatial models with SPDE random fields, see
`vignette("spatial-models")`. For multi-species community models, see
`vignette("multi-species")`. For a full treatment of the diagnostic toolkit,
see `vignette("diagnostics")`.


## The occupancy model

The standard single-species, single-season occupancy model has two levels.

**Ecological process (true state).** Each site $i = 1, \dots, N$ has a latent
occupancy state $z_i \in \{0, 1\}$ drawn from a Bernoulli distribution:

$$z_i \sim \text{Bernoulli}(\psi_i), \quad \text{logit}(\psi_i) = \mathbf{x}_i^\top \boldsymbol{\beta}$$

where $\psi_i$ is the occupancy probability at site $i$ and
$\mathbf{x}_i$ is a vector of site-level covariates (e.g., elevation, forest
cover) with regression coefficients $\boldsymbol{\beta}$.

**Observation process (detection).** At each site $i$, we conduct $J$ repeat
visits. The observed detection/non-detection $y_{ij}$ on visit $j$ depends on
both the true state and a detection probability $p_{ij}$:

$$y_{ij} \mid z_i \sim \text{Bernoulli}(z_i \cdot p_{ij}), \quad \text{logit}(p_{ij}) = \mathbf{w}_{ij}^\top \boldsymbol{\alpha}$$

where $\mathbf{w}_{ij}$ is a vector of visit-level covariates (e.g., survey
effort, weather) with coefficients $\boldsymbol{\alpha}$.

The key constraint is that when $z_i = 0$ (species absent), detection is
impossible: $y_{ij} = 0$ for all visits $j$, regardless of $p_{ij}$. When
$z_i = 1$ (species present), each visit is an independent Bernoulli trial with
success probability $p_{ij}$. A site where the species was never detected
across all visits could be either truly unoccupied or occupied but missed every
time --- the model uses the repeated visits to disentangle these two
explanations.

The full likelihood marginalizes over the latent $z_i$, producing a finite
mixture that standard INLA cannot fit directly. INLAocc handles this with an
EM algorithm: the E-step computes the posterior probability
$P(z_i = 1 \mid \mathbf{y}_i)$ for each site, and the M-step refits the
occupancy and detection sub-models via INLA using these posterior weights.
After convergence, a short Gibbs data augmentation phase corrects for the
coefficient attenuation inherent in the EM point-mass approximation, yielding
properly calibrated posterior summaries.


## Simulating data

We start by simulating occupancy data with known parameters so we can verify
that the model recovers them. The `simulate_occu()` function generates sites,
visits, covariates, and detection histories from the model described above.

```{r simulate}
library(INLAocc)

sim <- simulate_occu(
  N = 200, J = 4,
  beta_occ = c(0.5, -0.8, 0.8),
  beta_det = c(0, -0.5),
  seed = 42
)
```

This creates 200 sites with 4 visits each. The occupancy linear predictor has
an intercept of 0.5 and two covariate effects ($-0.8$ and $0.8$). The detection
linear predictor has an intercept of 0 (corresponding to detection probability
0.5 on the logit scale) and one covariate effect ($-0.5$). The function returns
a list with the formatted data (`sim$data`, an `occu_data` object) and the
true parameter values (`sim$truth`).

We can inspect the data with `summary()`:

```{r summary-data}
summary(sim$data)
```

The naive occupancy rate (proportion of sites where the species was detected at
least once) is a lower bound on the true occupancy rate --- some occupied sites
had the species go undetected across all four visits. A central goal of the
model is to estimate how many of those "all-zero" sites are actually occupied.

The `plot()` method for `occu_data` objects produces a panel of histograms
showing the distribution of covariates and the per-site detection frequency:

```{r plot-data, fig.height = 4}
plot(sim$data)
```

The detection frequency histogram (number of visits per site where the species
was detected) is the first thing to examine. A distribution concentrated at zero
means either the species is rare or detection probability is low --- the model
will try to disentangle these, but extremely sparse data (say, detections at
fewer than 10% of sites) makes parameter estimation unstable. At the other
extreme, if the species is detected on every visit at most sites, there is
little information to separate occupancy from detection, because the "all-zero"
sites that drive the occupancy estimate are rare.

Check the covariate histograms for extreme skew, outliers, or gaps. Covariates
that are strongly skewed or have a few extreme values can dominate the linear
predictor, pulling fitted probabilities toward 0 or 1 for those observations.
Centering and scaling covariates before fitting (which `simulate_occu()` does by
default) mitigates this. Also check for missing visits --- `NA` entries in the
detection matrix are handled by the likelihood, but sites with only one valid
visit out of four contribute less information about detection probability, and
many such sites can widen credible intervals.


## Fitting the basic model

The `occu()` function takes two formulas: the first specifies occupancy
covariates (right-hand side only), and the second specifies detection
covariates. Intercepts are included automatically in both.

```{r fit-basic}
fit <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim$data, verbose = 0)
```

The `summary()` method prints the full posterior summary:

```{r summary-fit}
summary(fit)
```

**Interpreting the output.** The true occupancy coefficients were
$\boldsymbol{\beta} = (0.5, -0.8, 0.8)$ and the true detection coefficients
were $\boldsymbol{\alpha} = (0, -0.5)$. Compare the posterior means to these
true values --- they should be close, and all 95% credible intervals should
contain the generating values.

The estimated mean occupancy probability is higher than the naive
detection rate, as expected. Accounting for imperfect detection reveals
sites that were occupied but the species went undetected across all four
visits.

The occupancy summary table reports posterior means, standard deviations, and
95% credible intervals for each coefficient on the logit scale. The intercept
($\beta_0$) sets the baseline occupancy probability when all covariates are at
zero: an intercept of 0.5 corresponds to $\text{logit}^{-1}(0.5) \approx 0.62$,
meaning about 62% of sites are occupied at the covariate mean. Each slope
coefficient gives the change in log-odds per unit increase in the covariate. A
coefficient of $-0.8$ means a one-unit increase drops the log-odds of occupancy
by 0.8, which is a substantial effect --- roughly halving the odds.

The detection intercept has a parallel interpretation. An intercept of 0
translates to $\text{logit}^{-1}(0) = 0.5$, meaning a 50% chance of detecting
the species on any single visit when detection covariates are at zero. This
baseline matters for survey design: if per-visit detection is 0.5 and you have 4
visits, the probability of detecting a present species at least once is
$1 - (1 - 0.5)^4 = 0.94$. Pushing detection above 0.7 per visit (through
better survey methods or more effort) would raise this to 0.99. Conversely, if
the detection intercept is strongly negative (say $-1.5$, giving $p \approx
0.18$), four visits yield only a 55% chance of detecting a present species ---
and the model must attribute many all-zero sites to missed detections rather
than true absence, widening the occupancy credible intervals.

Because INLA is deterministic, re-running this code produces identical
results --- there is no Monte Carlo variability to worry about.


## Checking the fit

Model checking is not optional. A model that converges is not necessarily a
model that fits the data well. INLAocc provides simulation-based diagnostics
through the `checkModel()` function and targeted hypothesis tests through
`testUniformity()` and `testDispersion()`.

Occupancy models pose a specific diagnostic challenge. The response variable is
binary (detected or not), so standard residual plots that assume continuous or
count data are misleading. Worse, the ecological state $z_i$ is latent --- you
never observe whether a site is truly occupied, only whether the species was
detected. This means you cannot compute traditional residuals that compare
observed vs. expected counts at a site. Simulation-based diagnostics sidestep
both problems: they generate new datasets from the fitted model and compare
properties of the simulated data to the observed data, without requiring
distributional assumptions beyond the model itself.

### Full diagnostic panel

The `checkModel()` function generates posterior predictive simulations and
produces a four-panel diagnostic plot:

```{r check-model, fig.height = 6}
checkModel(fit)
```

The four panels are:

1. **QQ plot of PIT residuals.** Probability integral transform (PIT) residuals
   should follow a uniform distribution if the model is correctly specified.
   On the QQ plot, this means the points should fall along the diagonal.
   Systematic departures indicate model misspecification --- an S-shaped curve
   suggests overdispersion, while a step pattern suggests zero-inflation.

2. **Residuals vs. fitted values.** The scaled residuals plotted against
   predicted detection rates should show no systematic trend. A funnel shape
   or curvature indicates that the variance structure or the functional form
   of the linear predictor is wrong.

3. **Dispersion test.** The observed residual variance is compared against the
   distribution of variances from simulated data sets generated under the
   fitted model. The observed value (red line) should fall within the
   simulated distribution. A dispersion ratio near 1.0 is ideal.

4. **Outlier test.** Extreme residuals are compared against their expected
   frequency under the model. An excess of outliers suggests that some sites
   behave very differently from the model's predictions.

### Formal hypothesis tests

For a quantitative assessment, use the individual test functions:

```{r test-uniformity}
testUniformity(fit)
```

A non-significant p-value means we cannot reject the hypothesis that the PIT
residuals are uniformly distributed --- no evidence of systematic
misspecification.

```{r test-dispersion}
testDispersion(fit)
```

A dispersion ratio close to 1 confirms that the model's assumed variance
matches the observed variance. Ratios substantially above 1 indicate
overdispersion; ratios below 1 indicate underdispersion.

### Posterior predictive check

For a Bayesian goodness-of-fit assessment, `ppcOccu()` compares a summary
statistic computed on the observed data to the same statistic computed on data
simulated from the posterior:

```{r ppc}
ppc <- ppcOccu(fit, fit.stat = "freeman-tukey", group = 1)
ppc$bayesian.p
```

A Bayesian p-value near 0.5 means the model generates data whose summary
statistics are centred on the observed values. Values near 0 or 1 signal
systematic misfit. The `group = 1` argument aggregates by site; `group = 2`
aggregates by visit.

In practice, the diagnostic workflow proceeds in layers. Start with
`checkModel()` for a visual overview --- the four-panel plot catches most forms
of misspecification at a glance. If something looks off, run the targeted tests
(`testUniformity()`, `testDispersion()`) to quantify the concern. Finally, use
`ppcOccu()` with different summary statistics (`"freeman-tukey"`, `"chi-square"`)
and grouping variables to probe specific hypotheses about where the model fails.


## Comparing true vs. estimated parameters

When working with simulated data, the most direct validation is to compare the
estimated coefficients against the known truth. This parameter recovery step
builds confidence that the model and its fitting algorithm are working
correctly.

```{r recovery}
truth <- data.frame(
  parameter = c("beta_occ[1]", "beta_occ[2]", "beta_occ[3]",
                "beta_det[1]", "beta_det[2]"),
  true      = c(0.5, -0.8, 0.8, 0, -0.5),
  estimated = c(coef(fit, "occupancy"), coef(fit, "detection"))
)
truth
```

The Gibbs data augmentation correction that runs after the EM algorithm is
responsible for this close recovery --- without it, the EM approximation alone
would attenuate the coefficients toward zero.


## Random effects

When data are grouped --- sites nested within regions, surveys conducted by
different observers, or species within communities --- random effects capture
the between-group variation that fixed effects cannot. In occupancy models,
random effects are common on both the occupancy and detection sides.

The mathematical extension adds a group-level random intercept to the linear
predictor:

$$\text{logit}(\psi_i) = \mathbf{x}_i^\top \boldsymbol{\beta} + u_{g[i]}, \quad u_g \sim \mathcal{N}(0, \sigma^2_u)$$

where $g[i]$ maps site $i$ to its group and $\sigma^2_u$ controls how much
groups differ after accounting for the fixed-effect covariates.

INLAocc uses standard mixed-model formula syntax. To add a random intercept
for region:

```{r fit-re}
# Add a grouping variable to the data
sim$data$occ.covs$region <- sample(1:5, 200, replace = TRUE)

fit_re <- occu(~ occ_x1 + occ_x2 + (1 | region), ~ det_x1,
               data = sim$data, verbose = 0)
summary(fit_re)
```

INLA parameterizes random effects in terms of precision (the inverse of
variance). High precision means low between-group variation --- the groups do
not differ substantially after accounting for the covariates. This is expected
here, since we simulated the data without any grouping structure.

To convert precision to a standard deviation, take $\sigma = 1 / \sqrt{\tau}$.
A precision of 4 corresponds to $\sigma = 0.5$, meaning groups differ by about
1 unit on the logit scale ($\pm 2\sigma$). What does that look like in
practice? If the mean occupancy is 0.5 (logit = 0), a group-level SD of 0.5
shifts the logit to roughly $-1$ and $+1$ at the extremes, giving occupancy
probabilities between about 0.27 and 0.73 --- meaningful ecological variation. A
SD of 0.1, by contrast, keeps all groups between 0.48 and 0.52, which is
negligible. When reporting results, always convert precision to SD for
interpretability. Precision values are useful for INLA's internal
parameterization but carry no ecological intuition.

The supported random effect syntaxes are:

- `(1 | group)` --- random intercept per level of `group`

- `(x | group)` --- correlated random intercept and slope for covariate `x`

- `(x || group)` --- uncorrelated random intercept and slope

- `(1 | a/b)` --- nested random effects (random intercept for `b` within `a`)

Random effects work on both the occupancy and detection sides of the model.
For example, `occu(~ elev + (1 | region), ~ (1 | observer), data = dat)`
includes a random intercept for region in occupancy and a random intercept for
observer in detection.


## Model comparison

Selecting among competing model structures is a standard step in any analysis.
INLAocc supports comparison via WAIC (Watanabe-Akaike Information Criterion),
AIC, and BIC. WAIC is the natural choice for Bayesian models fitted with INLA,
as it uses the full posterior predictive distribution rather than a point
estimate.

```{r compare}
m1 <- occu(~ 1, ~ 1, data = sim$data, verbose = 0)
m2 <- occu(~ occ_x1, ~ det_x1, data = sim$data, verbose = 0)
m3 <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim$data, verbose = 0)

compare_models(null = m1, simple = m2, full = m3, criterion = "waic")
```

The table ranks models by WAIC (lower is better). The `delta` column gives the
WAIC difference from the best model, and `weight` gives the Akaike-style model
weight. The full model carrying both covariates should be strongly preferred,
confirming that both `occ_x1` and `occ_x2` contribute meaningful information
about occupancy.

For situations where no single model is clearly dominant, `modelAverage()`
computes WAIC-weighted averages of predictions across models:

```{r model-average}
avg <- modelAverage(null = m1, simple = m2, full = m3, criterion = "waic")
head(avg$psi_hat)  # weighted-average occupancy probabilities
```

Model averaging is particularly useful when several models have similar WAIC
values and you want predictions that account for model uncertainty rather than
conditioning on a single "best" structure.

When interpreting the `delta` column (WAIC difference from the best model),
standard rules of thumb from Burnham and Anderson (2002) apply: a delta below 2
means the models are essentially indistinguishable given the data, and model
averaging is appropriate. A delta between 2 and 7 indicates moderate support for
the better model, though the weaker model may still contribute usefully in an
averaging framework. A delta above 10 is strong evidence that the better model
fits the data substantially more accurately, and the weaker model can be dropped
from consideration. These thresholds are guidelines, not bright lines ---
context and sample size matter.

### K-fold cross-validation

For out-of-sample validation, pass `k.fold` during fitting:

```{r kfold, eval = FALSE}
fit_cv <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim$data,
               k.fold = 5, verbose = 0)
fit_cv$k.fold
```

This holds out 1/5 of sites at a time, refits on the rest, and evaluates
predictions on the held-out sites. K-fold is slower than WAIC (the model is
refit $k$ times) but gives a direct estimate of out-of-sample predictive
performance.


## Prediction

### Marginal effects

To understand how a single covariate influences occupancy (or detection) while
holding other covariates at their means, use `marginal_effect()`:

```{r marginal-effect, fig.height = 4}
me <- marginal_effect(fit, "occ_x1")
plot(me)
```

This varies `occ_x1` across its observed range, fixes all other covariates at
their sample means, and computes predicted occupancy probabilities with 95%
credible intervals. The `plot()` method produces a ribbon plot with the
posterior mean and credible band. Because `occ_x1` has a negative coefficient,
the plot shows occupancy declining from left to right.

Marginal effects differ from raw coefficient interpretation in an important way.
The coefficients are on the logit scale, where a one-unit change always
corresponds to the same change in log-odds. But on the probability scale, the
effect is nonlinear. A unit increase in a covariate has the largest impact on
probability when the baseline is near 0.5 (the steepest part of the logistic
curve) and a diminishing impact as the baseline approaches 0 or 1. A coefficient
of $-0.8$ drops occupancy from 0.62 to 0.43 when the baseline logit is 0.5, but
only from 0.12 to 0.07 when the baseline logit is $-2$. Marginal effect plots
capture this nonlinearity directly, showing the actual probability change across
the covariate range rather than a constant slope.

### Prediction at new covariate values

For predictions at specific covariate combinations, use `predict()` with a
data frame of new values:

```{r predict-new}
new_sites <- data.frame(occ_x1 = c(-1, 0, 1), occ_x2 = c(0, 0, 0))
predict(fit, X.0 = new_sites, type = "occupancy")
```

The predicted occupancy decreases as `occ_x1` increases, reflecting the
negative effect of this covariate. All predictions are on the probability
scale (inverse-logit of the linear predictor), with quantile-based credible
intervals.

The `terms` argument offers a more concise syntax for marginal prediction
grids, similar to `ggeffects::ggpredict()`:

```{r predict-terms}
predict(fit, terms = "occ_x1 [-2:2 by=0.5]")
```


## S3 methods

Fitted `occu_inla` objects support the standard R model interface, making them
interoperable with tools you already know:

| Method | Description |
|:-------|:------------|
| `summary(fit)` | Full posterior summary with model fit statistics |
| `coef(fit)` | Posterior means of fixed effects |
| `confint(fit)` | Credible intervals for fixed effects |
| `fitted(fit)` | Fitted occupancy and detection probabilities |
| `residuals(fit)` | Residuals (Pearson or PIT) |
| `predict(fit, X.0 = ...)` | Prediction at new covariate values |
| `plot(fit)` | Four-panel diagnostic plot (EM trace, psi, p, z vs. naive) |
| `logLik(fit)` | Marginal log-likelihood |
| `nobs(fit)` | Number of sites |
| `vcov(fit)` | Posterior variance-covariance matrix |
| `simulate(fit)` | Simulate new detection histories from the fitted model |
| `update(fit, ~ . - occ_x2)` | Refit with a modified formula |
| `tidy(fit)` / `glance(fit)` | Tidy summaries (broom-compatible) |
| `ranef(fit)` | Extract random effect posteriors |


## What's next

This vignette covered the core single-species workflow. A recommended analysis
sequence: start with the intercept-only model to establish baseline occupancy
and detection rates, then add covariates guided by ecological hypotheses. Check
diagnostics at each step --- a model that fits well without covariates but fails
after adding them suggests a misspecified functional form (e.g., a quadratic
effect forced through a linear term). Add random effects only when the study
design includes genuine grouping structure, not as a generic variance sponge.
Compare models with WAIC, and report marginal effects rather than raw
coefficients for biological interpretation.

INLAocc supports substantially more complex models through the same `occu()`
interface:

- **Spatial models.** When sites are spatially correlated, add an SPDE random
  field by passing coordinates to the `spatial` argument. See
  `vignette("spatial-models")` for the full treatment, including mesh
  construction, spatially varying coefficients, and spatial prediction maps.

- **Multi-species models.** Community occupancy models that share information
  across species, with optional latent factor structure. See
  `vignette("multi-species")`.

- **Diagnostics deep dive.** The full suite of simulation-based diagnostics,
  including spatial residual analysis, temporal autocorrelation tests, and
  posterior predictive checks. See `vignette("diagnostics")`.

- **Migrating from spOccupancy.** If you are switching from an MCMC-based
  workflow, `vignette("spoccupancy-migration")` maps spOccupancy's function
  names, data formats, and output structures to their INLAocc equivalents.
