---
title: "Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

INLAocc fits hierarchical occupancy models using Integrated Nested Laplace
Approximations (INLA) instead of Markov chain Monte Carlo (MCMC). The package
provides a single entry point, `occu()`, that handles single-species,
multi-species, spatial, and temporal occupancy models through a familiar
formula interface.

Occupancy modeling addresses a fundamental problem in ecology: species are
present at some sites but may go undetected on any given survey visit. Without
accounting for this imperfect detection, naive estimates of species occurrence
are biased low. The occupancy model separates two processes: the ecological
state (is the species truly there?) and the observation process (did we detect
it?). This separation produces unbiased estimates of both occupancy and detection
probability.

INLA replaces MCMC sampling with a deterministic approximation to the
posterior distribution. In practice, this means no burn-in period, no thinning,
no convergence diagnostics, and identical results every time you run the model.
For occupancy models of moderate complexity, this translates to 5--20x
speedups over equivalent MCMC implementations while producing posterior
summaries that are nearly indistinguishable from long MCMC runs.

This vignette walks through the core workflow: simulating data with known
parameters, fitting a basic single-species model, checking the fit with
diagnostics, adding random effects, comparing competing models, making
predictions, and then extending to spatial and multi-species models. For the
full treatment of spatial models (mesh control, PC priors, spatially varying
coefficients), see `vignette("spatial-models")`. For the full treatment of
multi-species models (latent factors, JSDMs), see `vignette("multi-species")`.
For the diagnostic toolkit in depth, see `vignette("diagnostics")`.


## The occupancy model

The standard single-species, single-season occupancy model has two levels.

**Ecological process (true state).** Each site $i = 1, \dots, N$ has a latent
occupancy state $z_i \in \{0, 1\}$ drawn from a Bernoulli distribution:

$$z_i \sim \text{Bernoulli}(\psi_i), \quad \text{logit}(\psi_i) = \mathbf{x}_i^\top \boldsymbol{\beta}$$

where $\psi_i$ is the occupancy probability at site $i$ and
$\mathbf{x}_i$ is a vector of site-level covariates (e.g., elevation, forest
cover) with regression coefficients $\boldsymbol{\beta}$.

**Observation process (detection).** At each site $i$, we conduct $J$ repeat
visits. The observed detection/non-detection $y_{ij}$ on visit $j$ depends on
both the true state and a detection probability $p_{ij}$:

$$y_{ij} \mid z_i \sim \text{Bernoulli}(z_i \cdot p_{ij}), \quad \text{logit}(p_{ij}) = \mathbf{w}_{ij}^\top \boldsymbol{\alpha}$$

where $\mathbf{w}_{ij}$ is a vector of visit-level covariates (e.g., survey
effort, weather) with coefficients $\boldsymbol{\alpha}$.

The key constraint is that when $z_i = 0$ (species absent), detection is
impossible: $y_{ij} = 0$ for all visits $j$, regardless of $p_{ij}$. When
$z_i = 1$ (species present), each visit is an independent Bernoulli trial with
success probability $p_{ij}$. A site where the species was never detected
across all visits could be either truly unoccupied or occupied but missed every
time. The model uses the repeated visits to disentangle these two
explanations.

The likelihood that results from marginalizing over the latent $z_i$ is a
finite mixture of two components: the site is absent (probability $1 - \psi_i$,
all detections must be zero) or the site is occupied (probability $\psi_i$,
detections follow independent Bernoulli trials). This mixture does not factor
into separate occupancy and detection terms the way a standard GLM likelihood
does. The two processes are entangled: the detection data inform occupancy
(many non-detections lower the posterior probability of presence), and the
occupancy probability informs detection (if a site is likely absent, its
non-detections carry no information about $p$). Any fitting method must handle
this coupling. MCMC does it by augmenting the model with the latent $z_i$ and
sampling all unknowns jointly. INLA, which is designed for latent Gaussian
models, cannot handle the discrete mixture directly.

Identifiability comes from the repeated visits. With a single visit ($J = 1$),
a non-detection could mean the species is absent or that it is present but
undetected, and there is no way to distinguish these explanations from the
data alone. With $J > 1$ visits, the pattern of detections and non-detections
across visits provides leverage. If a species is detected at some visits but
not others at a site, this directly estimates $p$ (the proportion of visits
with detections at occupied sites). If a species is never detected at a site
despite many visits, and detection is estimated to be reasonably high from
other sites, the model can infer that the site is more likely absent than
occupied-but-missed. The information accumulates geometrically: the probability
of missing a present species on all $J$ visits is $(1 - p)^J$, which shrinks
quickly as $J$ or $p$ increases.

The difficulty of estimation depends on where detection falls. When per-visit
detection is high ($p > 0.5$), the model is easy to fit. Most occupied sites
are detected, the "all-zero" sites are mostly truly absent, and there is little
ambiguity. The occupancy estimate is close to the naive detection rate, and
credible intervals are narrow. When detection is low ($p < 0.15$), the model
becomes hard. Many occupied sites go undetected, so a large fraction of sites
have ambiguous status. The model must infer both $\psi$ and $p$ from data
where most of the occupancy signal is hidden, and the estimates are sensitive
to model assumptions. In the extreme case ($p \to 0$), the model is
unidentifiable again: if you never detect the species anywhere, you cannot
distinguish zero occupancy from nonzero occupancy with zero detection. Most
real surveys fall between these extremes, with detection rates of 0.15--0.5
per visit, where the model is identifiable but the estimates carry nontrivial
uncertainty.

The full likelihood marginalizes over the latent $z_i$, producing a finite
mixture that standard INLA cannot fit directly. INLAocc handles this with an
EM algorithm: the E-step computes the posterior probability
$P(z_i = 1 \mid \mathbf{y}_i)$ for each site, and the M-step refits the
occupancy and detection sub-models via INLA using these posterior weights.
After convergence, a short Gibbs data augmentation phase corrects for the
coefficient attenuation inherent in the EM point-mass approximation, yielding
properly calibrated posterior summaries.


## Simulating data

We start by simulating occupancy data with known parameters so we can verify
that the model recovers them. The `simulate_occu()` function generates sites,
visits, covariates, and detection histories from the model described above.

```{r simulate}
library(INLAocc)

sim <- simulate_occu(
  N = 200, J = 4,
  beta_occ = c(0.5, -0.8, 0.8),
  beta_det = c(0, -0.5),
  seed = 42
)
```

Simulation is the first step in any occupancy analysis workflow. Before fitting
a model to field data, you simulate from known parameter values to verify that
the model and code recover those values correctly. This catches formula errors
(wrong covariates, mismatched dimensions), data formatting issues (detection
matrix orientation, missing values), and convergence problems (non-identified
parameters, extreme starting values) before they contaminate real results.
Simulation also builds intuition about what the data should look like under
different parameter regimes. If your simulated data with $p = 0.2$ produces
detection histories that look nothing like your field data, your assumed
detection rate is probably wrong.

The parameters in the call above have specific ecological interpretations. An
occupancy intercept of 0.5 on the logit scale corresponds to
$\text{logit}^{-1}(0.5) \approx 0.62$, meaning 62% baseline occupancy. The
slope of $-0.8$ is a strong negative effect: a one-unit increase in the
covariate drops the log-odds of occupancy by 0.8, roughly halving the odds.
The second slope of $0.8$ is an equally strong positive effect. The detection
intercept of 0 corresponds to exactly 50% detection per visit at the covariate
mean, and the detection slope of $-0.5$ means detection degrades under
unfavorable conditions. These are not unusual values for real ecological
surveys; detection rates between 0.2 and 0.6 per visit and occupancy rates
between 0.3 and 0.7 are common for many vertebrate taxa.

This creates 200 sites with 4 visits each. The occupancy linear predictor has
an intercept of 0.5 and two covariate effects ($-0.8$ and $0.8$). The detection
linear predictor has an intercept of 0 (corresponding to detection probability
0.5 on the logit scale) and one covariate effect ($-0.5$). The function returns
a list with the formatted data (`sim$data`, an `occu_data` object) and the
true parameter values (`sim$truth`).

We can inspect the data with `summary()`:

```{r summary-data}
summary(sim$data)
```

The naive occupancy rate (proportion of sites where the species was detected at
least once) is a lower bound on the true occupancy rate, because some occupied sites
had the species go undetected across all four visits. A central goal of the
model is to estimate how many of those "all-zero" sites are actually occupied.

The `plot()` method for `occu_data` objects produces a panel of histograms
showing the distribution of covariates and the per-site detection frequency:

```{r plot-data, fig.height = 4}
plot(sim$data)
```

The detection frequency histogram (number of visits per site where the species
was detected) is the first thing to examine. A distribution concentrated at zero
means either the species is rare or detection probability is low. The model
will try to disentangle these, but extremely sparse data (say, detections at
fewer than 10% of sites) makes parameter estimation unstable. At the other
extreme, if the species is detected on every visit at most sites, there is
little information to separate occupancy from detection, because the "all-zero"
sites that drive the occupancy estimate are rare.

Check the covariate histograms for extreme skew, outliers, or gaps. Covariates
that are strongly skewed or have a few extreme values can dominate the linear
predictor, pulling fitted probabilities toward 0 or 1 for those observations.
Centering and scaling covariates before fitting (which `simulate_occu()` does by
default) mitigates this. Also check for missing visits: `NA` entries in the
detection matrix are handled by the likelihood, but sites with only one valid
visit out of four contribute less information about detection probability, and
many such sites can widen credible intervals.


## Fitting the basic model

The `occu()` function takes two formulas: the first specifies occupancy
covariates (right-hand side only), and the second specifies detection
covariates. Intercepts are included automatically in both.

```{r fit-basic}
fit <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim$data, verbose = 0)
```

The `summary()` method prints the full posterior summary:

```{r summary-fit}
summary(fit)
```

**Interpreting the output.** The true occupancy coefficients were
$\boldsymbol{\beta} = (0.5, -0.8, 0.8)$ and the true detection coefficients
were $\boldsymbol{\alpha} = (0, -0.5)$. Compare the posterior means to these
true values. They should be close, and all 95% credible intervals should
contain the generating values.

The estimated mean occupancy probability is higher than the naive
detection rate, as expected. Accounting for imperfect detection reveals
sites that were occupied but the species went undetected across all four
visits.

The occupancy summary table reports posterior means, standard deviations, and
95% credible intervals for each coefficient on the logit scale. The intercept
($\beta_0$) sets the baseline occupancy probability when all covariates are at
zero: an intercept of 0.5 corresponds to $\text{logit}^{-1}(0.5) \approx 0.62$,
meaning about 62% of sites are occupied at the covariate mean. Each slope
coefficient gives the change in log-odds per unit increase in the covariate. A
coefficient of $-0.8$ means a one-unit increase drops the log-odds of occupancy
by 0.8, a substantial effect that roughly halves the odds.

The detection intercept has a parallel interpretation. An intercept of 0
translates to $\text{logit}^{-1}(0) = 0.5$, meaning a 50% chance of detecting
the species on any single visit when detection covariates are at zero. This
baseline matters for survey design: if per-visit detection is 0.5 and you have 4
visits, the probability of detecting a present species at least once is
$1 - (1 - 0.5)^4 = 0.94$. Pushing detection above 0.7 per visit (through
better survey methods or more effort) would raise this to 0.99. Conversely, if
the detection intercept is strongly negative (say $-1.5$, giving $p \approx
0.18$), four visits yield only a 55% chance of detecting a present species,
and the model must attribute many all-zero sites to missed detections rather
than true absence, widening the occupancy credible intervals.

The credible intervals deserve attention. A 95% credible interval of, say,
$(-1.1, -0.5)$ for a slope coefficient means there is a 95% posterior
probability that the true coefficient falls in that range. The interval does
not contain zero, so there is strong evidence for a nonzero effect. When an
interval spans zero (say $(-0.3, 0.4)$), the data are consistent with
no effect, a weak positive effect, and a weak negative effect. The width of
the interval reflects how much information the data carry about that parameter.
Narrow intervals come from large sample sizes and high detection rates; wide
intervals come from small samples, low detection, or collinear covariates.

The estimated occupancy is higher than the naive detection rate (the
proportion of sites where the species was detected at least once). The gap
between these two numbers is the model's estimate of how many occupied sites
were missed entirely. In conservation contexts, this gap matters: the naive
rate underestimates the species' range, which could lead to underprotection.
If the estimated occupancy is 0.65 but the naive rate is 0.50, the model
says roughly 15% of sites are occupied but undetected, and those sites
may include habitat that would otherwise be classified as empty and
potentially cleared or developed.

Because INLA is deterministic, re-running this code produces identical
results, with no Monte Carlo variability to worry about.


## Checking the fit

Model checking is not optional. A model that converges is not necessarily a
model that fits the data well. INLAocc provides simulation-based diagnostics
through the `checkModel()` function and targeted hypothesis tests through
`testUniformity()` and `testDispersion()`.

Occupancy models pose a specific diagnostic challenge. The response variable is
binary (detected or not), so standard residual plots that assume continuous or
count data are misleading. Worse, the ecological state $z_i$ is latent: you
never observe whether a site is truly occupied, only whether the species was
detected. This means you cannot compute traditional residuals that compare
observed vs. expected counts at a site. Simulation-based diagnostics sidestep
both problems: they generate new datasets from the fitted model and compare
properties of the simulated data to the observed data, without requiring
distributional assumptions beyond the model itself.

### Full diagnostic panel

The `checkModel()` function generates posterior predictive simulations and
produces a four-panel diagnostic plot:

```{r check-model, fig.height = 6}
checkModel(fit)
```

The four panels are:

1. **QQ plot of PIT residuals.** Probability integral transform (PIT) residuals
   should follow a uniform distribution if the model is correctly specified.
   On the QQ plot, this means the points should fall along the diagonal.
   Systematic departures indicate model misspecification. An S-shaped curve
   (residuals above the diagonal at both tails) suggests overdispersion,
   meaning the data are more variable than the model predicts. In occupancy models,
   overdispersion often arises from unmodelled heterogeneity: sites within a
   group differ in ways that the covariates do not capture. For example, two
   forests with the same canopy cover may differ in understory density, and
   the species responds to understory. A step pattern (flat stretches in the
   QQ plot) suggests zero-inflation: more all-zero sites than the model
   expects. This happens when some portion of the landscape is structurally
   unsuitable (e.g., paved surfaces in a bird survey) and should have zero
   occupancy regardless of covariates. Adding a habitat-suitability filter
   or a zero-inflated occupancy model addresses this.

2. **Residuals vs. fitted values.** The scaled residuals plotted against
   predicted detection rates should show no systematic trend. A funnel shape
   or curvature indicates that the variance structure or the functional form
   of the linear predictor is wrong. A U-shape (residuals large at both
   extremes of the fitted range) often means the true relationship between a
   covariate and detection is nonlinear. Adding a quadratic term may help.

3. **Dispersion test.** The observed residual variance is compared against the
   distribution of variances from simulated data sets generated under the
   fitted model. The observed value (red line) should fall within the
   simulated distribution. A dispersion ratio near 1.0 is ideal. Ratios
   above 1.5 usually indicate missing covariates or unmodelled spatial
   structure. Adding a site-level random effect or a spatial term often
   brings the ratio back toward 1. Ratios below 0.7 suggest the model is
   overfitting: it predicts more variation than exists, which can happen
   with too many random effects relative to the number of groups.

4. **Outlier test.** Extreme residuals are compared against their expected
   frequency under the model. An excess of outliers suggests that some sites
   behave very differently from the model's predictions. Common causes:
   data entry errors (a species recorded at an impossible site), misidentified
   species, or genuinely unusual sites (a wetland bird at a desert site due
   to a temporary water source).

When diagnostics disagree (say the QQ plot looks acceptable but the
dispersion ratio is elevated), trust the dispersion test. The QQ plot is a
visual assessment that can mask moderate overdispersion when the sample size is
large (the points scatter tightly around the diagonal even with a ratio of
1.3). The dispersion test is a quantitative comparison that catches smaller
departures. Conversely, a clean dispersion ratio with a problematic QQ plot
usually means the issue is localized: a subset of sites depart from the model,
but the overall variance is fine. In that case, the outlier test will typically
flag the problematic sites.

### Formal hypothesis tests

For a quantitative assessment, use the individual test functions:

```{r test-uniformity}
testUniformity(fit)
```

A non-significant p-value means we cannot reject the hypothesis that the PIT
residuals are uniformly distributed, indicating no evidence of systematic
misspecification.

```{r test-dispersion}
testDispersion(fit)
```

A dispersion ratio close to 1 confirms that the model's assumed variance
matches the observed variance. Ratios substantially above 1 indicate
overdispersion; ratios below 1 indicate underdispersion.

### Posterior predictive check

For a Bayesian goodness-of-fit assessment, `ppcOccu()` compares a summary
statistic computed on the observed data to the same statistic computed on data
simulated from the posterior:

```{r ppc}
ppc <- ppcOccu(fit, fit.stat = "freeman-tukey", group = 1)
ppc$bayesian.p
```

A Bayesian p-value near 0.5 means the model generates data whose summary
statistics are centred on the observed values. Values near 0 or 1 signal
systematic misfit. The `group = 1` argument aggregates by site; `group = 2`
aggregates by visit.

In practice, the diagnostic workflow proceeds in layers. Start with
`checkModel()` for a visual overview, since the four-panel plot catches most forms
of misspecification at a glance. If something looks off, run the targeted tests
(`testUniformity()`, `testDispersion()`) to quantify the concern. Finally, use
`ppcOccu()` with different summary statistics (`"freeman-tukey"`, `"chi-square"`)
and grouping variables to probe specific hypotheses about where the model fails.


## Comparing true vs. estimated parameters

When working with simulated data, the most direct validation is to compare the
estimated coefficients against the known truth. This parameter recovery step
builds confidence that the model and its fitting algorithm are working
correctly.

```{r recovery}
truth <- data.frame(
  parameter = c("beta_occ[1]", "beta_occ[2]", "beta_occ[3]",
                "beta_det[1]", "beta_det[2]"),
  true      = c(0.5, -0.8, 0.8, 0, -0.5),
  estimated = c(coef(fit, "occupancy"), coef(fit, "detection"))
)
truth
```

The Gibbs data augmentation correction that runs after the EM algorithm is
responsible for this close recovery. Without it, the EM approximation alone
would attenuate the coefficients toward zero.


## Random effects

When data are grouped (sites nested within regions, surveys conducted by
different observers, or species within communities), random effects capture
the between-group variation that fixed effects cannot. In occupancy models,
random effects are common on both the occupancy and detection sides.

The mathematical extension adds a group-level random intercept to the linear
predictor:

$$\text{logit}(\psi_i) = \mathbf{x}_i^\top \boldsymbol{\beta} + u_{g[i]}, \quad u_g \sim \mathcal{N}(0, \sigma^2_u)$$

where $g[i]$ maps site $i$ to its group and $\sigma^2_u$ controls how much
groups differ after accounting for the fixed-effect covariates.

INLAocc uses standard mixed-model formula syntax. To add a random intercept
for region:

```{r fit-re}
# Add a grouping variable to the data
sim$data$occ.covs$region <- sample(1:5, 200, replace = TRUE)

fit_re <- occu(~ occ_x1 + occ_x2 + (1 | region), ~ det_x1,
               data = sim$data, verbose = 0)
summary(fit_re)
```

INLA parameterizes random effects in terms of precision (the inverse of
variance). High precision means low between-group variation: the groups do
not differ substantially after accounting for the covariates. This is expected
here, since we simulated the data without any grouping structure.

To convert precision to a standard deviation, take $\sigma = 1 / \sqrt{\tau}$.
A precision of 4 corresponds to $\sigma = 0.5$, meaning groups differ by about
1 unit on the logit scale ($\pm 2\sigma$). What does that look like in
practice? If the mean occupancy is 0.5 (logit = 0), a group-level SD of 0.5
shifts the logit to roughly $-1$ and $+1$ at the extremes, giving occupancy
probabilities between about 0.27 and 0.73, which is meaningful ecological variation. A
SD of 0.1, by contrast, keeps all groups between 0.48 and 0.52, which is
negligible. When reporting results, always convert precision to SD for
interpretability. Precision values are useful for INLA's internal
parameterization but carry no ecological intuition.

The supported random effect syntaxes are:

- `(1 | group)`: random intercept per level of `group`

- `(x | group)`: correlated random intercept and slope for covariate `x`

- `(x || group)`: uncorrelated random intercept and slope

- `(1 | a/b)`: nested random effects (random intercept for `b` within `a`)

Random effects work on both the occupancy and detection sides of the model.
For example, `occu(~ elev + (1 | region), ~ (1 | observer), data = dat)`
includes a random intercept for region in occupancy and a random intercept for
observer in detection.


## Model comparison

Selecting among competing model structures is a standard step in any analysis.
INLAocc supports comparison via WAIC (Watanabe-Akaike Information Criterion),
AIC, and BIC. WAIC is the natural choice for Bayesian models fitted with INLA,
as it uses the full posterior predictive distribution rather than a point
estimate.

```{r compare}
m1 <- occu(~ 1, ~ 1, data = sim$data, verbose = 0)
m2 <- occu(~ occ_x1, ~ det_x1, data = sim$data, verbose = 0)
m3 <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim$data, verbose = 0)

compare_models(null = m1, simple = m2, full = m3, criterion = "waic")
```

The table ranks models by WAIC (lower is better). The `delta` column gives the
WAIC difference from the best model, and `weight` gives the Akaike-style model
weight. The full model carrying both covariates should be strongly preferred,
confirming that both `occ_x1` and `occ_x2` contribute meaningful information
about occupancy.

For situations where no single model is clearly dominant, `modelAverage()`
computes WAIC-weighted averages of predictions across models:

```{r model-average}
avg <- modelAverage(null = m1, simple = m2, full = m3, criterion = "waic")
head(avg$psi_hat)  # weighted-average occupancy probabilities
```

Model averaging is particularly useful when several models have similar WAIC
values and you want predictions that account for model uncertainty rather than
conditioning on a single "best" structure.

WAIC approximates out-of-sample predictive accuracy from within-sample data.
It penalizes model complexity through the effective number of parameters
$p_{\text{WAIC}}$, which counts how many parameters the model actually uses
(as opposed to the nominal count). A model with 10 covariates but little data
support for most of them will have an effective parameter count well below 10,
because the posterior barely moves from the prior for those parameters. This
makes WAIC suitable for comparing models of different complexity on a level
playing field. Unlike DIC, WAIC is fully Bayesian and does not depend on a
single point estimate. Unlike cross-validation, it does not require refitting
the model multiple times. The cost is that WAIC relies on the posterior
predictive distribution being a good approximation to the true predictive
distribution, which can break down for very small datasets or heavily
misspecified models.

The choice between model averaging and single-model inference depends on the
goal. When the goal is prediction (forecasting occupancy at new sites or
under future conditions), model averaging is safer because it hedges against
choosing the wrong model. The averaged predictions incorporate uncertainty
about which model is correct, producing wider but better-calibrated prediction
intervals. When the goal is inference on specific coefficients (testing
whether elevation affects occupancy, estimating the magnitude of a land-use
effect), single-model reporting is clearer. Model-averaged coefficients
are a weighted mixture of estimates from models that may define the coefficient
differently (present in some models, absent in others), and the resulting
estimate lacks a clean interpretation.

When interpreting the `delta` column (WAIC difference from the best model),
standard rules of thumb from Burnham and Anderson (2002) apply: a delta below 2
means the models are essentially indistinguishable given the data, and model
averaging is appropriate. A delta between 2 and 7 indicates moderate support for
the better model, though the weaker model may still contribute usefully in an
averaging framework. A delta above 10 is strong evidence that the better model
fits the data substantially more accurately, and the weaker model can be dropped
from consideration. These thresholds are guidelines, not bright lines ---
context and sample size matter.

### K-fold cross-validation

For out-of-sample validation, pass `k.fold` during fitting:

```{r kfold}
fit_cv <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim$data,
               k.fold = 5, verbose = 0)
fit_cv$k.fold
```

This holds out 1/5 of sites at a time, refits on the rest, and evaluates
predictions on the held-out sites. K-fold is slower than WAIC (the model is
refit $k$ times) but gives a direct estimate of out-of-sample predictive
performance.


## Prediction

### Marginal effects

To understand how a single covariate influences occupancy (or detection) while
holding other covariates at their means, use `marginal_effect()`:

```{r marginal-effect, fig.height = 4}
me <- marginal_effect(fit, "occ_x1")
plot(me)
```

This varies `occ_x1` across its observed range, fixes all other covariates at
their sample means, and computes predicted occupancy probabilities with 95%
credible intervals. The `plot()` method produces a ribbon plot with the
posterior mean and credible band. Because `occ_x1` has a negative coefficient,
the plot shows occupancy declining from left to right.

The nonlinearity of the logit link means that the "effect size" on the
probability scale depends on where you are on the curve. Near the center
($\psi \approx 0.5$, logit $\approx 0$), the curve is steep: a unit change in
a covariate produces the largest change in probability. Near the extremes
($\psi$ close to 0 or 1), the curve flattens: the same unit change has a
smaller effect because the probability is already near its boundary. A
coefficient of $-0.8$ drops occupancy from 0.62 to 0.43 starting at a logit
of 0.5 (a 19 percentage point change), but only from 0.12 to 0.07 starting at
a logit of $-2$ (a 5 percentage point change). Marginal effect plots show
this directly: the slope of the curve varies across the covariate range, and
the credible band widens where data are sparse.

A second distinction is between prediction at observed sites and prediction at
new sites. Prediction at observed sites uses the fitted values from the model,
which interpolate within the range and distribution of the training data. This
is safe, since the model has seen data in this region and the predictions are
anchored. Prediction at new sites with covariate values outside the observed
range is extrapolation. The logit link constrains predictions to $[0, 1]$, so
you will never get a negative probability, but the extrapolated values rely
entirely on the assumed linear relationship on the logit scale continuing
beyond the data. If the true relationship is nonlinear (saturating, threshold,
or hump-shaped), extrapolation will be wrong. Report predictions beyond the
covariate range with appropriate caution.

Marginal effects differ from raw coefficient interpretation in an important way.
The coefficients are on the logit scale, where a one-unit change always
corresponds to the same change in log-odds. But on the probability scale, the
effect is nonlinear. A unit increase in a covariate has the largest impact on
probability when the baseline is near 0.5 (the steepest part of the logistic
curve) and a diminishing impact as the baseline approaches 0 or 1. A coefficient
of $-0.8$ drops occupancy from 0.62 to 0.43 when the baseline logit is 0.5, but
only from 0.12 to 0.07 when the baseline logit is $-2$. Marginal effect plots
capture this nonlinearity directly, showing the actual probability change across
the covariate range rather than a constant slope.

### Prediction at new covariate values

For predictions at specific covariate combinations, use `predict()` with a
data frame of new values:

```{r predict-new}
new_sites <- data.frame(occ_x1 = c(-1, 0, 1), occ_x2 = c(0, 0, 0))
predict(fit, X.0 = new_sites, type = "occupancy")
```

The predicted occupancy decreases as `occ_x1` increases, reflecting the
negative effect of this covariate. All predictions are on the probability
scale (inverse-logit of the linear predictor), with quantile-based credible
intervals.

The `terms` argument offers a more concise syntax for marginal prediction
grids, similar to `ggeffects::ggpredict()`:

```{r predict-terms}
predict(fit, terms = "occ_x1 [-2:2 by=0.5]")
```


## Spatial models

When sites are distributed across a landscape, their occupancy states may be
spatially correlated: nearby sites share environmental conditions that
covariates do not fully capture. Ignoring this spatial dependence inflates
confidence in the fixed-effect estimates and can bias predictions.

INLAocc handles spatial autocorrelation by adding a Gaussian random field to
the occupancy linear predictor. The field is approximated via the SPDE
(Stochastic Partial Differential Equations) approach, which constructs a
triangulation mesh over the study area and estimates the spatial correlation
structure from the data.

To fit a spatial model, pass coordinates to the `spatial` argument:

```{r spatial-sim}
sim_sp <- simulate_occu(N = 300, J = 4, n_occ_covs = 1, n_det_covs = 1,
                        beta_occ = c(0.5, -0.8),
                        beta_det = c(0, -0.5),
                        spatial_range = 0.3, seed = 42)
```

First, check whether spatial structure is present by fitting a non-spatial
model and testing the residuals:

```{r spatial-moran}
fit_nospatial <- occu(~ occ_x1, ~ det_x1, data = sim_sp$data, verbose = 0)
moranI(fit_nospatial)
```

A significant p-value (below 0.05) means the residuals are more spatially
clustered than expected under independence. The non-spatial model treats each
site as exchangeable, but in reality nearby sites share unmeasured conditions
(soil type, microclimate, disturbance history) that make their occupancy
states more similar than the covariates alone predict. This leftover
correlation violates the independence assumption and makes the standard errors
on the fixed effects too narrow.

Now fit the spatial model:

```{r spatial-fit}
fit_spatial <- occu(~ occ_x1, ~ det_x1, data = sim_sp$data,
                    spatial = sim_sp$data$coords, verbose = 0)
summary(fit_spatial)
```

The spatial output includes two additional parameters beyond the fixed effects.
The **Range** parameter is the distance at which spatial correlation drops to
roughly 0.1 (the "practical range" of the Matern field). If Range is estimated
at 0.3 in a unit square, sites closer than 0.3 coordinate units are strongly
correlated and sites farther apart are nearly independent. The **Stdev**
parameter quantifies how much spatial variation exists after accounting for the
covariates. A large Stdev relative to the fixed-effect magnitudes means the
spatial field dominates the occupancy surface; a small Stdev means the
covariates explain most of the pattern and the spatial component is a minor
correction.

Compare the fixed effects between the two fits. The spatial model's credible
intervals for `occ_x1` should be wider than those from the non-spatial model.
This is not a loss of precision; it is more honest uncertainty. The
non-spatial model underestimated uncertainty by ignoring the positive
correlation between nearby sites, which inflated the effective sample size.
The spatial model corrects this. Verify that the spatial field absorbed the
autocorrelation:

```{r spatial-moran-check}
moranI(fit_spatial)
```

A non-significant Moran's I (p-value above 0.05) confirms that the spatial
model captured the autocorrelation structure. The residuals now behave as if
they came from spatially independent sites, which is what the downstream
inference assumes.

Compare the two models:

```{r spatial-compare}
compare_models(non_spatial = fit_nospatial, spatial = fit_spatial, criterion = "waic")
```

The spatial model should have a lower WAIC, confirming that the spatial
component is justified by the data. A large delta WAIC (above 10) is typical
when the true data-generating process includes spatial correlation, since the
non-spatial model pays a heavy price for misattributing spatially structured
variation to noise.

### Mapping occupancy

The spatial model estimates a continuous occupancy surface across the study
area. The `occuMap()` function projects this surface onto a regular grid and
overlays the observed detection status at each site, producing a
publication-quality map in a single call:

```{r spatial-map, fig.height = 5}
occuMap(fit_spatial)
```

The color gradient shows predicted occupancy probability interpolated between
survey locations. Filled circles mark sites where the species was detected at
least once; open circles mark sites where it was never observed. Sites with
high predicted occupancy but no detections are candidates for additional
survey effort, because the model believes the species is likely present but
imperfect detection means it was missed.

The map also reveals spatial patterns that the covariates alone cannot
explain. If the occupancy surface varies smoothly across the landscape in
ways that do not track the included covariates, the spatial random field is
picking up unmeasured environmental gradients (soil type, microclimate,
historical land use) that influence occupancy. This is useful information:
it points toward covariates that could improve the model if they were
measured, and it flags regions where extrapolation from covariates alone
would be misleading.

Pass `type = "sd"` to map posterior uncertainty instead of the mean, or
`type = "all"` for a side-by-side panel of both. For full control over the
SPDE mesh, PC priors for the spatial range and variance, areal (CAR/BYM2)
models, and spatially varying coefficients, see
`vignette("spatial-models")`.


## Multi-species models

When multiple species are surveyed at the same sites, fitting each species
independently wastes information. A community occupancy model shares
information across species through a hierarchical prior: species-level
coefficients are drawn from a common distribution, which acts as a regularizer.
Rare species with few detections benefit most, since their estimates are pulled
toward the community mean rather than being driven to extreme values by sparse
data.

```{r ms-sim}
sim_ms <- simMsOcc(N = 200, J = 4, n_species = 10, seed = 42)
```

Fit the community model by adding `multispecies = TRUE`:

```{r ms-fit}
fit_ms <- occu(~ occ_x1, ~ det_x1, data = sim_ms$data,
               multispecies = TRUE, verbose = 0)
summary(fit_ms)
```

The summary reports two layers of output. The community-level means
($\mu_\beta$, $\mu_\alpha$) describe the response of a typical species in the
assemblage: a positive community mean for `occ_x1` means species in this
community tend to occur more frequently where `occ_x1` is high. The
community-level standard deviations ($\sigma_\beta$, $\sigma_\alpha$) describe
how much species differ from this average. A large $\sigma$ means species
respond heterogeneously to the covariate, some positively, some negatively.
Below the community summary, per-species estimates show each species' own
intercept and slope.

The shrinkage effect is visible in the per-species estimates. Species detected
at many sites have estimates close to their individual maximum-likelihood
values, because the data are informative enough that the community prior barely
pulls them. Species detected at only a handful of sites have estimates pulled
substantially toward the community mean. Without this shrinkage, a species
detected at 3 out of 200 sites would get an occupancy estimate near zero with
a huge standard error. The hierarchical model borrows strength from the other
9 species to produce a more stable estimate, at the cost of some bias toward
the community average. This bias-variance tradeoff is almost always favorable
when species have fewer than about 20 detections.

Estimated species richness at each site accounts for imperfect detection:

```{r ms-richness}
r <- richness(fit_ms)
head(r)
```

Richness here is the sum of estimated occupancy probabilities across species at
each site. A site where 7 species were detected but estimated richness is 8.3
has a detection shortfall of about 1.3 species; at least one species is
likely present but was not observed during any visit. Sites where the gap
between naive richness (count of detected species) and estimated richness is
large are candidates for additional survey effort.

A histogram of estimated richness across sites shows the spatial distribution
of diversity:

```{r ms-richness-hist, fig.height = 4}
hist(r$richness, breaks = 15, col = "steelblue", border = "white",
     main = "Estimated species richness", xlab = "Species richness",
     ylab = "Number of sites")
abline(v = mean(r$richness), lty = 2, lwd = 2, col = "firebrick")
```

The dashed line marks the mean estimated richness across sites. The spread
of the histogram reflects spatial heterogeneity in the community: sites in
favorable habitat support more species, while sites at the margins harbor
fewer. This distribution is already corrected for imperfect detection ---
the naive species count (number of species detected at least once) would
underestimate richness at every site, shifting the entire histogram to the
left.

A coefficient forest plot shows how species differ in their responses:

```{r ms-plot, fig.height = 6}
plot(fit_ms)
```

The forest plot orders species by their estimated occupancy intercept (or slope,
depending on the panel). Each horizontal line is a 95% credible interval for
one species' coefficient. Species whose intervals cross zero show no detectable
response to the covariate. The spread of the intervals around the community
mean (vertical dashed line) reflects the community-level standard deviation. A
tight cluster means species respond similarly; a wide spread means the
community-level covariate effect masks substantial species-level heterogeneity.

For latent factor models, spatial factor models, and JSDMs, see
`vignette("multi-species")`.


## Integrated models

When a species is monitored by more than one survey program (structured
point counts at a fixed grid of sites, camera traps running continuously at a
different set of locations, citizen-science checklists scattered across the
broader landscape), each program captures the same underlying occupancy
state through a different detection filter. Analysing the programs separately
wastes information. An integrated occupancy model links them through a shared
occupancy layer while estimating source-specific detection probabilities.

The `simIntOcc()` function generates data from two sources that differ in
detection and in the sites they cover:

```{r int-sim}
sim_int <- simIntOcc(
  N_total  = 200,
  n_data   = 2,
  J        = c(4, 3),
  n_shared = 30,
  beta_occ = c(0.5, 0.3),
  beta_det = list(c(0.2, -0.4), c(-0.5, 0.3)),
  seed     = 42
)
```

Source 1 conducts 4 visits per site with moderately high baseline detection
(intercept 0.2 on the logit scale, roughly 55%). Source 2 conducts only 3
visits with lower baseline detection (intercept $-0.5$, roughly 38%). The two
sources also respond differently to their detection covariates: source 1's
detection degrades with its covariate ($-0.4$), while source 2's detection
improves ($+0.3$). Thirty sites are surveyed by both programs, providing
direct calibration between the two detection processes.

Fit the integrated model by setting `integrated = TRUE` and passing a list
of per-source detection formulas:

```{r int-fit}
fit_int <- occu(~ occ_x1, det.formula = list(~ det_x1, ~ det_x1),
                data = sim_int$data, integrated = TRUE, verbose = 0)
summary(fit_int)
```

The output reports a single set of occupancy coefficients (the shared
ecological state) alongside separate detection summaries for each source.
Compare the detection intercepts: source 1 should be estimated near 0.2 and
source 2 near $-0.5$, reflecting their different survey protocols. The
occupancy estimate benefits from the combined information: sites surveyed
by both programs contribute detections from both, and sites surveyed by only
one program still borrow strength from the shared occupancy layer.

To see the benefit of integration, compare to a single-source analysis. Fit
source 1 alone:

```{r int-compare}
data_s1 <- list(
  y        = sim_int$data$y[[1]],
  occ.covs = sim_int$data$occ.covs[sim_int$data$sites[[1]], , drop = FALSE],
  det.covs = sim_int$data$det.covs[[1]]
)
fit_s1 <- occu(~ occ_x1, ~ det_x1, data = data_s1, verbose = 0)

cat(sprintf("Integrated occ_x1: %.3f (%.3f, %.3f)\n",
            coef(fit_int, "occupancy")[2],
            confint(fit_int, "occupancy")[2, 1],
            confint(fit_int, "occupancy")[2, 2]))
cat(sprintf("Source 1 only occ_x1: %.3f (%.3f, %.3f)\n",
            coef(fit_s1, "occupancy")[2],
            confint(fit_s1, "occupancy")[2, 1],
            confint(fit_s1, "occupancy")[2, 2]))
```

The integrated model's credible interval for the occupancy slope should be
narrower, because the extra information from source 2 tightens the estimate. The
gain is largest when the two sources cover complementary regions of the
landscape, because the integrated model can estimate occupancy at sites
surveyed by only one program while using the shared sites to calibrate the
detection processes against each other.

For the full treatment of integrated models, including spatial integration,
per-source random effects, and diagnostics for multi-source data, see
`vignette("integrated-models")`.


## A realistic simulation

The examples above use abstract covariates drawn from standard normal
distributions. A more realistic scenario helps build intuition for what these
numbers mean in a field setting. Consider a single-species survey for a
forest-dwelling bird across a fragmented landscape:

```{r realistic}
set.seed(7)
N_real <- 150
J_real <- 3

# Site-level covariates: proportion of forest cover (0-1) and elevation (m)
forest <- runif(N_real, 0, 1)
elev   <- rnorm(N_real, 800, 200)

# Visit-level covariate: survey duration in minutes (varies by visit)
duration <- matrix(rnorm(N_real * J_real, 60, 15), N_real, J_real)

# Standardize for modeling
forest_s   <- scale(forest)[, 1]
elev_s     <- scale(elev)[, 1]
duration_s <- (duration - mean(duration)) / sd(duration)

# Ecological truth: species prefers forests, avoids high elevation
psi_logit <- 0.2 + 1.2 * forest_s - 0.6 * elev_s
psi_true  <- 1 / (1 + exp(-psi_logit))
z_true    <- rbinom(N_real, 1, psi_true)

# Detection: improves with longer surveys
p_logit  <- -0.3 + 0.5 * duration_s
p_true   <- 1 / (1 + exp(-p_logit))
y_real   <- matrix(NA, N_real, J_real)
for (i in seq_len(N_real)) {
  for (j in seq_len(J_real)) {
    y_real[i, j] <- rbinom(1, 1, z_true[i] * p_true[i, j])
  }
}

real_data <- occu_format(
  y = y_real,
  occ.covs = data.frame(forest = forest_s, elev = elev_s),
  det.covs = list(duration = duration_s)
)

fit_real <- occu(~ forest + elev, ~ duration, data = real_data, verbose = 0)
summary(fit_real)
```

The occupancy coefficients should recover the generating values: a strong
positive effect of forest cover ($\beta \approx 1.2$) and a moderate
negative effect of elevation ($\beta \approx -0.6$). The detection intercept
of $-0.3$ (about 43% per visit at mean duration) and the positive duration
slope confirm that longer surveys improve detection, exactly as a field
ecologist would expect. This kind of sanity check (simulating from
parameters that match your study system, then verifying recovery) is worth
doing before fitting any real-data model. If the model cannot recover known
parameters under realistic conditions, it will not produce trustworthy
estimates from your actual survey.


## S3 methods

Fitted `occu_inla` objects support the standard R model interface, making them
interoperable with tools you already know:

| Method | Description |
|:-------|:------------|
| `summary(fit)` | Full posterior summary with model fit statistics |
| `coef(fit)` | Posterior means of fixed effects |
| `confint(fit)` | Credible intervals for fixed effects |
| `fitted(fit)` | Fitted occupancy and detection probabilities |
| `residuals(fit)` | Residuals (Pearson or PIT) |
| `predict(fit, X.0 = ...)` | Prediction at new covariate values |
| `plot(fit)` | Four-panel diagnostic plot (EM trace, psi, p, z vs. naive) |
| `logLik(fit)` | Marginal log-likelihood |
| `nobs(fit)` | Number of sites |
| `vcov(fit)` | Posterior variance-covariance matrix |
| `simulate(fit)` | Simulate new detection histories from the fitted model |
| `update(fit, ~ . - occ_x2)` | Refit with a modified formula |
| `tidy(fit)` / `glance(fit)` | Tidy summaries (broom-compatible) |
| `ranef(fit)` | Extract random effect posteriors |


## What's next

This vignette covered the core single-species workflow. A recommended analysis
sequence: start with the intercept-only model to establish baseline occupancy
and detection rates, then add covariates guided by ecological hypotheses. Check
diagnostics at each step. A model that fits well without covariates but fails
after adding them suggests a misspecified functional form (e.g., a quadratic
effect forced through a linear term). Add random effects only when the study
design includes genuine grouping structure, not as a generic variance sponge.
Compare models with WAIC, and report marginal effects rather than raw
coefficients for biological interpretation.

INLAocc supports substantially more complex models through the same `occu()`
interface:

- **Spatial models.** When sites are spatially correlated, add an SPDE random
  field by passing coordinates to the `spatial` argument. See
  `vignette("spatial-models")` for the full treatment, including mesh
  construction, spatially varying coefficients, and spatial prediction maps.

- **Multi-species models.** Community occupancy models that share information
  across species, with optional latent factor structure. See
  `vignette("multi-species")`.

- **Integrated models.** When multiple survey programs monitor the same species,
  combine them into a single analysis with source-specific detection. See
  `vignette("integrated-models")`.

- **Diagnostics deep dive.** The full suite of simulation-based diagnostics,
  including spatial residual analysis, temporal autocorrelation tests, and
  posterior predictive checks. See `vignette("diagnostics")`.

- **Migrating from spOccupancy.** If you are switching from an MCMC-based
  workflow, `vignette("spoccupancy-migration")` maps spOccupancy's function
  names, data formats, and output structures to their INLAocc equivalents.
