---
title: "Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Fitting a model

Simulate data and fit a single-species occupancy model:

```{r basic, eval = FALSE}
library(INLAocc)

sim <- simulate_occu(N = 200, J = 4, seed = 42)
fit <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim$data)
summary(fit)
```

First formula is occupancy, second is detection. If you omit the detection
formula, it defaults to `~ 1` with a warning.

## Exploring the data first

Before fitting, check what you're working with:

```{r explore, eval = FALSE}
summary(sim$data)
plot(sim$data)
```

`summary()` gives detection frequencies, naive occupancy and detection rates,
visit completeness, and per-visit detection rates. `plot()` produces a panel
with histograms and (if coordinates exist) a spatial detection map.

## Random effects

Random intercepts and slopes go directly in the formula:

```{r re, eval = FALSE}
sim$data$occ.covs$region <- sample(1:5, 200, replace = TRUE)

fit_re <- occu(
  ~ occ_x1 + (1 | region),
  ~ det_x1,
  data = sim$data
)
ranef(fit_re)
```

Also supported: `(x | group)` for random slopes, `(x || group)` for
uncorrelated slopes, and `(1 | a/b)` for nested grouping.

## Spatial models

Pass coordinates to add an SPDE spatial random effect:

```{r spatial, eval = FALSE}
sim_sp <- simulate_occu(N = 200, J = 4, spatial_range = 0.2, seed = 123)

fit_sp <- occu(~ occ_x1, ~ det_x1, data = sim_sp$data,
               spatial = sim_sp$data$coords)
summary(fit_sp)
```

The mesh is auto-scaled from the coordinate extent. To control resolution
(units match your CRS, so metres for UTM):

```{r mesh, eval = FALSE}
fit_sp <- occu(~ occ_x1, ~ det_x1, data = sim_sp$data,
               spatial = sim_sp$data$coords,
               spde.args = list(max.edge = c(5000, 15000)))
```

After fitting, extract the estimated spatial range:

```{r range, eval = FALSE}
spatialRange(fit_sp)
```

This returns the posterior mean, SD, and 95% credible interval for the SPDE
range parameter, in the same units as your coordinates.

## Diagnostics

`checkModel()` produces a diagnostic panel in one call:

```{r check, eval = FALSE}
checkModel(fit_sp)
```

Four panels: QQ plot of PIT residuals against Uniform(0,1), residuals vs
fitted, dispersion histogram, and a Moran's I correlogram (when coordinates
are available).

Each test is also available on its own:

```{r diag, eval = FALSE}
testUniformity(fit)     # KS test on PIT residuals
testDispersion(fit)     # over/underdispersion
testOutliers(fit)       # simulation envelope
testZeroInflation(fit)  # excess unoccupied sites
moranI(fit_sp)          # spatial autocorrelation in residuals
variogram(fit_sp)       # empirical semivariogram
```

All return `htest` objects with `$p.value`.

If you have DHARMa installed:

```{r dharma, eval = FALSE}
res <- dharma(fit)
plot(res)
```

## Model comparison and averaging

```{r compare, eval = FALSE}
m1 <- occu(~ 1, ~ 1, data = sim$data)
m2 <- occu(~ occ_x1, ~ det_x1, data = sim$data)
m3 <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim$data)

compare_models(null = m1, elev = m2, full = m3, criterion = "aic")
```

This returns AIC (or BIC/WAIC), delta, and model weights.

For model-averaged predictions:

```{r avg, eval = FALSE}
avg <- modelAverage(null = m1, elev = m2, full = m3, criterion = "aic")
avg$psi_hat   # weighted-average occupancy
avg$psi_se    # unconditional SEs (Burnham & Anderson eq. 4.9)
```

## Prediction

```{r predict, eval = FALSE}
# Vary one covariate over a grid, hold others at their mean
predict(fit, terms = "occ_x1 [-2:2 by=0.5]")

# Design-matrix prediction at new sites
predict(fit, X.0 = data.frame(occ_x1 = 0, occ_x2 = 1))

# Marginal effect curve
marginal_effect(fit, "occ_x1")
```

## Multi-species models

```{r ms, eval = FALSE}
sim_ms <- simMsOcc(N = 100, J = 3, n_species = 10, seed = 200)

fit_ms <- occu(~ occ_x1, ~ det_x1, data = sim_ms$data, multispecies = TRUE)
summary(fit_ms)
richness(fit_ms)
```

## Multi-season models

```{r temporal, eval = FALSE}
sim_t <- simTOcc(N = 50, J = 3, n_seasons = 5, seed = 300)

fit_t <- occu(~ 1, ~ 1, data = sim_t$data, temporal = "ar1")
summary(fit_t)
temporalCorr(fit_t)
```

## S3 methods

```{r s3, eval = FALSE}
coef(fit)               # posterior means
confint(fit)            # credible intervals
tidy(fit)               # broom-compatible data.frame
glance(fit)             # one-row model summary
AIC(fit); BIC(fit)      # information criteria
update(fit, ~ . - occ_x2)  # refit without occ_x2
```
