---
title: "Random Effects in Occupancy Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Random Effects in Occupancy Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

Ecological survey data are almost always grouped. Sites are clustered within
regions, visits are conducted by different observers, and species detections
depend on local conditions that vary across spatial or administrative units.
When these groups share unmeasured characteristics that affect occupancy or
detection, ignoring the grouping structure leads to underestimated uncertainty
and, in some cases, biased parameter estimates. Random effects provide a
principled way to account for this structure.

A random effect allows intercepts or slopes to vary across groups while
borrowing strength through partial pooling. Instead of estimating a completely
separate parameter for each group (fixed effects, which is wasteful with many
groups) or ignoring groups entirely (which is wrong when they matter), the
random effects model assumes group-level deviations are drawn from a common
distribution. Groups with little data are pulled toward the overall mean; groups
with abundant data are allowed to deviate more freely. This shrinkage is
automatic and data-driven.

INLAocc supports random effects in both the occupancy and detection sub-models.
The formula interface follows the same syntax as lme4 and glmmTMB:
`(1 | group)` for random intercepts, `(x | group)` for correlated random
slopes, `(x || group)` for uncorrelated slopes, and `(1 | a/b)` for nested
grouping. For finer control, the `occu_re()` function allows explicit
specification of the latent model and prior. All random effects are estimated
via INLA using penalised complexity (PC) priors that shrink toward the base
model (no random effect) unless the data provide evidence otherwise.


## The random effects model

### Random intercept

The standard occupancy model with a random intercept on the occupancy process
is:

$$
\text{logit}(\psi_i) = \mathbf{x}_i^\top \boldsymbol{\beta} + u_{g[i]}, \quad u_g \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2_u)
$$

where $g[i]$ maps site $i$ to its group (e.g., region, management unit, or
observer). The random effect $u_g$ shifts the baseline log-odds of occupancy
up or down for each group. With many groups, the individual $u_g$ estimates
are shrunk toward zero (the overall mean on the link scale), which is the
partial pooling property.

### Random slope

When the effect of a covariate varies across groups, the model becomes:

$$
\text{logit}(\psi_i) = \beta_0 + (\beta_1 + v_{g[i]}) x_{1i}, \quad v_g \stackrel{iid}{\sim} \mathcal{N}(0, \sigma^2_v)
$$

Here, $\beta_1$ is the average effect of $x_1$ across all groups, and $v_g$
captures group-specific deviations from that average. A large $\sigma^2_v$
means the covariate effect varies substantially across groups; a small
$\sigma^2_v$ means it is approximately constant.


## Simulating data with random effects

INLAocc's `simulate_occu()` generates site-level random effects. To create
grouped data suitable for demonstrating random intercepts, we simulate the data
and then assign sites to groups:

```{r simulate}
library(INLAocc)

sim <- simulate_occu(
  N = 300, J = 4,
  beta_occ = c(0.5, -0.8),
  beta_det = c(0, -0.5),
  n_occ_covs = 1, n_det_covs = 1,
  random_occ_sd = 1.0,
  seed = 42
)

# Assign sites to 10 groups
n_groups <- 10
sim$data$occ.covs$group <- rep(seq_len(n_groups), length.out = 300)
```

The true random effect standard deviation is 1.0, which is substantial ---
on the logit scale, a shift of $\pm 1$ moves occupancy probability from 0.62
to 0.27 or 0.73 when the baseline is at 0.5.

Explore the data to confirm the group structure:

```{r explore}
table(sim$data$occ.covs$group)
```


## Random intercepts

### Formula syntax

Random intercepts use the `(1 | group)` syntax inside the occupancy or
detection formula, exactly as in lme4:

```{r fit-re}
fit_re <- occu(~ occ_x1 + (1 | group), ~ det_x1, data = sim$data, verbose = 0)
summary(fit_re)
```

The summary includes a hyperparameters section reporting the precision of each
random effect. Precision is the inverse of variance: $\tau = 1 / \sigma^2$.
The estimated precision should imply a standard deviation close to the true
value of 1.0. High precision means little between-group variation; low
precision means groups differ substantially.

### Extracting random effects

The `ranef()` method extracts posterior summaries of the group-level deviations:

```{r ranef}
re <- ranef(fit_re)
re$occupancy
```

A positive random effect means that group has higher occupancy than the overall
average (after accounting for covariates); a negative value means lower
occupancy. These are analogous to BLUPs (best linear unbiased predictors) in
the frequentist mixed model literature, though in the Bayesian framework they
are full posterior distributions.

### Comparison: with vs. without random effects

Model comparison via WAIC confirms whether the grouping structure matters:

```{r compare}
fit_no_re <- occu(~ occ_x1, ~ det_x1, data = sim$data, verbose = 0)
compare_models(no_re = fit_no_re, with_re = fit_re, criterion = "waic")
```

When the grouping structure is real, the model with random effects typically has
a substantially lower WAIC. The effective number of parameters (`p_waic`)
increases modestly because partial pooling keeps the random effects from
consuming a full degree of freedom per group.


## Random slopes

### Correlated slopes: `(x | group)`

When both the intercept and a covariate slope vary by group, and these two
sources of variation may be correlated, use:

```{r slope-corr, eval = FALSE}
fit_slope <- occu(~ occ_x1 + (occ_x1 | group), ~ det_x1, data = sim$data, verbose = 0)
```

This specifies the joint model:

$$
\begin{pmatrix} u_g \\ v_g \end{pmatrix} \sim \mathcal{N}\left(\mathbf{0}, \begin{pmatrix} \sigma^2_u & \rho \sigma_u \sigma_v \\ \rho \sigma_u \sigma_v & \sigma^2_v \end{pmatrix}\right)
$$

where $u_g$ is the random intercept and $v_g$ the random slope for group $g$.
A negative correlation ($\rho < 0$) would mean that groups with high baseline
occupancy tend to have a weaker covariate effect, and vice versa.

### Uncorrelated slopes: `(x || group)`

If you do not expect --- or cannot estimate --- the correlation between intercept
and slope variation, use the double-bar syntax:

```{r slope-uncorr, eval = FALSE}
fit_slope_uncor <- occu(~ occ_x1 + (occ_x1 || group), ~ det_x1, data = sim$data, verbose = 0)
```

This fits independent random intercepts and random slopes (i.e., $\rho = 0$
by construction). The uncorrelated specification is computationally simpler and
requires fewer data to estimate reliably. In practice, if you have fewer than
about 20 groups, the correlation parameter is often poorly identified and the
uncorrelated model is the safer choice.


## Nested random effects

When sites are nested within higher-level units (e.g., plots within regions),
the formula syntax supports nested grouping with `/`:

```{r nested, eval = FALSE}
# Create a nested structure
sim$data$occ.covs$region <- rep(1:3, each = 100)
sim$data$occ.covs$site_id <- seq_len(300)

fit_nested <- occu(~ occ_x1 + (1 | region/site_id), ~ det_x1,
                   data = sim$data, verbose = 0)
```

The syntax `(1 | region/site_id)` expands internally to
`(1 | region) + (1 | region:site_id)`, fitting two variance components: one for
between-region variation and one for within-region, between-site variation. This
is the standard nested random effects parameterisation used in lme4.


## Detection random effects

Random effects on the detection process handle variation in observer skill,
survey conditions across time periods, or equipment differences:

```{r det-re, eval = FALSE}
# Simulate observer IDs (5 observers, assigned randomly to visits)
sim$data$det.covs$observer <- matrix(
  sample(1:5, 300 * 4, replace = TRUE), nrow = 300, ncol = 4
)

fit_det_re <- occu(~ occ_x1, ~ det_x1 + (1 | observer), data = sim$data, verbose = 0)
```

Extract detection-level random effects with:

```{r det-ranef, eval = FALSE}
ranef(fit_det_re, process = "detection")
```

Observers with positive random effects have higher detection rates; those with
negative values have lower rates. Accounting for observer heterogeneity prevents
it from being absorbed into the occupancy estimates, which would bias the
ecological conclusions.


## Using `occu_re()` for explicit specification

The formula syntax covers the most common cases. When you need more control ---
a non-default latent model, a custom prior, or a sum-to-zero constraint ---
use `occu_re()` directly and pass the result to the `occ.re` or `det.re`
argument of `occu()`:

```{r occu-re, eval = FALSE}
re_spec <- occu_re(
  type  = "intercept",
  group = "group",
  model = "iid",
  prior = list(prec = list(prior = "pc.prec", param = c(1, 0.05)))
)

fit_explicit <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                     occ.re = re_spec, verbose = 0)
```

The `model` argument accepts any INLA latent model name. For standard random
intercepts, `"iid"` (independent and identically distributed) is the default.
When groups have a natural ordering (e.g., years, age classes), structured
models are available:

- `"iid"` --- exchangeable (default). No ordering assumed.
- `"ar1"` --- first-order autoregressive. Adjacent groups are correlated.
- `"rw1"` --- random walk of order 1. Smooth trend across ordered groups.
- `"rw2"` --- random walk of order 2. Smoother than rw1.

```{r ar1, eval = FALSE}
# Temporal random effect with AR(1) structure
sim$data$occ.covs$year <- rep(2010:2019, each = 30)

re_temporal <- occu_re(
  type  = "intercept",
  group = "year",
  model = "ar1",
  prior = list(prec = list(prior = "pc.prec", param = c(1, 0.05)))
)

fit_temporal <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                     occ.re = re_temporal, verbose = 0)
```


## Priors on random effects

### Default PC prior

INLAocc uses penalised complexity (PC) priors for random effect precision by
default. The default is `P(sigma > 1) = 0.05`, meaning there is a 5% prior
probability that the standard deviation exceeds 1 on the linear predictor
(logit) scale. This is moderately conservative: it encodes a prior belief that
the random effect is unlikely to dominate the linear predictor, while remaining
flexible enough to let the data speak.

### Adjusting the prior

When the default is too tight or too loose, specify a custom prior through
`occu_re()`:

```{r priors}
# More permissive: P(sigma > 2) = 0.1
re_loose <- occu_re(
  type  = "intercept",
  group = "group",
  prior = list(prec = list(prior = "pc.prec", param = c(2, 0.1)))
)

# Tighter: P(sigma > 0.5) = 0.01
re_tight <- occu_re(
  type  = "intercept",
  group = "group",
  prior = list(prec = list(prior = "pc.prec", param = c(0.5, 0.01)))
)
```

Situations where adjusting the prior matters:

- **Few groups (<5).** The data alone cannot pin down the variance well. The
  prior has outsized influence, so ensure it reflects genuine prior knowledge.
- **Many sites per group (>100).** The likelihood dominates; the prior matters
  less. A wider prior is safe.
- **Strong prior knowledge.** If previous studies consistently report small
  between-region variation, a tighter prior is justified and improves
  estimation.
- **Covariates on different scales.** If covariates are not standardised, a
  random slope prior of $\sigma = 1$ might be too permissive or too tight
  depending on the covariate's range. Standardise covariates first, or adjust
  the prior to match.


## How many groups do you need?

Practical guidance for the number of groups:

| Groups   | Recommendation |
|----------|---------------------------------------------------------------|
| < 3      | Use fixed effects. Not enough groups to estimate a variance.  |
| 3--5     | Random effects are possible but the variance estimate will be noisy. Consider fixed effects unless partial pooling is specifically desired. |
| 5--20    | The sweet spot for random intercepts. Enough groups for reliable variance estimation; partial pooling provides clear benefits. |
| > 20     | Well-estimated. Can support random slopes and correlated structures. |

With few groups, the individual group-level estimates (the $u_g$) are still
partially pooled, which can be useful. But the variance parameter $\sigma^2$
itself is weakly identified, meaning it is sensitive to the prior and its
posterior is wide. With 3--5 groups, always conduct a sensitivity analysis by
fitting the model under two or three different priors and checking whether
conclusions change.


## Diagnostics for random effects

### Overall model check

```{r diagnostics}
checkModel(fit_re)
```

The `checkModel()` function runs a suite of posterior predictive checks. For
models with random effects, it verifies that the random effect structure
adequately captures the between-group variation. See `vignette("diagnostics")`
for a detailed walkthrough.

### Dispersion

Overdispersion in occupancy models often signals unmodelled heterogeneity. A
random effect can absorb this extra variation:

```{r dispersion}
testDispersion(fit_no_re)  # may show overdispersion
testDispersion(fit_re)     # RE should reduce it
```

If overdispersion persists after adding random effects, the grouping structure
may not capture the relevant source of heterogeneity, or additional covariates
may be needed.

### Identifiability

When random effects are complex (crossed random slopes, nested structures),
check that all components are identifiable:

```{r identifiability}
checkIdentifiability(fit_re)
```

This function flags random effect terms whose posterior variance is dominated
by the prior, indicating that the data provide little information about that
component.


## Practical guidance

1. **Start simple.** Fit the model without random effects first. Add them only
   if diagnostics (dispersion, residual patterns by group) suggest grouped
   variation that covariates do not capture.

2. **Scale continuous covariates.** Before fitting random slopes, centre and
   scale all continuous covariates. This stabilises estimation and makes the
   prior on the slope variance interpretable.

3. **Keep detection simple.** Random slopes on detection are rarely needed and
   are difficult to identify. Start with random intercepts if observer or
   temporal variation in detection is suspected.

4. **Prefer uncorrelated slopes when groups are few.** The correlation between
   random intercept and slope (`(x | group)`) requires substantial data to
   estimate. With fewer than 20 groups, `(x || group)` is usually the safer
   choice.

5. **Use structured models for ordered groups.** If your grouping factor has a
   natural ordering (years, age classes, elevation bands), an `"ar1"` or
   `"rw1"` model in `occu_re()` is more efficient than `"iid"` because it
   borrows strength from neighbouring groups.

6. **Convergence issues.** If the EM algorithm is slow to converge with random
   effects, try: (a) reducing model complexity (drop correlated slopes),
   (b) increasing damping (`damping = 0.5`), or (c) providing more informative
   priors on the random effect precision.

7. **Report both fixed and random.** When presenting results, report the fixed
   effect coefficients, the random effect variance ($\sigma^2$ or its posterior
   summary), and --- if relevant --- the group-level predictions from `ranef()`.

For the general model-fitting workflow, see `vignette("quickstart")`. For
spatial random effects (continuous Gaussian random fields rather than discrete
groups), see `vignette("spatial-models")`.
