---
title: "Spatial Occupancy Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Occupancy Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

Species detection/non-detection data are collected at sites distributed across
a landscape, and nearby sites tend to share unmeasured environmental conditions.
When this spatial dependence is ignored, occupancy models treat sites as
independent replicates --- an assumption that is almost always wrong. The
consequences are real: residual spatial autocorrelation inflates effective
sample size, producing confidence intervals that are too narrow and
fixed-effect estimates that may be biased. If a covariate of interest is
itself spatially structured, the bias can be substantial.

INLAocc handles spatial structure through the SPDE (Stochastic Partial
Differential Equations) approximation to a continuous Gaussian random field.
Rather than working with a dense $N \times N$
covariance matrix (which becomes intractable for large $N$), the SPDE approach
represents the spatial field as a piecewise linear function over a
triangulation mesh. This yields a sparse precision matrix that INLA can
exploit for fast, exact inference. The mesh is built automatically from site
coordinates, or you can construct it manually for full control.

This vignette covers the full spatial modelling workflow: fitting SPDE models,
tuning the mesh and PC priors, mapping occupancy, predicting at new locations,
checking spatial residuals, areal (CAR/BYM2) models for regional data,
spatially varying coefficients, and space-time extensions. For general model
diagnostics beyond the spatial context, see `vignette("diagnostics")`.


## The spatial occupancy model

The standard single-season occupancy model with a spatial random effect is:

$$z_i \sim \text{Bernoulli}(\psi_i), \quad \text{logit}(\psi_i) = \mathbf{x}_i^\top \boldsymbol{\beta} + w(\mathbf{s}_i)$$

$$y_{ij} \mid z_i \sim \text{Bernoulli}(z_i \cdot p_{ij}), \quad \text{logit}(p_{ij}) = \mathbf{w}_{ij}^\top \boldsymbol{\alpha}$$

where $z_i$ is the true (latent) occupancy state at site $i$, $\psi_i$ is the
occupancy probability, $y_{ij}$ is the observed detection/non-detection at
site $i$ during visit $j$, and $p_{ij}$ is the detection probability.

The spatial component $w(\mathbf{s})$ is a Gaussian random field (GRF) with
zero mean and Matern covariance function:

$$\text{Cov}(w(\mathbf{s}_1), w(\mathbf{s}_2)) = \sigma^2 \frac{2^{1-\nu}}{\Gamma(\nu)} \left(\kappa \|\mathbf{s}_1 - \mathbf{s}_2\|\right)^\nu K_\nu\left(\kappa \|\mathbf{s}_1 - \mathbf{s}_2\|\right)$$

The three parameters have intuitive interpretations:

- $\sigma^2$ is the **marginal variance** of the spatial field --- how much
  spatial variation there is on the logit-occupancy scale.
- $\kappa$ is a **scale parameter** that relates to the practical correlation
  range via $\rho = \sqrt{8\nu}/\kappa$. The practical range $\rho$ is the
  distance at which the Matern correlation drops to approximately 0.1.
- $\nu$ controls **smoothness**. For the 2D SPDE approximation, $\nu$ is
  fixed at 1 (corresponding to $\alpha = 2$), giving a once-differentiable
  field.

**The SPDE trick.** Rather than evaluating the dense covariance matrix, the
SPDE approach solves a stochastic PDE whose stationary solution has the Matern
covariance. The solution is approximated as a piecewise linear function over a
triangulation mesh, yielding a sparse precision matrix $\mathbf{Q}$ with
$O(n)$ non-zero entries (where $n$ is the number of mesh nodes). This
sparsity is what makes INLA fast: the Cholesky factorisation of
$\mathbf{Q}$ scales as $O(n^{3/2})$ in 2D, compared to $O(N^3)$ for the
dense approach.


## Why spatial matters --- a motivating example

To see what goes wrong when spatial structure is ignored, we simulate data
with a known spatial field, then fit a model without the spatial term.

```{r simulate-spatial}
library(INLAocc)

sim <- simulate_occu(N = 300, J = 4,
                     n_occ_covs = 1, n_det_covs = 1,
                     beta_occ = c(0.5, -0.8),
                     beta_det = c(0, -0.5),
                     spatial_range = 0.3, seed = 42)
```

The simulated data has 300 sites with 4 visits each. The true occupancy
intercept is 0.5, the effect of `occ_x1` is -0.8, and there is spatial
autocorrelation with a practical range of 0.3 (in coordinate units).

### Fitting without the spatial effect

```{r fit-nospatial}
fit_nospatial <- occu(~ occ_x1, ~ det_x1, data = sim$data, verbose = 0)
moranI(fit_nospatial)
```

If Moran's I is significantly positive with a p-value near zero, the residuals are
spatially autocorrelated. The non-spatial model has left systematic spatial
structure on the table. This does not just violate an assumption --- it
means the reported standard errors are too small and the 95% credible
intervals are too narrow to be trusted.


## Fitting the SPDE model

Adding the spatial effect is a single argument:

```{r fit-spatial}
fit <- occu(~ occ_x1, ~ det_x1, data = sim$data,
            spatial = sim$data$coords, verbose = 0)
summary(fit)
```

Several things to note:

- **Fixed effects should be recovered well.** The true occupancy intercept is 0.5
  and the true `occ_x1` effect is -0.8. Both credible intervals should contain
  the true values.
- **The spatial range should be recovered.** The true range was 0.3; the posterior
  should be close to this value. The model correctly identifies the scale of
  spatial correlation.
- **Intervals are wider than the non-spatial model.** This is correct
  behaviour: accounting for spatial dependence reduces the effective sample
  size, producing more honest uncertainty.
- **A substantial spatial SD** confirms that the spatial field
  explains meaningful variation in occupancy.

You can extract the spatial range and standard deviation directly:

```{r spatial-range}
spatialRange(fit)
```

### Verifying the spatial residuals

```{r morani-spatial}
moranI(fit)
```

Moran's I should have dropped substantially and should no longer be significant.
The spatial model has absorbed the autocorrelation.

The empirical semivariogram tells the same story:

```{r variogram-spatial}
variogram(fit)
```

A flat line at the sill confirms that no residual spatial trend remains. If
the variogram still shows an increasing trend at short distances, the mesh may
be too coarse or the range prior too restrictive --- see the next section.

For the full diagnostic panel (residuals, Moran's I, Q-Q plot, DHARMa):

```{r checkmodel}
checkModel(fit)
```


## Controlling the mesh

The automatic mesh works well for most datasets. INLAocc scales `max.edge` to
roughly 1/20th (inner) and 1/5th (outer) of the coordinate extent, with a
cutoff of `max.edge[1] / 5`. This is adequate when sites are reasonably
spread across the study area.

You may need manual control when:

- **Sites are clustered** (e.g., along roads or rivers) --- the auto-mesh may
  waste nodes in empty regions.
- **The study extent is very large** --- a coarser mesh saves computation.

- **You have strong prior knowledge** about the spatial range --- tighter PC
  priors can stabilise estimation.

Use `occu_spatial()` to pre-build the mesh:

```{r mesh-control}
sp <- occu_spatial(sim$data$coords,
                   max.edge = c(0.05, 0.2),
                   cutoff = 0.01,
                   prior.range = c(0.2, 0.5),
                   prior.sigma = c(1, 0.05))
sp
```

Then pass the pre-built object to `occu()`:

```{r fit-custom-mesh, eval = FALSE}
fit <- occu(~ occ_x1, ~ det_x1, data = sim$data, spatial = sp, verbose = 0)
```

### PC priors for the spatial field

The PC (penalizing complexity) priors deserve explanation because they
directly influence how much spatial structure the model captures.

**`prior.range = c(r0, p)`** encodes $P(\rho < r_0) = p$, where $\rho$ is
the practical correlation range. This sets a lower bound on the spatial scale:
if you believe the range is unlikely to be shorter than 0.2 (in your
coordinate units), set `prior.range = c(0.2, 0.5)`. A tighter prior
(e.g., `c(0.2, 0.01)`) strongly penalises very short-range fluctuations.

**`prior.sigma = c(s0, p)`** encodes $P(\sigma > s_0) = p$, where $\sigma$ is
the marginal standard deviation of the spatial field. This sets an upper bound
on the spatial variance: `prior.sigma = c(1, 0.05)` means you believe the
spatial SD is unlikely to exceed 1 on the logit scale.

The logic behind PC priors is shrinkage toward the **base model** --- in this
case, no spatial effect at all ($\sigma = 0$). This prevents the spatial field
from soaking up variation that properly belongs to fixed-effect covariates.
When in doubt, use weakly informative priors (the defaults) and let the data
speak.

**Units matter.** If your coordinates are in UTM metres, the range prior
should be in metres. If they are in decimal degrees, the prior is in degrees.
Rescaling coordinates to a [0, 1] box can simplify prior elicitation.

### Quick mesh tweaks via `spde.args`

For minor adjustments without building a full `occu_spatial()` object, pass
mesh arguments directly:

```{r spde-args}
fit <- occu(~ occ_x1, ~ det_x1, data = sim$data,
            spatial = sim$data$coords,
            spde.args = list(max.edge = c(0.05, 0.2)),
            verbose = 0)
```


## Occupancy maps

After fitting a spatial model, `occuMap()` produces publication-quality maps
by projecting the spatial field and fixed effects onto a regular grid:

```{r maps}
occuMap(fit)                          # occupancy probability
occuMap(fit, type = "sd")             # posterior uncertainty
occuMap(fit, type = "spatial")        # spatial random effect only
occuMap(fit, type = "all")            # 2-panel: psi + sd
```

Each map type serves a different purpose:

- **`"psi"`** (default): the predicted occupancy probability $\hat{\psi}$ at
  each grid cell, combining fixed effects and the spatial field. This is the
  primary output for conservation or management applications.
- **`"sd"`**: the posterior standard deviation of $\psi$. High-uncertainty
  regions are candidates for additional survey effort.
- **`"spatial"`**: the spatial random effect $w(\mathbf{s})$ alone, centred at
  zero. This reveals spatial patterns not explained by the covariates ---
  essentially, where occupancy is higher or lower than the covariates predict.
  Symmetric colour scale (blue-white-red) with zero as white.
- **`"all"`**: a side-by-side panel of `"psi"` and `"sd"`.

All maps overlay observed sites, coloured by whether the species was detected
(filled circles) or not (open circles). The returned object (invisible)
contains the gridded values for custom plotting.


## Spatial prediction at new locations

To predict occupancy at unsampled locations --- for instance, a regular grid
covering the study area --- use `predict_spatial()`:

```{r predict-spatial}
new_coords <- expand.grid(
  x = seq(0, 1, by = 0.02),
  y = seq(0, 1, by = 0.02)
)
new_covs <- data.frame(occ_x1 = 0)  # hold covariate at its mean

preds <- predict_spatial(fit, newcoords = new_coords, newocc.covs = new_covs)
head(preds)
```

The returned data frame contains the posterior mean occupancy (`psi`),
posterior standard deviation (`sd`), and 95% credible interval bounds
(`q0.025`, `q0.975`) at each prediction location. The spatial field is
projected from the mesh to the new coordinates using the SPDE basis
functions --- no re-fitting required.

Predictions respect the mesh boundary: locations far outside the convex hull
of the observed sites will have high uncertainty, reflecting extrapolation.


## Areal (CAR/BYM2) models

When sites are administrative regions, grid cells, or other areal units rather
than point locations, the SPDE approach is inappropriate. Instead, spatial
dependence is modelled through an adjacency structure: regions that share a
border are correlated.

INLAocc supports areal spatial models via `occu_areal()`:

```{r areal, eval = FALSE}
# Adjacency matrix: 1 if regions share a border, 0 otherwise
adj <- matrix(0, 100, 100)
# ... fill from shapefile or raster neighbourhood ...

areal_spec <- occu_areal(adj, model = "bym2")

fit_areal <- occu(~ occ_x1, ~ det_x1, data = my_data,
                  spatial = areal_spec, verbose = 0)
summary(fit_areal)
```

The **BYM2 model** (Riebler et al., 2016) decomposes the spatial effect into
two components:

- A **structured** component (conditional autoregressive / CAR): captures
  spatially smooth variation, where each region's value depends on its
  neighbours.
- An **unstructured** component (IID): captures region-specific noise that
  is not spatially structured.
- A **mixing parameter** $\phi \in [0, 1]$: controls the proportion of
  variance attributable to the structured vs. unstructured component. Values
  near 1 indicate strong spatial structure; near 0 means the variation is
  essentially independent across regions.

`occu_areal()` accepts three adjacency formats: a symmetric binary matrix, an
`spdep::nb` neighbourhood object, or a path to an INLA graph file. Use
whichever is most convenient for your workflow.

The simpler `"besag"` model (pure CAR, no unstructured component) is also
available but BYM2 is generally preferred because it handles disconnected
graphs and provides the structured/unstructured decomposition.


## Spatially varying coefficients (SVC)

A standard spatial model assumes the covariate effects $\boldsymbol{\beta}$
are constant across the study area, with the spatial field capturing residual
structure. But sometimes a covariate's effect genuinely varies in space ---
for example, elevation may matter more for occupancy in one region than
another. A spatially varying coefficient (SVC) model allows this:

$$\text{logit}(\psi_i) = \beta_0 + \beta_1 x_{1i} + \tilde{\beta}_1(\mathbf{s}_i) \, x_{1i}$$

where $\tilde{\beta}_1(\mathbf{s})$ is a spatial random field for the slope of
$x_1$. The total effect of $x_1$ at site $i$ is
$\beta_1 + \tilde{\beta}_1(\mathbf{s}_i)$: a global mean effect plus a
spatially varying departure.

In INLAocc, the `svc` argument specifies which occupancy coefficient varies
spatially, indexed by position in the design matrix (1 = intercept, 2 = first
covariate, etc.):

```{r svc-intercept}
# Spatially varying intercept (svc = 1)
fit_svc <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                spatial = sim$data$coords, svc = 1, verbose = 0)
summary(fit_svc)
```

```{r svc-covariate}
# Spatially varying effect of occ_x1 (svc = 2)
fit_svc2 <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                 spatial = sim$data$coords, svc = 2, verbose = 0)
summary(fit_svc2)
```

A spatially varying intercept (`svc = 1`) is equivalent to the standard SPDE
model --- the spatial field modulates the baseline occupancy. A spatially
varying slope (`svc = 2`) asks a more specific question: does the effect of
`occ_x1` change across space?

### Extracting SVC surfaces

To obtain posterior samples of the spatially varying coefficient at each mesh
node:

```{r svc-samples}
svc_draws <- getSVCSamples(fit_svc2)
str(svc_draws)
```

The returned object contains posterior samples of $\tilde{\beta}_1(\mathbf{s})$
at each site, which can be mapped to visualise where the covariate effect is
strongest. Combine with `occuMap(fit_svc2, type = "spatial")` to see the
spatial pattern.


## Space-time models

For multi-season data with spatial structure, INLAocc combines the SPDE
spatial field with a temporal process. The `temporal` argument specifies the
temporal correlation structure:

```{r spatiotemporal, eval = FALSE}
sim_st <- simTOcc(N = 100, J = 3, n_seasons = 5, seed = 300)

fit_st <- occu(~ 1, ~ 1, data = sim_st$data,
               spatial = sim_st$data$coords,
               temporal = "ar1", verbose = 0)
summary(fit_st)
```

This fits an AR(1) temporal process alongside the SPDE spatial field,
yielding a separable space-time model where the spatial field evolves
smoothly over seasons. The AR(1) correlation parameter quantifies temporal
persistence: values near 1 mean the spatial pattern is stable across
seasons; values near 0 mean each season's spatial field is nearly
independent.

For more on temporal modelling without the spatial component, see
`vignette("quickstart")`.


## Practical guidance

**Start simple.** Fit the non-spatial model first, then check `moranI()`. If
residual autocorrelation is non-significant (say, p > 0.05), the spatial term
may not be needed. Compare WAIC between spatial and non-spatial models to
confirm.

**Trust the auto-mesh.** The default mesh construction works for most
datasets (up to a few thousand sites). Only tune `max.edge`, `cutoff`, or
the PC priors when you have domain knowledge about the spatial range or
when diagnostics suggest problems.

**Match prior units to coordinate units.** If your coordinates are in UTM
metres, `prior.range = c(5000, 0.5)` means you believe the range is likely
at least 5 km. If coordinates are in decimal degrees, the prior is in
degrees. Rescaling to a unit square avoids confusion and makes the defaults
more portable.

**Watch for large N.** The SPDE approach scales well (the mesh has far fewer
nodes than sites), but for $N > 5000$ the EM iterations can be slow. Options
include coarsening the mesh (`max.edge` larger), subsampling sites, or
increasing the `cutoff` to reduce mesh density in data-rich areas.

**Use WAIC for model comparison.** A spatial model that does not improve WAIC
over the non-spatial version is adding complexity without payoff. The spatial
term should be justified by the data, not added by default.

**Check the spatial field map.** `occuMap(fit, type = "spatial")` shows the
spatial random effect alone. If this map shows sharp features that track a
known covariate you omitted, consider adding that covariate instead of
relying on the spatial field to absorb it. The spatial field should capture
genuinely unstructured residual variation, not proxy for missing covariates.

**Areal vs. SPDE.** Use SPDE when sites are point locations with known
coordinates. Use areal (CAR/BYM2) when sites are regions defined by
boundaries (counties, grid cells, management units). Do not use SPDE with
region centroids --- the adjacency structure is more appropriate.
