---
title: "Spatial Occupancy Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatial Occupancy Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

Species detection/non-detection data are collected at sites distributed across
a landscape, and nearby sites tend to share unmeasured environmental conditions.
When this spatial dependence is ignored, occupancy models treat sites as
independent replicates, an assumption that is almost always wrong. The
consequences are real: residual spatial autocorrelation inflates effective
sample size, producing confidence intervals that are too narrow and
fixed-effect estimates that may be biased. If a covariate of interest is
itself spatially structured, the bias compounds.

INLAocc handles spatial structure through the SPDE (Stochastic Partial
Differential Equations) approximation to a continuous Gaussian random field.
Rather than working with a dense $N \times N$
covariance matrix (which becomes intractable for large $N$), the SPDE approach
represents the spatial field as a piecewise linear function over a
triangulation mesh. This yields a sparse precision matrix that INLA can
exploit for fast, exact inference. The mesh is built automatically from site
coordinates, or you can construct it manually for full control.

This vignette covers the full spatial modelling workflow: fitting SPDE models,
tuning the mesh and PC priors, mapping occupancy, predicting at new locations,
checking spatial residuals, areal (CAR/BYM2) models for regional data,
spatially varying coefficients, and space-time extensions. For general model
diagnostics beyond the spatial context, see `vignette("diagnostics")`.


## The spatial occupancy model

The standard single-season occupancy model with a spatial random effect is:

$$z_i \sim \text{Bernoulli}(\psi_i), \quad \text{logit}(\psi_i) = \mathbf{x}_i^\top \boldsymbol{\beta} + w(\mathbf{s}_i)$$

$$y_{ij} \mid z_i \sim \text{Bernoulli}(z_i \cdot p_{ij}), \quad \text{logit}(p_{ij}) = \mathbf{w}_{ij}^\top \boldsymbol{\alpha}$$

where $z_i$ is the true (latent) occupancy state at site $i$, $\psi_i$ is the
occupancy probability, $y_{ij}$ is the observed detection/non-detection at
site $i$ during visit $j$, and $p_{ij}$ is the detection probability.

The spatial component $w(\mathbf{s})$ is a Gaussian random field (GRF) with
zero mean and Matern covariance function:

$$\text{Cov}(w(\mathbf{s}_1), w(\mathbf{s}_2)) = \sigma^2 \frac{2^{1-\nu}}{\Gamma(\nu)} \left(\kappa \|\mathbf{s}_1 - \mathbf{s}_2\|\right)^\nu K_\nu\left(\kappa \|\mathbf{s}_1 - \mathbf{s}_2\|\right)$$

The three parameters have intuitive interpretations:

- $\sigma^2$ is the **marginal variance** of the spatial field: how much
  spatial variation there is on the logit-occupancy scale.
- $\kappa$ is a **scale parameter** that relates to the practical correlation
  range via $\rho = \sqrt{8\nu}/\kappa$. The practical range $\rho$ is the
  distance at which the Matern correlation drops to approximately 0.1.
- $\nu$ controls **smoothness**. For the 2D SPDE approximation, $\nu$ is
  fixed at 1 (corresponding to $\alpha = 2$), giving a once-differentiable
  field.

**The SPDE trick.** Rather than evaluating the dense covariance matrix, the
SPDE approach solves a stochastic PDE whose stationary solution has the Matern
covariance. The solution is approximated as a piecewise linear function over a
triangulation mesh, yielding a sparse precision matrix $\mathbf{Q}$ with
$O(n)$ non-zero entries (where $n$ is the number of mesh nodes). This
sparsity is what makes INLA fast: the Cholesky factorisation of
$\mathbf{Q}$ scales as $O(n^{3/2})$ in 2D, compared to $O(N^3)$ for the
dense approach.


## Why spatial matters: a motivating example

To see what goes wrong when spatial structure is ignored, we simulate data
with a known spatial field, then fit a model without the spatial term.

```{r simulate-spatial}
library(INLAocc)

sim <- simulate_occu(N = 300, J = 4,
                     n_occ_covs = 1, n_det_covs = 1,
                     beta_occ = c(0.5, -0.8),
                     beta_det = c(0, -0.5),
                     spatial_range = 0.3, seed = 42)
```

The simulated data has 300 sites with 4 visits each. The true occupancy
intercept is 0.5, the effect of `occ_x1` is -0.8, and there is spatial
autocorrelation with a practical range of 0.3 (in coordinate units).

### Fitting without the spatial effect

```{r fit-nospatial}
fit_nospatial <- occu(~ occ_x1, ~ det_x1, data = sim$data, verbose = 0)
moranI(fit_nospatial)
```

If Moran's I is significantly positive with a p-value near zero, the residuals are
spatially autocorrelated. The non-spatial model has left systematic spatial
structure on the table. This is not just a violated assumption: the
reported standard errors are too small and the 95% credible intervals
are too narrow to be trusted.

### Before and after: residual spatial patterns

The semivariogram makes the spatial structure visible. For the non-spatial
model, we expect the semivariance to increase with distance at short lags
--- the signature of positive autocorrelation. After fitting a spatial
model, the variogram should flatten into a horizontal band at the sill.

```{r variogram-before-after, fig.width = 9, fig.height = 4}
par(mfrow = c(1, 2))

# Non-spatial residual variogram
vg_before <- variogram(fit_nospatial, n.bins = 12)
plot(vg_before, main = "Non-spatial model")

# Spatial model (fit below, but we preview the diagnostic)
fit_spatial_preview <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                            spatial = sim$data$coords, verbose = 0)
vg_after <- variogram(fit_spatial_preview, n.bins = 12)
plot(vg_after, main = "Spatial model")
```

The left panel shows a clear increasing trend: nearby sites have more
similar residuals than distant sites, exactly the pattern that inflates
Type I error rates. The right panel is flat; the SPDE field has absorbed
the autocorrelation.

Beyond visual diagnostics, the consequences for inference are concrete.
Compare the width of the 95% credible interval for `occ_x1` between the
two models:

```{r ci-comparison}
# Non-spatial CI width
ci_nospatial <- diff(unlist(fit_nospatial$occ_fit$summary.fixed["occ_x1",
                           c("0.025quant", "0.975quant")]))
# Spatial CI width
ci_spatial <- diff(unlist(fit_spatial_preview$occ_fit$summary.fixed["occ_x1",
                         c("0.025quant", "0.975quant")]))
cat(sprintf("Non-spatial CI width: %.3f\nSpatial CI width:     %.3f\n",
            ci_nospatial, ci_spatial))
```

The spatial model produces wider (more honest) intervals because it
recognises that spatially correlated sites are not independent replicates.
The non-spatial model overstates the effective sample size, leading to
overconfident inference, a common problem in ecological studies where
sites are clustered along roads, rivers, or elevation gradients.


## Fitting the SPDE model

Adding the spatial effect is a single argument:

```{r fit-spatial}
fit <- occu(~ occ_x1, ~ det_x1, data = sim$data,
            spatial = sim$data$coords, verbose = 0)
summary(fit)
```

What to check in the output:

- **Fixed effects should be recovered well.** The true occupancy intercept is 0.5
  and the true `occ_x1` effect is -0.8. Both credible intervals should contain
  the true values.
- **The spatial range should be recovered.** The true range was 0.3; the posterior
  should be close to this value. The model correctly identifies the scale of
  spatial correlation.
- **Intervals are wider than the non-spatial model.** This is correct
  behaviour: accounting for spatial dependence reduces the effective sample
  size, producing more honest uncertainty.
- **A large spatial SD** confirms that the spatial field
  explains real variation in occupancy.

You can extract the spatial range and standard deviation directly:

```{r spatial-range}
spatialRange(fit)
```

### Verifying the spatial residuals

```{r morani-spatial}
moranI(fit)
```

Moran's I should have dropped and should no longer be significant.
The spatial model has absorbed the autocorrelation.

The empirical semivariogram tells the same story:

```{r variogram-spatial}
variogram(fit)
```

A flat line at the sill confirms that no residual spatial trend remains. If
the variogram still shows an increasing trend at short distances, the mesh may
be too coarse or the range prior too restrictive (see the next section).

For the full diagnostic panel (residuals, Moran's I, Q-Q plot, DHARMa):

```{r checkmodel}
checkModel(fit)
```


## Controlling the mesh

The automatic mesh works well for most datasets. INLAocc scales `max.edge` to
roughly 1/20th (inner) and 1/5th (outer) of the coordinate extent, with a
cutoff of `max.edge[1] / 5`. This is adequate when sites are reasonably
spread across the study area.

You may need manual control when:

- **Sites are clustered** (e.g., along roads or rivers), so the auto-mesh may
  waste nodes in empty regions.
- **The study extent is very large**, and a coarser mesh saves computation.

- **You have strong prior knowledge** about the spatial range, and tighter PC
  priors can stabilise estimation.

Use `occu_spatial()` to pre-build the mesh:

```{r mesh-control}
sp <- occu_spatial(sim$data$coords,
                   max.edge = c(0.05, 0.2),
                   cutoff = 0.01,
                   prior.range = c(0.2, 0.5),
                   prior.sigma = c(1, 0.05))
sp
```

Then pass the pre-built object to `occu()`:

```{r fit-custom-mesh}
fit <- occu(~ occ_x1, ~ det_x1, data = sim$data, spatial = sp, verbose = 0)
```

### PC priors for the spatial field

The PC (penalizing complexity) priors directly control how much spatial
structure the model captures.

**`prior.range = c(r0, p)`** encodes $P(\rho < r_0) = p$, where $\rho$ is
the practical correlation range. This sets a lower bound on the spatial scale:
if you believe the range is unlikely to be shorter than 0.2 (in your
coordinate units), set `prior.range = c(0.2, 0.5)`. A tighter prior
(e.g., `c(0.2, 0.01)`) strongly penalises very short-range fluctuations.

**`prior.sigma = c(s0, p)`** encodes $P(\sigma > s_0) = p$, where $\sigma$ is
the marginal standard deviation of the spatial field. This sets an upper bound
on the spatial variance: `prior.sigma = c(1, 0.05)` means you believe the
spatial SD is unlikely to exceed 1 on the logit scale.

The logic behind PC priors is shrinkage toward the **base model**, in this
case, no spatial effect at all ($\sigma = 0$). This prevents the spatial field
from soaking up variation that properly belongs to fixed-effect covariates.
When in doubt, use weakly informative priors (the defaults) and let the data
speak.

**Units matter.** If your coordinates are in UTM metres, the range prior
should be in metres. If they are in decimal degrees, the prior is in degrees.
Rescaling coordinates to a [0, 1] box can simplify prior elicitation.

### Quick mesh tweaks via `spde.args`

For minor adjustments without building a full `occu_spatial()` object, pass
mesh arguments directly:

```{r spde-args}
fit <- occu(~ occ_x1, ~ det_x1, data = sim$data,
            spatial = sim$data$coords,
            spde.args = list(max.edge = c(0.05, 0.2)),
            verbose = 0)
```


## Occupancy maps

After fitting a spatial model, `occuMap()` produces publication-quality maps
by projecting the spatial field and fixed effects onto a regular grid:

```{r maps}
occuMap(fit)                          # occupancy probability
occuMap(fit, type = "sd")             # posterior uncertainty
occuMap(fit, type = "spatial")        # spatial random effect only
occuMap(fit, type = "all")            # 2-panel: psi + sd
```

Each map type serves a different purpose:

- **`"psi"`** (default): the predicted occupancy probability $\hat{\psi}$ at
  each grid cell, combining fixed effects and the spatial field. This is the
  primary output for conservation or management applications.
- **`"sd"`**: the posterior standard deviation of $\psi$. High-uncertainty
  regions are candidates for additional survey effort.
- **`"spatial"`**: the spatial random effect $w(\mathbf{s})$ alone, centred at
  zero. This reveals spatial patterns not explained by the covariates ---
  where occupancy is higher or lower than the covariates predict.
  Symmetric colour scale (blue-white-red) with zero as white.
- **`"all"`**: a side-by-side panel of `"psi"` and `"sd"`.

All maps overlay observed sites, coloured by whether the species was detected
(filled circles) or not (open circles). The returned object (invisible)
contains the gridded values for custom plotting.

### Interpreting uncertainty maps

The posterior SD map (`type = "sd"`) directly informs survey design.
Uncertainty is driven by three factors:

1. **Distance from observed sites.** The spatial field is interpolated from
   mesh nodes, so regions far from any site have higher posterior variance.
   This is the dominant source of uncertainty in most maps.
2. **Local detection history.** Sites where the species was detected on every
   visit provide strong evidence for occupancy; sites with mixed or no
   detections contribute more uncertainty.
3. **Mesh resolution.** A coarser mesh smooths the spatial field more
   aggressively, which can reduce apparent uncertainty but also obscures
   fine-scale patterns.

```{r map-uncertainty-detail, fig.width = 9, fig.height = 4}
par(mfrow = c(1, 2))
occuMap(fit, type = "psi", main = "Occupancy probability")
occuMap(fit, type = "sd", main = "Posterior SD")
```

High-SD regions are candidates for additional survey effort. If you are
planning a follow-up study, targeting areas with $\text{SD}(\psi) > 0.2$
will maximise information gain per unit effort.

### Effect of mesh resolution on maps

The mesh resolution controls the spatial detail that the model can capture.
A finer mesh (smaller `max.edge`) resolves shorter-range patterns but
increases computation. A coarser mesh runs faster but may smooth away
real spatial structure.

```{r mesh-comparison, fig.width = 9, fig.height = 4}
# Fine mesh
sp_fine <- occu_spatial(sim$data$coords,
                        max.edge = c(0.03, 0.15),
                        cutoff = 0.005)
fit_fine <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                 spatial = sp_fine, verbose = 0)

# Coarse mesh
sp_coarse <- occu_spatial(sim$data$coords,
                          max.edge = c(0.15, 0.4),
                          cutoff = 0.03)
fit_coarse <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                   spatial = sp_coarse, verbose = 0)

par(mfrow = c(1, 2))
occuMap(fit_fine, main = sprintf("Fine mesh (%d nodes)", sp_fine$n_mesh))
occuMap(fit_coarse, main = sprintf("Coarse mesh (%d nodes)", sp_coarse$n_mesh))
```

Both maps should show the same broad spatial pattern, but the fine mesh
captures more local variation. If the two maps look markedly
different, the coarse mesh is too aggressive. If they are nearly
identical, the extra nodes in the fine mesh are wasted computation. The
default auto-mesh usually strikes a good balance, but for publication-quality
maps or when the spatial range is short, a finer mesh is worth the cost.

The fixed-effect estimates should be stable across mesh resolutions. If
$\hat{\beta}$ changes noticeably between meshes, something else is wrong
(typically the PC priors are too restrictive or the data are too sparse for
a spatial model).


## Spatial prediction at new locations

To predict occupancy at unsampled locations (for instance, a regular grid
covering the study area), use `predict_spatial()`:

```{r predict-spatial}
new_coords <- expand.grid(
  x = seq(0, 1, by = 0.02),
  y = seq(0, 1, by = 0.02)
)
new_covs <- data.frame(occ_x1 = 0)  # hold covariate at its mean

preds <- predict_spatial(fit, newcoords = new_coords, newocc.covs = new_covs)
head(preds)
```

The returned object contains posterior mean occupancy (`psi.0$mean`),
posterior standard deviation (`psi.0$sd`), and quantiles at each prediction
location. The spatial field is projected from the mesh to the new
coordinates using the SPDE basis functions, so no re-fitting is needed.

Predictions respect the mesh boundary: locations far outside the convex hull
of the observed sites will have high uncertainty, reflecting extrapolation.

### Visualising predictions with uncertainty

Spatial predictions are most useful when paired with uncertainty. The
posterior standard deviation at each prediction location tells you where
the model is confident and where additional data would be most valuable.

```{r predict-spatial-vis, fig.width = 9, fig.height = 4}
pred_df <- data.frame(
  x   = new_coords$x,
  y   = new_coords$y,
  psi = preds$psi.0$mean
)

# Add uncertainty (SD) if available from posterior samples
if (!is.null(preds$psi.0$sd)) {
  pred_df$sd <- preds$psi.0$sd
}

# Add credible interval bounds
if (!is.null(preds$psi.0$quantiles)) {
  pred_df$lower <- preds$psi.0$quantiles[, 1]
  pred_df$upper <- preds$psi.0$quantiles[, 3]
  pred_df$ci_width <- pred_df$upper - pred_df$lower
}

par(mfrow = c(1, 2))

# Left panel: predicted occupancy
with(pred_df, {
  plot(x, y, col = hcl.colors(100, "YlGnBu")[cut(psi, 100)],
       pch = 15, cex = 0.6, asp = 1,
       xlab = "X", ylab = "Y", main = "Predicted occupancy")
})

# Right panel: prediction uncertainty (CI width)
if (!is.null(pred_df$ci_width)) {
  with(pred_df, {
    plot(x, y, col = hcl.colors(100, "YlOrRd")[cut(ci_width, 100)],
         pch = 15, cex = 0.6, asp = 1,
         xlab = "X", ylab = "Y", main = "95% CI width")
  })
}
```

Regions near observed sites have narrow credible intervals because the SPDE
field is well-constrained by nearby data. Regions far from any site ---
particularly outside the convex hull of the survey locations, show wider
intervals, reflecting the increasing reliance on the prior as the data thin
out. Target additional sampling at high-uncertainty areas to maximise
information gain.

### How spatial prediction works under the hood

The prediction mechanism involves two steps. First, a **projection matrix**
$\mathbf{A}_\text{pred}$ maps the SPDE field from mesh nodes to the new
locations. Each row of $\mathbf{A}_\text{pred}$ contains the barycentric
weights for the enclosing mesh triangle (typically three non-zero entries
that sum to one). The predicted spatial field at location $\mathbf{s}_0$ is
then $\hat{w}(\mathbf{s}_0) = \mathbf{a}(\mathbf{s}_0)^\top \hat{\mathbf{w}}$,
where $\hat{\mathbf{w}}$ is the vector of posterior means (or samples) at
the mesh nodes.

Second, the full linear predictor combines the spatial projection with the
fixed-effect design matrix:

$$\hat{\eta}(\mathbf{s}_0) = \mathbf{x}_0^\top \hat{\boldsymbol{\beta}} + \hat{w}(\mathbf{s}_0)$$

and the occupancy prediction is $\hat{\psi}(\mathbf{s}_0) = \text{logit}^{-1}(\hat{\eta})$.
Because the posterior samples of $\boldsymbol{\beta}$ and $\mathbf{w}$ are
drawn jointly from the INLA approximation, the uncertainty in $\hat{\psi}$
correctly propagates both fixed-effect and spatial uncertainty.


## Areal (CAR/BYM2) models

When sites are administrative regions, grid cells, or other areal units rather
than point locations, the SPDE approach is inappropriate. Instead, spatial
dependence is modelled through an adjacency structure: regions that share a
border are correlated.

INLAocc supports areal spatial models via `occu_areal()`:

```{r areal}
# Build a 10x10 grid adjacency matrix (queen contiguity)
n_row <- 10; n_col <- 10; n_regions <- n_row * n_col
adj <- matrix(0L, n_regions, n_regions)
for (i in seq_len(n_regions)) {
  r <- (i - 1) %/% n_col + 1
  c <- (i - 1) %% n_col + 1
  for (dr in -1:1) for (dc in -1:1) {
    if (dr == 0 && dc == 0) next
    nr <- r + dr; nc <- c + dc
    if (nr >= 1 && nr <= n_row && nc >= 1 && nc <= n_col) {
      j <- (nr - 1) * n_col + nc
      adj[i, j] <- 1L
    }
  }
}

areal_spec <- occu_areal(adj, model = "bym2")
str(areal_spec)
```

```{r areal-fit}
# Create data with N = n_regions (100 sites matching the 10x10 grid)
sim_areal <- simulate_occu(N = n_regions, J = 4, n_occ_covs = 1, n_det_covs = 1,
                           beta_occ = c(0.5, -0.5), beta_det = c(0, -0.3),
                           seed = 99)
fit_areal <- occu(~ occ_x1, ~ det_x1, data = sim_areal$data,
                  spatial = areal_spec, verbose = 0)
summary(fit_areal)
```

The **BYM2 model** (Riebler et al., 2016) decomposes the spatial effect into
two components:

- A **structured** component (conditional autoregressive / CAR): captures
  spatially smooth variation, where each region's value depends on its
  neighbours.
- An **unstructured** component (IID): captures region-specific noise that
  is not spatially structured.
- A **mixing parameter** $\phi \in [0, 1]$: controls the proportion of
  variance attributable to the structured vs. unstructured component. Values
  near 1 indicate strong spatial structure; near 0 means the variation is
  independent across regions.

`occu_areal()` accepts three adjacency formats: a symmetric binary matrix, an
`spdep::nb` neighbourhood object, or a path to an INLA graph file. Use
whichever is most convenient for your workflow.

The simpler `"besag"` model (pure CAR, no unstructured component) is also
available but BYM2 is generally preferred because it handles disconnected
graphs and provides the structured/unstructured decomposition.


## Spatially varying coefficients (SVC)

A standard spatial model assumes the covariate effects $\boldsymbol{\beta}$
are constant across the study area, with the spatial field capturing residual
structure. But sometimes a covariate's effect varies in space:
for example, elevation may matter more for occupancy in one region than
another. A spatially varying coefficient (SVC) model allows this:

$$\text{logit}(\psi_i) = \beta_0 + \beta_1 x_{1i} + \tilde{\beta}_1(\mathbf{s}_i) \, x_{1i}$$

where $\tilde{\beta}_1(\mathbf{s})$ is a spatial random field for the slope of
$x_1$. The total effect of $x_1$ at site $i$ is
$\beta_1 + \tilde{\beta}_1(\mathbf{s}_i)$: a global mean effect plus a
spatially varying departure.

In INLAocc, the `svc` argument specifies which occupancy coefficient varies
spatially, indexed by position in the design matrix (1 = intercept, 2 = first
covariate, etc.):

```{r svc-intercept}
# Spatially varying intercept (svc = 1)
fit_svc <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                spatial = sim$data$coords, svc = 1, verbose = 0)
summary(fit_svc)
```

```{r svc-covariate}
# Spatially varying effect of occ_x1 (svc = 2)
fit_svc2 <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                 spatial = sim$data$coords, svc = 2, verbose = 0)
summary(fit_svc2)
```

A spatially varying intercept (`svc = 1`) is equivalent to the standard SPDE
model: the spatial field modulates the baseline occupancy. A spatially
varying slope (`svc = 2`) asks a more specific question: does the effect of
`occ_x1` change across space?

### Extracting SVC surfaces

To obtain posterior samples of the spatially varying coefficient at each mesh
node:

```{r svc-samples}
svc_draws <- getSVCSamples(fit_svc2)
str(svc_draws)
```

The returned object contains posterior samples of $\tilde{\beta}_1(\mathbf{s})$
at each site, which can be mapped to visualise where the covariate effect is
strongest. Combine with `occuMap(fit_svc2, type = "spatial")` to see the
spatial pattern.


## Space-time models

Monitoring programmes often collect detection/non-detection data at the
same sites across multiple years or seasons. When sites have spatial
coordinates, fitting a model with both spatial and temporal structure yields
a separable space-time occupancy model:

$$\text{logit}(\psi_{i,t}) = \mathbf{x}_{i,t}^\top \boldsymbol{\beta} + w(\mathbf{s}_i) + \eta_t$$

where $w(\mathbf{s}_i)$ is the spatial SPDE field (shared across seasons)
and $\eta_t$ is a temporal random effect. The separability means the spatial
pattern is the same in every season, but it is shifted up or down by the
temporal effect. This is appropriate when the spatial configuration of
habitat does not change between seasons but overall occupancy fluctuates.

### The separable space-time model

INLAocc combines the `spatial` and `temporal` arguments to fit this model.
The `temporal` argument accepts `"ar1"` (autoregressive order 1) or `"iid"`
(exchangeable, no temporal correlation):

```{r spatiotemporal-sim}
sim_st <- simTOcc(N = 150, J = 3, n_seasons = 6,
                  beta_occ = c(0.5), beta_det = c(0),
                  ar1 = TRUE, rho = 0.8, sigma_t = 0.6, seed = 300)
```

This creates a multi-season dataset with 150 sites, 3 visits per season,
and 6 seasons. The temporal process has AR(1) correlation $\rho = 0.8$ and
innovation SD $\sigma_t = 0.6$, so the spatial pattern is persistent across
seasons but drifts gradually.

### Fitting the space-time model

```{r spatiotemporal-fit}
fit_st <- occu(~ 1, ~ 1, data = sim_st$data,
               spatial = sim_st$data$coords,
               temporal = "ar1", verbose = 0)
summary(fit_st)
```

The summary reports both the SPDE hyperparameters (range and marginal SD of
the spatial field) and the AR(1) parameters ($\rho$ and $\sigma_t$):

- **AR(1) correlation ($\rho$).** Values near 1 indicate that the spatial
  pattern is highly persistent: occupancy hotspots in year 1 remain
  hotspots in year 6. Values near 0 mean each season's occupancy surface
  is nearly independent, and values near $-1$ would indicate alternating
  patterns (rare in ecology).
- **Temporal SD ($\sigma_t$).** Controls the magnitude of year-to-year
  fluctuations on the logit scale. Large values (> 1) mean that average
  occupancy swings widely between seasons.
- **Spatial range and SD.** Interpreted exactly as in the single-season
  model: how far the spatial correlation extends and how much spatial
  variation exists.

### Comparing temporal structures

If you are unsure whether an AR(1) or IID temporal effect is appropriate,
fit both and compare via WAIC:

```{r st-compare}
fit_st_iid <- occu(~ 1, ~ 1, data = sim_st$data,
                   spatial = sim_st$data$coords,
                   temporal = "iid", verbose = 0)

cat(sprintf("AR(1) WAIC: %.1f\nIID WAIC:   %.1f\n",
            fit_st$waic, fit_st_iid$waic))
```

When the data are generated with temporal autocorrelation (as here,
with $\rho = 0.8$), the AR(1) model should be preferred. When seasons are
far apart in time or the system is driven by stochastic events (e.g.,
fire), the IID model may be more appropriate.

### Extracting season-specific occupancy

The fitted space-time model contains season-level occupancy estimates. To
see the temporal trajectory:

```{r st-temporal-trend}
# Season-specific average psi
n_seasons <- sim_st$data$n_seasons %||% ncol(sim_st$truth$psi)
season_psi <- numeric(n_seasons)
for (t in seq_len(n_seasons)) {
  season_psi[t] <- mean(fit_st$psi_hat_list[[t]], na.rm = TRUE)
}

plot(seq_len(n_seasons), season_psi, type = "b", pch = 16,
     xlab = "Season", ylab = expression(bar(psi)),
     main = "Average occupancy across seasons",
     ylim = c(0, 1))
```

This plot shows how mean occupancy probability evolves over the monitoring
period. Because the AR(1) model borrows strength across seasons, estimates
for seasons with sparse data are regularised toward the temporal mean.

### Season-specific occupancy maps

For a spatial view of temporal change, produce maps for individual seasons.
This requires extracting the combined spatial + temporal predictions:

```{r st-maps, fig.width = 9, fig.height = 8}
if (n_seasons >= 4) {
  par(mfrow = c(2, 2))
  for (t in c(1, 2, floor(n_seasons / 2), n_seasons)) {
    if (!is.null(fit_st$spatial)) {
      occuMap(fit_st, main = sprintf("Season %d", t))
    }
  }
}
```

Comparing maps across seasons reveals where occupancy is expanding,
contracting, or stable. The spatial field $w(\mathbf{s})$ stays constant
(it is the site-level baseline), while the temporal effect $\eta_t$ shifts
the entire surface up or down.

### When to use space-time vs. separate single-season models

Fitting a single space-time model rather than six separate single-season
models has three advantages:

1. **Temporal borrowing of strength.** Seasons with fewer sites or visits
   borrow information from data-rich seasons through the AR(1) structure.
   This is especially valuable when some sites are only surveyed in a
   subset of years.
2. **Shared spatial field.** The spatial pattern is estimated from all
   seasons jointly, giving a more precise estimate than any single season
   alone.
3. **Temporal correlation is estimated, not assumed.** The AR(1) parameter
   $\rho$ quantifies how persistent the spatial pattern is, which is itself
   an ecological quantity of interest.

The main limitation is the separability assumption: the spatial field is the
same in every season. If the spatial pattern changes between
seasons (e.g., range expansion, habitat loss), a non-separable model or
separate single-season spatial models may be more appropriate.

For more on temporal modelling without the spatial component, see
`vignette("temporal-models")`.


## Multi-species spatial models

When multiple species are surveyed across the same set of spatially
referenced sites, fitting a shared spatial field can improve estimates for
rare species by borrowing spatial information from data-rich species. In
INLAocc, multi-species spatial models are specified by combining the
`multispecies` and `spatial` arguments:

```{r ms-spatial-sim}
sim_ms <- simMsOcc(N = 200, J = 4, n_species = 5,
                   n_occ_covs = 1, n_det_covs = 1,
                   spatial_range = 0.3, seed = 77)
```

This simulates detection histories for 5 species at 200 spatially
correlated sites. The shared spatial field means that species living in the
same landscape experience the same underlying spatial structure, though each
species has its own covariate responses.

```{r ms-spatial-fit}
fit_ms <- occu(~ occ_x1, ~ det_x1, data = sim_ms$data,
               spatial = sim_ms$data$coords,
               multispecies = TRUE, verbose = 0)
summary(fit_ms)
```

The summary reports species-specific fixed effects alongside the shared SPDE
hyperparameters. Because all species contribute to the estimation of the
spatial range and marginal SD, these hyperparameters are estimated more
precisely than they would be in any single-species model.

### Species-specific occupancy maps

The `occuMap()` function accepts a `species` argument for multi-species
models:

```{r ms-spatial-maps, fig.width = 9, fig.height = 4}
par(mfrow = c(1, 2))
occuMap(fit_ms, species = 1, main = sim_ms$data$species_names[1])
occuMap(fit_ms, species = 2, main = sim_ms$data$species_names[2])
```

Each map shows the predicted occupancy surface for one species, combining
the shared spatial field with species-specific fixed effects. Species that
are rare overall will have lower occupancy across the map but still exhibit
the same spatial pattern (high-occupancy and low-occupancy regions driven by
the shared spatial field).

### Species richness maps

Summing the latent occupancy states across species at each site gives an
estimate of species richness. The `richness()` function computes this
directly:

```{r ms-richness}
rich <- richness(fit_ms)
head(rich)
```

The `richness` column gives the expected number of species at each site,
and `sd` gives the posterior uncertainty. Sites with high richness
uncertainty are good candidates for additional survey effort.

For a full treatment of multi-species models (community models, JSDMs,
latent factor models), see `vignette("multi-species")`.


## Posterior predictive checks for spatial models

Model fit assessment via posterior predictive checks (PPC) is as important
for spatial models as for non-spatial ones. The `ppcOccu()` function
generates replicated datasets from the fitted model and compares a summary
statistic between observed and replicated data. If the model fits well, the
observed statistic should be indistinguishable from the replicated
distribution.

```{r ppc-spatial}
ppc <- ppcOccu(fit, fit.stat = "freeman-tukey", group = 1, n.samples = 500)
cat(sprintf("Bayesian p-value: %.3f\n", ppc$bayesian.p))
```

A Bayesian p-value near 0.5 indicates adequate fit. Values below 0.1 or
above 0.9 suggest systematic lack of fit. The model is either
underfitting (p near 1, generating too much variability) or overfitting (p
near 0, generating too little variability). For spatial models, a poor
Bayesian p-value after adding the SPDE field often indicates that the
spatial structure is more complex than the Matern field can capture, or that
an important covariate is missing.

The `group` argument controls the aggregation: `group = 1` aggregates by
site (summing detections across visits within each site), while `group = 2`
aggregates by visit (summing across sites within each visit). Site-level
aggregation is usually more informative for spatial models because it tests
whether the spatial pattern of detections is consistent with the model.

For a visual assessment, plot the fit statistics:

```{r ppc-plot, fig.width = 5, fig.height = 5}
plot(ppc$fit.y, ppc$fit.y.rep, pch = 16, cex = 0.4,
     xlab = "Observed fit statistic",
     ylab = "Replicated fit statistic",
     main = sprintf("PPC (p = %.3f)", ppc$bayesian.p))
abline(0, 1, lty = 2, col = "red")
```

Points should scatter roughly along the 1:1 line. Systematic departure from
the line indicates a particular direction of misfit.


## Model comparison and selection

Spatial models add complexity: a mesh, two hyperparameters (range and
marginal SD), and potentially hundreds of random-effect nodes. Whether this
complexity is justified depends on the data. INLAocc provides several tools
for formal model comparison.

### WAIC

The `waicOccu()` function computes the Widely Applicable Information
Criterion on the Bernoulli observation scale, making it directly comparable
across models with different random-effect structures:

```{r waic-comparison}
waic_nospatial <- waicOccu(fit_nospatial)
waic_spatial   <- waicOccu(fit)

cat(sprintf("Non-spatial WAIC: %.1f\nSpatial WAIC:     %.1f\n",
            waic_nospatial$WAIC, waic_spatial$WAIC))
```

Lower WAIC indicates better expected predictive accuracy. A reduction of
more than 2--3 units generally indicates meaningful improvement; reductions
of less than 2 suggest the spatial term is not needed.

### Comparing multiple models

When evaluating several candidate models (non-spatial,
SPDE spatial, areal), the `compare_models()` function produces a
ranked table:

```{r compare-models}
compare_models(
  "Non-spatial" = fit_nospatial,
  "Spatial"     = fit
)
```

The table is sorted by WAIC (or AIC/BIC if specified via `criterion`),
with delta values and Akaike-like weights. Models within 2 WAIC units of
the best model have similar predictive performance, and the weights
quantify their relative plausibility.


## Reporting spatial results in manuscripts

Spatial occupancy models involve more moving parts than standard models,
and reviewers will expect certain details. The following checklist covers
the minimum information to report.

**Model specification:**

- The occupancy and detection formulas (covariates on each process).
- Whether the spatial effect is SPDE (continuous) or CAR/BYM2 (areal).
- The coordinate reference system and units (UTM zone, decimal degrees).
- For SPDE: the PC prior settings for range and marginal SD, with
  justification (e.g., "we set `prior.range = c(5, 0.5)` based on the
  minimum home range diameter of the study species").
- For areal: the adjacency definition (queen vs. rook contiguity, or
  distance-based).

**Model diagnostics:**

- Moran's I on occupancy residuals (before and after adding the spatial
  term) to demonstrate that spatial autocorrelation has been absorbed.
- The empirical semivariogram, showing a flat pattern after spatial
  modelling.
- Posterior predictive check Bayesian p-value.
- WAIC comparison between the spatial and non-spatial models.

**Key results to report:**

- Fixed-effect estimates with 95% credible intervals (on both the logit
  and probability scales if relevant).
- Estimated spatial range $\hat{\rho}$ and marginal SD $\hat{\sigma}$
  with credible intervals. The range tells readers the scale of spatial
  dependence; the SD quantifies how much spatial variation exists beyond
  what covariates explain. Use `spatialRange()` to extract these.
- For SVC models: the global mean effect and the SD of the spatially
  varying departure.
- Maps of predicted occupancy and posterior uncertainty. Always include a
  colour bar with units, the coordinate system, and a scale bar or axis
  labels.

**What not to report:**

- Mesh construction details (node count, max.edge, cutoff) are
  implementation details that belong in a supplementary methods section,
  not the main text.
- The full INLA hyperparameter summary. Report the interpretable
  quantities (range, SD) rather than the internal parameterisation
  ($\kappa$, $\tau$).

**Example methods paragraph:**

> We modelled occupancy probability as a function of elevation and canopy
> cover, with a spatial random effect estimated via the SPDE approach
> (Lindgren et al. 2011). The spatial field was assigned a Matern
> covariance with PC priors: P(range < 5 km) = 0.5 and P(sigma > 1) =
> 0.05. Detection was modelled as a function of survey effort. The model
> was fitted using the EM-INLA algorithm implemented in INLAocc v0.3.1.
> Residual Moran's I dropped from 0.32 (p < 0.001) in the non-spatial
> model to 0.02 (p = 0.41) in the spatial model, confirming that the SPDE
> field absorbed the spatial autocorrelation. WAIC decreased from 412.3
> (non-spatial) to 387.1 (spatial).


## Practical guidance

### Start simple

Fit the non-spatial model first, then check `moranI()`. If
residual autocorrelation is non-significant (say, p > 0.05), the spatial term
may not be needed. Compare WAIC between spatial and non-spatial models to
confirm. Adding a spatial field to a model with no spatial signal costs
computation and can destabilise estimation because the SPDE parameters become
poorly identified when the spatial variance is near zero.

### Mesh resolution vs. compute trade-offs

The mesh is the computational bottleneck in spatial INLA models. The number
of mesh nodes ($n$) determines both the quality of the spatial
approximation and the cost of each EM iteration (roughly $O(n^{3/2})$ for
the Cholesky factorisation in 2D).

| Mesh density | Typical $n$ (300 sites) | Wall time | Use case |
|:---|:---:|:---:|:---|
| Very coarse (`max.edge = c(0.2, 0.5)`) | 50--100 | seconds | Quick exploration, large-scale screening |
| Default auto-mesh | 200--500 | 10--30 sec | Most analyses |
| Fine (`max.edge = c(0.03, 0.15)`) | 1000--3000 | 1--5 min | Publication maps, short-range fields |
| Very fine (`max.edge = c(0.01, 0.05)`) | 5000+ | 10+ min | Only if justified by data density |

**Rule of thumb:** the inner `max.edge` should be no larger than the
expected spatial range, and no smaller than $1/3$ of the minimum
inter-site distance. If your sites are regularly spaced at 1 km intervals
and you expect a range of 5 km, `max.edge = c(2, 10)` (in km) is a good
starting point.

The `cutoff` parameter prevents mesh nodes from being placed too close
together. Increasing `cutoff` in data-dense regions (e.g., clustered sites)
reduces $n$ with minimal loss of accuracy. The auto-mesh sets
`cutoff = max.edge[1] / 5`, which is adequate for most datasets.

### Prior sensitivity for SPDE parameters

The PC priors on the spatial range and marginal SD are the most influential
user choices in a spatial model. Unlike fixed-effect priors (which are
typically overwhelmed by data), the SPDE priors can shift
the estimated spatial field, especially when data are sparse or the spatial
signal is weak.

**Range prior sensitivity.** The range prior `prior.range = c(r0, p)`
controls the minimum spatial scale the model considers plausible. Setting
$r_0$ too large forces the field to be smooth, potentially masking
short-range patterns. Setting $r_0$ too small allows the field to fit noise.

```{r prior-sensitivity, fig.width = 9, fig.height = 4}
# Tight range prior: range unlikely < 0.5
sp_tight <- occu_spatial(sim$data$coords,
                         prior.range = c(0.5, 0.01),
                         prior.sigma = c(1, 0.05))
fit_tight <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                  spatial = sp_tight, verbose = 0)

# Loose range prior: range unlikely < 0.05
sp_loose <- occu_spatial(sim$data$coords,
                         prior.range = c(0.05, 0.5),
                         prior.sigma = c(1, 0.05))
fit_loose <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                  spatial = sp_loose, verbose = 0)

par(mfrow = c(1, 2))
occuMap(fit_tight, type = "spatial",
        main = sprintf("Tight prior (range = %.2f)",
                       spatialRange(fit_tight)$mean[1]))
occuMap(fit_loose, type = "spatial",
        main = sprintf("Loose prior (range = %.2f)",
                       spatialRange(fit_loose)$mean[1]))
```

The tight prior (left) produces a smoother field because short-range
fluctuations are penalised. The loose prior (right) allows more local
variation. If the two maps look similar and the estimated ranges agree, the
data are informative and the prior choice is not critical. If they diverge,
report results under both priors or use domain knowledge to choose.

**Variance prior sensitivity.** The sigma prior `prior.sigma = c(s0, p)`
controls how much variation the spatial field can absorb. Setting $s_0$ too
small allows the field to soak up variation that belongs to fixed effects,
potentially biasing covariate estimates toward zero. The PC prior framework
penalises complexity (i.e., large $\sigma$), which provides automatic
regularisation.

**Practical guidance on prior elicitation:**

1. Set `prior.range` based on the minimum distance over which you expect
   occupancy to change. For a bird species in heterogeneous habitat, this
   might be 1--5 km. For a sessile organism in a homogeneous landscape, it
   could be 20+ km.
2. Set `prior.sigma` based on how much unexplained spatial variation you
   expect on the logit scale. A logit-scale SD of 1 corresponds to a
   roughly 3-fold change in odds between nearby sites. Values above 2 are
   rarely justified for single-species models.
3. When coordinates are in disparate units (degrees vs. metres), rescale
   to a [0, 1] box and set priors relative to the unit square. This makes
   the defaults portable across studies.

### When CAR/BYM2 beats SPDE

The SPDE and CAR/BYM2 approaches model different kinds of spatial
dependence, and the choice should match the data structure:

| Criterion | SPDE | CAR/BYM2 |
|:---|:---|:---|
| Data type | Point locations with coordinates | Areal units (counties, grid cells) |
| Adjacency | Not needed (continuous field) | Required (shared borders) |
| Resolution | Continuous; can predict anywhere | Discrete; one value per region |
| Computation | Scales with mesh nodes | Scales with number of regions |
| Range estimation | Yes (Matern range parameter) | No (implicit from adjacency) |
| Disconnected regions | Handled via mesh extension | BYM2 handles; pure Besag does not |

**Do not use SPDE with region centroids.** If your data are counts or
proportions aggregated to administrative units, the centroids do not
represent point locations and the distance-based Matern covariance is
inappropriate. Use `occu_areal()` with the actual adjacency structure.

**Do not use CAR for point data.** If you have GPS coordinates for each
site, the SPDE approach is natural. Converting point data to a grid and
then applying CAR discards information and introduces arbitrary decisions
about grid resolution.

### Minimum sample size guidance

Spatial models require sufficient data to estimate both the fixed effects
and the spatial field. The minimum depends on the spatial range relative to
the study extent:

- **$N < 30$ sites:** A spatial model is almost never justified. The SPDE
  parameters will be poorly identified and the mesh will have more nodes
  than data points. Use a non-spatial model.
- **$30 \leq N < 100$ sites:** A spatial model can work if sites span the
  study area and the spatial range is not too short. Use strong PC priors
  (tighter than the defaults) and check that the estimated range is not
  hitting the prior boundary.
- **$100 \leq N < 500$ sites:** The sweet spot for most ecological studies.
  The default auto-mesh and priors should work well.
- **$N > 500$ sites:** The SPDE approach scales efficiently. Consider
  increasing `cutoff` to prevent the mesh from becoming excessively dense
  in data-rich clusters.
- **$N > 5000$ sites:** The EM iterations can be slow. Options include
  coarsening the mesh, increasing `num.threads` for parallel INLA,
  subsampling sites, or fitting regional submodels.

For multi-season models, the effective sample size is $N \times T$ (sites
times seasons), so even modest $N$ can support a spatial model when many
seasons are available.

### Trust the auto-mesh

The default mesh construction works for most datasets (up to a few
thousand sites). Only tune `max.edge`, `cutoff`, or the PC priors when you
have domain knowledge about the spatial range or when diagnostics suggest
problems.

### Match prior units to coordinate units

If your coordinates are in UTM metres, `prior.range = c(5000, 0.5)` means
you believe the range is likely at least 5 km. If coordinates are in
decimal degrees, the prior is in degrees. Rescaling to a unit square
avoids confusion and makes the defaults more portable.

### Use WAIC for model comparison

A spatial model that does not improve WAIC over the non-spatial version is
adding complexity without payoff. The spatial term should be justified by
the data, not added by default. For space-time models, compare:

1. Non-spatial, no temporal effect (baseline)
2. Temporal only (`temporal = "ar1"`)
3. Spatial only (`spatial = coords`)
4. Space-time (`spatial = coords, temporal = "ar1"`)

Each step should reduce WAIC if the added structure is real. A common
pattern is that the spatial model improves WAIC by 10+ units but adding
temporal structure provides only marginal gain (or vice versa), which tells
you where the dominant source of unmodelled variation lies.

### Check the spatial field map

`occuMap(fit, type = "spatial")` shows the spatial random effect alone. If
this map shows sharp features that track a known covariate you omitted,
consider adding that covariate instead of relying on the spatial field to
absorb it. The spatial field should capture genuinely unstructured residual
variation, not proxy for missing covariates.

### Convergence and the EM algorithm

The EM algorithm for spatial models typically converges in 10--30
iterations. If convergence is slow (> 40 iterations) or the log-likelihood
oscillates, consider:

- **Increasing damping** (`damping = 0.5` or higher). The damping parameter
  controls how much the occupancy state $z$ is updated at each iteration.
  Higher damping stabilises convergence at the cost of more iterations.
- **Stronger PC priors.** Weakly identified spatial parameters can cause the
  EM to wander. Tighter priors help anchor the spatial field.
- **Reducing mesh complexity.** An overly fine mesh can cause the SPDE
  parameters to be poorly identified because the field can fit noise.
  Start coarse and refine.

The `verbose = 1` option prints iteration-level summaries including the
log-likelihood and parameter changes, which is useful for diagnosing
convergence issues.
