---
title: "Spatially Varying Coefficient Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatially Varying Coefficient Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

Standard occupancy models assume that a covariate has the same effect
everywhere: one slope, applied uniformly across all sites. In practice,
ecological relationships are often spatially non-stationary. Elevation might
drive occupancy strongly in one region but have little effect in another. Forest
cover might be positively associated with occurrence near the range centre but
negatively associated at range edges where the species is already marginal. When
a constant-coefficient model is forced onto data with this kind of spatial
heterogeneity, the global estimate is a compromise that accurately describes no
region in particular.

Spatially varying coefficient (SVC) models address this by allowing one or more
covariate effects to change continuously across space. Instead of a single slope
$\beta_1$, each site $i$ gets its own coefficient $\beta_1(\mathbf{s}_i)$,
smoothed by a Gaussian spatial process. The global (mean) effect is still
estimated --- it answers "on average, what is the relationship?" --- but the
spatial surface of deviations around that mean reveals *where* the relationship
is stronger, weaker, or reversed.

This vignette covers fitting SVC occupancy models in INLAocc, extracting and
visualising the varying coefficients, combining SVCs with temporal structure,
multi-species extensions, and practical guidelines for when SVCs are warranted
and how to interpret them. For the foundations of spatial occupancy modelling
(SPDE mesh construction, PC priors, spatial residual checks), see
`vignette("spatial-models")`. For general model diagnostics and comparison
tools, see `vignette("diagnostics")`.


## The SVC occupancy model

The occupancy submodel with a spatially varying coefficient on one covariate is:

$$\text{logit}(\psi_i) = \beta_0 + \beta_1(\mathbf{s}_i) \cdot x_{1i} + w(\mathbf{s}_i)$$

where the spatially varying slope decomposes as:

$$\beta_1(\mathbf{s}_i) = \bar{\beta}_1 + \tilde{\beta}_1(\mathbf{s}_i)$$

- $\bar{\beta}_1$ is the **global (mean) effect** of $x_1$ across all sites.

- $\tilde{\beta}_1(\mathbf{s}_i) \sim \text{GP}(0, C_\beta(\cdot))$ is the
  **spatial deviation** at site $i$, drawn from a Gaussian process with its own
  Matern covariance function $C_\beta$.
- $w(\mathbf{s}_i)$ is a **separate spatial random effect** (the intercept
  spatial field) that captures residual spatial variation in baseline occupancy.

The SVC therefore has two distinct spatial fields: the intercept field
$w(\mathbf{s}_i)$ and the coefficient field $\tilde{\beta}_1(\mathbf{s}_i)$.
Each has its own range and variance parameters, controlling how smoothly they
vary across space and how much variation they carry.

The detection submodel is standard --- no SVC is applied to detection:

$$\text{logit}(p_{ij}) = \mathbf{w}_{ij}^\top \boldsymbol{\alpha}$$

The spatial smoothing via the Gaussian process prevents overfitting to
individual sites. Sites close together share information about the local
coefficient value, which is particularly important when per-site sample sizes
(number of visits) are modest.

Spatial non-stationarity is common in ecology. Elevation effects on bird
occupancy change with latitude because the treeline shifts: 1000 m is alpine
tundra in the Alps but montane forest in the tropics. Temperature effects on
plant occurrence differ between continental and maritime climates because
humidity mediates the physiological response. Land-use intensity has different
consequences depending on the regional species pool: converting forest to
agriculture in a region with many forest specialists causes steep occupancy
declines, while the same conversion in a region dominated by generalists has
less effect. In each case, the slope of the covariate-occupancy relationship
changes across space because of an interaction with geography that the analyst
may not have measured or may not know about in advance.

The SVC captures these patterns without requiring the analyst to specify the
interaction. Instead of writing `occupancy ~ elevation * latitude`, which
assumes a specific parametric form for the interaction, the SVC lets the data
determine where and how the elevation effect changes. This is useful when the
interaction is nonlinear, when the interacting variable is unknown, or when
multiple interacting variables operate simultaneously. The cost is that the SVC
is harder to interpret than a simple interaction term, because the spatial
surface of deviations does not come with a label saying "this is latitude" or
"this is humidity." The analyst must examine the surface and propose
explanations.


## Simulating SVC data

```{r simulate-data}
library(INLAocc)

# Simulate spatial occupancy data (generates a spatial random effect)
sim <- simulate_occu(N = 300, J = 4,
                     beta_occ = c(0.5, -0.8),
                     beta_det = c(0, -0.5),
                     n_occ_covs = 1, n_det_covs = 1,
                     spatial_range = 0.3,
                     seed = 42)
```

`simulate_occu()` generates a spatial random effect on the intercept but not a
true SVC. To create data where a covariate effect genuinely varies across space,
construct the varying coefficient manually:

```{r simulate-svc-manual}
# Extract site coordinates
coords <- sim$data$coords

# True SVC: effect of occ_x1 varies with latitude
# Stronger negative effect in the south, weaker (approaching zero) in the north
svc_true <- -0.8 + 1.5 * (coords[, 2] - 0.5)

# Verify: effect at southern edge (y = 0) is -0.8 + 1.5*(-0.5) = -1.55
#          effect at northern edge (y = 1) is -0.8 + 1.5*(0.5)  = -0.05
range(svc_true)
```

This gradient --- a covariate that matters strongly in the south but barely at
all in the north --- is exactly the kind of pattern an SVC model should recover.


## Fitting the SVC model

### Spatially varying intercept

The simplest SVC places spatial variation on the intercept itself. This is
distinct from the standard spatial random effect: the intercept SVC interacts
with the constant "1" in the design matrix, so it is functionally equivalent to
a spatial intercept field, but parameterised through the SVC machinery.

```{r fit-svc-intercept, eval = FALSE}
fit_svc_int <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                    spatial = sim$data$coords, svc = 1, verbose = 0)
summary(fit_svc_int)
```

`svc = 1` means the first occupancy coefficient (the intercept) varies
spatially. The global intercept $\bar{\beta}_0$ is the spatial mean, and each
site gets a deviation $\tilde{\beta}_0(\mathbf{s}_i)$.

### Spatially varying slope

More commonly, the goal is to let a covariate's effect vary. Here, the slope of
`occ_x1` changes across space:

```{r fit-svc-slope}
fit_svc <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                spatial = sim$data$coords, svc = 2, verbose = 0)
summary(fit_svc)
```

`svc = 2` targets the second occupancy coefficient --- the slope of `occ_x1`.
The model estimates both a global slope $\bar{\beta}_1$ and a spatial surface of
deviations $\tilde{\beta}_1(\mathbf{s}_i)$.

Reading this output:

- **Global effect of occ_x1**: the mean effect on the logit scale. Across all
  sites on average, a one-unit increase in `occ_x1` changes log-odds of
  occupancy by this amount.
- **Spatial Component**: the intercept spatial field captures residual spatial
  variation in baseline occupancy after accounting for covariates.
- **SVC Component**: the coefficient spatial field. The SVC range controls how
  smoothly the coefficient changes across space (larger range = smoother). The
  SVC standard deviation measures how much the coefficient varies across space.

The global effect is the "average" relationship between the covariate and
occupancy, integrated across the entire study area. It answers the same
question as a standard (non-SVC) model would. The SVC range tells you the
spatial scale over which the relationship changes. A short range means the
coefficient shifts rapidly from site to site, implying fine-grained spatial
heterogeneity in the ecological process. A long range means broad, regional
gradients where the coefficient changes slowly. If the range is much larger
than the study extent, the coefficient is approximately constant and the SVC
adds little beyond what a fixed slope provides.

The SVC standard deviation measures total spatial variation in the coefficient.
Compare it to the magnitude of the global effect. If the global effect is -0.8
and the SVC SD is 0.1, the coefficient ranges from roughly -0.9 to -0.7 across
space, and the relationship is qualitatively the same everywhere. If the SVC SD
is 1.2, the coefficient ranges from about -2.0 to +0.4, meaning the
relationship reverses sign in parts of the study area. That is a qualitatively
different situation and warrants careful ecological interpretation. When the SVC
SD is small relative to the global effect, consider dropping the SVC in favor
of the simpler constant-coefficient model.

### Multiple SVCs

When theory suggests that more than one covariate has a spatially varying
effect, pass a vector to `svc`:

```{r fit-svc-multi}
sim2 <- simulate_occu(N = 300, J = 4, n_occ_covs = 2, n_det_covs = 1,
                       beta_occ = c(0.5, -0.8, 0.4),
                       beta_det = c(0, -0.5),
                       spatial_range = 0.3, seed = 43)
fit_svc_multi <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim2$data,
                      spatial = sim2$data$coords, svc = c(1, 2), verbose = 0)
```

Both the intercept and the `occ_x1` slope vary spatially. Each SVC adds a
separate spatial field to the model, so computational cost scales with the
number of SVCs. Use multiple SVCs sparingly and only when each has a clear
ecological justification.


## Extracting SVC surfaces

`getSVCSamples()` returns the posterior summary of the site-level coefficient:
the total effect $\beta_1(\mathbf{s}_i) = \bar{\beta}_1 + \tilde{\beta}_1(\mathbf{s}_i)$
at each site.

```{r extract-svc}
svc_all <- getSVCSamples(fit_svc)
svc_samples <- svc_all[[1]]  # extract the data.frame for the first SVC covariate
head(svc_samples)
```

Each row gives the posterior mean, standard deviation, and 95% credible interval
for the total effect of `occ_x1` at that site. Sites where the interval
excludes zero have a "significant" local effect; sites where it spans zero may
have no local effect even if the global effect is non-zero.

### Visualising the SVC surface

A map of site-level coefficients reveals the spatial pattern:

```{r plot-svc-surface}
svc_df <- data.frame(
  x = sim$data$coords[, 1],
  y = sim$data$coords[, 2],
  beta1 = svc_samples$mean
)

# Base R visualisation
col_idx <- cut(svc_df$beta1, breaks = 100, labels = FALSE)
plot(svc_df$x, svc_df$y,
     col = hcl.colors(100, "Blue-Red 3")[col_idx],
     pch = 19, cex = 1.2,
     xlab = "Easting", ylab = "Northing",
     main = "Spatially varying effect of occ_x1")
```

When examining an SVC map, look for spatial structure that matches known
geography. A north-south gradient in the coefficient might indicate an
interaction with temperature or latitude. East-west variation near a coastline
might reflect maritime influence on the ecological relationship. Patchy
variation at scales smaller than the spacing between sites usually indicates
noise rather than signal: the SVC is interpolating between sparse data points
and the resulting pattern is not ecologically interpretable. If the map looks
like static, the SVC may be overfitting, and the constant-coefficient model is
likely a better choice.

For publication-quality maps, use `ggplot2`:

```{r plot-svc-ggplot}
library(ggplot2)

ggplot(svc_df, aes(x = x, y = y, colour = beta1)) +
  geom_point(size = 2.5) +
  scale_colour_gradient2(low = "#2166AC", mid = "grey90", high = "#B2182B",
                         midpoint = 0, name = expression(beta[1](s))) +
  coord_equal() +
  labs(title = "Spatially varying effect of occ_x1",
       x = "Easting", y = "Northing") +
  theme_minimal(base_size = 13)
```

### Mapping uncertainty

The posterior standard deviation reveals where the SVC is well-estimated versus
uncertain:

```{r plot-svc-uncertainty}
svc_df$sd <- svc_samples$sd

ggplot(svc_df, aes(x = x, y = y, colour = sd)) +
  geom_point(size = 2.5) +
  scale_colour_viridis_c(option = "magma", name = "Posterior SD") +
  coord_equal() +
  labs(title = "Uncertainty in spatially varying effect",
       x = "Easting", y = "Northing") +
  theme_minimal(base_size = 13)
```

Regions with few sites or low detection probability will have wider credible
intervals. If the uncertainty map shows uniformly high SD, the data may not
support an SVC --- consider dropping back to a constant coefficient.

### Posterior predictive check

Verify model adequacy after fitting the SVC:

```{r ppc-svc}
ppc <- ppcOccu(fit_svc, fit.stat = "freeman-tukey", group = 1)
ppc$bayesian.p
```

A Bayesian p-value near 0.5 confirms that the SVC model generates data consistent with observations.

### Parameter recovery

When working with simulated data, compare the estimated SVC surface against the known truth:

```{r recovery-svc}
# Correlation between true and estimated site-level coefficients
cor(svc_true, svc_samples$mean)
```

A high correlation between the true spatially varying coefficient and the estimated surface indicates good recovery. The correlation will not be 1.0 because: (a) estimation uncertainty at each site, (b) the Gaussian process smooths over sharp local variation, and (c) the SPDE mesh approximation introduces discretisation error. With more sites or higher detection probability, the recovery improves.


## Model comparison: SVC vs. constant coefficient

An SVC should only be retained if it meaningfully improves model fit. Always
compare against simpler alternatives:

```{r model-comparison}
# Baseline: no spatial effect
m_base <- occu(~ occ_x1, ~ det_x1, data = sim$data, verbose = 0)

# Spatial intercept only (no SVC)
m_spatial <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                  spatial = sim$data$coords, verbose = 0)

# SVC on intercept
m_svc_int <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                  spatial = sim$data$coords, svc = 1, verbose = 0)

# SVC on occ_x1 slope
m_svc_slope <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                    spatial = sim$data$coords, svc = 2, verbose = 0)

compare_models(base = m_base,
               spatial = m_spatial,
               svc_int = m_svc_int,
               svc_slope = m_svc_slope,
               criterion = "waic")
```

Interpretation:

- **base to spatial**: a WAIC improvement confirms spatial structure matters.

- **spatial to svc_int**: any improvement indicates spatial variation in
  baseline occupancy beyond what the standard spatial field captures.
- **spatial to svc_slope**: a substantial improvement indicates the effect of
  `occ_x1` genuinely varies across space.

If the SVC model does not improve WAIC over the spatial intercept model, the
constant coefficient is sufficient and the additional complexity is not
justified. A spatially varying intercept captures spatial variation in *baseline*
occupancy. An SVC on a slope captures spatial variation in *how the covariate
affects occupancy*. These are different ecological hypotheses --- compare them
explicitly rather than assuming one subsumes the other.

Each step in this comparison tests a specific ecological hypothesis. Going from
base to spatial tests "is there unexplained spatial structure in occupancy?"
Going from spatial to SVC-intercept tests "does baseline occupancy vary beyond
what the covariates and a single spatial field explain?" Going from spatial to
SVC-slope tests "does the effect of the covariate change across space?" The
WAIC differences between these nested models answer each question
quantitatively. A large WAIC drop at one step but not another tells you which
kind of spatial complexity the data support.


## Prediction at new locations

SVC predictions at unsampled sites incorporate the spatially varying coefficient. Pass new coordinates and covariates to `predict_spatial()`:

```{r predict-svc}
new_coords <- as.matrix(expand.grid(x = seq(0, 1, by = 0.05),
                                     y = seq(0, 1, by = 0.05)))
new_covs <- data.frame(occ_x1 = 0)

preds <- predict_spatial(fit_svc, newcoords = new_coords,
                         newocc.covs = new_covs)
head(preds)
```

The predictions reflect both the spatial intercept field and the SVC. Even with `occ_x1` held at 0, occupancy varies across space because of the intercept spatial field. To see the SVC's contribution, compare predictions at `occ_x1 = 0` vs `occ_x1 = 1` across the same grid --- the difference will vary spatially, which is the SVC in action.


## SVC + temporal models

For multi-season data where the covariate effect might vary in space but the
temporal dynamics also matter, combine SVC with temporal structure:

```{r fit-svc-temporal}
sim_t <- simTOcc(N = 100, J = 3, n_seasons = 5, seed = 300)

fit_svc_t <- occu(~ occ_x1, ~ det_x1, data = sim_t$data,
                  spatial = sim_t$data$coords,
                  temporal = "ar1", svc = 2, verbose = 0)
summary(fit_svc_t)
```

This model combines AR(1) temporal autocorrelation with a spatially varying
slope on `occ_x1`. The SVC captures spatial non-stationarity in the
covariate--occupancy relationship, while the AR(1) component captures temporal
persistence in occupancy states. This is the most complex single-species model
available in INLAocc and should be used only when both spatial non-stationarity
and temporal autocorrelation are ecologically plausible.

A sensible workflow is to first establish whether the SVC is warranted using
spatial-only models, then add temporal structure to the preferred spatial
specification. Do not add both simultaneously on a first attempt --- it becomes
difficult to diagnose which component is absorbing which signal.


## Multi-species SVC

When multiple species are surveyed at the same sites, each species can have its
own spatially varying coefficient:

```{r fit-ms-svc, eval = FALSE}
sim_ms <- simMsOcc(N = 200, J = 4, n_species = 5,
                   spatial_range = 0.3, seed = 123)

fit_ms_svc <- occu(~ occ_x1, ~ det_x1, data = sim_ms$data,
                   multispecies = TRUE,
                   spatial = sim_ms$data$coords, svc = 2, verbose = 0)
```

Each species gets its own SVC field --- the effect of `occ_x1` varies spatially
*and* differs across species. A generalist might show a flat SVC (constant
effect everywhere), while a specialist might show strong spatial variation. The
species-level SVC surfaces from `getSVCSamples()` can be compared to understand
how species respond differently to the same environmental gradient across space.

This is computationally demanding: each species adds a separate spatial field.
With 5 species and `svc = 2`, the model estimates 5 SVC fields plus 5 intercept
spatial fields (10 spatial fields total). Use only with adequate sample sizes
(200+ sites, 3+ visits per site per species).


## Practical guidelines for interpretation

### Don't assign SVCs arbitrarily

SVCs should reflect a specific ecological hypothesis: "the effect of $X$ varies
across space because of [mechanism]." Plausible mechanisms include:

- **Local adaptation**: species respond differently to a resource in different
  parts of their range.
- **Interaction with an unmeasured gradient**: temperature modulates the effect
  of precipitation, and temperature varies spatially.
- **Range boundary effects**: a habitat covariate matters more at the range edge
  where the species is at its physiological limit.

Without a mechanistic rationale, an SVC may be overfitting to noise or absorbing
the signal of an omitted covariate.

### Start simple, compare up

The recommended workflow for deciding whether an SVC is needed:

1. Fit the model without any spatial effect.

2. Add a spatial intercept (captures unmeasured spatial variation in baseline
   occupancy).
3. Add an SVC on the hypothesised covariate.

4. Compare all three via WAIC using `compare_models()`.

If the SVC does not improve on the spatial intercept model, the covariate effect
is likely constant across space, and the simpler model should be preferred.

### Visualise the SVC surface

Always map the site-level coefficient estimates. Diagnostic patterns to look
for:

- **Smooth gradients**: likely real ecological non-stationarity (e.g., a
  temperature effect that varies with latitude). This is the ideal case ---
  the SVC is capturing a genuine spatial process.
- **Patchy or noisy pattern**: possibly overfitting, or confounding with another
  spatially structured process. Consider whether an omitted covariate could
  explain the pattern.
- **Sign changes**: the covariate has opposite effects in different regions.
  This can be ecologically meaningful (positive effect at range centre, negative
  at range edge) or a sign of confounding. Investigate carefully before
  reporting.
- **Uniform surface with high uncertainty**: the data do not support an SVC.
  The constant-coefficient model is sufficient.

### Consider the spatial scale

The SVC range parameter controls how smoothly the coefficient varies. Its
relationship to the study extent determines what the model can resolve:
\vspace{0.5em}

- **SVC range >> study extent**: the effect is nearly constant across the study
  area. The SVC is not needed; a fixed coefficient suffices.
- **SVC range << inter-site distance**: the spatial process varies faster than
  the sampling can resolve. Site-level estimates will be noisy and poorly
  constrained.
- **SVC range $\approx$ study extent / 5--10**: the ideal regime. The spatial
  variation is smooth enough to estimate with reasonable precision but fast
  enough to reveal meaningful heterogeneity.

If the posterior for SVC range piles up against the upper boundary of the prior,
it is signalling that the effect is effectively constant. If it piles up near
zero, the data may be too sparse to support an SVC.

### Prior sensitivity

SVC spatial variance and range priors matter more than in standard spatial
models, because the SVC field competes with the intercept spatial field for
signal. The default PC (penalised complexity) priors in INLAocc are calibrated
to shrink toward the constant-coefficient model (i.e., zero SVC variance). This
is conservative by design: the SVC must earn its complexity by improving the
marginal likelihood.

If results are sensitive to the choice of PC prior parameters, the data are
likely marginal for supporting an SVC. Report the sensitivity analysis alongside
the primary results.

In practice, this means fitting the SVC under at least two prior
specifications. For example, change the PC prior from $P(\sigma > 1) = 0.05$
to $P(\sigma > 2) = 0.1$ and refit. If the SVC surface changes substantially
in magnitude or spatial pattern, the data do not contain enough information to
pin down the SVC, and the posterior is being driven by the prior rather than
the likelihood. Report both results side by side. If the surface is stable
across reasonable prior choices, the SVC is well-identified and the conclusions
do not depend on subjective prior decisions.

### Identifiability concerns

With both an intercept spatial field $w(\mathbf{s}_i)$ and an SVC field
$\tilde{\beta}_1(\mathbf{s}_i)$, the model has two spatial processes that can
partially trade off against each other. This trade-off becomes problematic when:

- The SVC covariate $x_1$ is itself spatially structured (e.g., elevation,
  which varies smoothly across space). The intercept field and the SVC field
  may both try to capture the same spatial signal, leading to poorly identified
  variance components.
- The data are sparse relative to the spatial scale of variation. Few sites
  means few data points to separate the two spatial processes.

Use `checkIdentifiability()` to flag potential issues:

```{r check-identifiability}
checkIdentifiability(fit_svc)
```

If identifiability warnings appear, consider whether the SVC covariate is too
collinear with the spatial structure of the data. Options include:
residualising the covariate against a spatial smooth before fitting, using
stronger PC priors to regularise the SVC, or accepting the spatial intercept
model and reporting the global coefficient only.


## Summary

- SVCs decompose a covariate effect into a global mean plus a spatial surface of
  deviations. They answer "where does this relationship hold, and where does it
  break down?"
- Fit with `occu(..., spatial = coords, svc = k)` where `k` indexes the
  occupancy covariate. Extract surfaces with `getSVCSamples()`.
- SVCs are the most complex single-species model in INLAocc. Use only with
  ecological justification and adequate sample size (200+ sites).
- Always compare to simpler alternatives: no spatial effect, spatial intercept
  only, constant coefficient with spatial intercept.
- Visualise the SVC surface. If it looks like noise, it probably is.

- SVC + temporal is slow. Establish that the SVC is warranted in a
  spatial-only model before adding temporal structure.
- Prior sensitivity and identifiability checks are essential, not optional.

The full decision tree for SVC analysis: (1) fit the base model without spatial
effects, (2) check Moran's I on residuals for spatial autocorrelation, (3) add
a spatial intercept field, (4) hypothesize which covariate might have a
spatially varying effect based on ecological reasoning, (5) fit the SVC model,
(6) compare all models via WAIC, (7) visualize the SVC surface and check
whether the spatial pattern is interpretable, (8) run a prior sensitivity
analysis to confirm the results are stable. If the SVC does not improve WAIC,
or the surface looks like noise, or the results shift under different priors,
fall back to the spatial intercept model and report the global coefficient.
