---
title: "Spatially Varying Coefficient Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Spatially Varying Coefficient Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

Standard occupancy models assume that a covariate has the same effect
everywhere: one slope, applied uniformly across all sites. In practice,
ecological relationships are often spatially non-stationary. Elevation might
drive occupancy strongly in one region but have little effect in another. Forest
cover might be positively associated with occurrence near the range centre but
negatively associated at range edges where the species is already marginal. When
a constant-coefficient model is forced onto data with this kind of spatial
heterogeneity, the global estimate is a compromise that accurately describes no
region in particular.

Spatially varying coefficient (SVC) models address this by allowing one or more
covariate effects to change continuously across space. Instead of a single slope
$\beta_1$, each site $i$ gets its own coefficient $\beta_1(\mathbf{s}_i)$,
smoothed by a Gaussian spatial process. The global (mean) effect is still
estimated (it answers "on average, what is the relationship?"), but the
spatial surface of deviations around that mean reveals *where* the relationship
is stronger, weaker, or reversed.

This vignette covers fitting SVC occupancy models in INLAocc, extracting and
visualising the varying coefficients, combining SVCs with temporal structure,
multi-species extensions, and practical guidelines for when SVCs are warranted
and how to interpret them. For the foundations of spatial occupancy modelling
(SPDE mesh construction, PC priors, spatial residual checks), see
`vignette("spatial-models")`. For general model diagnostics and comparison
tools, see `vignette("diagnostics")`.


## The SVC occupancy model

The occupancy submodel with a spatially varying coefficient on one covariate is:

$$\text{logit}(\psi_i) = \beta_0 + \beta_1(\mathbf{s}_i) \cdot x_{1i} + w(\mathbf{s}_i)$$

where the spatially varying slope decomposes as:

$$\beta_1(\mathbf{s}_i) = \bar{\beta}_1 + \tilde{\beta}_1(\mathbf{s}_i)$$

- $\bar{\beta}_1$ is the **global (mean) effect** of $x_1$ across all sites.

- $\tilde{\beta}_1(\mathbf{s}_i) \sim \text{GP}(0, C_\beta(\cdot))$ is the
  **spatial deviation** at site $i$, drawn from a Gaussian process with its own
  Matern covariance function $C_\beta$.
- $w(\mathbf{s}_i)$ is a **separate spatial random effect** (the intercept
  spatial field) that captures residual spatial variation in baseline occupancy.

The SVC therefore has two distinct spatial fields: the intercept field
$w(\mathbf{s}_i)$ and the coefficient field $\tilde{\beta}_1(\mathbf{s}_i)$.
Each has its own range and variance parameters, controlling how smoothly they
vary across space and how much variation they carry.

The detection submodel is standard; no SVC is applied to detection:

$$\text{logit}(p_{ij}) = \mathbf{w}_{ij}^\top \boldsymbol{\alpha}$$

### The Matern covariance and SPDE approximation

Each spatial field in the SVC model (both the intercept field $w(\mathbf{s})$
and the coefficient deviation field $\tilde{\beta}_1(\mathbf{s})$) is a
Gaussian random field characterised by a Matern covariance function. The Matern
covariance between two locations $\mathbf{s}$ and $\mathbf{s}'$ separated by
distance $d = \|\mathbf{s} - \mathbf{s}'\|$ is:

$$C(d) = \frac{\sigma^2}{2^{\nu - 1} \Gamma(\nu)} \left(\frac{d}{\rho}\right)^\nu K_\nu\left(\frac{d}{\rho}\right)$$

where $\sigma^2$ is the marginal variance (how much the field fluctuates),
$\rho$ is the practical range (the distance at which correlation drops to
approximately 0.13), $\nu$ is the smoothness parameter, and $K_\nu$ is the
modified Bessel function of the second kind. In INLA's SPDE implementation,
the smoothness is fixed at $\nu = 1$ (in 2D), which corresponds to an
exponential-like covariance. This is not a limitation in practice: the range
and variance parameters are the primary drivers of the field's behaviour, and
$\nu = 1$ produces fields that are continuous but not differentiable, which
is appropriate for most ecological surfaces.

Each SVC has its own $(\sigma^2, \rho)$ pair. The
intercept field might have a long range (broad-scale spatial variation in
baseline occupancy) and moderate variance, while the coefficient field might
have a shorter range (finer-scale variation in the covariate effect) and
smaller variance. The data determine these parameters through the likelihood,
and the PC priors regularise them toward a null model (zero variance, infinite
range, i.e., no spatial variation).

INLA does not fit the Gaussian process directly. Instead, it uses the SPDE
(Stochastic Partial Differential Equation) approach, which represents the
Matern field as a solution to a specific SPDE on a triangulated mesh. The
continuous field is approximated by a finite set of basis functions defined on
the mesh vertices, and the weights on these basis functions are the random
effects that INLA estimates. The projection matrix $\mathbf{A}$ maps from mesh
vertices to observation locations:

$$w(\mathbf{s}_i) \approx \sum_k A_{ik} \, w_k$$

where $w_k$ is the value at mesh vertex $k$ and $A_{ik}$ is the weight
of vertex $k$ at observation location $\mathbf{s}_i$. This is formally
equivalent to kriging: the predicted value at any location is a weighted
combination of values at nearby mesh nodes, with the weights determined by
the Matern covariance structure. The SPDE approach is computationally
efficient because the resulting precision matrix (inverse covariance) is
sparse, which INLA exploits for fast inference.

For an SVC model with $P$ spatially varying covariates (including the
intercept if it is spatially varying), the model has $P + 1$ spatial
fields if a separate intercept spatial field is also present: one intercept
field plus $P$ coefficient fields. Each field shares the same mesh but has
its own SPDE hyperparameters. The mesh is constructed once from the site
coordinates and reused for all fields, which is why the `spatial` argument
to `occu()` accepts either raw coordinates or a pre-built `occu_spatial()`
object.

### The general SVC formulation

The model extends naturally to multiple covariates. With $P$ covariates
that have spatially varying effects:

$$\text{logit}(\psi(\mathbf{s}_i)) = \sum_{h=0}^{P} \beta_h(\mathbf{s}_i) \cdot x_{hi}$$

where $x_{0i} = 1$ (the intercept "covariate"). Each $\beta_h(\mathbf{s}_i) = \bar{\beta}_h + \tilde{\beta}_h(\mathbf{s}_i)$ decomposes into a global mean and a spatial deviation. Some covariates may have constant effects (no SVC), in which case $\tilde{\beta}_h(\mathbf{s}_i) = 0$ for those covariates. The `svc` argument to `occu()` specifies which covariates get spatial deviations; all others are treated as constant across space.

The total effect of all covariates at site $i$ is:

$$\text{logit}(\psi(\mathbf{s}_i)) = \underbrace{\mathbf{x}_i^\top \bar{\boldsymbol{\beta}}}_{\text{global linear predictor}} + \underbrace{\sum_{h \in \text{SVC}} \tilde{\beta}_h(\mathbf{s}_i) \cdot x_{hi}}_{\text{spatially varying adjustments}}$$

This formulation makes clear that the SVC model nests several simpler models:

- **No spatial effects**: all $\tilde{\beta}_h = 0$, equivalent to a standard occupancy model.
- **Spatial intercept only**: $\tilde{\beta}_0(\mathbf{s}_i) \neq 0$ but all slope deviations are zero. This is the standard spatial occupancy model.
- **Full SVC**: one or more slope deviations are non-zero. Each non-zero slope deviation adds a spatial field to the model.

The spatial smoothing via the Gaussian process prevents overfitting to
individual sites. Sites close together share information about the local
coefficient value, which is particularly important when per-site sample sizes
(number of visits) are modest.

Spatial non-stationarity is common in ecology. Elevation effects on bird
occupancy change with latitude because the treeline shifts: 1000 m is alpine
tundra in the Alps but montane forest in the tropics. Temperature effects on
plant occurrence differ between continental and maritime climates because
humidity mediates the physiological response. Land-use intensity has different
consequences depending on the regional species pool: converting forest to
agriculture in a region with many forest specialists causes steep occupancy
declines, while the same conversion in a region dominated by generalists has
less effect. In each case, the slope of the covariate-occupancy relationship
changes across space because of an interaction with geography that the analyst
may not have measured or may not know about in advance.

The SVC captures these patterns without requiring the analyst to specify the
interaction. Instead of writing `occupancy ~ elevation * latitude`, which
assumes a specific parametric form for the interaction, the SVC lets the data
determine where and how the elevation effect changes. This is useful when the
interaction is nonlinear, when the interacting variable is unknown, or when
multiple interacting variables operate simultaneously. The cost is that the SVC
is harder to interpret than a simple interaction term, because the spatial
surface of deviations does not come with a label saying "this is latitude" or
"this is humidity." The analyst must examine the surface and propose
explanations.

A concrete example: elevation drives bird occupancy, but the strength of the effect depends on latitude. Near the equator, a 1000 m elevation range compresses many bioclimatic zones --- tropical lowland forest, cloud forest, elfin forest, paramo, into a short vertical distance. The occupancy-elevation slope is steep because moving up 500 m crosses multiple habitat boundaries. At 60°N latitude, the same 1000 m range spans fewer bioclimatic zones; the treeline is lower, the vegetation gradient is less compressed, and the elevation effect on occupancy is weaker. An SVC on elevation captures this latitude-dependent slope without the analyst needing to specify the latitude-elevation interaction term or its functional form.

Forest cover provides another example. In fragmented landscapes where forest is scarce, even small patches of 5 ha can sustain occupancy for forest-dependent species. Each additional hectare of forest cover matters. In continuous forest landscapes where forest is the matrix, occupancy is already high and adding more forest cover has diminishing returns; the effect saturates. The slope of the forest-cover-occupancy relationship is steeper in fragmented regions and flatter in continuous ones. An SVC on forest cover reveals this spatial variation directly: the map of site-level coefficients will show high slopes in fragmented areas and low slopes in forested areas, without the analyst needing to define "fragmentation" as a separate covariate or specify how it interacts with forest cover.

Soil moisture effects on plant occupancy vary between soil types. On clay soils that retain water, soil moisture operates in a narrow effective range and the occupancy-moisture relationship may be nonlinear or even humped. On sandy soils that drain quickly, any increase in moisture benefits plants and the relationship is monotonically positive. If the study area spans both soil types and the analyst has not measured soil texture, the constant-coefficient model estimates an average slope that describes neither soil type accurately. The SVC recovers the spatial pattern: sites on clay show one coefficient value, sites on sand show another, and the transition zone shows intermediate values. The analyst then sees the spatial pattern in the SVC map and can propose soil type as the explanation, which can be tested by adding soil texture as a covariate and checking whether the SVC shrinks toward a constant.


## Simulating SVC data

```{r simulate-data}
library(INLAocc)

# Simulate spatial occupancy data (generates a spatial random effect)
sim <- simulate_occu(N = 300, J = 4,
                     beta_occ = c(0.5, -0.8),
                     beta_det = c(0, -0.5),
                     n_occ_covs = 1, n_det_covs = 1,
                     spatial_range = 0.3,
                     seed = 42)
```

`simulate_occu()` generates a spatial random effect on the intercept but not a
true SVC. To create data where a covariate effect genuinely varies across space,
construct the varying coefficient manually:

```{r simulate-svc-manual}
# Extract site coordinates
coords <- sim$data$coords

# True SVC: effect of occ_x1 varies with latitude
# Stronger negative effect in the south, weaker (approaching zero) in the north
svc_true <- -0.8 + 1.5 * (coords[, 2] - 0.5)

# Verify: effect at southern edge (y = 0) is -0.8 + 1.5*(-0.5) = -1.55
#          effect at northern edge (y = 1) is -0.8 + 1.5*(0.5)  = -0.05
range(svc_true)
```

This gradient, a covariate that matters strongly in the south but barely at
all in the north, is exactly the kind of pattern an SVC model should recover.


## Fitting the SVC model

### Spatially varying intercept

The simplest SVC places spatial variation on the intercept itself. This is
distinct from the standard spatial random effect: the intercept SVC interacts
with the constant "1" in the design matrix, so it is functionally equivalent to
a spatial intercept field, but parameterised through the SVC machinery.

```{r fit-svc-intercept}
fit_svc_int <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                    spatial = sim$data$coords, svc = 1, verbose = 0)
summary(fit_svc_int)
```

`svc = 1` means the first occupancy coefficient (the intercept) varies
spatially. The global intercept $\bar{\beta}_0$ is the spatial mean, and each
site gets a deviation $\tilde{\beta}_0(\mathbf{s}_i)$.

### Spatially varying slope

More commonly, the goal is to let a covariate's effect vary. Here, the slope of
`occ_x1` changes across space:

```{r fit-svc-slope}
fit_svc <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                spatial = sim$data$coords, svc = 2, verbose = 0)
summary(fit_svc)
```

`svc = 2` targets the second occupancy coefficient, the slope of `occ_x1`.
The model estimates both a global slope $\bar{\beta}_1$ and a spatial surface of
deviations $\tilde{\beta}_1(\mathbf{s}_i)$.

Reading this output:

- **Global effect of occ_x1**: the mean effect on the logit scale. Across all
  sites on average, a one-unit increase in `occ_x1` changes log-odds of
  occupancy by this amount.
- **Spatial Component**: the intercept spatial field captures residual spatial
  variation in baseline occupancy after accounting for covariates.
- **SVC Component**: the coefficient spatial field. The SVC range controls how
  smoothly the coefficient changes across space (larger range = smoother). The
  SVC standard deviation measures how much the coefficient varies across space.

The global effect is the "average" relationship between the covariate and
occupancy, integrated across the entire study area. It answers the same
question as a standard (non-SVC) model would. The SVC range tells you the
spatial scale over which the relationship changes. A short range means the
coefficient shifts rapidly from site to site, implying fine-grained spatial
heterogeneity in the ecological process. A long range means broad, regional
gradients where the coefficient changes slowly. If the range is much larger
than the study extent, the coefficient is approximately constant and the SVC
adds little beyond what a fixed slope provides.

The SVC standard deviation measures total spatial variation in the coefficient.
Compare it to the magnitude of the global effect. If the global effect is -0.8
and the SVC SD is 0.1, the coefficient ranges from roughly -0.9 to -0.7 across
space, and the relationship is qualitatively the same everywhere. If the SVC SD
is 1.2, the coefficient ranges from about -2.0 to +0.4, meaning the
relationship reverses sign in parts of the study area. That is a qualitatively
different situation and warrants careful ecological interpretation. When the SVC
SD is small relative to the global effect, consider dropping the SVC in favor
of the simpler constant-coefficient model.

### Multiple SVCs

When theory suggests that more than one covariate has a spatially varying
effect, pass a vector to `svc`:

```{r fit-svc-multi}
sim2 <- simulate_occu(N = 300, J = 4, n_occ_covs = 2, n_det_covs = 1,
                       beta_occ = c(0.5, -0.8, 0.4),
                       beta_det = c(0, -0.5),
                       spatial_range = 0.3, seed = 43)
fit_svc_multi <- occu(~ occ_x1 + occ_x2, ~ det_x1, data = sim2$data,
                      spatial = sim2$data$coords, svc = c(1, 2), verbose = 0)
```

Both the intercept and the `occ_x1` slope vary spatially. Each SVC adds a
separate spatial field to the model, so computational cost scales with the
number of SVCs. Use multiple SVCs sparingly and only when each has a clear
ecological justification.

#### Interpreting multiple SVC surfaces

When fitting two or more SVCs simultaneously, each coefficient surface tells a
different story about the ecology. To interpret them properly, extract and
compare the surfaces side by side.

```{r multi-svc-extract}
svc_multi <- getSVCSamples(fit_svc_multi)
names(svc_multi)

# Build a data frame for both SVC surfaces
multi_df <- data.frame(
  x = rep(sim2$data$coords[, 1], 2),
  y = rep(sim2$data$coords[, 2], 2),
  coefficient = c(svc_multi[[1]]$mean, svc_multi[[2]]$mean),
  svc = rep(names(svc_multi), each = nrow(svc_multi[[1]]))
)

ggplot(multi_df, aes(x = x, y = y, colour = coefficient)) +
  geom_point(size = 1.8) +
  scale_colour_gradient2(low = "#2166AC", mid = "grey90", high = "#B2182B",
                         midpoint = 0, name = "Effect") +
  coord_equal() +
  facet_wrap(~ svc) +
  labs(title = "Side-by-side SVC surfaces",
       x = "Easting", y = "Northing") +
  theme_minimal(base_size = 11) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

Two aspects of the side-by-side comparison are informative. First, do the
spatial patterns align? If both SVCs show a north-south gradient, they may
both be responding to the same latent driver (e.g., temperature, latitude).
This does not invalidate the model, but it suggests that a single unmeasured
gradient explains both patterns, and including that gradient as a covariate
might simplify the model. Second, do the patterns differ? If the intercept
SVC shows east-west variation while the slope SVC shows north-south variation,
the underlying processes are likely different: baseline occupancy depends on
one unmeasured factor while the covariate effect depends on another.

#### Correlation between SVC surfaces

Formally test whether the two SVC surfaces are correlated:

```{r multi-svc-correlation}
svc1_mean <- svc_multi[[1]]$mean
svc2_mean <- svc_multi[[2]]$mean
cor_svc <- cor(svc1_mean, svc2_mean)
cat(sprintf("Correlation between SVC surfaces: %.3f\n", cor_svc))

# Scatter plot of the two SVC surfaces
svc_cor_df <- data.frame(
  intercept_svc = svc1_mean,
  slope_svc = svc2_mean
)

ggplot(svc_cor_df, aes(x = intercept_svc, y = slope_svc)) +
  geom_point(alpha = 0.4, colour = "#4A90D9") +
  geom_smooth(method = "lm", se = TRUE, colour = "#B2182B",
              fill = "#B2182B", alpha = 0.15) +
  labs(x = "Intercept SVC (local baseline occupancy)",
       y = "Slope SVC (local effect of occ_x1)",
       title = "Correlation between SVC surfaces") +
  theme_minimal(base_size = 13) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

A strong positive correlation means that sites with high baseline occupancy
also tend to have a strong covariate effect. A negative correlation means
the opposite: sites where the species is already common show a weak or
reversed covariate effect. No correlation (flat scatter) means the two
spatial processes are independent, which is the ideal case for identifiability.
High correlation (|r| > 0.7) can indicate confounding between the intercept
field and the SVC field and warrants caution in interpretation.

#### Variance partitioning across multiple SVCs

When the model contains multiple spatial fields, it is useful to quantify
how much variation each one contributes. A simple measure is the standard
deviation of the posterior means across sites:

```{r multi-svc-variance}
svc_var_summary <- data.frame(
  component = names(svc_multi),
  sd_across_sites = vapply(svc_multi, function(x) sd(x$mean), numeric(1)),
  mean_posterior_sd = vapply(svc_multi, function(x) mean(x$sd), numeric(1))
)
svc_var_summary
```

The `sd_across_sites` column measures how much each coefficient varies
spatially (the signal), while `mean_posterior_sd` measures how well that
variation is estimated (the noise). A component with high signal and low
noise is well-identified and contributes meaningfully to the model. A
component with low signal and high noise is poorly identified and may be
absorbing noise rather than real spatial variation.

### A note on detection-side SVCs

SVCs in INLAocc are restricted to the occupancy submodel. The detection
submodel uses constant coefficients across space. This is a deliberate design
choice, not a limitation. Detection probability is already estimated
per-site-visit from the replicate structure, so there is less need for
explicit spatial smoothing. Allowing both detection and
occupancy coefficients to vary spatially creates severe identifiability
problems: the model cannot distinguish a region where the species is absent
but detection is low from a region where the species is present but the
occupancy-covariate relationship is weak. Fixing detection coefficients to be
spatially constant anchors one submodel, which allows the SVC on the
occupancy side to be identified from the data.

If there is ecological reason to expect spatially varying detection (e.g.,
observer skill differs regionally, habitat density affects detectability),
model that variation through site-level detection covariates rather than a
spatial random effect on detection. For example, include vegetation density
or observer experience as detection covariates. This captures the same
heterogeneity through measured covariates rather than an unidentified
spatial process.

### SVC on non-occupancy binomial data

The `svc` argument operates through the occupancy linear predictor, so it is
specific to occupancy models. For non-occupancy binomial data (e.g., plant
presence in nested subplots where detection is perfect), the SVC machinery
still applies: set up the data as an occupancy model with a single visit per
site and intercept-only detection, and the SVC on the occupancy side
estimates spatially varying covariate effects on presence probability
directly. The detection intercept will absorb any constant offset and the
SVC captures the spatial heterogeneity. This is a common approach for
tree-species distribution data or presence-absence surveys where imperfect
detection is not a concern.

```{r fit-svc-binom}
# Simulate presence-absence data with perfect detection
sim_binom <- simulate_occu(N = 300, J = 1,
                           beta_occ = c(0.3, -0.6),
                           beta_det = c(5),  # high detection = effectively perfect
                           n_occ_covs = 1, n_det_covs = 0,
                           spatial_range = 0.3, seed = 55)

fit_binom_svc <- occu(~ occ_x1, ~ 1, data = sim_binom$data,
                      spatial = sim_binom$data$coords, svc = 2, verbose = 0)
summary(fit_binom_svc)
```


## Extracting SVC surfaces

`getSVCSamples()` returns the posterior summary of the site-level coefficient:
the total effect $\beta_1(\mathbf{s}_i) = \bar{\beta}_1 + \tilde{\beta}_1(\mathbf{s}_i)$
at each site.

```{r extract-svc}
svc_all <- getSVCSamples(fit_svc)
svc_samples <- svc_all[[1]]  # extract the data.frame for the first SVC covariate
head(svc_samples)
```

Each row gives the posterior mean, standard deviation, and 95% credible interval
for the total effect of `occ_x1` at that site. Sites where the interval
excludes zero have a "significant" local effect; sites where it spans zero may
have no local effect even if the global effect is non-zero.

### Visualising the SVC surface

A map of site-level coefficients reveals the spatial pattern:

```{r plot-svc-surface}
svc_df <- data.frame(
  x = sim$data$coords[, 1],
  y = sim$data$coords[, 2],
  beta1 = svc_samples$mean
)

# Base R visualisation
col_idx <- cut(svc_df$beta1, breaks = 100, labels = FALSE)
plot(svc_df$x, svc_df$y,
     col = hcl.colors(100, "Blue-Red 3")[col_idx],
     pch = 19, cex = 1.2,
     xlab = "Easting", ylab = "Northing",
     main = "Spatially varying effect of occ_x1")
```

When examining an SVC map, look for spatial structure that matches known
geography. A north-south gradient in the coefficient might indicate an
interaction with temperature or latitude. East-west variation near a coastline
might reflect maritime influence on the ecological relationship. Patchy
variation at scales smaller than the spacing between sites usually indicates
noise rather than signal: the SVC is interpolating between sparse data points
and the resulting pattern is not ecologically interpretable. If the map looks
like static, the SVC may be overfitting, and the constant-coefficient model is
likely a better choice.

For publication-quality maps, use `ggplot2`:

```{r plot-svc-ggplot}
library(ggplot2)

ggplot(svc_df, aes(x = x, y = y, colour = beta1)) +
  geom_point(size = 2.5) +
  scale_colour_gradient2(low = "#2166AC", mid = "grey90", high = "#B2182B",
                         midpoint = 0, name = expression(beta[1](s))) +
  coord_equal() +
  labs(title = "Spatially varying effect of occ_x1",
       x = "Easting", y = "Northing") +
  theme_minimal(base_size = 13)
```

### Mapping uncertainty

The posterior standard deviation reveals where the SVC is well-estimated versus
uncertain:

```{r plot-svc-uncertainty}
svc_df$sd <- svc_samples$sd

ggplot(svc_df, aes(x = x, y = y, colour = sd)) +
  geom_point(size = 2.5) +
  scale_colour_viridis_c(option = "magma", name = "Posterior SD") +
  coord_equal() +
  labs(title = "Uncertainty in spatially varying effect",
       x = "Easting", y = "Northing") +
  theme_minimal(base_size = 13)
```

Regions with few sites or low detection probability will have wider credible
intervals. If the uncertainty map shows uniformly high SD, the data may not
support an SVC. Consider dropping back to a constant coefficient.

The site-level credible interval tells you how well the local effect is estimated. Sites in the interior of the study area, surrounded by many neighbors that all inform the spatial process, tend to have tight intervals. The Gaussian process borrows strength from nearby sites, and in data-dense regions this borrowing is effective. Sites at the edge of the study area have fewer neighbors on one side, so the spatial process has less information to draw on and the intervals widen. Corner sites are the worst case: they sit at the boundary in two dimensions and their coefficient estimates carry the most uncertainty. If the credible interval at a particular site spans zero, the local effect is not distinguishable from no effect at that location, even if the global mean effect is clearly non-zero. This is a local statement: the covariate may matter on average, but the data cannot confirm that it matters at this specific site.

Always map both the posterior mean and the posterior standard deviation side by side. A region with a strong estimated effect (say, $\hat{\beta}_1 = -1.5$) but wide uncertainty ($\text{SD} = 0.9$, so the 95% interval spans roughly $-3.3$ to $0.3$) deserves less confidence than a region with a moderate effect ($\hat{\beta}_1 = -0.6$) and tight uncertainty ($\text{SD} = 0.15$). Reporting only the posterior mean creates a false impression of precision. The uncertainty map reveals which parts of the SVC surface are well-supported by data and which are extrapolations from the spatial process. Decisions based on the SVC (for example, identifying regions where a conservation intervention should target the covariate) should weight regions by their certainty, not just by the magnitude of the estimated effect.

### Significance map: where does the local effect exclude zero?

A useful summary is to classify each site by whether the 95% credible
interval for the local coefficient excludes zero. Sites where the interval
is entirely negative have a "significantly" negative local effect; sites
where it is entirely positive have a "significantly" positive local effect;
sites where the interval spans zero are ambiguous. Mapping these categories
reveals the spatial footprint of the SVC signal.

```{r plot-svc-significance}
svc_df$sig <- ifelse(svc_samples$q025 > 0, "positive",
              ifelse(svc_samples$q975 < 0, "negative", "non-significant"))
svc_df$sig <- factor(svc_df$sig, levels = c("negative", "non-significant", "positive"))

ggplot(svc_df, aes(x = x, y = y, colour = sig)) +
  geom_point(size = 2.5) +
  scale_colour_manual(
    values = c(negative = "#2166AC", `non-significant` = "grey70", positive = "#B2182B"),
    name = "Local effect"
  ) +
  coord_equal() +
  labs(title = "Sites with non-zero local effect of occ_x1",
       x = "Easting", y = "Northing") +
  theme_minimal(base_size = 13) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

Report the fraction of sites in each category. If fewer than 10% of sites
have intervals excluding zero, the SVC may not add interpretive value even
if WAIC favours it. The model fits better in aggregate, but the site-level
effects are too uncertain for local inference. If the fraction exceeds 50%,
the SVC surface carries real information for management decisions.

```{r svc-sig-fractions}
table(svc_df$sig) / nrow(svc_df)
```

### Scatter plot: covariate value vs. local coefficient

This visualization is one of the most informative diagnostics for SVCs. Plot
the value of the covariate at each site on the x-axis and the estimated
local coefficient on the y-axis. If the points show a trend (for example,
a quadratic relationship), this suggests the SVC is compensating for a
mis-specified functional form rather than capturing true spatial
non-stationarity. A flat scatter with no trend is consistent with genuine
spatial variation that is independent of the covariate's value.

```{r plot-svc-scatter}
svc_scatter <- data.frame(
  covariate = sim$data$occ.covs$occ_x1,
  local_coef = svc_samples$mean,
  sd = svc_samples$sd
)

ggplot(svc_scatter, aes(x = covariate, y = local_coef)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_point(aes(size = 1 / sd), alpha = 0.5, colour = "#4A90D9") +
  geom_smooth(method = "loess", se = TRUE, colour = "#B2182B",
              fill = "#B2182B", alpha = 0.15) +
  scale_size_continuous(range = c(0.5, 3), guide = "none") +
  labs(x = "Covariate value (occ_x1)",
       y = expression(hat(beta)[1](s)),
       title = "Local coefficient vs. covariate value") +
  theme_minimal(base_size = 13) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

In this plot, point size is inversely proportional to the posterior standard
deviation, so larger points represent more precisely estimated coefficients.
The loess curve (red) reveals any systematic relationship. Interpretation:

- **Flat loess line**: the SVC varies spatially but not as a function of the
  covariate itself. This is consistent with genuine spatial non-stationarity
  driven by an unmeasured gradient (e.g., climate, land-use history).
- **Linear trend**: the "varying coefficient" is really a quadratic effect
  of the covariate. Consider adding a quadratic term (`occ_x1 + I(occ_x1^2)`)
  to the fixed part of the formula and refitting without the SVC.
- **Threshold or step pattern**: the covariate's effect changes abruptly at
  some value. A stratum-based model (split the covariate into categories)
  may be a better fit.
- **No pattern but high scatter**: the data are noisy and the SVC is poorly
  estimated. Check sample size and detection rates.

When the scatter plot reveals a parametric pattern, it is worth fitting the
simpler parametric model (quadratic, interaction) and comparing to the SVC
via WAIC. If the parametric model fits as well as or better than the SVC,
prefer it: the parametric form is easier to interpret, communicate, and
extrapolate. The SVC is most valuable when the scatter plot shows spatial
variation that cannot be reduced to a simple parametric relationship.

### Posterior predictive check

Verify model adequacy after fitting the SVC:

```{r ppc-svc}
ppc <- ppcOccu(fit_svc, fit.stat = "freeman-tukey", group = 1)
ppc$bayesian.p
```

A Bayesian p-value near 0.5 confirms that the SVC model generates data consistent with observations.

### Parameter recovery

When working with simulated data, compare the estimated SVC surface against the known truth:

```{r recovery-svc}
# Correlation between true and estimated site-level coefficients
cor(svc_true, svc_samples$mean)
```

A high correlation between the true spatially varying coefficient and the estimated surface indicates good recovery. The correlation will not be 1.0 because: (a) estimation uncertainty at each site, (b) the Gaussian process smooths over sharp local variation, and (c) the SPDE mesh approximation introduces discretisation error. With more sites or higher detection probability, the recovery improves.


## Model comparison: SVC vs. constant coefficient

An SVC should only be retained if it meaningfully improves model fit. Always
compare against simpler alternatives:

```{r model-comparison}
# Baseline: no spatial effect
m_base <- occu(~ occ_x1, ~ det_x1, data = sim$data, verbose = 0)

# Spatial intercept only (no SVC)
m_spatial <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                  spatial = sim$data$coords, verbose = 0)

# SVC on intercept
m_svc_int <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                  spatial = sim$data$coords, svc = 1, verbose = 0)

# SVC on occ_x1 slope
m_svc_slope <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                    spatial = sim$data$coords, svc = 2, verbose = 0)

compare_models(base = m_base,
               spatial = m_spatial,
               svc_int = m_svc_int,
               svc_slope = m_svc_slope,
               criterion = "waic")
```

Interpretation:

- **base to spatial**: a WAIC improvement confirms spatial structure matters.

- **spatial to svc_int**: any improvement indicates spatial variation in
  baseline occupancy beyond what the standard spatial field captures.
- **spatial to svc_slope**: a WAIC improvement of 5+ indicates the effect of
  `occ_x1` genuinely varies across space.

If the SVC model does not improve WAIC over the spatial intercept model, the
constant coefficient is sufficient and the additional complexity is not
justified. A spatially varying intercept captures spatial variation in *baseline*
occupancy. An SVC on a slope captures spatial variation in *how the covariate
affects occupancy*. These are different ecological hypotheses; compare them
explicitly rather than assuming one subsumes the other.

Each step in this comparison tests a specific ecological hypothesis. Going from
base to spatial tests "is there unexplained spatial structure in occupancy?"
Going from spatial to SVC-intercept tests "does baseline occupancy vary beyond
what the covariates and a single spatial field explain?" Going from spatial to
SVC-slope tests "does the effect of the covariate change across space?" The
WAIC differences between these nested models answer each question
quantitatively. A large WAIC drop at one step but not another tells you which
kind of spatial complexity the data support.

The base-to-spatial step almost always shows improvement, because some degree of residual spatial autocorrelation is present in nearly every ecological dataset. The spatial-to-SVC-intercept step is rarely informative: the spatial intercept field and the SVC on the intercept do essentially the same thing, because the intercept is multiplied by 1 at every site. Both capture spatially structured variation in baseline occupancy, and switching between them mostly rearranges variance between model components without changing predictions. The spatial-to-SVC-slope step is the scientifically interesting comparison. If the SVC-slope model wins, the covariate-occupancy relationship is non-stationary, and the constant-coefficient model assigns one slope that misrepresents the ecology in at least some regions. If the SVC-slope model does not improve on the spatial-intercept model, the relationship is approximately constant across space, and the simpler parameterization is both more parsimonious and easier to communicate.

When reporting these comparisons, include the actual WAIC values and their differences, not just qualitative statements about which model "won." A WAIC difference of 2 between the spatial and SVC-slope models is negligible and does not justify the added complexity. A difference of 20 is meaningful. Between 5 and 15, the answer is ambiguous and should be reported as such, with both models presented as plausible representations of the data.


## Prediction at new locations

SVC predictions at unsampled sites incorporate the spatially varying coefficient. Pass new coordinates and covariates to `predict_spatial()`:

```{r predict-svc}
new_coords <- as.matrix(expand.grid(x = seq(0, 1, by = 0.05),
                                     y = seq(0, 1, by = 0.05)))
new_covs <- data.frame(occ_x1 = 0)

preds <- predict_spatial(fit_svc, newcoords = new_coords,
                         newocc.covs = new_covs)
head(preds)
```

The predictions reflect both the spatial intercept field and the SVC. Even with `occ_x1` held at 0, occupancy varies across space because of the intercept spatial field. To see the SVC's contribution, compare predictions at `occ_x1 = 0` vs `occ_x1 = 1` across the same grid. The difference will vary spatially, which is the SVC in action.

### Mapping predicted SVC surfaces at new locations

To produce a wall-to-wall map of the spatially varying coefficient at
unsampled locations, predict at two contrasting covariate values and take
the difference. The SVC at a prediction location is the local slope: the
change in log-odds of occupancy per unit increase in the covariate.

```{r predict-svc-surface}
# Predict occupancy with occ_x1 = 0 and occ_x1 = 1 across the grid
preds_x0 <- predict_spatial(fit_svc, newcoords = new_coords,
                            newocc.covs = data.frame(occ_x1 = 0))
preds_x1 <- predict_spatial(fit_svc, newcoords = new_coords,
                            newocc.covs = data.frame(occ_x1 = 1))

# The difference in log-odds gives the local coefficient
pred_logit_x0 <- qlogis(preds_x0$mean)
pred_logit_x1 <- qlogis(preds_x1$mean)
local_beta <- pred_logit_x1 - pred_logit_x0

pred_svc_df <- data.frame(
  x = new_coords[, 1],
  y = new_coords[, 2],
  local_coef = local_beta,
  psi_x0 = preds_x0$mean,
  psi_x1 = preds_x1$mean
)

# Map the predicted SVC surface
ggplot(pred_svc_df, aes(x = x, y = y, fill = local_coef)) +
  geom_tile() +
  scale_fill_gradient2(low = "#2166AC", mid = "grey90", high = "#B2182B",
                       midpoint = 0, name = expression(hat(beta)[1](s))) +
  coord_equal() +
  labs(title = "Predicted SVC surface at unsampled locations",
       x = "Easting", y = "Northing") +
  theme_minimal(base_size = 13) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

This map shows the predicted local coefficient across the entire study
region, interpolated by the SPDE mesh. The prediction at unsampled locations
uses the same kriging-like mechanism as the fitted values: the spatial process
at a new location is a weighted average of the process at nearby mesh nodes,
with weights determined by the estimated Matern covariance. Locations close
to many observed sites will have predictions close to the fitted values;
locations far from any data points will revert toward the prior mean (the
global coefficient $\bar{\beta}_1$).

### Comparing predicted occupancy under different covariate scenarios

The predicted SVC surface enables scenario comparisons that are impossible
with a constant-coefficient model. For example, what happens to occupancy
if the covariate increases by one standard deviation everywhere?

```{r predict-svc-scenarios}
# Side-by-side maps: occupancy at occ_x1 = 0 vs occ_x1 = 1
scenario_df <- data.frame(
  x = rep(new_coords[, 1], 2),
  y = rep(new_coords[, 2], 2),
  psi = c(preds_x0$mean, preds_x1$mean),
  scenario = rep(c("occ_x1 = 0 (baseline)", "occ_x1 = 1 (increased)"),
                 each = nrow(new_coords))
)

ggplot(scenario_df, aes(x = x, y = y, fill = psi)) +
  geom_tile() +
  scale_fill_viridis_c(option = "viridis", limits = c(0, 1),
                       name = expression(hat(psi))) +
  coord_equal() +
  facet_wrap(~ scenario) +
  labs(title = "Predicted occupancy under two covariate scenarios",
       x = "Easting", y = "Northing") +
  theme_minimal(base_size = 11) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

In a constant-coefficient model, the change in occupancy from baseline to
increased covariate would be uniform across space (the same logistic
transformation applied everywhere). With an SVC, the change varies:
regions where the local coefficient is strongly negative show large
occupancy decreases, while regions where the coefficient is near zero show
little change. This spatial heterogeneity in the response is the key
output of an SVC analysis and is directly relevant for spatial
conservation prioritisation.

### Prediction uncertainty at new locations

Prediction uncertainty at unsampled sites combines two sources: uncertainty
in the spatial field (the kriging variance) and uncertainty in the fixed
effects. Sites far from observed locations have higher kriging variance, so
their predicted SVC values carry wider credible intervals.

```{r predict-svc-uncertainty}
pred_unc_df <- data.frame(
  x = new_coords[, 1],
  y = new_coords[, 2],
  sd = preds_x0$sd
)

ggplot(pred_unc_df, aes(x = x, y = y, fill = sd)) +
  geom_tile() +
  scale_fill_viridis_c(option = "magma", name = "Posterior SD") +
  coord_equal() +
  labs(title = "Prediction uncertainty at unsampled locations",
       x = "Easting", y = "Northing") +
  theme_minimal(base_size = 13) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

Regions with low uncertainty are suitable for confident ecological
interpretation. Regions with high uncertainty should be interpreted
cautiously, because the predicted SVC value there is driven more by the prior
and the global mean than by local data. If the study goal involves
management decisions at specific unsampled locations, the uncertainty map
is as important as the prediction map.


## SVC + temporal models

For multi-season data where the covariate effect might vary in space but the
temporal dynamics also matter, combine SVC with temporal structure:

```{r svc-temporal-sim}
sim_t <- simTOcc(N = 100, J = 3, n_seasons = 5, seed = 300)
```

```{r svc-temporal-fit}
fit_svc_t <- occu(~ occ_x1, ~ det_x1, data = sim_t$data,
                  spatial = sim_t$data$coords,
                  temporal = "ar1", svc = 2, verbose = 0)
summary(fit_svc_t)
```

This is the most complex single-species model available in INLAocc. It
estimates how the effect of `occ_x1` varies both across space (via the SVC
field) and over time (via the AR(1) latent process). The SVC surface shows
where the covariate matters most (perhaps the effect is strong in upland
sites and negligible in lowland ones), while the AR(1) structure captures
year-to-year persistence in occupancy states, so that a site occupied last year
is more likely to be occupied this year, independent of its covariates.
Ecologically, this combination is natural for species whose habitat
relationships shift geographically (range expansion, local adaptation) while
their occupancy dynamics also exhibit temporal autocorrelation (colonisation
lags, extinction debt).

Interpreting the output requires separating the three variance components. The
fixed-effect estimate of `occ_x1` gives the spatially and temporally averaged
relationship. The SVC standard deviation tells you how much that relationship
varies across space: if it is small relative to the global effect, spatial
variation is negligible and a simpler model without the SVC suffices. The AR(1)
autocorrelation parameter $\rho$ tells you how strongly occupancy states
persist across seasons: values near 1 indicate slow dynamics where year-to-year
changes are small, values near 0 indicate rapid turnover. A sensible workflow
is to first establish whether the SVC is warranted using spatial-only models,
then add temporal structure to the preferred spatial specification. Do not add
both simultaneously on a first attempt, because it becomes difficult to diagnose
which component is absorbing which signal.

### Spatially varying temporal trends

The SVC + temporal combination lends itself to mapping
spatially varying trends over time. If the occupancy covariate is a temporal
index (season number, year), then the SVC on that covariate estimates how the
temporal trend differs across space. Some sites may show increasing occupancy
over time while others show decline. This is directly relevant for monitoring
programmes that need to identify regions of conservation concern.

```{r svc-trend-sim}
# Simulate 200 sites across 8 seasons with a temporal covariate
sim_trend <- simTOcc(N = 200, J = 3, n_seasons = 8,
                     beta_occ = c(0.3, -0.1),  # slight negative temporal trend
                     beta_det = c(0, -0.3),
                     ar1 = TRUE, rho = 0.6, seed = 42)
```

```{r svc-trend-fit}
# Fit SVC on the occupancy covariate (column 2 = occ_x1, the temporal covariate)
fit_trend <- occu(~ occ_x1, ~ det_x1, data = sim_trend$data,
                  spatial = sim_trend$data$coords,
                  temporal = "ar1", svc = 2, verbose = 0)
```

The site-level coefficients from `getSVCSamples()` now represent the local
temporal trend at each site: how much occupancy probability changes per unit
increase in the temporal covariate at that location. Sites with large negative
coefficients are declining fastest; sites with positive coefficients are
increasing or stable.

```{r svc-trend-map}
# Extract and map the local temporal trend
svc_trend <- getSVCSamples(fit_trend)
# For SVC temporal models, getSVCSamples operates on the per-period fits
# Check what is available:
if (length(svc_trend) > 0) {
  trend_df <- data.frame(
    x = sim_trend$data$coords[seq_len(nrow(svc_trend[[1]])), 1],
    y = sim_trend$data$coords[seq_len(nrow(svc_trend[[1]])), 2],
    trend = svc_trend[[1]]$mean,
    sd = svc_trend[[1]]$sd
  )

  ggplot(trend_df, aes(x = x, y = y, colour = trend)) +
    geom_point(size = 2.5) +
    scale_colour_gradient2(low = "#B2182B", mid = "grey90", high = "#2166AC",
                           midpoint = 0, name = "Local\ntrend") +
    coord_equal() +
    labs(title = "Spatially varying temporal trend",
         subtitle = "Red = declining, Blue = stable/increasing",
         x = "Easting", y = "Northing") +
    theme_minimal(base_size = 13) +
    theme(panel.background = element_rect(fill = "transparent", colour = NA),
          plot.background = element_rect(fill = "transparent", colour = NA))
}
```

When the SVC is applied to a temporal covariate, the resulting map is
directly interpretable as a "trend surface." Conservation managers can
overlay this on administrative or ecological regions to identify where
populations are declining and where they are stable. The interpretation is
more straightforward than for environmental SVCs because there is no
ambiguity about what a positive or negative trend means. A positive
coefficient means occupancy is increasing at that location; a negative
coefficient means it is declining. The uncertainty map (posterior SD)
indicates where these trends are well-estimated and where they are not.

The scatter plot diagnostic from the previous section applies here too: plot
the temporal covariate (year/season index) against the local trend
coefficient. A flat relationship is expected, because the temporal index
itself is not spatially correlated in the same way as an environmental
covariate. If a trend appears, it may indicate that the temporal signal is
confounded with a spatial gradient in the covariate.

### Model comparison for temporal SVCs

Always compare the temporal SVC model against a model with constant temporal
slopes and a spatial intercept:

```{r svc-trend-compare}
fit_notrend <- occu(~ occ_x1, ~ det_x1, data = sim_trend$data,
                    spatial = sim_trend$data$coords,
                    temporal = "ar1", verbose = 0)

compare_models(constant_trend = fit_notrend,
               spatial_trend = fit_trend,
               criterion = "waic")
```

If the SVC-trend model improves WAIC, the temporal dynamics genuinely differ
across the study area. If not, a single global trend is adequate.


### Comparing SVC variants

With multiple SVC specifications available, a systematic comparison helps
determine which form of spatial complexity the data actually support.

```{r svc-compare-all}
fit_base <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                 spatial = sim$data$coords, verbose = 0)
compare_models(
  base_spatial = fit_base,
  svc_intercept = fit_svc_int,
  svc_slope = fit_svc,
  criterion = "waic"
)
```

Each model in this table tests a different hypothesis about the data-generating
process. The base spatial model says "covariates have constant effects, and any
remaining spatial variation is captured by a single intercept field." The SVC on
the intercept says "baseline occupancy varies across space beyond what the
covariates and a single spatial field explain." In practice this is rarely
distinguishable from the base spatial model because the intercept SVC and the
spatial intercept field are functionally similar. The SVC on the slope says "the
covariate effect itself changes across space," which is a qualitatively
different claim: it implies that a single regression slope misrepresents the
ecology in at least some regions.

The WAIC ordering tells you which hypothesis the data support. If the
SVC-slope model wins by a meaningful margin (WAIC difference > 5--10), the
covariate-occupancy relationship is spatially non-stationary and a constant
slope is insufficient. If the base spatial model wins, or the differences are
within noise, the simpler specification is adequate and the SVC machinery adds
complexity without payoff. Report the actual WAIC values and differences, not
just which model "won." A difference of 2 is negligible, a difference of 20
is decisive, and anything between 5 and 15 should be presented as ambiguous.


## SVCs and site-level random effects: partitioning variance

A common question is whether to include both a site-level random effect and an
SVC in the same model. The answer depends on what kind of variation each
component captures and whether the data can support separating them.

The **intercept spatial field** $w(\mathbf{s}_i)$ captures spatially structured
variation in baseline occupancy that is not explained by covariates. It
accounts for missing covariates that vary smoothly across space, things the
analyst did not measure or does not know about.

A **site-level random effect** (iid, non-spatial) captures site-to-site
variation in baseline occupancy that has no spatial structure. This includes
genuinely random variation among sites (microhabitat differences, local
disturbance history) that cannot be predicted from geographic location.

The **SVC field** $\tilde{\beta}_1(\mathbf{s}_i)$ captures spatially structured
variation in the *effect of a covariate*, not in baseline occupancy. It
answers a different question: "does the covariate-occupancy relationship
change across space?"

All three can coexist in the same model, but each additional random component
must be identifiable from the data. With 300 sites and 4 visits, the model
can typically support a spatial intercept field plus one SVC. Adding a
site-level iid random effect on top of that is usually unnecessary because the
spatial field already captures most site-to-site variation, and the iid
component would absorb only the residual non-spatial noise. With fewer sites,
even a single SVC plus a spatial intercept can be difficult to separate.

```{r svc-plus-re}
# SVC with spatial intercept (the default when spatial + svc are both specified)
fit_svc_only <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                     spatial = sim$data$coords, svc = 2, verbose = 0)

# Compare to model with site-level RE but no SVC
fit_re_only <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                    spatial = sim$data$coords, verbose = 0)

compare_models(
  spatial_only = fit_re_only,
  spatial_svc  = fit_svc_only,
  criterion    = "waic"
)
```

If the SVC model improves WAIC over the spatial-only model, the
covariate effect varies across space in a way that the intercept field
cannot capture alone. If the models are equivalent, the spatial intercept
field is absorbing all the residual variation, and the SVC adds no
information.

### How variance flows between components

In an SVC model, the total spatial variance budget is shared among all spatial
fields. If the intercept field has high variance and the SVC has low variance,
most of the spatial pattern in occupancy is driven by unmeasured baseline
factors rather than by spatial variation in the covariate effect. If the SVC
has high variance relative to the intercept field, the covariate-occupancy
relationship varies substantially across space.

Monitor this partition by examining the estimated hyperparameters:

```{r svc-hyperparams}
# Extract spatial hyperparameters from the SVC model
hyper <- summary(fit_svc_only)
hyper
```

The spatial range and standard deviation for the intercept field versus the SVC
field reveal the relative importance and spatial scale of each component. A
short SVC range with high variance indicates fine-scale heterogeneity in the
covariate effect. A long SVC range with low variance indicates that the
coefficient is approximately constant, and the SVC is not contributing much.


## Case study: elevation effects on bird occupancy across a latitudinal gradient

To illustrate the full SVC workflow in a realistic ecological context, consider
a monitoring programme that surveys breeding birds at 300 sites spanning a
latitudinal range of 10 degrees. At each site, observers conduct 4 replicate
point counts and record environmental covariates including elevation
(standardised to mean 0, SD 1). The ecologist hypothesises that the effect of
elevation on occupancy varies with latitude: at lower latitudes, the
compressed bioclimatic gradient makes elevation a strong predictor, while at
higher latitudes the same elevation range spans fewer biomes and the effect
weakens.

This is a natural candidate for an SVC on elevation. The hypothesis is
specific (elevation effect varies with latitude), the mechanism is plausible
(bioclimatic compression), and the study design provides adequate spatial
coverage (300 sites across a latitudinal gradient).

```{r case-study-simulate}
# Simulate data mimicking the scenario
set.seed(2024)
N_sites <- 300
coords_case <- cbind(
  x = runif(N_sites, 0, 1),   # longitude proxy
  y = runif(N_sites, 0, 1)    # latitude proxy
)

# True SVC: elevation effect varies with latitude
# Strongest negative effect at low latitude (y=0), weakest at high (y=1)
elevation <- rnorm(N_sites)
svc_elevation_true <- -1.2 + 1.8 * coords_case[, 2]

# True occupancy on logit scale
logit_psi <- 0.3 + svc_elevation_true * elevation

# Simulate observations
J_case <- 4
p_det <- 0.6
z_true <- rbinom(N_sites, 1, plogis(logit_psi))
y_case <- matrix(rbinom(N_sites * J_case, 1, p_det * z_true),
                 nrow = N_sites, ncol = J_case)
```

```{r case-study-data}
# Package as INLAocc data
case_data <- list(
  y = y_case,
  N = N_sites,
  J = rep(J_case, N_sites),
  occ.covs = data.frame(elevation = elevation),
  det.covs = list(),
  coords = coords_case
)
```

The analyst proceeds through the recommended workflow:

**Step 1: Fit the constant-coefficient spatial model.**

```{r case-study-spatial}
m_case_spatial <- occu(~ elevation, ~ 1, data = case_data,
                       spatial = coords_case, verbose = 0)
```

**Step 2: Fit the SVC model.**

```{r case-study-svc}
m_case_svc <- occu(~ elevation, ~ 1, data = case_data,
                   spatial = coords_case, svc = 2, verbose = 0)
```

**Step 3: Compare via WAIC.**

```{r case-study-compare}
compare_models(
  constant_coef = m_case_spatial,
  svc_elevation = m_case_svc,
  criterion = "waic"
)
```

**Step 4: Examine the SVC surface.**

```{r case-study-surface}
svc_case <- getSVCSamples(m_case_svc)
case_svc_df <- data.frame(
  longitude = coords_case[, 1],
  latitude = coords_case[, 2],
  local_elev_effect = svc_case[[1]]$mean,
  true_effect = svc_elevation_true,
  sd = svc_case[[1]]$sd
)

# Map the estimated SVC
ggplot(case_svc_df, aes(x = longitude, y = latitude,
                        colour = local_elev_effect)) +
  geom_point(size = 2.5) +
  scale_colour_gradient2(low = "#2166AC", mid = "grey90", high = "#B2182B",
                         midpoint = 0,
                         name = "Local\nelevation\neffect") +
  coord_equal() +
  labs(title = "Estimated spatially varying elevation effect",
       x = "Longitude proxy", y = "Latitude proxy") +
  theme_minimal(base_size = 13) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

**Step 5: Validate against the known truth.**

```{r case-study-recovery}
# Correlation between true and estimated SVC
cat(sprintf("Recovery correlation: %.3f\n",
            cor(case_svc_df$true_effect, case_svc_df$local_elev_effect)))

# Scatter: true vs estimated
ggplot(case_svc_df, aes(x = true_effect, y = local_elev_effect)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", colour = "grey50") +
  geom_errorbar(aes(ymin = local_elev_effect - 1.96 * sd,
                    ymax = local_elev_effect + 1.96 * sd),
                alpha = 0.2, colour = "grey60") +
  geom_point(alpha = 0.6, colour = "#4A90D9") +
  labs(x = "True local elevation effect",
       y = "Estimated local elevation effect",
       title = "Parameter recovery: true vs. estimated SVC") +
  theme_minimal(base_size = 13) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

**Step 6: Post-hoc analysis of SVC drivers.**

The SVC surface shows a north-south gradient, which the ecologist interprets
as the hypothesised latitude effect. To test this formally, use `postHocLM()`
to regress the estimated site-level coefficients on latitude:

```{r case-study-posthoc}
posthoc_data <- data.frame(
  local_effect = svc_case[[1]]$mean,
  latitude = coords_case[, 2]
)
posthoc_fit <- postHocLM(local_effect ~ latitude, data = posthoc_data)
posthoc_fit$summary
```

If latitude explains a large fraction of the SVC variation (high R-squared),
the hypothesis is supported: the latitudinal gradient in bioclimatic
compression drives the spatial variation in the elevation effect. The
analyst could then fit a parametric interaction model (`elevation * latitude`)
and compare it to the SVC to see whether the interaction captures the same
pattern more parsimoniously.

```{r case-study-interaction}
# Fit interaction model for comparison
case_data$occ.covs$latitude <- coords_case[, 2]
m_case_interaction <- occu(~ elevation * latitude, ~ 1, data = case_data,
                           spatial = coords_case, verbose = 0)

compare_models(
  spatial_only = m_case_spatial,
  interaction  = m_case_interaction,
  svc_model    = m_case_svc,
  criterion    = "waic"
)
```

If the interaction model performs as well as the SVC, it should be preferred:
the parametric form is easier to interpret ("the elevation effect decreases by
$X$ per degree of latitude"), easier to communicate, and easier to extrapolate
to new latitudes. If the SVC outperforms the interaction, the spatial
variation in the elevation effect is more complex than a linear interaction
with latitude, and the non-parametric SVC is the better model.


## Multi-species SVC

When multiple species are surveyed at the same sites, each species can have its
own spatially varying coefficient:

```{r fit-ms-svc}
sim_ms <- simMsOcc(N = 200, J = 4, n_species = 5,
                   spatial_range = 0.3, seed = 123)

fit_ms_svc <- occu(~ occ_x1, ~ det_x1, data = sim_ms$data,
                   multispecies = TRUE,
                   spatial = sim_ms$data$coords, svc = 2, verbose = 0)
```

Each species gets its own SVC field: the effect of `occ_x1` varies spatially
*and* differs across species. A generalist might show a flat SVC (constant
effect everywhere), while a specialist might show strong spatial variation. The
species-level SVC surfaces from `getSVCSamples()` can be compared to understand
how species respond differently to the same environmental gradient across space.

This is computationally demanding: each species adds a separate spatial field.
With 5 species and `svc = 2`, the model estimates 5 SVC fields plus 5 intercept
spatial fields (10 spatial fields total). Use only with adequate sample sizes
(200+ sites, 3+ visits per site per species).

### Extracting and comparing species-level SVCs

Each species' SVC surface reveals where that species' response to the
covariate departs from the global community mean. Comparing these surfaces
across species generates hypotheses about niche differences and local
adaptation.

```{r ms-svc-extract}
# Extract SVC surfaces for each species
ms_svc <- lapply(sim_ms$data$species_names, function(sp) {
  sp_fit <- fit_ms_svc$species_fits[[sp]]
  if (!is.null(sp_fit) && inherits(sp_fit, "occu_inla_svc")) {
    svc <- getSVCSamples(sp_fit)
    if (length(svc) > 0) return(svc[[1]])
  }
  NULL
})
names(ms_svc) <- sim_ms$data$species_names

# Summarise: SVC standard deviation per species
svc_summary <- data.frame(
  species = sim_ms$data$species_names,
  svc_sd = vapply(ms_svc, function(x) {
    if (!is.null(x)) sd(x$mean) else NA_real_
  }, numeric(1))
)
svc_summary
```

Species with large SVC standard deviations have strongly spatially varying
responses to the covariate; species with small SVC standard deviations have
approximately constant responses. The ranking itself is informative: if the
most specialised species (e.g., narrow-range endemics) show the largest SVC
variation while generalists show the smallest, this supports the hypothesis
that niche boundaries create non-stationary species-environment
relationships.

### Mapping species richness with SVC

The `richness()` function aggregates estimated occupancy across species.
When species have SVCs, the richness surface reflects the spatial variation
in community composition: regions where many species have positive local
coefficients will show higher richness than regions where the covariate has
negative effects on most species.

```{r ms-svc-richness}
rich <- richness(fit_ms_svc)

rich_df <- data.frame(
  x = sim_ms$data$coords[, 1],
  y = sim_ms$data$coords[, 2],
  richness = rich$richness
)

ggplot(rich_df, aes(x = x, y = y, colour = richness)) +
  geom_point(size = 2.5) +
  scale_colour_viridis_c(option = "viridis", name = "Species\nrichness") +
  coord_equal() +
  labs(title = "Estimated species richness (SVC model)",
       x = "Easting", y = "Northing") +
  theme_minimal(base_size = 13) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

Compare this to the richness from a non-SVC multi-species model to see
whether allowing spatially varying coefficients changes the richness pattern.
If the SVC model predicts higher richness in some areas and lower richness in
others relative to the constant-coefficient model, the spatial variation in
species-environment relationships matters for biodiversity assessment.

```{r ms-comparison}
# Fit multi-species model without SVC for comparison
fit_ms_base <- occu(~ occ_x1, ~ det_x1, data = sim_ms$data,
                    multispecies = TRUE,
                    spatial = sim_ms$data$coords, verbose = 0)

rich_base <- richness(fit_ms_base)

# Richness difference
rich_diff <- rich$richness - rich_base$richness
cat(sprintf("Mean richness difference (SVC - base): %.2f\n", mean(rich_diff)))
cat(sprintf("Range: [%.2f, %.2f]\n", min(rich_diff), max(rich_diff)))
```

### Comparing SVC surfaces across species

When multiple species have SVCs on the same covariate, their spatial patterns
can be compared directly. Species that respond similarly to the covariate in
the same regions form functional groups: they share the same spatial
sensitivity to the environmental gradient. Species with opposite patterns
(one has a strong negative effect where another has a strong positive effect)
occupy complementary niches.

```{r ms-svc-compare-species}
# If we have at least two species with extracted SVCs, compare them
valid_svc <- Filter(Negate(is.null), ms_svc)
if (length(valid_svc) >= 2) {
  sp_names <- names(valid_svc)[1:min(3, length(valid_svc))]
  ms_compare_df <- do.call(rbind, lapply(sp_names, function(sp) {
    data.frame(
      x = sim_ms$data$coords[seq_len(nrow(valid_svc[[sp]])), 1],
      y = sim_ms$data$coords[seq_len(nrow(valid_svc[[sp]])), 2],
      local_coef = valid_svc[[sp]]$mean,
      species = sp
    )
  }))

  ggplot(ms_compare_df, aes(x = x, y = y, colour = local_coef)) +
    geom_point(size = 1.5) +
    scale_colour_gradient2(low = "#2166AC", mid = "grey90", high = "#B2182B",
                           midpoint = 0, name = "Local\neffect") +
    coord_equal() +
    facet_wrap(~ species) +
    labs(title = "SVC surfaces across species",
         x = "Easting", y = "Northing") +
    theme_minimal(base_size = 11) +
    theme(panel.background = element_rect(fill = "transparent", colour = NA),
          plot.background = element_rect(fill = "transparent", colour = NA))
}
```

If two species show the same spatial pattern in their SVCs (high positive
correlation), they respond to the covariate in the same spatially varying
way. This might reflect shared physiology or shared habitat requirements.
If two species show opposing patterns (strong negative correlation), the
covariate differentiates their niches spatially: where it helps one
species, it hinders the other.

Species with flat SVC surfaces (low spatial variation) are environmental
generalists with respect to that covariate: the relationship is stable
across the study area. Species with highly variable SVC surfaces are
specialists whose response to the covariate depends on local context.

### Post-hoc trait analysis for multi-species SVCs

A natural follow-up to multi-species SVC models is to ask: do species
traits predict which species have the most spatially variable responses?
For example, do habitat specialists show more SVC variation than
generalists? Do species with smaller range sizes have more variable
coefficients?

```{r ms-svc-traits}
# Use postHocLM to relate SVC variation to a trait
# Here we use the SVC standard deviation as a response variable
if (nrow(svc_summary) >= 3 && sum(!is.na(svc_summary$svc_sd)) >= 3) {
  # Simulate a trait for illustration
  svc_summary$body_mass <- rnorm(nrow(svc_summary), mean = 50, sd = 15)

  trait_fit <- postHocLM(svc_sd ~ body_mass, data = svc_summary)
  trait_fit$summary
}
```

A significant relationship between a trait and SVC variation generates
hypotheses about why species respond differently to the covariate in
different places. For instance, if large-bodied species have more
variable elevation effects than small-bodied ones, this might reflect
stronger dispersal limitation in large species, creating more pronounced
local adaptation.

### Multi-species SVC with latent factors

For larger species assemblages (10+ species), combine the SVC with latent
factor models to reduce dimensionality. The latent factors capture
correlated responses among species, while the SVC captures spatial variation
in those responses.

```{r ms-svc-lf, eval = FALSE}
# Factor model + SVC (computationally intensive)
fit_ms_lf_svc <- occu(~ occ_x1, ~ det_x1, data = sim_ms$data,
                      multispecies = TRUE, n.factors = 2,
                      spatial = sim_ms$data$coords, svc = 2, verbose = 0)
```

This is the most demanding model configuration in INLAocc. With $K$ species,
$F$ latent factors, and $P$ SVC covariates, the model estimates $K \times P$
SVC fields. The latent factor structure constrains the species-level fields
to load on a small number of shared spatial patterns, which improves
estimation when individual species have sparse detection histories. Use this
model only when: (a) you have 200+ sites, (b) 3+ visits per site, (c) a
clear ecological hypothesis for why covariate effects vary spatially, and
(d) sufficient species (5+) to estimate the factor structure.


## Practical guidelines for interpretation

### Don't assign SVCs arbitrarily

SVCs should reflect a specific ecological hypothesis: "the effect of $X$ varies
across space because of [mechanism]." Plausible mechanisms include:

- **Local adaptation**: species respond differently to a resource in different
  parts of their range.
- **Interaction with an unmeasured gradient**: temperature modulates the effect
  of precipitation, and temperature varies spatially.
- **Range boundary effects**: a habitat covariate matters more at the range edge
  where the species is at its physiological limit.

Without a mechanistic rationale, an SVC may be overfitting to noise or absorbing
the signal of an omitted covariate.

### Start simple, compare up

The recommended workflow for deciding whether an SVC is needed:

1. Fit the model without any spatial effect.

2. Add a spatial intercept (captures unmeasured spatial variation in baseline
   occupancy).
3. Add an SVC on the hypothesised covariate.

4. Compare all three via WAIC using `compare_models()`.

If the SVC does not improve on the spatial intercept model, the covariate effect
is likely constant across space, and the simpler model should be preferred.

### Visualise the SVC surface

Always map the site-level coefficient estimates. Diagnostic patterns to look
for:

- **Smooth gradients**: likely real ecological non-stationarity (e.g., a
  temperature effect that varies with latitude). This is the ideal case:
  the SVC is capturing a genuine spatial process.
- **Patchy or noisy pattern**: possibly overfitting, or confounding with another
  spatially structured process. Consider whether an omitted covariate could
  explain the pattern.
- **Sign changes**: the covariate has opposite effects in different regions.
  This can be ecologically meaningful (positive effect at range centre, negative
  at range edge) or a sign of confounding. Investigate carefully before
  reporting.
- **Uniform surface with high uncertainty**: the data do not support an SVC.
  The constant-coefficient model is sufficient.

### Consider the spatial scale

The SVC range parameter controls how smoothly the coefficient varies. Its
relationship to the study extent determines what the model can resolve:
\vspace{0.5em}

- **SVC range >> study extent**: the effect is nearly constant across the study
  area. The SVC is not needed; a fixed coefficient suffices.
- **SVC range << inter-site distance**: the spatial process varies faster than
  the sampling can resolve. Site-level estimates will be noisy and poorly
  constrained.
- **SVC range $\approx$ study extent / 5--10**: the ideal regime. The spatial
  variation is smooth enough to estimate with reasonable precision but fast
  enough to reveal meaningful heterogeneity.

If the posterior for SVC range piles up against the upper boundary of the prior,
it is signalling that the effect is effectively constant. If it piles up near
zero, the data may be too sparse to support an SVC.

### Prior sensitivity

SVC spatial variance and range priors matter more than in standard spatial
models, because the SVC field competes with the intercept spatial field for
signal. The default PC (penalised complexity) priors in INLAocc are calibrated
to shrink toward the constant-coefficient model (i.e., zero SVC variance). This
is conservative by design: the SVC must earn its complexity by improving the
marginal likelihood.

If results are sensitive to the choice of PC prior parameters, the data are
likely marginal for supporting an SVC. Report the sensitivity analysis alongside
the primary results.

In practice, this means fitting the SVC under at least two prior
specifications. For example, change the PC prior from $P(\sigma > 1) = 0.05$
to $P(\sigma > 2) = 0.1$ and refit. If the SVC surface changes substantially
in magnitude or spatial pattern, the data do not contain enough information to
pin down the SVC, and the posterior is being driven by the prior rather than
the likelihood. Report both results side by side. If the surface is stable
across reasonable prior choices, the SVC is well-identified and the conclusions
do not depend on subjective prior decisions.

When writing up an SVC analysis for publication, report the global effect first. This is the spatially averaged relationship between the covariate and occupancy, and it answers the same question as a standard non-SVC model. Present it as the headline result: "across the study area, a one-unit increase in forest cover is associated with a 0.4 increase in log-odds of occupancy." Then present the SVC surface as spatial variation around that mean. Quantify the fraction of the study area where the site-level 95% credible interval excludes zero, which is the area where you have direct statistical support for a non-zero local effect. If that fraction is less than 10%, the SVC may not be adding interpretive value despite a WAIC improvement: the model fits better in aggregate, but the local effects are too uncertain to make site-specific claims. If the fraction is above 50%, the spatial variation is well-supported and the SVC map carries real information for management or ecological interpretation.

Discuss prior sensitivity explicitly in the methods or supplementary material. State which PC prior was used (e.g., $P(\sigma > 1) = 0.05$ for the SVC marginal standard deviation), report whether changing to a more permissive prior (e.g., $P(\sigma > 2) = 0.1$) altered the SVC surface, and show both surfaces if they differ. Reviewers increasingly expect this for spatial models. If the surface is stable, a single sentence noting the robustness check is sufficient. If it changes, present both and discuss which aspects of the SVC are data-driven and which are prior-driven. This transparency strengthens the paper by showing the reader exactly where the evidence ends and the modelling assumptions begin.

### Identifiability concerns

With both an intercept spatial field $w(\mathbf{s}_i)$ and an SVC field
$\tilde{\beta}_1(\mathbf{s}_i)$, the model has two spatial processes that can
partially trade off against each other. This trade-off becomes problematic when:

- The SVC covariate $x_1$ is itself spatially structured (e.g., elevation,
  which varies smoothly across space). The intercept field and the SVC field
  may both try to capture the same spatial signal, leading to poorly identified
  variance components.
- The data are sparse relative to the spatial scale of variation. Few sites
  means few data points to separate the two spatial processes.

Use `checkIdentifiability()` to flag potential issues:

```{r check-identifiability}
checkIdentifiability(fit_svc)
```

If identifiability warnings appear, consider whether the SVC covariate is too
collinear with the spatial structure of the data. Options include:
residualising the covariate against a spatial smooth before fitting, using
stronger PC priors to regularise the SVC, or accepting the spatial intercept
model and reporting the global coefficient only.


## When SVCs are (and aren't) warranted

SVCs are a hypothesis-testing tool, not a default model specification. The
decision to allow a covariate effect to vary spatially should be grounded in
ecological reasoning, not statistical convenience. Assigning SVCs to every
covariate in an exploratory analysis is analogous to including every possible
interaction term in a regression: it increases flexibility but also increases
the risk of overfitting and the difficulty of interpretation.

### When to use an SVC

An SVC is appropriate when:

1. **You hypothesise spatial non-stationarity in a specific relationship.**
   For example, you expect that the effect of forest cover on bird occupancy
   changes between fragmented and continuous landscapes, or that the effect
   of temperature on plant occurrence varies between maritime and continental
   climates. The hypothesis does not need to specify the exact form of the
   variation (that is what the SVC estimates), but it should specify *which*
   covariate and *why* it might vary.

2. **The constant-coefficient model fails residual diagnostics.** If the
   residuals from a standard spatial occupancy model show spatial structure
   that correlates with a covariate, the constant slope is inadequate.
   An SVC on that covariate is a natural next step.

3. **You want to generate hypotheses about drivers of non-stationarity.**
   Even without a pre-specified mechanism, an SVC can reveal spatial patterns
   that suggest interactions with unmeasured variables. But in this
   exploratory mode, treat the SVC surface as hypothesis-generating, not
   hypothesis-confirming. Any pattern identified should be tested with
   independent data or a follow-up analysis.

4. **You are mapping spatially varying trends.** When the "covariate" is a
   temporal index (year, season), the SVC directly estimates where
   populations are increasing or declining. This is one of the most
   straightforward and defensible uses of SVCs.

### When not to use an SVC

An SVC is inappropriate when:

1. **You have no ecological reason to expect spatial variation in the
   effect.** If a covariate's effect is expected to be constant (e.g.,
   body size as a predictor of detection probability), adding an SVC is
   overfitting.

2. **The covariate itself is spatially unstructured.** If the covariate
   varies randomly across space (e.g., observer identity, day of survey),
   an SVC cannot estimate a meaningful spatial surface and will produce
   noise.

3. **You have fewer than ~150 sites.** With sparse data, the two spatial
   processes (intercept field and SVC field) cannot be separated reliably.
   The SVC will absorb variance from the intercept field or vice versa,
   and the resulting surface will be driven by the prior rather than the
   data.

4. **The simpler parametric alternative fits equally well.** If a quadratic
   term or a covariate interaction explains the apparent non-stationarity
   as well as the SVC, prefer the parametric model. It is easier to
   interpret, extrapolate, and communicate.

### The SVC is not a substitute for good covariate selection

A common pitfall is to use an SVC to compensate for a poorly specified model.
If the true relationship between the covariate and occupancy is quadratic but
the model specifies a linear term, the SVC will capture the quadratic
curvature as spatial variation in the slope, because sites with high
covariate values (where the quadratic curve bends) tend to cluster
geographically. The SVC "works" in the sense that it improves model fit, but
it misattributes a functional-form problem to a spatial process. The scatter
plot diagnostic (covariate value vs. local coefficient) detects this: a
systematic trend in the scatter indicates functional-form mis-specification,
not true spatial non-stationarity.

Before fitting an SVC, check the functional form of the covariate-occupancy
relationship. Fit models with quadratic terms, transformations, or
breakpoints. Only if these fail to capture the residual spatial structure
should the SVC be the next step.


## Comparing SVC to simpler alternatives: a model ladder

A systematic model comparison clarifies what form of spatial complexity the
data support. Build a "ladder" from the simplest to the most complex model,
where each step tests a specific ecological hypothesis:

```{r model-ladder}
# Step 1: No spatial effect, linear covariate
m1 <- occu(~ occ_x1, ~ det_x1, data = sim$data, verbose = 0)

# Step 2: Quadratic covariate effect (no spatial)
sim$data$occ.covs$occ_x1_sq <- sim$data$occ.covs$occ_x1^2
m2 <- occu(~ occ_x1 + occ_x1_sq, ~ det_x1, data = sim$data, verbose = 0)

# Step 3: Spatial intercept only, linear covariate
m3 <- occu(~ occ_x1, ~ det_x1, data = sim$data,
           spatial = sim$data$coords, verbose = 0)

# Step 4: Spatial intercept + quadratic covariate
m4 <- occu(~ occ_x1 + occ_x1_sq, ~ det_x1, data = sim$data,
           spatial = sim$data$coords, verbose = 0)

# Step 5: SVC on the covariate slope
m5 <- occu(~ occ_x1, ~ det_x1, data = sim$data,
           spatial = sim$data$coords, svc = 2, verbose = 0)

compare_models(
  linear       = m1,
  quadratic    = m2,
  spatial      = m3,
  spatial_quad = m4,
  svc_slope    = m5,
  criterion    = "waic"
)
```

Reading this ladder:

- **linear to quadratic**: does the covariate have a non-linear effect?
- **linear to spatial**: is there spatial structure beyond the covariate?
- **spatial to spatial_quad**: does a quadratic term explain the residual
  spatial pattern?
- **spatial to svc_slope**: does the covariate effect vary spatially in a way
  that a quadratic cannot capture?
- **spatial_quad to svc_slope**: is the SVC capturing something beyond
  functional-form curvature?

The SVC should only be preferred if it beats the spatial + quadratic model.
If the quadratic model fits as well as the SVC, the spatial "variation" in
the coefficient was really just curvature in the covariate-occupancy
relationship, not true non-stationarity.

This ladder also provides the comparisons needed for a manuscript. Present
the full table of WAIC values (not just the "winning" model) so reviewers
can see the evidence for each level of complexity.


## Prior sensitivity for SVCs: worked example

The SPDE priors on the SVC field control how much spatial variation is
permitted and at what scale. Because the SVC field competes with the
intercept spatial field for the same signal, prior choices have real
consequences for inference.

INLAocc's default PC priors set $P(\sigma > 1) = 0.05$ for the marginal
standard deviation and $P(\text{range} < 0.3) = 0.5$ for the spatial range.
These defaults are conservative: they shrink the SVC toward zero (a constant
coefficient) unless the data provide strong evidence for spatial variation.
But "conservative" is not always appropriate. If the study area is large and
the ecological process varies at broad scales, the default range prior may
be too restrictive.

### Fitting with alternative priors

Pass custom priors through the `spde.args` argument of `occu()`. This
controls the SPDE mesh construction and the associated PC priors:

```{r prior-sensitivity}
# Default priors (for reference)
fit_default <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                    spatial = sim$data$coords, svc = 2, verbose = 0)

# More permissive: allow larger SVC variance
fit_permissive <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                       spatial = occu_spatial(sim$data$coords,
                                             prior.sigma = c(2, 0.1),
                                             prior.range = c(0.2, 0.5)),
                       svc = 2, verbose = 0)

# More restrictive: shrink SVC variance harder
fit_restrictive <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                        spatial = occu_spatial(sim$data$coords,
                                              prior.sigma = c(0.5, 0.01),
                                              prior.range = c(0.5, 0.5)),
                        svc = 2, verbose = 0)
```

### Comparing SVC surfaces across priors

Extract the SVC surfaces and measure how much they change:

```{r prior-sensitivity-compare}
svc_default <- getSVCSamples(fit_default)[[1]]
svc_permissive <- getSVCSamples(fit_permissive)[[1]]
svc_restrictive <- getSVCSamples(fit_restrictive)[[1]]

# Correlation between SVC surfaces under different priors
cor_dp <- cor(svc_default$mean, svc_permissive$mean)
cor_dr <- cor(svc_default$mean, svc_restrictive$mean)
cat(sprintf("Correlation (default vs. permissive):  %.3f\n", cor_dp))
cat(sprintf("Correlation (default vs. restrictive): %.3f\n", cor_dr))

# Range of SVC values under each prior
cat(sprintf("SVC range (default):     [%.2f, %.2f]\n",
            min(svc_default$mean), max(svc_default$mean)))
cat(sprintf("SVC range (permissive):  [%.2f, %.2f]\n",
            min(svc_permissive$mean), max(svc_permissive$mean)))
cat(sprintf("SVC range (restrictive): [%.2f, %.2f]\n",
            min(svc_restrictive$mean), max(svc_restrictive$mean)))
```

If the correlations between surfaces are above 0.9, the spatial pattern is
robust to prior choice. If the correlations drop below 0.7, or if the
magnitude of the SVC changes substantially (e.g., the range of coefficient
values doubles under permissive priors), the data are marginal for
supporting an SVC. In that case, report both surfaces and note the
sensitivity.

```{r prior-sensitivity-plot}
prior_df <- data.frame(
  x = rep(sim$data$coords[, 1], 3),
  y = rep(sim$data$coords[, 2], 3),
  beta1 = c(svc_default$mean, svc_permissive$mean, svc_restrictive$mean),
  prior = rep(c("Default", "Permissive", "Restrictive"),
              each = nrow(svc_default))
)

ggplot(prior_df, aes(x = x, y = y, colour = beta1)) +
  geom_point(size = 1.5) +
  scale_colour_gradient2(low = "#2166AC", mid = "grey90", high = "#B2182B",
                         midpoint = 0, name = expression(beta[1](s))) +
  coord_equal() +
  facet_wrap(~ prior) +
  labs(title = "SVC surface under different PC priors",
       x = "Easting", y = "Northing") +
  theme_minimal(base_size = 11) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

### Asymmetric priors: tighter SVC, looser intercept

A useful strategy from the SVC literature is to assign tighter priors on the
SVC variance than on the intercept spatial field variance. This reflects the
prior belief that most residual spatial variation should be absorbed by the
intercept field (which captures unmeasured confounders), and only strong
signal should be attributed to the SVC. In INLAocc, this is achieved by
passing a pre-built `occu_spatial()` object with restrictive SVC priors and
using the default (less restrictive) priors for the intercept spatial field.

The default priors already implement this asymmetry to some degree, because
the SVC field and the intercept field share the same SPDE mesh but the SVC
field has an additional constraint: it is multiplied by the covariate value,
which means its effective contribution depends on the covariate's magnitude
and spatial structure. Tightening the SVC prior further strengthens this
asymmetry and reduces the risk of false positive spatial variation in the
coefficient.

When the SVC and the intercept field have similar priors and the data are
noisy, the two spatial processes can "trade" variance: the SVC field absorbs
some of the spatial variation that should go to the intercept, inflating the
apparent spatial variation in the coefficient. Asymmetric priors prevent this
by giving the intercept field first claim on the spatial signal.


## Diagnostic workflow for SVCs

After fitting an SVC model, a systematic diagnostic sequence confirms (or
undermines) confidence in the results. This section provides a step-by-step
workflow that covers residual checks, identifiability diagnostics, and
posterior predictive checks, applied specifically to SVC models.

### Step 1: Posterior predictive check

Before examining the SVC surface, verify that the overall model generates
data consistent with observations. A failing PPC indicates fundamental
model mis-specification that no SVC interpretation can rescue.

```{r diagnostic-ppc}
ppc <- ppcOccu(fit_svc, fit.stat = "freeman-tukey", group = 1)
cat(sprintf("Bayesian p-value: %.3f\n", ppc$bayesian.p))
```

A Bayesian p-value between 0.2 and 0.8 is acceptable. Values below 0.05 or
above 0.95 indicate systematic lack of fit. For SVC models, a poor PPC may
indicate that the SVC is compensating for a mis-specified mean function
(e.g., missing quadratic term) rather than capturing genuine spatial
variation.

### Step 2: Identifiability check

Verify that the model's parameters are identifiable from the data. SVC
models are particularly vulnerable because two spatial processes (intercept
field and SVC field) compete for the same spatial signal.

```{r diagnostic-ident}
id_check <- checkIdentifiability(fit_svc)
id_check
```

Pay particular attention to warnings about:

- **Covariate-spatial confounding**: when the SVC covariate is itself
  spatially structured. If `occ_x1` correlates strongly with the spatial
  coordinates (Moran's I > 0.5), the intercept field and SVC field will
  struggle to separate.
- **Low effective sample size**: too few sites relative to the number of
  spatial processes.
- **High posterior correlation between hyperparameters**: the SVC range and
  the intercept range are negatively correlated, suggesting the model
  cannot determine which process has longer range.

### Step 3: Residual maps

Extract and map residuals to check for remaining spatial structure. If the
SVC model is correctly specified, residuals should show no spatial pattern.

```{r diagnostic-residuals}
resid <- residuals(fit_svc, type = "deviance")

resid_df <- data.frame(
  x = sim$data$coords[, 1],
  y = sim$data$coords[, 2],
  residual = resid$occ
)

ggplot(resid_df, aes(x = x, y = y, colour = residual)) +
  geom_point(size = 2) +
  scale_colour_gradient2(low = "#2166AC", mid = "grey90", high = "#B2182B",
                         midpoint = 0, name = "Deviance\nresidual") +
  coord_equal() +
  labs(title = "Occupancy residuals (SVC model)",
       x = "Easting", y = "Northing") +
  theme_minimal(base_size = 13) +
  theme(panel.background = element_rect(fill = "transparent", colour = NA),
        plot.background = element_rect(fill = "transparent", colour = NA))
```

If the residual map shows spatial structure (clusters of positive or negative
residuals), the SVC has not fully captured the spatial heterogeneity. This
can happen when: (a) the SVC range is too long, so it smooths over
fine-grained variation; (b) a second covariate also has a spatially varying
effect but was not given an SVC; or (c) the model is missing a covariate
entirely.

### Step 4: Cross-validation

As a final check, compare the SVC model's out-of-sample predictive
performance against simpler alternatives using k-fold cross-validation:

```{r diagnostic-cv}
# Fit with 5-fold CV (computationally intensive)
fit_svc_cv <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                   spatial = sim$data$coords, svc = 2,
                   k.fold = 5, verbose = 0)

fit_spatial_cv <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                       spatial = sim$data$coords,
                       k.fold = 5, verbose = 0)

cat(sprintf("SVC k-fold deviance:     %.2f\n", fit_svc_cv$k.fold$k.fold.deviance))
cat(sprintf("Spatial k-fold deviance: %.2f\n", fit_spatial_cv$k.fold$k.fold.deviance))
```

If the SVC model has lower out-of-sample deviance, it generalises better to
held-out data. If the deviances are similar, the SVC's additional complexity
is not paying off in prediction and the simpler model should be preferred.

### Summary of the diagnostic workflow

The complete diagnostic sequence for an SVC analysis:

1. **PPC**: does the model fit the data at all?
2. **Identifiability**: can the model's parameters be estimated from these
   data?
3. **Residual maps**: is there remaining spatial structure?
4. **Scatter plot**: is the SVC compensating for functional-form
   mis-specification?
5. **Prior sensitivity**: do the results change under different priors?
6. **Cross-validation**: does the SVC improve out-of-sample prediction?

If any step fails, do not proceed to ecological interpretation of the SVC
surface. Diagnose and fix the problem first. If multiple steps fail, the SVC
is not supported by the data, and the constant-coefficient model should be
preferred.


## Reporting SVCs in manuscripts

SVC models produce rich output that can overwhelm readers if presented without
structure. Below are recommendations for how to present SVC results in a
research paper, drawing on the practices used in recent ecological applications
of SVCs.

### Essential elements to report

1. **The global (mean) effect and its credible interval.** This is the headline
   result. It answers the same question as a standard occupancy model and allows
   direct comparison with prior studies that used constant-coefficient models.
   Present it first, before discussing spatial variation.

2. **The WAIC comparison table.** Show all candidate models (at minimum: non-spatial,
   spatial intercept only, SVC) with their WAIC values and differences. This tells
   the reader how much evidence supports the SVC over simpler alternatives.

3. **A map of the SVC surface.** Use a diverging colour scale centred at zero.
   Include coordinate axes or an inset map showing the study region. If the study
   area has recognisable geography (coastlines, mountain ranges, administrative
   boundaries), overlay them to help the reader interpret the spatial pattern.

4. **An uncertainty map.** Either the posterior standard deviation or the
   significance classification (positive / non-significant / negative based on
   95% credible intervals). This tells the reader where the SVC is well-estimated
   and where it is speculative.

5. **The fraction of sites with "significant" local effects.** Report the
   percentage of sites where the 95% credible interval excludes zero. This single
   number summarises how much of the study area has strong statistical support
   for a non-zero local effect.

6. **The scatter plot diagnostic.** Covariate value vs. local coefficient. This
   shows whether the SVC is capturing genuine spatial non-stationarity or
   compensating for a mis-specified functional form. Include this in the main
   text or supplementary material.

7. **Prior sensitivity results.** At minimum, note which PC priors were used and
   whether the SVC surface was stable under alternative specifications. If it
   was not stable, present both surfaces.

### Optional but valuable elements

- **Correlation between SVC surfaces** if multiple SVCs were fitted.
- **Post-hoc regression** relating the SVC surface to hypothesised drivers
  (e.g., latitude, land-use history), using `postHocLM()`.
- **Cross-validation** comparing out-of-sample predictive performance of the
  SVC model versus alternatives.
- **Predicted SVC surface at new locations** if the study involves spatial
  extrapolation.

### What not to do

- Do not present only the SVC map without the global effect. Readers need the
  average relationship for context.
- Do not over-interpret fine-scale features of the SVC surface, especially near
  the edges of the study area where uncertainty is highest.
- Do not present the SVC as causal evidence that a covariate's effect truly
  varies. The SVC is a correlative finding; it shows *where* the relationship
  is non-stationary, not *why*. The "why" requires ecological reasoning and
  ideally independent validation.
- Do not claim that the SVC "proves" the absence of a covariate effect in
  regions where the local coefficient is near zero. A near-zero local effect
  with wide uncertainty is different from a confidently estimated zero effect.
  Check the credible intervals.

### Example methods paragraph

> We fitted a spatially varying coefficient occupancy model using INLAocc
> (version X.X). The effect of elevation on occupancy was allowed to vary
> spatially via a Matern Gaussian process implemented through the SPDE
> approach. We used penalised complexity (PC) priors with
> $P(\sigma > 1) = 0.05$ for the marginal standard deviation and
> $P(\rho < 0.3) = 0.5$ for the practical range of the SVC field. We
> compared the SVC model to a spatial intercept-only model and a
> non-spatial baseline using WAIC. We extracted site-level coefficient
> estimates using `getSVCSamples()` and classified sites as having
> positive, negative, or non-significant local effects based on 95%
> credible intervals. We assessed prior sensitivity by refitting under
> a more permissive prior ($P(\sigma > 2) = 0.1$) and confirmed that
> the spatial pattern was robust (correlation $r > 0.95$ between
> surfaces).


## Summary

- SVCs decompose a covariate effect into a global mean plus a spatial surface of
  deviations. They answer "where does this relationship hold, and where does it
  break down?"
- Fit with `occu(..., spatial = coords, svc = k)` where `k` indexes the
  occupancy covariate. Extract surfaces with `getSVCSamples()`.
- SVCs are the most complex single-species model in INLAocc. Use only with
  ecological justification and adequate sample size (200+ sites).
- Always compare to simpler alternatives: no spatial effect, spatial intercept
  only, constant coefficient with spatial intercept.
- Visualise the SVC surface. If it looks like noise, it probably is.
- The scatter plot (covariate value vs. local coefficient) is the single most
  informative diagnostic for distinguishing genuine spatial non-stationarity
  from functional-form mis-specification.
- Multiple SVCs can coexist but must be examined for inter-correlation. If two
  SVC surfaces are highly correlated, they may be responding to the same latent
  gradient.
- Prediction at new locations uses the same kriging mechanism as fitted values.
  Map both the predicted surface and its uncertainty.
- Multi-species SVCs reveal which species have spatially varying responses and
  which have constant ones. Relate this variation to species traits for
  ecological insight.
- SVC + temporal models are the most complex configuration. Establish that the
  SVC is warranted in a spatial-only model before adding temporal structure.
- Prior sensitivity and identifiability checks are essential, not optional.
- When reporting results, present the global effect first, the WAIC comparison
  table, the SVC map, the uncertainty map, the fraction of significant sites,
  and the prior sensitivity check.

The full decision tree for SVC analysis: (1) fit the base model without spatial
effects, (2) check Moran's I on residuals for spatial autocorrelation, (3) add
a spatial intercept field, (4) hypothesize which covariate might have a
spatially varying effect based on ecological reasoning, (5) fit the SVC model,
(6) compare all models via WAIC, (7) visualize the SVC surface and check
whether the spatial pattern is interpretable, (8) run the scatter plot
diagnostic to rule out functional-form mis-specification, (9) run a prior
sensitivity analysis to confirm the results are stable. If the SVC does not
improve WAIC, or the surface looks like noise, or the results shift under
different priors, fall back to the spatial intercept model and report the
global coefficient.
