---
title: "Multi-Season Occupancy Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Season Occupancy Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
```

## Introduction

Many ecological studies span multiple years or seasons. A species that occupies
a site in one year may go locally extinct the next, and previously unoccupied
sites may be colonized. Single-season occupancy models treat each site as a
snapshot, discarding the temporal information embedded in repeated surveys across
years. When the data contain temporal structure, ignoring it wastes information
at best and produces biased estimates at worst --- occupancy trends are
confounded with detection variation, and standard errors are too narrow because
the model treats non-independent observations as independent.

Multi-season (dynamic) occupancy models address this by decomposing temporal
variation into a structured process. In the classical formulation (MacKenzie
et al. 2003), colonization and extinction rates are modelled explicitly. INLAocc
takes a different approach: it models the temporal dynamics as an AR(1) random
effect on occupancy, which captures the same persistence structure (high
occupancy tends to follow high occupancy) without requiring separate
colonization/extinction parameters. This is more parsimonious when the goal is
inference on occupancy probability and its drivers rather than the transition
rates themselves.

INLAocc fits multi-season models by stacking all site-period combinations into a
joint occupancy model with an AR(1) temporal random effect. Detection is fit
per-period, so each season gets its own detection parameters --- survey
conditions in July 2019 need not resemble those in July 2023. The temporal
random effect $\eta_t$ captures shared year-to-year fluctuation: good years
(favourable weather, high resource availability) raise occupancy across all
sites, and bad years lower it. Fixed effects on occupancy and detection are
estimated simultaneously, with the AR(1) structure absorbing residual temporal
autocorrelation that would otherwise inflate standard errors.


## The multi-season occupancy model

For site $i$ at time period $t$, the occupancy process is:

$$z_{it} \sim \text{Bernoulli}(\psi_{it}), \quad \text{logit}(\psi_{it}) = \mathbf{x}_i^\top \boldsymbol{\beta} + \eta_t$$

where $z_{it}$ is the true (latent) occupancy state and $\psi_{it}$ is the
occupancy probability. The temporal random effect follows an AR(1) process:

$$\eta_t = \rho \cdot \eta_{t-1} + \epsilon_t, \quad \epsilon_t \sim \mathcal{N}(0, \sigma^2_\eta)$$

where $\rho \in (-1, 1)$ is the temporal autocorrelation parameter and
$\sigma^2_\eta$ is the innovation variance.

The observation process (within each period) is:

$$y_{ijt} \mid z_{it} \sim \text{Bernoulli}(z_{it} \cdot p_{ijt}), \quad \text{logit}(p_{ijt}) = \mathbf{w}_{ijt}^\top \boldsymbol{\alpha}$$

where $y_{ijt}$ is the detection/non-detection outcome at site $i$, visit $j$,
period $t$, and $p_{ijt}$ is the detection probability.

Three features of this formulation are worth highlighting:

- **$\eta_t$ captures shared temporal trends.** All sites experience the same
  temporal random effect in a given period. This represents "good years vs. bad
  years" for the species --- shared environmental conditions that raise or lower
  occupancy across the landscape.

- **$\rho$ controls persistence.** A high $\rho$ (e.g., 0.8) means occupancy
  patterns persist across periods: if this year is good, next year is likely good
  too. A low $\rho$ means each year is nearly independent of the last.

- **Detection varies across periods.** The detection covariates
  $\mathbf{w}_{ijt}$ can include period-specific variables (e.g., survey date,
  observer effort, weather during surveys). Each period's detection model is
  estimated from that period's data, so changing survey protocols across years
  are handled naturally.

What does $\rho = 0.7$ mean in ecological terms? It means that if occupancy is
above average this year, about 70% of that excess carries over to the next year.
A site network where most occupied sites remain occupied and most empty sites
remain empty produces high $\rho$. This is typical for species with strong site
fidelity or slow population dynamics --- think territorial birds, long-lived
amphibians, or perennial plants. A low $\rho$ (say 0.2) indicates that knowing
this year's occupancy tells you little about next year: the species may be
nomadic, or local populations may be driven by stochastic events like flooding or
fire.

The innovation standard deviation $\sigma_\eta$ captures the magnitude of
year-to-year environmental stochasticity after accounting for temporal
persistence. A large $\sigma_\eta$ with high $\rho$ means the system has strong
shocks that then persist (a bad year stays bad for several years). A large
$\sigma_\eta$ with low $\rho$ means large but transient fluctuations. The
classical colonization-extinction parameterization (MacKenzie et al. 2003) models
the same persistence through explicit transition probabilities: colonization
$\gamma$ and extinction $\epsilon$. Both formulations capture temporal
dependence, but the AR(1) is more parsimonious when colonization and extinction
rates are not themselves the inferential target. If you need to estimate how
quickly empty sites are recolonized, the dynamic parameterization is more
natural. If you need to estimate occupancy trends and their drivers, the AR(1) is
typically sufficient and avoids identifiability issues that arise when both
$\gamma$ and $\epsilon$ interact with covariates.


## Data structure

Multi-season data in INLAocc uses a 3D array with dimensions
sites $\times$ periods $\times$ visits:

```
y[site, period, visit]
dim: N × T × J
```

Each entry is 1 (detected), 0 (not detected), or `NA` (visit not conducted).
Here is an example for a single site surveyed over 5 years with 3 visits per
year:

```{r data-example, eval = FALSE}
# 100 sites, surveyed over 5 years, 3 visits per year
dim(y)
#> [1] 100   5   3

y[1, , ]  # site 1, all years:
#>      [,1] [,2] [,3]
#> [1,]    1    0    1    # year 1: detected twice
#> [2,]    0    0    0    # year 2: never detected
#> [3,]    1    1    0    # year 3: detected twice
#> [4,]    0    0   NA    # year 4: detected zero, one visit missing
#> [5,]    1    1    1    # year 5: detected all visits
```

Covariates follow the same dimensional logic:

- **`occ.covs`**: Site-level covariates can be either a length-$N$ vector
  (static across periods, e.g., elevation) or an $N \times T$ matrix
  (time-varying, e.g., yearly temperature anomaly). Static covariates are
  recycled across all periods internally.

- **`det.covs`**: Visit-level covariates are $N \times T \times J$ arrays.
  For example, Julian survey date varies by site, year, and visit.

- **`coords`**: An $N \times 2$ matrix of site coordinates, constant across
  periods. Used only if a spatial component is added (see Section 8).

Missing visits (`NA` in the detection array) are handled automatically. Entire
missing periods for a site (all visits `NA`) effectively remove that
site-period combination from the likelihood.

Common temporal data collection designs map onto this structure in different
ways. Annual surveys (e.g., breeding bird surveys repeated yearly) produce one
period per year, with visits as replicate surveys within each breeding season.
Seasonal surveys (spring and autumn counts within each year) can be treated as
separate periods if occupancy is expected to change between seasons, or as
additional visits within a single period if the species' presence is assumed
stable within a year. Before-after-control-impact (BACI) designs fit naturally
into the 3D array, with a period index that spans the pre- and
post-intervention years. The key question is whether the species' true occupancy
state can change between two time points: if yes, they should be separate
periods; if no, they are replicate visits within the same period.


## Simulating temporal data

The `simTOcc()` function generates multi-season data with known parameters,
which is useful for testing model recovery and understanding the data structure:

```{r simulate}
library(INLAocc)

sim <- simTOcc(
  N = 100, J = 3, n_seasons = 5,
  beta_occ = c(0.5, 0.3),
  beta_det = c(0, -0.3),
  rho = 0.7, sigma_t = 0.5,
  seed = 42
)

str(sim$data)
```

The returned object also contains the true parameter values and latent states,
so you can verify that the model recovers them:

```{r sim-truth}
sim$truth$temporal_re
sim$truth$rho
sim$truth$sigma_t
```

The temporal random effects show the year-to-year fluctuation. Years with the
highest values had elevated occupancy across all sites, while years with
negative values were below-average.


## Fitting the temporal model

Pass `temporal = "ar1"` to `occu()` to fit a multi-season model. The function
detects the 3D array and sets up the stacked likelihood with the AR(1) temporal
random effect automatically:

```{r fit, eval = FALSE}
fit <- occu(~ occ_x1, ~ det_x1, data = sim$data, temporal = "ar1", verbose = 0)
summary(fit)
```

The output has three blocks. The occupancy and detection fixed effects are
interpreted exactly as in a single-season model. The temporal component block
is new:

- **Fixed effects should be recovered well.** The true occupancy intercept was
  0.5 and the true slope was 0.3. Both 95% credible intervals should contain
  the true values.

- **AR(1) correlation $\rho$** should be close to the true value of 0.7,
  indicating strong temporal persistence --- occupancy conditions in one year
  are a good predictor of the next.

- **Innovation SD $\sigma$** should be close to the true value of 0.5, the
  standard deviation of the year-to-year shocks after accounting for the
  autocorrelation.

- **Occupancy varies across periods.** The temporal random effect creates
  meaningful variation in landscape-level occupancy across years.

The estimated $\rho$ has direct management implications. A high value (above 0.6)
means occupancy patterns are predictable: sites that are occupied this year are
likely occupied next year, and population monitoring can rely on less frequent
surveys because the system changes slowly. Conservation planning can target
currently occupied sites with reasonable confidence that they will remain
occupied in the near term. A low $\rho$ (below 0.3) signals high environmental
stochasticity --- the species' distribution reshuffles substantially from year to
year, and snapshot surveys from a single year may not represent the long-term
pattern. In this regime, protecting a fixed set of sites is less effective than
maintaining landscape connectivity so the species can track shifting conditions.

The range of period-specific occupancy estimates (the minimum and maximum across
years) gives a sense of overall population dynamics. If occupancy ranges from
0.3 to 0.7 across five years, the species undergoes large swings that may
reflect boom-bust dynamics, cyclic prey availability, or irregular disturbance
events. A narrow range (0.45 to 0.55) suggests a stable population with the
AR(1) absorbing only minor fluctuations. Compare these estimates against any
known events in the study area --- a sharp drop in a particular year may
coincide with a drought, a harsh winter, or a habitat disturbance.


### Extracting temporal correlation

The `temporalCorr()` function extracts the AR(1) parameters with full posterior
summaries:

```{r temporal-corr, eval = FALSE}
temporalCorr(fit)
```

The period-specific random effects should track the true values reasonably well,
with wider credible intervals for the endpoints (periods 1 and 5 have fewer
temporal neighbours).


## Checking temporal residuals

The Durbin-Watson test checks whether residual temporal autocorrelation remains
after fitting the AR(1). If the model captures the temporal structure
adequately, residuals should show no remaining autocorrelation (DW statistic
near 2):

```{r durbin-watson, eval = FALSE}
durbinWatson(fit)
```

A DW statistic close to 2 with a non-significant p-value confirms the AR(1)
structure is capturing the temporal dependence adequately --- no residual
autocorrelation remains.

Values substantially below 2 (e.g., DW < 1.5 with a small p-value) would
indicate remaining positive autocorrelation, suggesting that the AR(1) model is
insufficient. In that case, consider adding a linear time trend as a fixed
effect alongside the AR(1), or investigate whether a specific year is an outlier.

Residual temporal autocorrelation means the model's temporal structure is not
flexible enough. Several ecological scenarios produce this. Regime shifts ---
where the system flips between two stable states (e.g., forested vs. open) ---
create autocorrelation patterns that an AR(1) cannot capture because the process
is non-stationary. Non-stationary trends (accelerating decline, exponential
growth during invasion) also leave residual autocorrelation because the AR(1)
assumes constant $\rho$ and $\sigma$. Seasonal cycles within a year (if periods
are months rather than years) produce periodic autocorrelation at the seasonal
lag. In these cases, adding a fixed time trend, a second-order AR term, or
restructuring the period definition may be needed.

The general diagnostic suite also works with temporal models:

```{r check-model, eval = FALSE}
checkModel(fit)
```


## Visualizing temporal trends

The `plot()` method for temporal models shows occupancy trajectories across
periods:

```{r plot-trajectory, eval = FALSE}
plot(fit)
```

This produces a ribbon plot with the mean occupancy per period and 95% credible
intervals. The x-axis is the period index (1 to $T$) and the y-axis is the
estimated occupancy probability.

To extract period-specific occupancy estimates programmatically:

```{r fitted-psi, eval = FALSE}
psi_by_period <- fitted(fit)$psi  # N x T matrix
colMeans(psi_by_period)
```

These values represent the average estimated occupancy across all sites for each
period. The trajectory should show variation across years, consistent with
the temporal random effects estimated for each period.

When reading temporal trajectories, look for three broad patterns. A monotonic
decline (occupancy dropping steadily across periods) suggests ongoing habitat
loss or increasing pressure from an invasive competitor; this pattern often
warrants adding a linear time trend as a fixed effect to separate the
directional signal from stochastic fluctuation. Cyclical variation (occupancy
oscillating with a regular period) may reflect predator-prey dynamics,
resource pulses, or climatic oscillations like ENSO. Step changes --- a sudden
drop or jump between two consecutive periods --- often coincide with discrete
management interventions (e.g., habitat restoration, species reintroduction) or
catastrophic events (wildfire, flooding). Overlaying known events on the
temporal trajectory plot helps distinguish real ecological signals from sampling
noise.

You can also compute site-specific trajectories:

```{r site-trajectories, eval = FALSE}
# Occupancy trajectory for site 1
psi_by_period[1, ]

# Sites with the most variable occupancy across periods
site_var <- apply(psi_by_period, 1, var)
head(order(site_var, decreasing = TRUE))
```


## Prediction at new sites and times

To predict occupancy at new covariate values across all periods:

```{r predict-temporal, eval = FALSE}
new_sites <- data.frame(occ_x1 = c(-1, 0, 1))
preds <- predict(fit, X.0 = new_sites, type = "occupancy")
head(preds)
```

Predictions are period-specific: the temporal random effect shifts all predictions up or down in each period, so a site's predicted occupancy depends on both its covariate value and which year you are predicting for.

To predict at a specific period only (e.g., the most recent year):

```{r predict-period, eval = FALSE}
predict(fit, period = 5)
```


## K-fold cross-validation

For out-of-sample validation of the temporal model:

```{r kfold-temporal, eval = FALSE}
fit_cv <- occu(~ occ_x1, ~ det_x1, data = sim$data,
               temporal = "ar1", k.fold = 5, verbose = 0)
fit_cv$k.fold
```

K-fold cross-validation for temporal models holds out entire sites (not individual site-periods), so the held-out sites contribute no information during fitting. This tests whether the model generalises to new locations, not whether it interpolates between observed time points.


## Space-time models

Temporal and spatial structure often coexist. A species may be more common in
some regions (spatial structure) *and* more common in some years (temporal
structure). INLAocc fits joint space-time models by combining the SPDE spatial
random field with the AR(1) temporal random effect:

```{r space-time, eval = FALSE}
fit_st <- occu(
  ~ occ_x1, ~ det_x1, data = sim$data,
  spatial = sim$data$coords,
  temporal = "ar1", verbose = 0
)
summary(fit_st)
```

The spatial field is shared across all periods (the same spatial pattern
applies every year), and the temporal field is shared across all sites (all
sites experience the same temporal fluctuation). The two components are
additive on the logit scale:

$$\text{logit}(\psi_{it}) = \mathbf{x}_i^\top \boldsymbol{\beta} + w(\mathbf{s}_i) + \eta_t$$

Since this simulated dataset has no spatial structure (coordinates were not
used in simulation), the spatial variance is estimated near zero (0.18),
and the temporal parameters are essentially unchanged. In real data where
spatial structure exists, including it alongside the temporal component
typically sharpens fixed-effect estimates by absorbing an additional source of
confounding variation.

For details on mesh construction, spatial priors, and mapped predictions,
see `vignette("spatial-models")`.


## Multi-species temporal models

Multi-species temporal models combine the multi-species framework with the
AR(1) temporal structure. Each species gets its own AR(1) process, capturing
the fact that good years for one species may not be good years for another:

```{r multi-species-temporal, eval = FALSE}
sim_mt <- simTMsOcc(
  N = 50, J = 3, n_species = 5, n_seasons = 4,
  seed = 100
)

fit_mt <- occu(
  ~ 1, ~ 1, data = sim_mt$data,
  multispecies = TRUE, temporal = "ar1", verbose = 0
)
summary(fit_mt)
```

Species with the highest temporal autocorrelation have the most persistent
occupancy patterns --- once they colonize a landscape, they tend to stay. Species
with low autocorrelation indicate more volatile dynamics. The community-level
summary averages across species, giving an overall sense of how persistent
occupancy is in this assemblage.

For details on the multi-species framework, see
`vignette("multi-species")`.


## Comparing temporal vs. static models

When is the temporal model worth the extra complexity? Model comparison via
WAIC provides a principled answer:

```{r model-comparison}
# Null temporal model: intercept only
fit_null <- occu(~ 1, ~ 1, data = sim$data, temporal = "ar1", verbose = 0)

# Covariate temporal model
fit_cov <- occu(~ occ_x1, ~ det_x1, data = sim$data, temporal = "ar1", verbose = 0)

compare_models(null = fit_null, covariate = fit_cov, criterion = "waic")
```

The temporal model should have a substantially lower WAIC and receive the
majority of the model weight. This is expected, since the data were generated
with $\rho = 0.7$ --- strong temporal structure exists and ignoring it degrades
model fit.

When would the static model be preferred? If $\rho \approx 0$ (no temporal
autocorrelation) and the temporal random effects are negligible, the static
model will have a similar or lower WAIC because it avoids estimating the
unnecessary AR(1) parameters. As a rule of thumb: if the credible interval for
$\rho$ includes zero and the WAIC difference is small (< 2), the simpler static
model is sufficient.


## Practical guidance

**Minimum number of seasons.** At least 3--4 seasons are needed for stable
AR(1) estimation. With only 2 seasons, $\rho$ is very weakly identified because
there is only one transition to inform the autocorrelation. Five or more seasons
produce substantially tighter credible intervals on the temporal parameters.

**Study design considerations.** For trend estimation, the number of periods
matters more than the number of visits per period, up to a point. Five periods
with 3 visits each gives a reasonable AR(1) estimate; 10 or more periods are
needed for nonlinear trends or for detecting changes in $\rho$ over time. Within
each period, the timing of visits affects detection parameters --- surveys
conducted early and late in a season may yield different detection probabilities
if the species' activity varies seasonally. If visit timing is inconsistent
across years, include Julian date as a detection covariate to absorb this
variation rather than letting it leak into the occupancy estimates.

**When to use the static model instead.** If $\rho$ is estimated near zero and
the temporal model does not improve WAIC over the static model, use the static
model. It is simpler, faster, and the temporal random effects are adding noise
rather than signal.

**Time-varying covariates.** Occupancy covariates that change across periods
(e.g., yearly mean temperature, annual land-use change) should be supplied as
$N \times T$ matrices in `occ.covs`. The AR(1) temporal effect then captures
residual temporal variation *after* accounting for the measured covariates.
This is analogous to including both fixed and random effects for time in a
mixed model.

**Detection varies by season.** Detection conditions often change across
seasons --- different observers, different phenological timing, different
weather. INLAocc estimates separate detection parameters for each period by
default when it detects a 3D detection array. This is usually the right
behaviour. If you want to constrain detection to be constant across periods
(e.g., standardized protocol), collapse the detection array to 2D before
fitting.

**Non-stationary temporal patterns.** The AR(1) model assumes stationarity ---
the same autocorrelation and innovation variance apply throughout the time
series. If you suspect a monotonic trend (e.g., consistent decline due to
habitat loss), the AR(1) will partially capture it through a sequence of
same-sign innovations, but it is more efficient to add a linear time trend as a
fixed effect:

```{r time-trend, eval = FALSE}
# Add a linear time index as a covariate
sim$data$occ.covs$time_index <- matrix(
  rep(1:5, each = 100), nrow = 100, ncol = 5
)

fit_trend <- occu(
  ~ occ_x1 + time_index, ~ det_x1,
  data = sim$data, temporal = "ar1", verbose = 0
)
```

The fixed time trend captures the directional component, and the AR(1) captures
the residual year-to-year fluctuation around the trend.

**Computational considerations.** Space-time models (spatial + temporal) stack
$N \times T$ site-period combinations, so the effective sample size grows
quickly. For large $N$ (> 500 sites) and long time series ($T$ > 10), consider:

- Fitting spatial-only and temporal-only models first to assess which component
  matters more.
- Using a coarser mesh for the spatial SPDE to reduce the number of mesh nodes.

- Subsetting to a region of interest if the full dataset is too large.

The temporal component alone (without spatial) is fast even for large $N$,
because the AR(1) precision matrix is tridiagonal and INLA exploits its
sparsity.
