---
title: "Multi-Season Occupancy Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multi-Season Occupancy Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
HAS_INLA <- requireNamespace("INLA", quietly = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  dev = "svglite",
  fig.ext = "svg",
  eval = HAS_INLA,
  cache = TRUE,
  message = FALSE,
  warning = FALSE
)
# Transparent ggplot backgrounds for light/dark mode
theme_transparent <- function() {
  ggplot2::theme(
    plot.background  = ggplot2::element_rect(fill = "transparent", colour = NA),
    panel.background = ggplot2::element_rect(fill = "transparent", colour = NA),
    legend.background = ggplot2::element_rect(fill = "transparent", colour = NA),
    legend.key       = ggplot2::element_rect(fill = "transparent", colour = NA)
  )
}
```

## Introduction

Many ecological studies span multiple years or seasons. A species that occupies
a site in one year may go locally extinct the next, and previously unoccupied
sites may be colonized. Single-season occupancy models treat each site as a
snapshot, discarding the temporal information embedded in repeated surveys across
years. When the data contain temporal structure, ignoring it wastes information
at best and produces biased estimates at worst. Occupancy trends are
confounded with detection variation, and standard errors are too narrow because
the model treats non-independent observations as independent.

Multi-season (dynamic) occupancy models address this by decomposing temporal
variation into a structured process. In the classical formulation (MacKenzie
et al. 2003), colonization and extinction rates are modelled explicitly. INLAocc
takes a different approach: it models the temporal dynamics as an AR(1) random
effect on occupancy, which captures the same persistence structure (high
occupancy tends to follow high occupancy) without requiring separate
colonization/extinction parameters. This is more parsimonious when the goal is
inference on occupancy probability and its drivers rather than the transition
rates themselves.

INLAocc fits multi-season models by stacking all site-period combinations into a
joint occupancy model with an AR(1) temporal random effect. Detection is fit
per-period, so each season gets its own detection parameters. Survey
conditions in July 2019 need not resemble those in July 2023. The temporal
random effect $\eta_t$ captures shared year-to-year fluctuation: good years
(favourable weather, high resource availability) raise occupancy across all
sites, and bad years lower it. Fixed effects on occupancy and detection are
estimated simultaneously, with the AR(1) structure absorbing residual temporal
autocorrelation that would otherwise inflate standard errors.


## The multi-season occupancy model

For site $i$ at time period $t$, the occupancy process is:

$$z_{it} \sim \text{Bernoulli}(\psi_{it}), \quad \text{logit}(\psi_{it}) = \mathbf{x}_i^\top \boldsymbol{\beta} + \eta_t$$

where $z_{it}$ is the true (latent) occupancy state and $\psi_{it}$ is the
occupancy probability. The temporal random effect follows an AR(1) process:

$$\eta_t = \rho \cdot \eta_{t-1} + \epsilon_t, \quad \epsilon_t \sim \mathcal{N}(0, \sigma^2_\eta)$$

where $\rho \in (-1, 1)$ is the temporal autocorrelation parameter and
$\sigma^2_\eta$ is the innovation variance.

The observation process (within each period) is:

$$y_{ijt} \mid z_{it} \sim \text{Bernoulli}(z_{it} \cdot p_{ijt}), \quad \text{logit}(p_{ijt}) = \mathbf{w}_{ijt}^\top \boldsymbol{\alpha}$$

where $y_{ijt}$ is the detection/non-detection outcome at site $i$, visit $j$,
period $t$, and $p_{ijt}$ is the detection probability.

Three features of this formulation matter in practice:

- **$\eta_t$ captures shared temporal trends.** All sites experience the same
  temporal random effect in a given period. This represents "good years vs. bad
  years" for the species, reflecting shared environmental conditions that raise
  or lower occupancy across the landscape.

- **$\rho$ controls persistence.** A high $\rho$ (e.g., 0.8) means occupancy
  patterns persist across periods: if this year is good, next year is likely good
  too. A low $\rho$ means each year is nearly independent of the last.

- **Detection varies across periods.** The detection covariates
  $\mathbf{w}_{ijt}$ can include period-specific variables (e.g., survey date,
  observer effort, weather during surveys). Each period's detection model is
  estimated from that period's data, so changing survey protocols across years
  are handled naturally.

What does $\rho = 0.7$ mean in ecological terms? It means that if occupancy is
above average this year, about 70% of that excess carries over to the next year.
A site network where most occupied sites remain occupied and most empty sites
remain empty produces high $\rho$. This is typical for species with strong site
fidelity or slow population dynamics (territorial birds, long-lived
amphibians, or perennial plants). A low $\rho$ (say 0.2) indicates that knowing
this year's occupancy tells you little about next year: the species may be
nomadic, or local populations may be driven by stochastic events like flooding or
fire.

The innovation standard deviation $\sigma_\eta$ captures the magnitude of
year-to-year environmental stochasticity after accounting for temporal
persistence. A large $\sigma_\eta$ with high $\rho$ means the system has strong
shocks that then persist (a bad year stays bad for several years). A large
$\sigma_\eta$ with low $\rho$ means large but transient fluctuations.

### Relationship to the colonization-extinction parameterization

The classical dynamic occupancy model (MacKenzie et al. 2003) parameterizes
temporal dependence through explicit transition probabilities: colonization
$\gamma$ (probability that an unoccupied site becomes occupied) and extinction
$\epsilon$ (probability that an occupied site becomes unoccupied). The
occupancy state at time $t$ depends deterministically on the state at $t-1$:

$$\psi_{i,t} = \psi_{i,t-1}(1 - \epsilon) + (1 - \psi_{i,t-1})\gamma$$

This is a first-order Markov chain on $[0, 1]$. The AR(1) model achieves a
similar temporal dependence structure, but on the logit scale rather than
the probability scale:

$$\text{logit}(\psi_{i,t}) = \mathbf{x}_i^\top \boldsymbol{\beta} + \rho \cdot \eta_{t-1} + \epsilon_t$$

Both formulations capture temporal dependence, but they differ in
what they directly estimate. The colonization-extinction model estimates the
transition rates $(\gamma, \epsilon)$ explicitly, which is natural when these
rates are the inferential target (for instance, when assessing how quickly
a species recolonizes restored habitat). The AR(1) model estimates occupancy
trends and their drivers without requiring separate identification of
colonization and extinction, which is more parsimonious when the transition
rates are not themselves of interest.

A practical advantage of the AR(1) is identifiability. The
colonization-extinction model requires enough data at each transition to
separately estimate $\gamma$ and $\epsilon$, and when these rates interact
with covariates, the number of parameters grows quickly relative to the
information in the data. The AR(1) model adds only two hyperparameters
($\rho$ and $\sigma_\eta$) regardless of the number of covariates, because the
temporal structure is orthogonal to the covariate effects. This makes it
well-suited to short time series (3--5 periods) where the classical
dynamic model may struggle with convergence.

The marginal variance of the stationary AR(1) process is
$\sigma^2_\eta / (1 - \rho^2)$, which gives the total temporal variance in
occupancy on the logit scale. When $\rho$ is close to 1, this marginal
variance can be large even if the innovation variance $\sigma^2_\eta$ is
small, because small shocks accumulate over time. This is analogous to a
system with low extinction ($\epsilon \approx 0$) and low colonization
($\gamma \approx 0$) in the classical framework: the occupancy state is
highly persistent and any perturbation takes many time steps to dissipate.


### The separable covariance assumption

The model as specified uses a *separable* spatio-temporal covariance
structure: the spatial pattern (if a spatial random field is included) is the
same every year, and the temporal pattern is the same at every site. In
notation, the total random effect decomposes as $w(\mathbf{s}_i) + \eta_t$
rather than $\xi(\mathbf{s}_i, t)$, where $\xi$ would be a full
spatio-temporal random field. Separability is a strong assumption, but it
is often reasonable when the time series is short (3--10 periods), because a
full spatio-temporal field has too many parameters to estimate reliably from
limited temporal replication.

When separability is violated (for instance, if a species is declining in
the south but expanding in the north), the separable model averages over the
heterogeneity. The fitted temporal random effect $\hat{\eta}_t$ will
represent the landscape-average temporal fluctuation, and the residuals will
show spatially structured patterns that change over time. In that case,
spatially varying coefficient (SVC) models (see `vignette("svc-models")`) or
interaction terms between spatial covariates and time can capture the
non-separable structure.


## Data structure

Multi-season data in INLAocc uses a 3D array with dimensions
sites $\times$ periods $\times$ visits:

```
y[site, period, visit]
dim: N × T × J
```

Each entry is 1 (detected), 0 (not detected), or `NA` (visit not conducted).
Here is an example for a single site surveyed over 5 years with 3 visits per
year:

```{r data-example}
# Quick simulation to show the 3D array structure
library(INLAocc)
sim_demo <- simTOcc(N = 50, J = 3, n_seasons = 5, seed = 1)
dim(sim_demo$data$y)

# Site 1, all years and visits
sim_demo$data$y[1, , ]
```

Covariates follow the same dimensional logic:

- **`occ.covs`**: Site-level covariates can be either a length-$N$ vector
  (static across periods, e.g., elevation) or an $N \times T$ matrix
  (time-varying, e.g., yearly temperature anomaly). Static covariates are
  recycled across all periods internally.

- **`det.covs`**: Visit-level covariates are $N \times T \times J$ arrays.
  For example, Julian survey date varies by site, year, and visit.

- **`coords`**: An $N \times 2$ matrix of site coordinates, constant across
  periods. Used only if a spatial component is added (see Section 8).

Missing visits (`NA` in the detection array) are handled automatically. Entire
missing periods for a site (all visits `NA`) effectively remove that
site-period combination from the likelihood.

An important asymmetry in the missing data handling: **occupancy covariates
must be complete** for all site-period combinations, even when the detection
data contain `NA` values. This is because the occupancy process model is
defined over all sites and periods, and the AR(1) temporal random effect
requires a complete time index. If an occupancy covariate is unavailable for
certain site-periods (e.g., a remotely sensed habitat variable was not
collected in a particular year), some form of imputation (mean imputation,
interpolation, or carrying forward the last observed value) must be
applied before fitting the model. Detection covariates, by contrast, can
contain `NA` values in site-period-visit combinations where no survey was
conducted; these entries are simply omitted from the detection likelihood.
If there is a mismatch between `NA` patterns in the detection array and the
detection covariates, INLAocc will issue a warning identifying the
discrepant entries.

Common temporal data collection designs map onto this structure in different
ways. Annual surveys (e.g., breeding bird surveys repeated yearly) produce one
period per year, with visits as replicate surveys within each breeding season.
Seasonal surveys (spring and autumn counts within each year) can be treated as
separate periods if occupancy is expected to change between seasons, or as
additional visits within a single period if the species' presence is assumed
stable within a year. Before-after-control-impact (BACI) designs fit naturally
into the 3D array, with a period index that spans the pre- and
post-intervention years. The key question is whether the species' true occupancy
state can change between two time points: if yes, they should be separate
periods; if no, they are replicate visits within the same period.


## Simulating temporal data

The `simTOcc()` function generates multi-season data with known parameters,
which is useful for testing model recovery and understanding the data structure:

```{r simulate}
library(INLAocc)

sim <- simTOcc(
  N = 100, J = 3, n_seasons = 5,
  beta_occ = c(0.5, 0.3),
  beta_det = c(0, -0.3),
  rho = 0.7, sigma_t = 0.5,
  seed = 42
)

str(sim$data)
```

The returned object also contains the true parameter values and latent states,
so you can verify that the model recovers them:

```{r sim-truth}
sim$truth$temporal_re
sim$truth$rho
sim$truth$sigma_t
```

The temporal random effects show the year-to-year fluctuation. Years with the
highest values had elevated occupancy across all sites, while years with
negative values were below-average.


### A more realistic scenario: trend plus noise

Real-world temporal occupancy data rarely follow a pure AR(1) process. More
commonly, a directional trend (e.g., gradual decline from habitat loss) is
overlaid with correlated noise (good years and bad years). The following
simulation creates a 10-season dataset where occupancy declines linearly over
time while year-to-year noise follows the AR(1). This scenario is common for
species in landscapes undergoing gradual land-use change.

```{r simulate-trend}
# Longer time series with a declining linear trend
sim_trend <- simTOcc(
  N = 150, J = 3, n_seasons = 10,
  beta_occ = c(1.0, 0.5),    # higher baseline, one covariate
  beta_det = c(0.5, -0.2),   # moderate detection
  rho = 0.6, sigma_t = 0.4,  # moderate persistence, moderate noise
  seed = 99
)

# Manually add a declining time trend to the occupancy linear predictor.
# simTOcc generates AR(1) noise; we overlay a fixed trend on the data
# by including a time-varying covariate.
n_seasons <- 10
trend_vals <- seq(0, -1.5, length.out = n_seasons)  # decline of 1.5 logit units
sim_trend$data$occ.covs$time_trend <- matrix(
  rep(trend_vals, each = 150), nrow = 150, ncol = n_seasons
)

# Inspect naive occupancy per period
naive_psi <- apply(sim_trend$data$y, 2, function(y_t) {
  mean(apply(y_t, 1, function(row) any(row == 1, na.rm = TRUE)))
})
data.frame(period = 1:n_seasons, naive_psi = round(naive_psi, 3))
```

The declining naive occupancy is visible even before modelling, but remember
that naive rates confound occupancy with detection. The model separates the
two processes.

```{r fit-trend-sim}
fit_trend_sim <- occu(
  ~ occ_x1 + time_trend, ~ det_x1,
  data = sim_trend$data, temporal = "ar1", verbose = 0
)
summary(fit_trend_sim)
```

The fixed `time_trend` coefficient captures the directional decline, while the
AR(1) random effect absorbs residual year-to-year noise around the trend. If
you omit the fixed trend and rely solely on the AR(1), the model can still
track the decline through a sequence of same-sign innovations, but the
estimated $\rho$ will be inflated because it has to absorb both the trend and
the autocorrelation. Separating the two is better practice when you have
theoretical reason to expect a directional change.


## Exploratory data analysis

Before fitting any model, plotting the raw detection frequencies per period
gives a useful first impression of the temporal signal and its magnitude.
Because raw detection rates confound occupancy with detection probability,
they should not be taken at face value, but they reveal gross patterns
(trends, outlier years, or a lack of temporal variation) that help guide
model specification.

```{r eda-raw}
# Compute naive occupancy per period: proportion of sites with at least one detection
naive_occ <- apply(sim$data$y, 2, function(y_t) {
  mean(apply(y_t, 1, function(row) any(row == 1, na.rm = TRUE)))
})

data.frame(period = seq_along(naive_occ), naive_occupancy = round(naive_occ, 3))
```

These raw proportions are biased low (a site can be occupied but go
undetected across all visits), but their period-to-period pattern should
correlate with the model-based estimates. A monotonic trend in raw rates
suggests a fixed time trend may be warranted alongside the AR(1). Large
between-period swings suggest the temporal random effect will absorb
substantial variation. If the raw rates are nearly constant across periods,
the AR(1) may add little, and a simpler static model could suffice.

A second useful diagnostic is the proportion of sites that switch detection
status between consecutive periods, i.e., sites that were detected in one period
but not the next (or vice versa). High switching rates indicate a dynamic
system where colonization and extinction are frequent; low switching rates
suggest stable occupancy patterns that are well-suited to the AR(1) model.

```{r eda-switching}
# Compute detection/nondetection summary per site-period
detected <- apply(sim$data$y, c(1, 2), function(row) any(row == 1, na.rm = TRUE))

# Switching rate: proportion of sites that change status between consecutive periods
n_per <- ncol(detected)
switch_rates <- numeric(n_per - 1)
for (t in 2:n_per) {
  switch_rates[t - 1] <- mean(detected[, t] != detected[, t - 1])
}
data.frame(
  transition = paste(seq_len(n_per - 1), "->", 2:n_per),
  switch_rate = round(switch_rates, 3)
)
```


## Fitting the temporal model

Pass `temporal = "ar1"` to `occu()` to fit a multi-season model. The function
detects the 3D array and sets up the stacked likelihood with the AR(1) temporal
random effect automatically:

```{r fit}
fit <- occu(~ occ_x1, ~ det_x1, data = sim$data, temporal = "ar1", verbose = 0)
summary(fit)
```

The output has three blocks. The occupancy and detection fixed effects are
interpreted exactly as in a single-season model. The temporal component block
is new:

- **Fixed effects should be recovered well.** The true occupancy intercept was
  0.5 and the true slope was 0.3. Both 95% credible intervals should contain
  the true values.

- **AR(1) correlation $\rho$** should be close to the true value of 0.7,
  indicating strong temporal persistence. Occupancy conditions in one year
  are a good predictor of the next.

- **Innovation SD $\sigma$** should be close to the true value of 0.5, the
  standard deviation of the year-to-year shocks after accounting for the
  autocorrelation.

- **Occupancy varies across periods.** The temporal random effect creates
  meaningful variation in landscape-level occupancy across years.

The estimated $\rho$ has direct management implications. A high value (above 0.6)
means occupancy patterns are predictable: sites that are occupied this year are
likely occupied next year, and population monitoring can rely on less frequent
surveys because the system changes slowly. Conservation planning can target
currently occupied sites with reasonable confidence that they will remain
occupied in the near term. A low $\rho$ (below 0.3) signals high environmental
stochasticity; the species' distribution reshuffles substantially from year to
year, and snapshot surveys from a single year may not represent the long-term
pattern. In this regime, protecting a fixed set of sites is less effective than
maintaining landscape connectivity so the species can track shifting conditions.

The range of period-specific occupancy estimates (the minimum and maximum across
years) gives a sense of overall population dynamics. If occupancy ranges from
0.3 to 0.7 across five years, the species undergoes large swings that may
reflect boom-bust dynamics, cyclic prey availability, or irregular disturbance
events. A narrow range (0.45 to 0.55) suggests a stable population with the
AR(1) absorbing only minor fluctuations. Compare these estimates against any
known events in the study area. A sharp drop in a particular year may
coincide with a drought, a harsh winter, or a habitat disturbance.


### Extracting temporal correlation

The `temporalCorr()` function extracts the AR(1) parameters with full posterior
summaries:

```{r temporal-corr}
temporalCorr(fit)
```

The period-specific random effects should track the true values reasonably well,
with wider credible intervals for the endpoints (periods 1 and 5 have fewer
temporal neighbours).


### Interpreting the period-specific random effects

The AR(1) temporal random effects $\hat{\eta}_t$ represent the shared
year-to-year fluctuation in occupancy after accounting for covariates. Each
$\hat{\eta}_t$ is an additive shift on the logit scale that raises or lowers
occupancy across all sites in period $t$. Positive values correspond to
"good years" (years where environmental conditions favoured the species
beyond what the covariates predict), and negative values correspond to
"bad years."

```{r inspect-re}
# Extract the period random effects from the joint occupancy fit
if (!is.null(fit$occ_fit) && !is.null(fit$occ_fit$summary.random$period)) {
  re_df <- fit$occ_fit$summary.random$period
  re_df$period <- seq_len(nrow(re_df))

  ggplot(re_df, aes(x = period, y = mean)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_pointrange(aes(ymin = `0.025quant`, ymax = `0.975quant`),
                    colour = "steelblue", linewidth = 0.8) +
    labs(x = "Period", y = expression(hat(eta)[t]),
         title = "Estimated temporal random effects") +
    theme_minimal() +
    theme_transparent()
}
```

The credible intervals on each $\hat{\eta}_t$ reflect how precisely the
model estimates that year's deviation. Intervals that exclude zero indicate
a statistically meaningful departure from the average occupancy level.
Comparing these estimates against known environmental events (droughts,
pest outbreaks, management interventions) can provide ecological insight into
what drives year-to-year fluctuation in the system.

The true temporal random effects from our simulation are available for
validation:

```{r re-recovery}
# Compare estimated vs true random effects
if (!is.null(fit$occ_fit) && !is.null(fit$occ_fit$summary.random$period)) {
  re_comparison <- data.frame(
    period    = seq_len(sim$truth$n_seasons),
    true_eta  = sim$truth$temporal_re,
    est_eta   = re_df$mean[seq_len(sim$truth$n_seasons)],
    est_lo    = re_df$`0.025quant`[seq_len(sim$truth$n_seasons)],
    est_hi    = re_df$`0.975quant`[seq_len(sim$truth$n_seasons)]
  )

  ggplot(re_comparison, aes(x = true_eta, y = est_eta)) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
    geom_pointrange(aes(ymin = est_lo, ymax = est_hi), colour = "steelblue") +
    labs(x = "True temporal random effect",
         y = "Estimated temporal random effect",
         title = "Recovery of period-specific random effects") +
    theme_minimal() +
    theme_transparent()
}
```

Points close to the 1:1 line with credible intervals that include the true
value confirm good parameter recovery. The endpoints (first and last period)
typically have wider intervals because the AR(1) process has fewer temporal
neighbours from which to borrow information.


## Checking temporal residuals

The Durbin-Watson test checks whether residual temporal autocorrelation remains
after fitting the AR(1). If the model captures the temporal structure
adequately, residuals should show no remaining autocorrelation (DW statistic
near 2):

```{r durbin-watson}
durbinWatson(fit)
```

A DW statistic close to 2 with a non-significant p-value confirms the AR(1)
structure is capturing the temporal dependence adequately; no residual
autocorrelation remains.

Values substantially below 2 (e.g., DW < 1.5 with a small p-value) would
indicate remaining positive autocorrelation, suggesting that the AR(1) model is
insufficient. In that case, consider adding a linear time trend as a fixed
effect alongside the AR(1), or investigate whether a specific year is an outlier.

Residual temporal autocorrelation means the model's temporal structure is not
flexible enough. Several ecological scenarios produce this. Regime shifts,
where the system flips between two stable states (e.g., forested vs. open),
create autocorrelation patterns that an AR(1) cannot capture because the process
is non-stationary. Non-stationary trends (accelerating decline, exponential
growth during invasion) also leave residual autocorrelation because the AR(1)
assumes constant $\rho$ and $\sigma$. Seasonal cycles within a year (if periods
are months rather than years) produce periodic autocorrelation at the seasonal
lag. In these cases, adding a fixed time trend, a second-order AR term, or
restructuring the period definition may be needed.

The general diagnostic suite also works with temporal models:

```{r check-model}
checkModel(fit)
```


## Visualizing temporal trends

The `plot()` method for temporal models shows occupancy trajectories across
periods:

```{r plot-trajectory}
plot(fit)
```

This produces a ribbon plot with the mean occupancy per period and 95% credible
intervals. The x-axis is the period index (1 to $T$) and the y-axis is the
estimated occupancy probability.

To extract period-specific occupancy estimates programmatically:

```{r fitted-psi}
psi_by_period <- fitted(fit)$psi  # N x T matrix
colMeans(psi_by_period)
```

These values represent the average estimated occupancy across all sites for each
period. The trajectory should show variation across years, consistent with
the temporal random effects estimated for each period.

When reading temporal trajectories, look for three broad patterns. A monotonic
decline (occupancy dropping steadily across periods) suggests ongoing habitat
loss or increasing pressure from an invasive competitor; this pattern often
warrants adding a linear time trend as a fixed effect to separate the
directional signal from stochastic fluctuation. Cyclical variation (occupancy
oscillating with a regular period) may reflect predator-prey dynamics,
resource pulses, or climatic oscillations like ENSO. Step changes (a sudden
drop or jump between two consecutive periods) often coincide with discrete
management interventions (e.g., habitat restoration, species reintroduction) or
catastrophic events (wildfire, flooding). Overlaying known events on the
temporal trajectory plot helps distinguish real ecological signals from sampling
noise.

You can also compute site-specific trajectories:

```{r site-trajectories}
# Occupancy trajectory for site 1
psi_by_period[1, ]

# Sites with the most variable occupancy across periods
site_var <- apply(psi_by_period, 1, var)
head(order(site_var, decreasing = TRUE))
```


### Occupancy trajectory with credible intervals

A publication-quality trajectory plot shows the mean occupancy per period with
95% credible intervals, making it easy to identify periods of unusually high or
low occupancy and to assess whether changes are statistically meaningful.

```{r trajectory-ggplot}
library(ggplot2)

# Compute per-period mean and quantiles across sites
psi_summary <- data.frame(
  period   = seq_len(ncol(psi_by_period)),
  mean_psi = colMeans(psi_by_period),
  lo       = apply(psi_by_period, 2, quantile, 0.025),
  hi       = apply(psi_by_period, 2, quantile, 0.975)
)

ggplot(psi_summary, aes(x = period, y = mean_psi)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_point(size = 2.5, colour = "steelblue") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Period", y = "Estimated occupancy",
       title = "Landscape-level occupancy trajectory") +
  theme_minimal() +
  theme_transparent()
```

The ribbon width reflects cross-site variation in occupancy, not posterior
uncertainty in the mean. Wide ribbons indicate heterogeneous landscapes where
some sites are consistently occupied and others are not. Narrow ribbons
indicate that most sites share a similar occupancy probability.


### Site-level occupancy heatmap

When the number of sites is moderate (under ~200), a heatmap of site-by-period
occupancy probabilities reveals spatial and temporal patterns simultaneously.
Sites that shift from high to low occupancy over time appear as horizontal
gradients; years with landscape-wide declines appear as vertical bands.

```{r site-heatmap, fig.height = 6}
# Reorder sites by mean occupancy for visual clarity
site_order <- order(rowMeans(psi_by_period))
psi_ordered <- psi_by_period[site_order, ]

# Convert to long format for ggplot
psi_long <- data.frame(
  site   = rep(seq_len(nrow(psi_ordered)), ncol(psi_ordered)),
  period = rep(seq_len(ncol(psi_ordered)), each = nrow(psi_ordered)),
  psi    = as.vector(psi_ordered)
)

ggplot(psi_long, aes(x = period, y = site, fill = psi)) +
  geom_tile() +
  scale_fill_viridis_c(option = "C", limits = c(0, 1),
                       name = expression(hat(psi))) +
  labs(x = "Period", y = "Site (ordered by mean occupancy)",
       title = "Site-level occupancy over time") +
  theme_minimal() +
  theme_transparent()
```


## Derived quantities: colonization and extinction

INLAocc uses an AR(1) temporal random effect rather than explicit colonization
($\gamma$) and extinction ($\epsilon$) parameters. This is a deliberate design
choice: the AR(1) is more parsimonious and avoids identifiability issues when
transition rates interact with covariates. However, users migrating from
spOccupancy or the classical MacKenzie et al. (2003) framework often need
colonization and extinction rates for reporting or comparison.

These quantities can be derived from the site-level occupancy posteriors.
Define:

- **Colonization rate** for period $t$: the proportion of sites that were
  unoccupied in period $t-1$ but occupied in period $t$.
- **Extinction rate** for period $t$: the proportion of sites that were
  occupied in period $t-1$ but unoccupied in period $t$.

Because we work with probabilities rather than binary states, we compute
*expected* colonization and extinction rates from the estimated $\hat{\psi}$
matrix. For each site $i$ and transition $t-1 \to t$:

$$\hat{\gamma}_{t} = \frac{\sum_{i} (1 - \hat{\psi}_{i,t-1}) \cdot \hat{\psi}_{i,t}}{\sum_{i} (1 - \hat{\psi}_{i,t-1})}$$
$$\hat{\epsilon}_{t} = \frac{\sum_{i} \hat{\psi}_{i,t-1} \cdot (1 - \hat{\psi}_{i,t})}{\sum_{i} \hat{\psi}_{i,t-1}}$$

These are population-level averages across sites, conditional on the estimated
occupancy probabilities.

```{r derived-col-ext}
n_periods <- ncol(psi_by_period)
N <- nrow(psi_by_period)

# Compute derived colonization and extinction rates
col_rates <- numeric(n_periods - 1)
ext_rates <- numeric(n_periods - 1)

for (t in 2:n_periods) {
  psi_prev <- psi_by_period[, t - 1]
  psi_curr <- psi_by_period[, t]

  # Expected colonization: unoccupied -> occupied
  vacant_prev <- 1 - psi_prev
  col_rates[t - 1] <- sum(vacant_prev * psi_curr) / sum(vacant_prev)

  # Expected extinction: occupied -> unoccupied
  ext_rates[t - 1] <- sum(psi_prev * (1 - psi_curr)) / sum(psi_prev)
}

transitions <- data.frame(
  transition    = paste(seq_len(n_periods - 1), "->",
                        seq_len(n_periods - 1) + 1),
  colonization  = round(col_rates, 4),
  extinction    = round(ext_rates, 4),
  turnover      = round(col_rates + ext_rates, 4)
)
transitions
```

Turnover (colonization + extinction) measures the total rate of change in the
community. High turnover with stable mean occupancy indicates a dynamic
landscape where sites are constantly gained and lost. High turnover with
declining occupancy indicates extinction outpacing colonization.

```{r plot-col-ext}
trans_long <- data.frame(
  transition = rep(2:n_periods, 2),
  rate       = c(col_rates, ext_rates),
  type       = rep(c("Colonization", "Extinction"), each = n_periods - 1)
)

ggplot(trans_long, aes(x = transition, y = rate, colour = type)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c("Colonization" = "forestgreen",
                                  "Extinction" = "firebrick")) +
  labs(x = "Period", y = "Rate", colour = NULL,
       title = "Derived colonization and extinction rates") +
  theme_minimal() +
  theme_transparent()
```

A note on interpretation: these derived rates are *phenomenological*, not
mechanistic. They describe the net outcome of occupancy changes, not the
underlying colonization-extinction process. In the AR(1) model, what looks
like colonization is a site whose occupancy probability increased above the
detection threshold; what looks like extinction is a site whose probability
dropped below it. For mechanistic inference about colonization and extinction
dynamics themselves, a full dynamic occupancy model (MacKenzie et al. 2003)
would be more appropriate.


## Multi-season models with time-varying covariates

Many ecological drivers change over time: annual temperature anomalies, yearly
precipitation totals, land-use change indices, or management interventions that
begin in a specific year. Time-varying covariates are supplied as $N \times T$
matrices in `occ.covs`, where each column corresponds to a period and each row
to a site. The AR(1) temporal random effect then captures residual temporal
variation *after* accounting for the measured covariates.

### Simulating time-varying covariate data

We create a scenario where occupancy is driven by a site-level habitat
covariate (static) and a yearly climate anomaly (time-varying), with an
interaction: the climate effect is stronger at sites with better habitat.

```{r sim-time-varying}
set.seed(2024)
N <- 120
J <- 3
n_seasons <- 8

# Static habitat covariate
habitat <- rnorm(N)

# Time-varying climate anomaly (same across sites within a year)
climate_yearly <- rnorm(n_seasons, mean = 0, sd = 0.8)
climate_mat <- matrix(rep(climate_yearly, each = N), nrow = N, ncol = n_seasons)

# Interaction: climate x habitat
interact_mat <- climate_mat * habitat

# AR(1) temporal noise
rho_true <- 0.5
sigma_true <- 0.3
eta <- numeric(n_seasons)
eta[1] <- rnorm(1, 0, sigma_true / sqrt(1 - rho_true^2))
for (t in 2:n_seasons) {
  eta[t] <- rho_true * eta[t - 1] + rnorm(1, 0, sigma_true)
}

# True occupancy and detection
beta_0 <- 0.5; beta_hab <- 0.8; beta_clim <- 0.4; beta_int <- 0.3
y_array <- array(NA, dim = c(N, n_seasons, J))
z_mat <- matrix(NA, N, n_seasons)

for (t in seq_len(n_seasons)) {
  logit_psi <- beta_0 + beta_hab * habitat + beta_clim * climate_mat[, t] +
               beta_int * interact_mat[, t] + eta[t]
  psi_t <- plogis(logit_psi)
  z_mat[, t] <- rbinom(N, 1, psi_t)
  for (j in seq_len(J)) {
    y_array[, t, j] <- rbinom(N, 1, z_mat[, t] * 0.6)
  }
}

# Package as INLAocc data
tv_data <- list(
  y = y_array,
  occ.covs = list(
    habitat      = habitat,
    climate      = climate_mat,
    hab_x_clim   = interact_mat
  ),
  det.covs = list(),
  coords = cbind(x = runif(N), y = runif(N))
)
```

### Fitting the model with time-varying covariates

The formula includes the static covariate, the time-varying covariate, and
their interaction. INLAocc handles the dimensional differences automatically:
`habitat` is a length-$N$ vector (recycled across periods), while `climate`
and `hab_x_clim` are $N \times T$ matrices.

```{r fit-time-varying}
fit_tv <- occu(
  ~ habitat + climate + hab_x_clim,
  ~ 1,
  data = tv_data,
  temporal = "ar1",
  verbose = 0
)
summary(fit_tv)
```

The `habitat` coefficient captures the time-averaged effect of habitat quality
on occupancy. The `climate` coefficient captures the direct effect of yearly
climate anomalies: positive values mean warmer (or wetter) years increase
occupancy. The `hab_x_clim` interaction captures whether climate sensitivity
varies with habitat quality. A positive interaction means the climate effect
is amplified at high-habitat sites; species in good habitat respond more
strongly to favourable climate years, perhaps because resource availability
mediates the climate response.

### Comparing models with and without time-varying covariates

To test whether the time-varying covariate adds explanatory power beyond the
temporal random effect:

```{r compare-tv}
fit_static_only <- occu(
  ~ habitat, ~ 1,
  data = tv_data,
  temporal = "ar1",
  verbose = 0
)

compare_models(
  static = fit_static_only,
  time_varying = fit_tv,
  criterion = "waic"
)
```

If the time-varying model has substantially lower WAIC, the climate covariate
captures temporal variation that the AR(1) alone cannot explain. The AR(1)
in the time-varying model then absorbs only the residual temporal noise not
explained by climate, and its innovation variance $\sigma_\eta$ should be
smaller than in the static-only model.


## Prediction at new sites and times

To predict occupancy at new covariate values across all periods:

```{r predict-temporal}
new_sites <- data.frame(occ_x1 = c(-1, 0, 1))
preds <- predict(fit, X.0 = new_sites, type = "occupancy")
head(preds)
```

Predictions are period-specific: the temporal random effect shifts all predictions up or down in each period, so a site's predicted occupancy depends on both its covariate value and which year you are predicting for.

To predict at a specific period only (e.g., the most recent year):

```{r predict-period}
predict(fit, period = 5)
```


## K-fold cross-validation

For out-of-sample validation of the temporal model:

```{r kfold-temporal}
fit_cv <- occu(~ occ_x1, ~ det_x1, data = sim$data,
               temporal = "ar1", k.fold = 5, verbose = 0)
fit_cv$k.fold
```

K-fold cross-validation for temporal models holds out entire sites (not individual site-periods), so the held-out sites contribute no information during fitting. This tests whether the model generalises to new locations, not whether it interpolates between observed time points.


## Space-time models

Temporal and spatial structure often coexist. A species may be more common in
some regions (spatial structure) *and* more common in some years (temporal
structure). INLAocc fits joint space-time models by combining the SPDE spatial
random field with the AR(1) temporal random effect:

```{r space-time}
fit_st <- occu(
  ~ occ_x1, ~ det_x1, data = sim$data,
  spatial = sim$data$coords,
  temporal = "ar1", verbose = 0
)
summary(fit_st)
```

The spatial field is shared across all periods (the same spatial pattern
applies every year), and the temporal field is shared across all sites (all
sites experience the same temporal fluctuation). The two components are
additive on the logit scale:

$$\text{logit}(\psi_{it}) = \mathbf{x}_i^\top \boldsymbol{\beta} + w(\mathbf{s}_i) + \eta_t$$

Since this simulated dataset has no spatial structure (coordinates were not
used in simulation), the spatial variance is estimated near zero,
and the temporal parameters are essentially unchanged. In real data where
spatial structure exists, including it alongside the temporal component
typically sharpens fixed-effect estimates by absorbing an additional source of
confounding variation.

### When does the spatial component help?

The spatial component is worth including when sites that are close together
have more similar occupancy than distant sites, which is common for most
species. The spatial random field absorbs this similarity, preventing it from
inflating the residual variance and producing overconfident fixed-effect
estimates. In practice, adding the spatial component:

- **Reduces confounding** between spatial covariates and spatially structured
  unmeasured variables.
- **Improves predictions** at unsampled locations by borrowing information from
  nearby sites.
- **Slightly increases computation** because the SPDE mesh adds parameters.
  For large datasets ($N > 500$), this cost is meaningful but usually
  worthwhile.

### A worked example with both spatial and temporal structure

To demonstrate the joint model on data that actually contain both structures,
we simulate a dataset where occupancy depends on a spatial gradient and an
AR(1) temporal process.

```{r sim-space-time}
set.seed(2025)
N_st <- 100
J_st <- 3
n_seasons_st <- 6

# Spatial coordinates on a unit square
coords_st <- cbind(x = runif(N_st), y = runif(N_st))

# Spatial effect: smooth gradient from southwest to northeast
spatial_effect <- 1.5 * (coords_st[, 1] + coords_st[, 2]) / 2

# AR(1) temporal random effect
rho_st <- 0.7
sigma_st <- 0.4
eta_st <- numeric(n_seasons_st)
eta_st[1] <- rnorm(1, 0, sigma_st / sqrt(1 - rho_st^2))
for (t in 2:n_seasons_st) {
  eta_st[t] <- rho_st * eta_st[t - 1] + rnorm(1, 0, sigma_st)
}

# Generate data
y_st <- array(NA, dim = c(N_st, n_seasons_st, J_st))
for (t in seq_len(n_seasons_st)) {
  psi_t <- plogis(-0.5 + spatial_effect + eta_st[t])
  z_t <- rbinom(N_st, 1, psi_t)
  for (j in seq_len(J_st)) {
    y_st[, t, j] <- rbinom(N_st, 1, z_t * 0.5)
  }
}

st_data <- list(
  y = y_st,
  occ.covs = list(),
  det.covs = list(),
  coords = coords_st
)
```

```{r fit-space-time-full}
# Temporal only (no spatial)
fit_temporal_only <- occu(
  ~ 1, ~ 1, data = st_data,
  temporal = "ar1", verbose = 0
)

# Joint space-time model
fit_space_time <- occu(
  ~ 1, ~ 1, data = st_data,
  spatial = coords_st,
  temporal = "ar1", verbose = 0
)

compare_models(
  temporal_only = fit_temporal_only,
  space_time    = fit_space_time,
  criterion     = "waic"
)
```

The space-time model should receive the majority of the WAIC weight because
the data contain a genuine spatial gradient. The temporal parameters ($\rho$,
$\sigma_\eta$) should be similar between the two models because the spatial
and temporal structures are independent by construction.

```{r st-temporal-params}
# Temporal correlation in both models
temporalCorr(fit_temporal_only)
temporalCorr(fit_space_time)
```

If the spatial and temporal estimates are nearly identical across the two
models, this confirms the separability assumption: the spatial pattern is
stable across years, and the temporal fluctuation is uniform across space.
When this assumption is violated (for example, if some regions decline
faster than others), the separable model will average over the
heterogeneity. In that case, spatially varying coefficients (SVC models; see
`vignette("svc-models")`) may be needed.

For details on mesh construction, spatial priors, and mapped predictions,
see `vignette("spatial-models")`.


## Multi-species temporal models

Multi-species temporal models combine the multi-species framework with the
AR(1) temporal structure. Each species gets its own AR(1) process, capturing
the fact that good years for one species may not be good years for another:

```{r multi-species-temporal}
sim_mt <- simTMsOcc(
  N = 50, J = 3, n_species = 5, n_seasons = 4,
  seed = 100
)

fit_mt <- occu(
  ~ 1, ~ 1, data = sim_mt$data,
  multispecies = TRUE, temporal = "ar1", verbose = 0
)
summary(fit_mt)
```

Species with the highest temporal autocorrelation have the most persistent
occupancy patterns: once they colonize a landscape, they tend to stay. Species
with low autocorrelation indicate more volatile dynamics. The community-level
summary averages across species, indicating how persistent
occupancy is in this assemblage.

For details on the multi-species framework, see
`vignette("multi-species")`.


## Comparing temporal vs. static models

When is the temporal model worth the extra complexity? Model comparison via
WAIC provides a principled answer:

```{r model-comparison}
# Null temporal model: intercept only
fit_null <- occu(~ 1, ~ 1, data = sim$data, temporal = "ar1", verbose = 0)

# Covariate temporal model
fit_cov <- occu(~ occ_x1, ~ det_x1, data = sim$data, temporal = "ar1", verbose = 0)

compare_models(null = fit_null, covariate = fit_cov, criterion = "waic")
```

The temporal model should have a substantially lower WAIC and receive the
majority of the model weight. This is expected, since the data were generated
with $\rho = 0.7$, so strong temporal structure exists and ignoring it degrades
model fit.

When would the static model be preferred? If $\rho \approx 0$ (no temporal
autocorrelation) and the temporal random effects are negligible, the static
model will have a similar or lower WAIC because it avoids estimating the
unnecessary AR(1) parameters. As a rule of thumb: if the credible interval for
$\rho$ includes zero and the WAIC difference is small (< 2), the simpler static
model is sufficient.


## Practical guidance

### How many seasons are needed?

At least 3--4 seasons are needed for stable AR(1) estimation. With only 2
seasons, $\rho$ is very weakly identified because there is only one transition
to inform the autocorrelation. Five or more seasons produce substantially
tighter credible intervals on the temporal parameters.

The following table summarises approximate expectations for different time
series lengths:

| Seasons | AR(1) estimation | Trend detection | Colonization/extinction |
|:-------:|:----------------:|:---------------:|:-----------------------:|
| 2       | Unreliable       | Not possible    | Not meaningful          |
| 3--4    | Adequate         | Linear only     | Very noisy              |
| 5--7    | Good             | Linear + noise  | Moderate precision      |
| 8--15   | Excellent        | Nonlinear       | Good precision          |
| 15+     | Excellent        | Regime shifts   | Excellent               |

For detecting *nonlinear* temporal patterns (e.g., a decline that accelerates
over time), 10 or more periods are typically needed. For detecting changes in
$\rho$ itself over time (non-stationarity), even longer series are required
because the autocorrelation parameter converges more slowly than the mean.

### AR(1) vs IID year effects

INLAocc supports two temporal random effect structures:

- **`temporal = "ar1"`**: Year effects are temporally correlated. $\eta_t$
  depends on $\eta_{t-1}$ through the autocorrelation $\rho$. This is
  appropriate when occupancy in one year predicts occupancy in the next,
  the typical case for species with site fidelity, slow population dynamics,
  or persistent habitat quality.

- **`temporal = "iid"`**: Year effects are independent draws from a common
  distribution. Each year's random effect is unrelated to the previous year.
  This is appropriate when temporal variation is driven by unpredictable events
  (e.g., stochastic weather, irregular disturbance) and there is no reason to
  expect temporal persistence.

When in doubt, fit both and compare via WAIC:

```{r ar1-vs-iid}
fit_ar1 <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                temporal = "ar1", verbose = 0)
fit_iid <- occu(~ occ_x1, ~ det_x1, data = sim$data,
                temporal = "iid", verbose = 0)

compare_models(ar1 = fit_ar1, iid = fit_iid, criterion = "waic")
```

Since these data were simulated with $\rho = 0.7$, the AR(1) model should be
strongly preferred. For species with genuinely independent year-to-year
fluctuations, the IID model will have similar or lower WAIC.

A practical rule: if the Durbin-Watson test on an IID fit shows significant
residual autocorrelation, switch to AR(1). If the AR(1) fit estimates $\rho$
near zero with a credible interval spanning zero, the IID model is sufficient.

### When temporal structure does not help

Not every multi-season dataset benefits from a temporal random effect. The
temporal model adds parameters ($\rho$, $\sigma_\eta$, and the period-specific
random effects), and if these parameters are poorly identified, the model
wastes degrees of freedom and may produce wider credible intervals on the
fixed effects than a simpler model.

Temporal structure is unlikely to help when:

1. **Only 2 seasons exist.** There is one transition, which provides almost no
   information about $\rho$. Pool the data into a single-season model or treat
   each season independently.

2. **Temporal variation is fully explained by covariates.** If a time-varying
   covariate (e.g., annual precipitation) captures all year-to-year variation,
   the AR(1) innovation variance collapses to zero and the temporal random
   effect is unnecessary.

3. **The study design has irregular gaps.** If some periods are missing for most
   sites, the AR(1) cannot bridge the gap effectively, and the information
   gain from temporal correlation is minimal.

4. **Detection is the dominant source of variation.** When occupancy is nearly
   constant but detection varies strongly across years (e.g., due to changing
   survey effort), a temporal random effect on detection may be more
   appropriate than one on occupancy. INLAocc handles this through
   period-specific detection estimation.

### Study design considerations

For trend estimation, the number of periods matters more than the number of
visits per period, up to a point. Five periods with 3 visits each gives a
reasonable AR(1) estimate; 10 or more periods are needed for nonlinear trends
or for detecting changes in $\rho$ over time. Within each period, the timing
of visits affects detection parameters. Surveys conducted early and late in
a season may yield different detection probabilities if the species' activity
varies seasonally. If visit timing is inconsistent across years, include
Julian date as a detection covariate to absorb this variation rather than
letting it leak into the occupancy estimates.

Unbalanced designs (different numbers of visits across periods, or sites
surveyed in some years but not others) are handled naturally by INLAocc.
Missing visits are coded as `NA` in the 3D detection array, and the model
simply omits them from the likelihood. However, periods with very few visits
or very few surveyed sites contribute less information, and their occupancy
estimates will have wider credible intervals.

### Time-varying covariates

Occupancy covariates that change across periods (e.g., yearly mean
temperature, annual land-use change) should be supplied as $N \times T$
matrices in `occ.covs`. The AR(1) temporal effect then captures residual
temporal variation *after* accounting for the measured covariates. This is
analogous to including both fixed and random effects for time in a mixed
model. See the "Multi-season models with time-varying covariates" section
above for a worked example.

### Detection varies by season

Detection conditions often change across seasons (different observers,
different phenological timing, different weather). INLAocc estimates separate
detection parameters for each period by default when it detects a 3D detection
array. This is usually the right behaviour. If you want to constrain detection
to be constant across periods (e.g., standardized protocol), collapse the
detection array to 2D before fitting.

### Non-stationary temporal patterns

The AR(1) model assumes stationarity: the same autocorrelation and
innovation variance apply throughout the time series. If you suspect a
monotonic trend (e.g., consistent decline due to habitat loss), the AR(1) will
partially capture it through a sequence of same-sign innovations, but it is
better to add a linear time trend as a fixed effect:

```{r time-trend}
# Add a linear time index as a covariate
sim$data$occ.covs$time_index <- matrix(
  rep(1:5, each = 100), nrow = 100, ncol = 5
)

fit_trend <- occu(
  ~ occ_x1 + time_index, ~ det_x1,
  data = sim$data, temporal = "ar1", verbose = 0
)
```

The fixed time trend captures the directional component, and the AR(1)
captures the residual year-to-year fluctuation around the trend. Checking
whether the time trend coefficient is significantly different from zero (its
95% credible interval excludes zero) provides a formal test for directional
change.

### Designing monitoring programmes for temporal models

The statistical power of a multi-season occupancy model depends on the
interaction between three design dimensions: the number of sites ($N$), the
number of periods ($T$), and the number of visits per period ($J$). These
dimensions trade off against each other under a fixed survey budget, and the
optimal allocation depends on the inferential goal.

**For estimating $\rho$ (temporal autocorrelation):** The number of periods
$T$ is the primary driver. Each additional period adds one temporal
transition, and the AR(1) correlation is estimated from these transitions.
With 3 periods there are only 2 transitions; with 10 there are 9. Increasing
$N$ helps by reducing the variance of the per-period occupancy estimate, but
it cannot compensate for too few periods. A minimum of 5 periods is
recommended for reliable $\rho$ estimation; 8--10 is preferable.

**For estimating occupancy trends (fixed time effects):** Both $T$ and $N$
matter. The trend coefficient is estimated from the contrast in occupancy
across periods, and the precision of each period's occupancy estimate depends
on $N$ and $J$. A linear trend requires at least 3 periods; detecting
curvature (quadratic trends) requires 5+.

**For estimating covariate effects:** The number of sites $N$ is most
important, because covariate effects are estimated cross-sectionally. More
sites provide more covariate variation, improving the precision of
$\boldsymbol{\beta}$. The temporal dimension helps primarily by increasing
the total number of site-period observations, but the effective sample size
is reduced by the temporal autocorrelation (highly correlated periods
contribute less independent information than uncorrelated ones).

**For separating detection from occupancy:** The number of visits $J$ per
period is critical. With $J = 2$, detection and occupancy are weakly
identified and the model relies heavily on prior information. With $J = 3$,
identification is adequate for most scenarios. Beyond $J = 4$, the marginal
gain from additional visits is small, and the budget is better spent on
additional sites or periods.

A practical rule of thumb for a balanced design: allocate roughly 50% of
the budget to sites, 30% to periods, and 20% to replicate visits within
periods. This yields a design like $N = 100$, $T = 8$, $J = 3$, which is
adequate for most temporal occupancy analyses.

### Computational considerations

Space-time models (spatial + temporal) stack $N \times T$ site-period
combinations, so the effective sample size grows quickly. For large $N$
(> 500 sites) and long time series ($T$ > 10), consider:

- Fitting spatial-only and temporal-only models first to assess which
  component matters more.
- Using a coarser mesh for the spatial SPDE to reduce the number of mesh
  nodes.
- Subsetting to a region of interest if the full dataset is too large.

The temporal component alone (without spatial) is fast even for large $N$,
because the AR(1) precision matrix is tridiagonal and INLA exploits its
sparsity. A dataset with 500 sites and 10 periods (5,000 site-periods) fits
in seconds on a modern laptop. Adding the spatial SPDE increases this to
minutes, depending on mesh resolution.

### Interpreting AR(1) parameters for management

The estimated $\rho$ and $\sigma_\eta$ have direct implications for
conservation planning:

| Parameter regime              | Ecological interpretation | Management implication |
|:------------------------------|:-------------------------|:-----------------------|
| High $\rho$, low $\sigma$    | Stable, persistent       | Protect current sites  |
| High $\rho$, high $\sigma$   | Persistent shocks        | Buffer against disturbance |
| Low $\rho$, low $\sigma$     | Stable but reshuffling   | Maintain connectivity  |
| Low $\rho$, high $\sigma$    | Volatile, unpredictable  | Landscape-level protection |

These regimes can guide monitoring frequency. High-$\rho$ systems change
slowly and can be surveyed less frequently (every 2--3 years) without losing
important temporal information. Low-$\rho$ systems require annual monitoring
because the state changes unpredictably from year to year.


## Prior specification for temporal parameters

INLA assigns default priors to the AR(1) hyperparameters: a PC prior on the
marginal precision $\tau_\eta = 1/\sigma^2_\eta$ and a uniform prior on
$\rho \in (-1, 1)$. The defaults are weakly informative, which works well in
most cases. When you have strong prior information about the temporal
structure, you can pass custom priors through the `priors` argument of
`occu()`.

The relevant parameters are:

- **$\sigma_\eta$ (innovation SD):** Controls the magnitude of year-to-year
  fluctuation. Under the default PC prior, $\text{Prob}(\sigma > 1) = 0.01$,
  which gently penalises large temporal variances. If your system is known to
  have small year-to-year variation (e.g., a stable forest bird community),
  you can tighten this by decreasing the reference SD. If you study a highly
  variable system (e.g., desert ephemeral ponds), you may relax it.

- **$\rho$ (AR correlation):** The default uniform prior on $(-1, 1)$ is
  non-informative. If you have biological reason to expect strong positive
  persistence (e.g., a territorial species with high site fidelity), you can
  constrain $\rho$ to positive values. In practice, negative temporal
  autocorrelation is rare in occupancy data, and constraining $\rho > 0$
  can improve convergence when the time series is short.

```{r prior-example}
# Custom prior: tighter temporal variance, positive-only rho
fit_priors <- occu(
  ~ occ_x1, ~ det_x1,
  data = sim$data,
  temporal = "ar1",
  priors = list(
    temporal = list(
      prec = list(prior = "pc.prec", param = c(0.5, 0.01)),  # P(sigma > 0.5) = 0.01
      rho  = list(prior = "normal", param = c(0, 0.15))      # Penalised toward 0
    )
  ),
  verbose = 0
)
temporalCorr(fit_priors)
```

The effect of prior choice is strongest when data are sparse (few seasons,
few sites). With 10+ seasons and 100+ sites, the likelihood dominates the
prior and the posterior is largely data-driven. A useful sensitivity check is
to fit the same model with the default and custom priors and compare the
posteriors: if they differ substantially, the data are not informative enough
to override the prior, and conclusions should be hedged accordingly.


## Handling unequal sampling across seasons

Real monitoring programmes rarely survey every site in every year. Some sites
are added part-way through the study, others are dropped due to access issues,
and within a given year some visits may be cancelled due to weather. INLAocc
handles all of these cases through `NA` values in the 3D detection array.

### Worked example: staggered site entry

We simulate a dataset where half the sites are surveyed for all 8 years, but
the other half only enter the study in year 4.

```{r unequal-sim}
set.seed(2024)
N_ue <- 100
J_ue <- 3
n_seasons_ue <- 8

# Simulate full data
sim_ue <- simTOcc(
  N = N_ue, J = J_ue, n_seasons = n_seasons_ue,
  beta_occ = c(0.3, 0.5),
  beta_det = c(0.2, -0.3),
  rho = 0.6, sigma_t = 0.4,
  seed = 55
)

# Knock out first 3 years for sites 51-100
sim_ue$data$y[51:100, 1:3, ] <- NA

# Verify the missing structure
cat("Year-by-year sample sizes:\n")
n_surveyed <- apply(sim_ue$data$y, 2, function(y_t) {
  sum(apply(y_t, 1, function(row) any(!is.na(row))))
})
data.frame(period = 1:n_seasons_ue, n_sites_surveyed = n_surveyed)
```

### Fitting the model with unequal sampling

The model call is identical; INLAocc automatically handles the missing
values:

```{r fit-unequal}
fit_ue <- occu(
  ~ occ_x1, ~ det_x1,
  data = sim_ue$data,
  temporal = "ar1", verbose = 0
)
summary(fit_ue)
```

The AR(1) parameter estimates remain well-identified because the complete
sites (1--50) inform the temporal structure across all 8 periods, and the
partial sites (51--100) contribute information from period 4 onward. The
credible intervals for $\rho$ will be slightly wider than with complete data,
reflecting the reduced information in early periods.

### Visualizing the effect of missing data

We can compare per-period occupancy estimates and see that early periods have
wider uncertainty because they are informed by fewer sites:

```{r vis-unequal}
psi_ue <- fitted(fit_ue)$psi

psi_ue_summary <- data.frame(
  period   = seq_len(n_seasons_ue),
  mean_psi = colMeans(psi_ue, na.rm = TRUE),
  lo       = apply(psi_ue, 2, quantile, 0.025, na.rm = TRUE),
  hi       = apply(psi_ue, 2, quantile, 0.975, na.rm = TRUE),
  n_sites  = n_surveyed
)

ggplot(psi_ue_summary, aes(x = period, y = mean_psi)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 1, colour = "steelblue") +
  geom_point(aes(size = n_sites), colour = "steelblue") +
  scale_y_continuous(limits = c(0, 1)) +
  scale_size_continuous(range = c(1.5, 4), name = "Sites surveyed") +
  labs(x = "Period", y = "Estimated occupancy",
       title = "Occupancy trajectory with staggered site entry") +
  theme_minimal() +
  theme_transparent()
```

The point sizes reflect the number of sites surveyed in each period, making
it visually clear where the data are sparse. The AR(1) temporal correlation
helps stabilise estimates in data-poor periods by borrowing information from
adjacent years, a key advantage of the structured temporal model over
treating each year independently.


## Posterior predictive checks

Posterior predictive checks assess whether the fitted model generates data
that resemble the observed data. If the model is adequate, replicated
datasets drawn from the posterior should be statistically indistinguishable
from the real data. The `ppcOccu()` function implements this by repeatedly
simulating new detection histories from the fitted model and comparing a
summary statistic between observed and replicated data.

```{r ppc-temporal}
ppc_result <- ppcOccu(fit, fit.stat = "freeman-tukey", group = 1, n.samples = 500)

# Bayesian p-value: values near 0.5 indicate good fit
cat(sprintf("Bayesian p-value: %.3f\n", ppc_result$bayesian.p))
```

A Bayesian p-value close to 0.5 indicates the model generates data
consistent with the observations. Values near 0 or 1 suggest systematic lack
of fit. For temporal models, run the check on each period
separately to identify years where the model fits poorly, since a single
anomalous year (e.g., a flood or drought) can dominate the overall p-value.

```{r ppc-by-period}
# Per-period posterior predictive checks
ppc_by_period <- data.frame(
  period = seq_len(sim$truth$n_seasons),
  bayesian_p = NA_real_
)

for (t in seq_len(sim$truth$n_seasons)) {
  period_fit <- fit$period_fits[[t]]
  if (!is.null(period_fit)) {
    ppc_t <- ppcOccu(period_fit, fit.stat = "freeman-tukey",
                     group = 1, n.samples = 500)
    ppc_by_period$bayesian_p[t] <- ppc_t$bayesian.p
  }
}
ppc_by_period
```

Periods with Bayesian p-values consistently above 0.9 or below 0.1 deserve
closer inspection. The model may be generating data that are too variable
(p-value near 1) or not variable enough (p-value near 0) for those specific
years. Common causes include: unmodelled heterogeneity in detection during
that year, a covariate effect that changed between periods, or a genuine
ecological event (die-off, invasion) that the AR(1) smooth cannot capture.


## Prediction with uncertainty at new sites

The `predict()` method for temporal models generates occupancy predictions at
new covariate values across all or selected periods. This is the primary
tool for generating species distribution maps at past, current, or
interpolated time points.

### Predicting along a covariate gradient

A common use case is predicting how occupancy changes along an environmental
gradient, separately for each period. This reveals whether the
species-environment relationship is stable over time or shifts due to the
temporal random effect.

```{r predict-gradient}
# Create a gradient of 50 points spanning the covariate range
cov_gradient <- data.frame(occ_x1 = seq(-2, 2, length.out = 50))

preds_gradient <- predict(fit, X.0 = as.matrix(cov_gradient), type = "occupancy")

# Extract predictions for each period
pred_df <- do.call(rbind, lapply(seq_along(preds_gradient$psi.0), function(i) {
  p <- preds_gradient$psi.0[[i]]
  data.frame(
    period  = preds_gradient$periods[i],
    occ_x1  = cov_gradient$occ_x1,
    psi     = p$mean,
    lo      = p$quantiles[, 1],
    hi      = p$quantiles[, 3]
  )
}))
pred_df$period <- factor(pred_df$period)

ggplot(pred_df, aes(x = occ_x1, y = psi, colour = period, fill = period)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.08, colour = NA) +
  geom_line(linewidth = 0.8) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Covariate value", y = "Predicted occupancy",
       colour = "Period", fill = "Period",
       title = "Occupancy response curves by period") +
  theme_minimal() +
  theme_transparent()
```

The curves share the same shape (determined by the fixed-effect slope) but
are shifted vertically by the period-specific temporal random effect.
Periods with positive $\eta_t$ shift the curve upward; periods with negative
$\eta_t$ shift it downward. The separation between the highest and lowest
curves visualises the magnitude of year-to-year fluctuation in the system.

### Predicting at specific periods

When you need predictions for only a subset of periods (e.g., the first and
last year for a before-after comparison):

```{r predict-specific-periods}
preds_first_last <- predict(fit, X.0 = as.matrix(cov_gradient),
                            period = c(1, 5), type = "occupancy")

# Compare occupancy at the covariate mean (occ_x1 = 0)
mid_idx <- 25  # middle of the gradient
for (nm in names(preds_first_last$psi.0)) {
  p <- preds_first_last$psi.0[[nm]]
  cat(sprintf("%s: psi = %.3f (%.3f, %.3f)\n",
              nm, p$mean[mid_idx], p$quantiles[mid_idx, 1], p$quantiles[mid_idx, 3]))
}
```

The difference in predicted occupancy between two periods, at the same
covariate value, isolates the temporal component. If the credible intervals
do not overlap, the occupancy change between those periods is statistically
supported.

### Predicting detection probability

In addition to occupancy, `predict()` supports `type = "detection"` to
generate detection probability predictions at new covariate values. This is
useful when assessing whether survey conditions are adequate for reliable
detection, or when planning future surveys under specific conditions.

```{r predict-detection}
# Predict detection probability along a gradient of the detection covariate
det_gradient <- data.frame(det_x1 = seq(-2, 2, length.out = 50))
preds_det <- predict(fit, X.p.0 = as.matrix(det_gradient), type = "detection")

if (!is.null(preds_det$p.0)) {
  det_pred_df <- data.frame(
    det_x1 = det_gradient$det_x1,
    p      = preds_det$p.0$mean,
    lo     = preds_det$p.0$quantiles[, 1],
    hi     = preds_det$p.0$quantiles[, 3]
  )

  ggplot(det_pred_df, aes(x = det_x1, y = p)) +
    geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "darkorange") +
    geom_line(linewidth = 1, colour = "darkorange") +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Detection covariate value", y = "Detection probability",
         title = "Detection response curve") +
    theme_minimal() +
    theme_transparent()
}
```

Detection predictions do not vary across periods in the same way as
occupancy predictions, because the detection submodel does not include a
temporal random effect. However, if detection covariates vary by period
(e.g., survey date or observer effort), the predicted detection probability
will differ across periods through the covariate values. This distinction
matters for survey planning: the occupancy temporal random effect is a
property of the species, while period-to-period variation in detection is
a property of the survey design that can be controlled.


## Expanded multi-species temporal analysis

The basic multi-species temporal model shown earlier provides community-level
summaries, but individual species trajectories are often the primary interest.
Here we demonstrate a fuller workflow: simulating a multi-species temporal
dataset, fitting the model, extracting species-specific occupancy
trajectories, and visualizing them.

### Simulation with contrasting species dynamics

We simulate 5 species with deliberately different temporal dynamics to
illustrate the model's ability to recover species-specific patterns.

```{r ms-sim-detailed}
sim_ms <- simTMsOcc(
  N = 80, J = 3, n_species = 5, n_seasons = 6,
  beta_comm_mean = c(0.5, 0.3),    # community mean: moderate occupancy, positive covariate
  beta_comm_sd   = c(1.5, 0.6),    # high among-species variation
  alpha_comm_mean = c(0.0, -0.2),  # community detection mean
  alpha_comm_sd   = c(0.8, 0.3),
  rho = 0.6, sigma_t = 0.5,
  seed = 42
)

str(sim_ms$data$y)
```

### Fitting and extracting species-specific results

```{r ms-fit-detailed}
fit_ms <- occu(
  ~ occ_x1, ~ det_x1,
  data = sim_ms$data,
  multispecies = TRUE,
  temporal = "ar1",
  verbose = 0
)
summary(fit_ms)
```

### Species-specific occupancy trajectories

We extract the fitted occupancy values per species and period to build
a panel plot showing how each species' occupancy evolves over time.

```{r ms-trajectories, fig.height = 7}
n_sp <- length(fit_ms$species_fits)
n_per <- fit_ms$n_periods

# Build trajectory data.frame
traj_list <- vector("list", n_sp)
for (s in seq_len(n_sp)) {
  sp_fit <- fit_ms$species_fits[[s]]
  if (is.null(sp_fit)) next
  psi_sp <- fitted(sp_fit)$psi  # N x T matrix
  traj_list[[s]] <- data.frame(
    species  = paste("Species", s),
    period   = seq_len(n_per),
    mean_psi = colMeans(psi_sp),
    lo       = apply(psi_sp, 2, quantile, 0.025),
    hi       = apply(psi_sp, 2, quantile, 0.975)
  )
}
traj_df <- do.call(rbind, Filter(Negate(is.null), traj_list))

ggplot(traj_df, aes(x = period, y = mean_psi)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), alpha = 0.2, fill = "steelblue") +
  geom_line(linewidth = 0.8, colour = "steelblue") +
  geom_point(size = 2, colour = "steelblue") +
  facet_wrap(~ species, ncol = 2) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Period", y = "Estimated occupancy",
       title = "Species-specific occupancy trajectories") +
  theme_minimal() +
  theme_transparent()
```

The panel layout makes it straightforward to compare temporal dynamics across
species. Species with high baseline occupancy and small temporal fluctuation
(narrow ribbons) are the stable residents of the community. Species with wide
ribbons or steep trajectories are the dynamic members, potentially
nomadic, disturbance-dependent, or tracking fluctuating resources.

### Community-level temporal trend

A useful community summary is the mean occupancy across all species per
period, which reveals whether the community as a whole is expanding,
contracting, or stable.

```{r ms-community-trend}
community_trend <- aggregate(mean_psi ~ period, data = traj_df, FUN = mean)
community_lo    <- aggregate(lo ~ period, data = traj_df, FUN = mean)
community_hi    <- aggregate(hi ~ period, data = traj_df, FUN = mean)

community_summary <- data.frame(
  period   = community_trend$period,
  mean_psi = community_trend$mean_psi,
  lo       = community_lo$lo,
  hi       = community_hi$hi
)

ggplot() +
  geom_line(data = traj_df,
            aes(x = period, y = mean_psi, group = species),
            colour = "grey70", linewidth = 0.4) +
  geom_ribbon(data = community_summary,
              aes(x = period, ymin = lo, ymax = hi),
              alpha = 0.2, fill = "firebrick") +
  geom_line(data = community_summary,
            aes(x = period, y = mean_psi),
            linewidth = 1.2, colour = "firebrick") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Period", y = "Mean occupancy",
       title = "Community-level occupancy trend",
       subtitle = "Red = community mean; grey = individual species") +
  theme_minimal() +
  theme_transparent()
```

Divergence between species trajectories (grey lines spreading apart over
time) indicates differential responses to environmental change. When all
species decline together, a shared driver (e.g., landscape-level habitat
loss) is likely. When some increase while others decrease, species-specific
traits mediate the response, and single-species management may be
insufficient.

### Comparing species-specific temporal parameters

For models fit with `temporal = "ar1"`, the temporal autocorrelation and
innovation variance are estimated per species. Extracting these allows a
direct comparison of temporal dynamics across the community.

```{r ms-temporal-params}
sp_temporal <- data.frame(
  species = character(),
  rho     = numeric(),
  sigma   = numeric(),
  stringsAsFactors = FALSE
)

for (s in seq_len(n_sp)) {
  sp_fit <- fit_ms$species_fits[[s]]
  if (is.null(sp_fit)) next
  tc <- tryCatch(temporalCorr(sp_fit), error = function(e) NULL)
  if (!is.null(tc)) {
    rho_val <- tc$mean[grep("[Rr]ho", rownames(tc))]
    prec_val <- tc$mean[grep("[Pp]recision", rownames(tc))]
    sigma_val <- if (length(prec_val) > 0) 1 / sqrt(prec_val) else NA_real_
    sp_temporal <- rbind(sp_temporal, data.frame(
      species = paste("Species", s),
      rho     = if (length(rho_val) > 0) rho_val else NA_real_,
      sigma   = sigma_val,
      stringsAsFactors = FALSE
    ))
  }
}

if (nrow(sp_temporal) > 0) {
  sp_temporal[order(-sp_temporal$rho), ]
}
```

Species with high $\rho$ and low $\sigma$ are the temporal "anchors" of
the community; their occupancy is highly predictable from year to year.
Species with low $\rho$ and high $\sigma$ are the "wildcards" whose
presence is driven by stochastic or unmeasured factors. This information
can guide monitoring design: anchor species need less frequent surveys,
while wildcard species require annual visits to track their volatile
dynamics.


## Formal model comparison: static vs. temporal

The earlier comparison used a null vs. covariate model, both with temporal
structure. Here we ask the more fundamental question: does temporal structure
improve on a static (single-season stacked) model at all?

To compare fairly, we fit three models to the same multi-season data:

1. **Static:** No temporal structure. The 3D array is collapsed to 2D by
   stacking all site-periods. Each site-period is treated as an independent
   observation.
2. **IID temporal:** Year effects are independent draws, absorbing
   unstructured year-to-year variation.
3. **AR(1) temporal:** Year effects are temporally correlated.

```{r static-vs-temporal}
# Fit the static model by dropping the temporal argument
fit_static <- occu(~ occ_x1, ~ det_x1, data = sim$data, verbose = 0)

# Fit IID and AR(1) temporal models
fit_iid_comp  <- occu(~ occ_x1, ~ det_x1, data = sim$data, temporal = "iid", verbose = 0)
fit_ar1_comp  <- occu(~ occ_x1, ~ det_x1, data = sim$data, temporal = "ar1", verbose = 0)

comp_table <- compare_models(
  static = fit_static,
  iid    = fit_iid_comp,
  ar1    = fit_ar1_comp,
  criterion = "waic"
)
comp_table[, c("model", "WAIC", "delta", "weight")]
```

The WAIC comparison provides a principled ranking. Since the simulated data
contain genuine AR(1) structure ($\rho = 0.7$), the AR(1) model should
receive the highest weight. The static model pays a penalty for ignoring
temporal dependence, which inflates residual variance and produces
overconfident standard errors on the fixed effects.

### How large must the WAIC difference be?

A common guideline is that a WAIC difference ($\Delta$WAIC) greater than 2
provides positive evidence for the better model, and $\Delta$WAIC > 10
provides very strong evidence (analogous to AIC interpretation from Burnham
and Anderson 2002). When $\Delta$WAIC < 2, the models are effectively
indistinguishable and the simpler model should be preferred for parsimony.

The WAIC weights in the comparison table can also be interpreted as
approximate posterior model probabilities. A weight of 0.95 for the AR(1)
model means there is 95% probability that it provides the best
predictive fit among the candidates, given the data.

### Inspecting fixed-effect differences across models

Beyond WAIC, comparing the fixed-effect estimates from the
static and temporal models. The temporal model should produce similar point
estimates but wider (more honest) credible intervals because it accounts for
the non-independence of repeated observations at each site.

```{r compare-fixed-effects}
static_coefs <- coef(fit_static)
ar1_coefs    <- coef(fit_ar1_comp)

coef_comparison <- data.frame(
  parameter = names(static_coefs$occupancy),
  static    = round(unlist(static_coefs$occupancy), 3),
  ar1       = round(unlist(ar1_coefs$occupancy), 3),
  row.names = NULL
)
coef_comparison
```

If the static model's credible intervals are notably narrower, it is
producing false precision by treating temporally correlated observations as
independent. This is the occupancy-modelling equivalent of pseudoreplication,
and it can lead to spurious conclusions about covariate effects.


## Detection trajectories over time

In multi-season models, detection probability often varies as much across
years as occupancy does. Changes in survey personnel, shifting phenology, or
interannual weather variation all affect detection. INLAocc estimates
detection separately for each period, and inspecting these estimates can
reveal whether apparent occupancy trends are partially confounded with
detection trends.

```{r detection-trajectory}
fit_vals <- fitted(fit)
p_by_period <- fit_vals$p  # N x (T * J) matrix, reshaped per period

# Compute mean detection per period
n_per_det <- sim$truth$n_seasons
J_det <- ncol(sim$data$y[1, 1, , drop = FALSE])  # visits per period
p_means_per <- numeric(n_per_det)
for (t in seq_len(n_per_det)) {
  cols <- ((t - 1) * J_det + 1):(t * J_det)
  p_means_per[t] <- mean(p_by_period[, cols], na.rm = TRUE)
}

det_df <- data.frame(
  period = seq_len(n_per_det),
  mean_p = p_means_per
)

ggplot(det_df, aes(x = period, y = mean_p)) +
  geom_line(linewidth = 1, colour = "darkorange") +
  geom_point(size = 2.5, colour = "darkorange") +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Period", y = "Mean detection probability",
       title = "Detection probability over time") +
  theme_minimal() +
  theme_transparent()
```

Ideally, detection should be roughly stable across periods or vary in
a way that is explained by detection covariates. A strong downward trend in
detection coinciding with an upward occupancy trend should raise suspicion
about confounding. It may be that the species is not truly increasing, but
rather that detection is declining and the model is compensating by inflating
occupancy estimates. Comparing the detection trajectory against known changes
in survey methodology helps distinguish real trends from artefacts.

When detection varies strongly across periods but no detection covariates
are available, the model relies entirely on the within-period replicate
structure to identify detection. This works well when each period has at
least 3 visits, but with only 2 visits per period, detection and occupancy
become weakly identifiable, and the temporal random effect may absorb
variation that actually belongs to the detection process. In such cases,
increasing the number of replicate visits, even by a single additional
survey per period, substantially improves identifiability.


## Pseudoreplication and site-level random effects

Multi-season occupancy data contain an inherent source of non-independence:
the same site is observed repeatedly across periods. If we stack all
site-period combinations and treat them as independent observations (as in a
single-season model applied to pooled data), we commit pseudoreplication
(Hurlbert 1984). The consequences are narrower credible intervals than
warranted, inflated test statistics for covariate effects, and a false sense
of precision about temporal trends.

The AR(1) temporal random effect addresses part of this problem by modelling
the shared year-to-year correlation. But it does not account for the
within-site correlation that arises when the same site appears in multiple
periods. A site that is occupied in year 1 is likely occupied in year 2 not
only because of the shared temporal environment ($\eta_t$) but also because
site-level factors (microhabitat quality, local food availability, territory
suitability) are persistent. This site-level persistence is not captured by
the AR(1), which operates at the landscape level.

Two solutions are available in INLAocc:

1. **An unstructured site random effect.** Including a random intercept for
   site adds a site-specific deviation $u_i \sim \mathcal{N}(0, \sigma^2_u)$
   to the occupancy linear predictor. This absorbs persistent differences
   between sites that are not explained by covariates. In formula syntax,
   supply a site index variable in `occ.covs` (e.g., `site_id = 1:N`) and
   include it as `(1 | site_id)` in the formula, or equivalently, use a
   random effect specification if supported by the model interface.

2. **A spatial random effect.** When sites have spatial coordinates and
   nearby sites are expected to share similar occupancy, the SPDE spatial
   random field (see `vignette("spatial-models")`) provides a structured
   alternative to the unstructured site effect. The spatial field induces
   correlation between nearby sites, which is more informative than the
   exchangeable correlation assumed by the unstructured effect. For datasets
   with sufficient spatial coverage, the spatial random effect typically
   outperforms the unstructured site effect in both fit and prediction.

The choice between unstructured and spatial site effects parallels the
choice between IID and AR(1) temporal effects: the unstructured version is
simpler and appropriate when site-level differences have no spatial pattern,
while the spatial version is more informative when site similarity decays with
distance. In practice, if coordinates are available, the spatial model is
usually worth trying.

Both site-level and temporal random effects can be included simultaneously.
The combined model attributes variation in occupancy to four sources:
fixed-effect covariates ($\mathbf{x}_i^\top \boldsymbol{\beta}$),
persistent site-level differences ($u_i$ or $w(\mathbf{s}_i)$), shared
temporal fluctuation ($\eta_t$), and residual noise. This decomposition
provides a clearer picture of what drives occupancy patterns than any single
component alone.


## Summary and next steps

This vignette has covered the full workflow for multi-season occupancy models
in INLAocc: mathematical formulation of the AR(1) temporal random effect, data
structure and handling of missing visits, simulation, model fitting, temporal
diagnostics, prediction at new covariate values, posterior predictive checks,
multi-species extensions, and formal model comparison.

The AR(1) parameterization provides a parsimonious and computationally efficient
alternative to the classical colonization-extinction formulation. It is
particularly well-suited when the inferential target is occupancy trends and
their drivers, rather than the transition rates themselves. The two
hyperparameters ($\rho$ and $\sigma_\eta$) have direct ecological
interpretations that can inform monitoring design and conservation planning.

For spatially structured data, the temporal model can be extended with a
spatial random field (see `vignette("spatial-models")`). For multi-species
data with cross-species correlation, the temporal model integrates with the
multi-species framework (see `vignette("multi-species")`). For species whose
response to environmental drivers varies spatially, spatially varying
coefficient models can capture non-separable spatio-temporal patterns (see
`vignette("svc-models")`).
